### 2018-10-31 article/ms.tex
foo

\documentclass[twocolumn]{aastex61}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{color}
\usepackage{comment}

\newcommand\teff{T_{\rm eff}}
\newcommand\logg{\log{g}}
\newcommand\feh{[\rm{Fe}/\rm{H}]}

\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\package}[1]{\texttt{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}

\newcommand{\Gaia}{\project{Gaia}}
\newcommand{\gaia}{\project{gaia}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}

\received{2018 XX XX}
\revised{2018 XX XX}
\accepted{2018 XX XX}

\newcommand{\vcpath}{vc.tex}

\IfFileExists{\vcpath}{\input{\vcpath}}{
	\newcommand{\giturl}{UNKNOWN}
	\newcommand{\gitslug}{UNKNOWN}
	\newcommand{\githash}{UNKNOWN}
	\newcommand{\gitdate}{UNKNOWN}
	\newcommand{\gitauthor}{UNKNOWN}
}

\submitjournal{AAS Journals}

\shorttitle{MCFA}
\shortauthors{Casey et al.}

\begin{document}

\title{MCFA on Galah}

\correspondingauthor{Andrew R. Casey}
\email{andrew.casey@monash.edu}

\author[0000-0003-0174-0564]{Andrew R. Casey}
\affiliation{School of Physics \& Astronomy, 
			 Monash University,
			 Wellington Rd, Clayton 3800, Victoria, Australia}
\affiliation{Faculty of Information Technology, 
			 Monash University, 
			 Wellington Rd, Clayton 3800, Victoria, Australia}

\author{John Lattanzio}
\affiliation{School of Physics \& Astronomy, 
			 Monash University,
			 Wellington Rd, Clayton 3800, Victoria, Australia}

\author{Aldeida Aleti}
\affiliation{Faculty of Information Technology, 
			 Monash University, 
			 Wellington Rd, Clayton 3800, Victoria, Australia}

\author{David Dowe}
\affiliation{Faculty of Information Technology, 
			 Monash University, 
			 Wellington Rd, Clayton 3800, Victoria, Australia}

\author{the GALAH team}

\begin{abstract}
foo
\end{abstract}

\keywords{methods: statistical}

\section{Introduction} \label{sec:intro}

\section{Model} \label{sec:model}

\section{Results} \label{sec:results}

\section{Discussion} \label{sec:discussion}

\section{Conclusion} \label{sec:conclusion}

\software{
	\package{Astropy} \citep{astropy:v1,astropy:v2},
    \package{IPython} \citep{ipython},
    \package{matplotlib} \citep{mpl},
    \package{numpy} \citep{numpy},
    \package{scipy} \citep{scipy},
    \package{Stan} \citep{stan},
    \package{CosmoHub} \citep{cosmohub},
    \package{TensorFlow} \citep{tensorflow}
    \package{Jupyter Notebooks} \citep{jupyter-notebooks}
}

\bibliographystyle{aasjournal}
\bibliography{mcfa}

\end{document}

### 2018-10-31 mcfa/mcfa.py
make faster

@@ -9,6 +9,7 @@ from inspect import signature as inspect_signature
 from scipy import linalg
 from scipy.special import logsumexp
 from sklearn.cluster import KMeans
+from time import  time
 
 logger = logging.getLogger(__name__)
 
@@ -17,7 +18,7 @@ class MCFA(object):
 
     r""" A mixture of common factor analyzers model. """
 
-    def __init__(self, n_components, n_latent_factors, max_iter=500, n_init=5, 
+    def __init__(self, n_components, n_latent_factors, max_iter=500, n_init=25, 
                  tol=1e-5, verbose=1, random_seed=None, **kwargs):
         r"""
         A mixture of common factor analyzers model.
@@ -289,20 +290,14 @@ class MCFA(object):
                 # = log(det(I + A @ omega @ U))
                 _, logdetD = np.linalg.slogdet(I_D + A @ omega_ @ U)
 
-            precision = _compute_precision_cholesky_full(cov)
-            diff = np.dot(X, precision) - np.dot(mu, precision)
-            dist = np.sum(np.square(diff), axis=1)
-
             log_prob[:, i] = -0.5 * dist - 0.5 * logdetD + log_prob_constants
 
         weighted_log_prob = log_prob + np.log(pi)
-        log_likelihood = logsumexp(weighted_log_prob, axis=1)
+        log_prob = logsumexp(weighted_log_prob, axis=1)
         with np.errstate(under="ignore"):
-            log_tau = weighted_log_prob - log_likelihood[:, np.newaxis]
+            log_tau = weighted_log_prob - log_prob[:, np.newaxis]
 
-        tau = np.exp(log_tau)
-
-        return (sum(log_likelihood), tau)
+        return (np.sum(log_prob), np.exp(log_tau))
 
 
     def maximization(self, X, tau, pi, A, xi, omega, psi):
@@ -389,9 +384,10 @@ class MCFA(object):
             A1 += np.atleast_2d(np.sum(X * tau_, axis=0)).T @ xi_.T \
                 + X.T @ tY_Axi_i.T @ gamma
             A2 += (omega[:, :, i] + xi[:, [i]] @ xi[:, [i]].T) * ti[i]
-            Di += np.sum(self._X2 * tau_, axis=0)
+        
 
         A = A1 @ linalg.solve(A2, I_J)
+        Di = np.sum(self._X2.T @ tau[np.newaxis], axis=2)
         psi = (Di - np.sum((A @ A2) * A, axis=1)) / N
         
         return (pi, A, xi, omega, psi)
[…]

### 2018-10-31 article/ms.tex
some methods

@@
 \usepackage{bm}
-\usepackage{amsmath}
+\usepackage{amsmath,amsfonts,amssymb}
 \usepackage{color}
@@
 \newcommand{\todo}[1]{\textcolor{red}{#1}}
+
+\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
+\renewcommand{\vec}[1]{\vect{#1}}
+
+\newcommand{\weight}{\pi}
+\newcommand{\data}{\textbf{Y}}
+\newcommand{\vecdata}{\vec\data}
+
+\newcommand{\nextstep}{^\textrm{(t+1)}}
+\newcommand{\thisstep}{^\textrm{(t)}}
+\newcommand{\transpose}{^\intercal}
+\newcommand{\eye}{\textbf{I}}
+
+\newcommand{\factorloads}{\textbf{L}}
+\newcommand{\factorscores}{\textbf{S}}
+
+\newcommand{\scoremeans}{\vec\xi}
+\newcommand{\scorecovs}{\vec\Omega}
 
@@
 
-\section{Model} \label{sec:model}
+\section{Methods} \label{sec:methods}
+
+% Describe how we will attempt to define a generative model for the data, from the latent space.
+
+In this work the data $\vecdata$ is a $n \times d$ matrix 
+where $n$ is the number of stars and $d$ is the number of
+chemical abundances measured for each star. We assume a
+generative model for the data 
+\begin{equation}
+	\vecdata = \factorloads\factorscores + \vec{e}
+	\label{eq:generative-model}
+\end{equation}
+
+\noindent{}where $\factorloads$ is a $p \times d$ matrix of factor
+loads that is common to all data points, and the factor scores
+\begin{equation}
+	\factorscores \sim \mathcal{N}(\vec\xi_k, \vec\Omega_k)
+\end{equation}
+\noindent{}is a \todo{$n \times p$} matrix where the entries
+are drawn from $k$ multivariate normal distributions.
+We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \vec\psi\right)$
+is independent of the latent space, and $\vec\psi$ is a
+diagonal matrix of $d$ entries.
+
+In this model each data point can be represented as being drawn
+from a mixture of multivariate normal components, except the components
+are \emph{clustered in the latent space} $\factorscores$ and projected
+into the data space by the factor loads $\factorloads$. 
+We assume that the latent space is lower dimensionality than the
+data space (e.g., $p < d$).
+Within the context of stellar abundances, the factor loads
+$\factorloads$ can be thought of as the \emph{mean} yields
+of nucleosynthetic
+events (e.g., $s$-process yields from AGB stars), and the
+factor scores are analogous to the relative counts of those 
+nucleosynthetic events. The clustering in factor scores
+achieves the same as a clustering procedure in data space,
+except we simultaneously estimate the factor loads
+(nucleosynthetic yields) that are common to all data points 
+(stars). Within this framework a rare nucleosynthetic event
+can still be described as a `common factor load' $\factorloads_i$, 
+but its rarity would be represented by associated factor
+scores being zero for most stars and thus have no contribution
+to the observed abundances.
+
+Without loss of generality the density of the data $\vecdata$ can be modelled as
+\begin{equation}
+	f(\textbf{y}; \vec\Psi) = \sum_{k=1}^{K}\weight_{k}\phi(\textbf{y};\factorloads\scoremeans_k, \factorloads\scorecovs_k\factorloads\transpose + \eye\vec\psi)
+\end{equation}
+\noindent{}given $p$ common factor loadings and $K$ components
+clustered in the latent (factor score) space. Here the parameter
+vector
+$\vec\Psi$ includes $\{\factorloads,\vec\pi,\scoremeans,\scorecovs,\vec\psi\}$, and $\phi(\textbf{y};\vec\mu, \vec\Sigma)$
+describes the density of a multivariate gaussian distribution with
+mean $\vec\mu$ and covariance matrix $\vec\Sigma$,
+and $\weight_k$ describes the relative weighting of the $k$th
+component in latent space and $\sum\weight_k = 1$.
+The log likelihood is then given by
+\begin{equation}
+	\log\mathcal{L}(\vecdata|\vec\Psi) = \sum_{k=1}^{K}\log{f(\vecdata;\vec\Psi)} \quad .
+\end{equation}
+
+The model described by Equation~\ref{eq:generative-model} is indeterminate:
+there is no unique solution for the factor loads $\factorloads$ and scores
+$\factorscores$. However, an accurate estimate of the parameter vector $\vec\Psi$ can be obtained by the expectation-maximization algorithm \citep{EM}.
+
+\subsection{Initialisation}
+
+We use the K-means++ algorithm to initially assign data points
+randomly to $K$ clusters (in data space). This provides a so-called
+hard initial association where each data point is wholly assigned
+to a single component. Specifically we calculate the $n \times K$
+responsibility matrix $\vec\tau$, where a entry of 1 indicates the
+$n$th data point is associated with the $k$th cluster, and an
+entry of 0 indicates no association such that
+\begin{equation}
+	\weight_k = \frac{1}{N}\sum_{n=1}^{N}\tau_{nk} \quad .
+\end{equation}
+We then chose 1,000
+random data points and factorize that matrix using singular-value
+decomposition. We take the principal $p$ eigenvectors as the 
+initial guess for the factor loads $\factorloads$, and calculate
+an initial estimate of the factor score means
+\begin{equation}
+	\scoremeans_k = \frac{1}{N}\sum_{n=1}^{N}\tau_{nk}\vecdata_{n}\factorloads
+\end{equation}
+\noindent{}and the covariance matrices of the clusters in latent space
+\begin{equation}
+	\scorecovs_k = \textrm{cov}{\left(\tau_{nk}\vecdata_{n}\right)} \quad .
+\end{equation}
+
+\subsection{Expectation-Maximization}
+
+At the expectation step we evaluate the log likelihood
+given the model parameters $\vec\Psi$, and we re-calculate the responsibility
+matrix whose entries are the posterior probability
+that the $n$th data point is associated to the $k$th component,
+given the data $\vecdata$ and the current estimate of the 
+parameter vector $\vec\Psi$:
+\begin{equation}
+	\tau_{nk} = \frac{\weight_k\phi(\vecdata_n;\factorloads\scoremeans_k, \factorloads\scorecovs_k\factorloads\transpose + \eye\vec\psi)}{\sum_{g=1}^{G}\weight_g\phi(\vecdata_n;\factorloads\scoremeans_g, \factorloads\scorecovs_g\factorloads\transpose + \eye\vec\psi)} \quad .
+\end{equation}
+
+% SVD complexity is O(n^3)
+% K-means++ complexity is O(log(k))
+
+% Discuss connection to PCA, FA, MFA.
+
+% Include macros for \vec, \data, etc.
+At the maximization step we update our estimates of the parameters,
+conditioned on the data $y$ and the responsibility matrix $\tau$.
+
+\begin{equation}
+	\weight_k\nextstep = \frac{1}{N} \sum_{n=1}^{N}\tau_{nk}
+	% Include proof.
+\end{equation}
+
+\begin{eqnarray}
+	\vec\mu_{k}\nextstep = \vec\mu_{k}\thisstep + \vec\gamma\transpose \frac{1}{N\weight_k}\sum_{n=1}^{N}{(\vecdata\transpose - \vec{}A\vec\mu_{k}\thisstep)\vec\tau_{k}}
+\end{eqnarray}
+
+% The intermediate stpe for phi:
+\begin{equation}
+	\vec\phi = N\sum_{k=1}^{K}\weight_k\left[\vec\Omega_k\nextstep + \vec\mu_k\nextstep(\vec\mu_k\nextstep)\transpose\right]
+\end{equation}
+
+% And the update of the psi.
+\begin{equation}
+	\vec\psi\nextstep = \frac{1}{N}\left[\sum_{k=1}^{K}\vec\data^2\vec\tau_{k} - (\vec{}A\nextstep \vec\phi)\vec{}A\right]
+\end{equation}
+
+\section{Experiments} \label{sec:experiments}
+
+\subsection{Toy model with generated data} \label{sec:experiment-toy-model}
+
+\subsection{GALAH} \label{sec:experiment-galah}
 

### 2018-11-01 mcfa/mcfa.py
minor

@@ -157,6 +157,8 @@ class MCFA(object):
         assignments = _initial_assignments(X, self.n_components, self.n_init)
         best_theta, best_log_likelihood = (None, -np.inf)
 
+        exceptions = []
+
         for i, assignment in enumerate(assignments):
             try:
                 params = _initial_parameters(X, 
@@ -164,7 +166,8 @@ class MCFA(object):
                                              self.n_latent_factors,
                                              assignment)
 
-            except ValueError:
+            except ValueError as exception:
+                exceptions.append(exception)
                 if self.verbose > 0:
                     logger.exception("Exception in initializing:")
 
@@ -178,7 +181,8 @@ class MCFA(object):
                     if log_likelihood > best_log_likelihood:
                         best_theta, best_log_likelihood = (theta, log_likelihood)
 
-                except:
+                except Exception as exception:
+                    exceptions.append(exception)
                     if self.verbose > 0:
                         logger.exception("Exception in running E-M step from "\
                                          "initialisation point:")
@@ -358,7 +362,6 @@ class MCFA(object):
         inv_D_A = inv_D @ A
 
         ti = np.sum(tau, axis=0).astype(float)
-        pi = ti / N
         
         I_J = np.eye(self.n_latent_factors)
 
@@ -385,11 +388,12 @@ class MCFA(object):
                 + X.T @ tY_Axi_i.T @ gamma
             A2 += (omega[:, :, i] + xi[:, [i]] @ xi[:, [i]].T) * ti[i]
         
-
         A = A1 @ linalg.solve(A2, I_J)
         Di = np.sum(self._X2.T @ tau[np.newaxis], axis=2)
         psi = (Di - np.sum((A @ A2) * A, axis=1)) / N
         
+        pi = ti / N
+
         return (pi, A, xi, omega, psi)
 
 

### 2018-11-01 test_mcfa.py

@@ -6,13 +6,13 @@ from sklearn.decomposition import FactorAnalysis
 from mcfa import mcfa, utils
 
 
-n_features = 8
-n_components = 30
+n_features = 15
+n_components = 2
 n_latent_factors = 3
 n_samples = 10000
 
-omega_scale = 1e-1
-noise_scale = 1e-1
+omega_scale = 1
+noise_scale = 1
 random_seed = 42
 
 X, truth = utils.generate_data(n_samples=n_samples,

### 2018-11-01 article/ms.tex
add linear algebra for m-step

@@
 \newcommand{\factorscores}{\textbf{S}}
+\newcommand{\specificvariance}{\vec{D}}
 
@@
 \noindent{}where $\factorloads$ is a $p \times d$ matrix of factor
-loads that is common to all data points, and the factor scores
+loads that is common to all data points, and the factor scores for
+the $i$th data point
 \begin{equation}
-	\factorscores \sim \mathcal{N}(\vec\xi_k, \vec\Omega_k)
+	\factorscores_i \sim \mathcal{N}(\vec\xi_k, \vec\Omega_k)
 \end{equation}
-\noindent{}is a \todo{$n \times p$} matrix where the entries
-are drawn from $k$ multivariate normal distributions.
-We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \vec\psi\right)$
-is independent of the latent space, and $\vec\psi$ is a
+\noindent{}are drawn from the $k$th multivariate  normal distribution.
+The factor scores for all data points $\factorscores$ is then a 
+\todo{$n \times p$} matrix.
+ We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \specificvariance\right)$
+is independent of the latent space, and $\specificvariance$ is a
 diagonal matrix of $d$ entries.
@@
 scores being zero for most stars and thus have no contribution
-to the observed abundances.
+to the observed abundances. The factor loads and factor scores 
+cannot be expressly interpreted like this because they have no
+physical meaning, but this description of relative yields and 
+rates should help build intuition for the model parameters,
+and provide an astrophysical context.
+
+One benefit of using a mixture of common factor analysers is
+for modelling stellar chemical abundances is the scaling with
+computational cost. If we considered data sets of order $3\times10^7$
+entries (e.g., 30 chemical abundances for $10^6$ stars) purely as a
+clustering problem, then even the most efficient clustering
+algorithms would incur a significant cumulative computational 
+overhead by searching the parameter space for the number of
+clusters, and the optimal model parameters given that number
+of components. However, because the mixture of factor analyzers
+approach assumes that there is a \emph{lower dimensional latent 
+space} in which the data are clustered, and that clustering is 
+projected into real space by common factor loads, the 
+dimensionality of the clustering problem is reduced from 
+$N \times D$ to $N \times J$. This reduces cost both through
+faster execution of each E-M step, and on average fewer E-M steps
+required to reach a specified convergence threshold.
+
+From a statistical standpoint, the primary advantage to using
+a mixture of factor analysers is that we can simultaneously
+estimate latent factors (e.g., infer nucleosynthetic 
+yields) and perform clustering (e.g., chemical tagging) 
+within a statistically consistent framework. That is to say
+that we have a generative model for the data that includes
+analogues of nucleosynthetic yields and star formation,
+and the parameters of this model can be estimated consistently
+given some data.
 
@@
 \begin{equation}
-	f(\textbf{y}; \vec\Psi) = \sum_{k=1}^{K}\weight_{k}\phi(\textbf{y};\factorloads\scoremeans_k, \factorloads\scorecovs_k\factorloads\transpose + \eye\vec\psi)
+	f(\textbf{y}; \vec\Psi) = \sum_{k=1}^{K}\weight_{k}\phi(\textbf{y};\factorloads\scoremeans_k, \factorloads\scorecovs_k\factorloads\transpose + \eye\specificvariance)
 \end{equation}
@@
 vector
-$\vec\Psi$ includes $\{\factorloads,\vec\pi,\scoremeans,\scorecovs,\vec\psi\}$, and $\phi(\textbf{y};\vec\mu, \vec\Sigma)$
+$\vec\Psi$ includes $\{\factorloads,\vec\pi,\scoremeans,\scorecovs,\specificvariance\}$, and $\phi(\textbf{y};\vec\mu, \vec\Sigma)$
 describes the density of a multivariate gaussian distribution with
@@
 \begin{equation}
-	\tau_{nk} = \frac{\weight_k\phi(\vecdata_n;\factorloads\scoremeans_k, \factorloads\scorecovs_k\factorloads\transpose + \eye\vec\psi)}{\sum_{g=1}^{G}\weight_g\phi(\vecdata_n;\factorloads\scoremeans_g, \factorloads\scorecovs_g\factorloads\transpose + \eye\vec\psi)} \quad .
+	\tau_{nk} = \frac{\weight_k\phi(\vecdata_n;\factorloads\scoremeans_k, \factorloads\scorecovs_k\factorloads\transpose + \eye\specificvariance)}{\sum_{g=1}^{G}\weight_g\phi(\vecdata_n;\factorloads\scoremeans_g, \factorloads\scorecovs_g\factorloads\transpose + \eye\specificvariance)} \quad .
 \end{equation}
 
-% SVD complexity is O(n^3)
-% K-means++ complexity is O(log(k))
+% The complexity of SVD is O(n^3), and
+% the complexity of K-means++ is O(log(k)). If we knew the complexity of
+% k-means++ relative to N, then we would be able to make better decisions
+% as to how to initialise: SVD first and then K-means in latent space?
 
@@
 At the maximization step we update our estimates of the parameters,
-conditioned on the data $y$ and the responsibility matrix $\tau$.
-
+conditioned on the data $y$ and the responsibility matrix $\vec\tau$.
+Initially we update the relative weights $\vec\weight\nextstep$ given
+the responsibility matrix $\vec\tau$
 \begin{equation}
-	\weight_k\nextstep = \frac{1}{N} \sum_{n=1}^{N}\tau_{nk}
+	\weight_k\nextstep = \frac{1}{N} \sum_{n=1}^{N}\tau_{nk} \quad .
 	% Include proof.
@@
 
+%\noindent{}In the equations that follow (Equation~\ref{eq:X} to \ref{eq:Y})
+%when we refer to any value in $\{\factorloads,\vec\pi,\scoremeans,\scorecovs,\specificvariance\}$ for brevity we are referring to the current estimate of that
+%parameter (e.g., $\factorloads\thisstep$). The updated estimates are marked as such
+%(e.g., $\scoremeans\nextstep$).
+
+The updated estimates of the mean factor scores 
+$\scoremeans\nextstep$ are given by
 \begin{eqnarray}
-	\vec\mu_{k}\nextstep = \vec\mu_{k}\thisstep + \vec\gamma\transpose \frac{1}{N\weight_k}\sum_{n=1}^{N}{(\vecdata\transpose - \vec{}A\vec\mu_{k}\thisstep)\vec\tau_{k}}
+	\scoremeans_{k}\nextstep = \scoremeans_{k}\thisstep + \frac{\vec{G}\transpose(\vecdata\transpose - \factorloads\thisstep\scoremeans_{k}\thisstep)\vec\tau_{k}}{N\weight_k\nextstep}
+\end{eqnarray}
+\noindent{}where:
+\begin{eqnarray}
+	\vec{W} &=& (\scorecovs_{k}\thisstep)^{-1}\eye \\
+	\vec{V} &=& \left(\specificvariance\thisstep\right)^{-1} \\
+	\vec{C} &=& (\vec{W} + (\factorloads\thisstep)\transpose\vec{V}\factorloads\thisstep)^{-1}\eye \\
+	\vec{G} &=& \left[\vec{V} - \vec{V}\factorloads\thisstep\vec{C}\left(\vec{V}\factorloads\thisstep\right)\transpose\right]\factorloads\thisstep\scorecovs_k\thisstep \quad .
 \end{eqnarray}
 
-% The intermediate stpe for phi:
+The covariance matrices of the components of factor scores $\scorecovs\nextstep$
+are updated next,
 \begin{equation}
-	\vec\phi = N\sum_{k=1}^{K}\weight_k\left[\vec\Omega_k\nextstep + \vec\mu_k\nextstep(\vec\mu_k\nextstep)\transpose\right]
+	\scorecovs_k\nextstep = \left(\eye - \vec{G}\transpose\factorloads\thisstep\right)\scorecovs_k\thisstep + \frac{\vec{G}\transpose\vec{Z}\left(\vec{Z}\vec\tau_k\transpose\right)\transpose\vec{G}}{N\weight_k\nextstep}
+\end{equation}
+\noindent{}where
+\begin{eqnarray}
+	\vec{Z} &=& \vecdata\transpose - \factorloads\thisstep\scoremeans_k\thisstep \quad .
+\end{eqnarray}
+
+After some linear algebra, updated estimates of the common factor loads $\factorloads\thisstep$
+can be found from
+\begin{equation}
+	\factorloads\nextstep = \factorloads_{1}\left(\factorloads_{2}^{-1}\eye\right)
+\end{equation}
+\noindent{}where:
+\begin{eqnarray}
+	\factorloads_1 &=& \sum_{k=1}^{K}\left[ \vec\tau_k\transpose\vecdata\left(\scoremeans_k\thisstep\right)\transpose + 
+	\vec{G}\transpose\vec\tau_k\vec{Z}\transpose\vec{G}\right] \\
+	%\left(\scoremeans_k\thisstep\vecdata\transpose\vec\tau_k\right)\transpose \left(\vec{G}\transpose\vec{Z}\vec\tau_k\transpose\vec{G}\right)\transpose\right]  \\
+	\factorloads_2 &=& N\sum_{k=1}^{K}\left[\weight_k\nextstep\left(\scorecovs_k\nextstep + \scoremeans_k\nextstep\left(\scoremeans_k\nextstep\right)\transpose\right)\right]
+\end{eqnarray}
+
+Finally, the updated estimate of the specific variances $\specificvariance\nextstep$ is given
+by
+\begin{equation}
+	\specificvariance\nextstep = \frac{1}{N}\left[\sum^{K}_{k=1}\vec\tau_k\transpose\left(\vecdata\odot\vecdata\right) - \sum_{j=1}^{J}\left(\factorloads\nextstep\factorloads_2\right)\odot\factorloads\nextstep\right]
 \end{equation}
 
-% And the update of the psi.
-\begin{equation}
-	\vec\psi\nextstep = \frac{1}{N}\left[\sum_{k=1}^{K}\vec\data^2\vec\tau_{k} - (\vec{}A\nextstep \vec\phi)\vec{}A\right]
-\end{equation}
+\noindent{}where $\odot$ denotes is the entrywise (Hadamard) product.
 
@@
 \subsection{Toy model with generated data} \label{sec:experiment-toy-model}
+
+We generated a data set with $\todo{N = 100000}$ data points, each with
+$D = 15$ dimensions, and assumed that those data are generated by
+latent space of $J = 5$ ($J \times D$) factor loads, and there are $K = 20$
+clusters in the latent space. The relative weights $\vec\weight$
+are drawn from a multinomial distribution and the means of the clusters
+in factor scores $\scoremeans$ are drawn from a standard normal
+distribution.
+The covariance matrices in factor scores $\scorecovs$ are assumed to be \todo{low rank}, where the diagonal entries are drawn from a gamma distribution $\scorecovs_{k,i,i} \sim \Gamma(1)$.
+The variance in
+each dimension $\specificvariance$ are also drawn $\specificvariance \sim \Gamma(1)$.
+The $i$th data point (which belongs to the $k$th cluster) is then
+generated by drawing $\factorscores_{i} \sim \mathcal{N}(\scoremeans_k,\scorecovs_k)$, projecting by the factor loads, and adding variance.
+
+Latent factor models are by definition indeterminate in that there are
+multiple solutions of the parameter vector $\vec\Psi$ which provide
+an identical data-generating process. This can be trivially shown by
+the fact that the factors loads and scores can be rotated with some rotation
+matrix $\textbf{R}$ and generate the exact same data. As a consequence,
+we cannot strictly evaluate our model fitting procedure by comparing
+how closely the inferred factor loads (or scores) match the true 
+variables. This is further complicated by \todo{orientation and order}
+of the latent factors: even if the inferred factor loads were very close
+to the true factor loads, they could be ordered differently, or have
+their signs reversed. Given these issues, we evaluate the search as
+being successful if the log probability given the estimated model
+parameters is within convergence tolerance (or better) than the
+log probability given the data and the true parameters.
+
+We fit this generated data using the expectation-maximization algorithm
+as described in the previous Section. After \todo{X} iterations
+(and a serial processing time of \todo{X} seconds) we
+reached our convergence threshold: the log probability increased by
+less than \todo{X} with each iteration. For this experiment we specified
+the number of latent factors $J$ and the number of clustering components
+$K$ to include, but below we repeat this experiment where the number
+of latent factors and clustering components is not known.
+
+We show a corner plot of the data in Figure~\ref{fig:toy-model-data}.
+In this figure we have over-plotted \todo{100000} samples generated by the
+estimated factor loads $\factorloads_{est}$ and the clustering in factor 
+scores $\scoremeans_{est}$ and $\scorecovs_{est}$ (weighted by $\vec\weight_{est}$)
+-- without adding variance $\specificvariance$ -- in order to illustrate that the
+estimated model parameters can realistically generate the data.
+This is despite the estimated factor loads being slightly different
+than the true factor loads (Figure~\ref{fig:toy-model-factor-loads}),
+a consequence of the uniqueness of latent factor models. The estimated
+latent space of factor scores is shown in Figure~\ref{fig:toy-model-factor-scores}
+where the estimated clustering of latent scores is shown.
+% Despite not matching perfectly, it generates the data? property of uniqueness?
 
@@
 
-\section{Conclusion} \label{sec:conclusion}
+\section{Conclusions} \label{sec:conclusion}
+
+\acknowledgements 
+\todo{
+\begin{itemize}
+	\item ARC DP
+	\item GALAH acknowledgement
+	\item Gaia acknowledgement
+\end{itemize}
+}
 

### 2018-11-04 article/ms.tex
boom

@@
 \newcommand{\gaia}{\project{gaia}}
+\newcommand{\Galah}{\project{Galah}}
+\newcommand{\GALAH}{\project{GALAH}}
 \newcommand{\todo}[1]{\textcolor{red}{#1}}
@@
 
+% TODO: Is D a diagonal vector, or a matrix?! Make it consistent.
+
 % Describe how we will attempt to define a generative model for the data, from the latent space.
@@
 \todo{$n \times p$} matrix.
- We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \specificvariance\right)$
+ We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \eye\specificvariance\right)$
 is independent of the latent space, and $\specificvariance$ is a
@@
 
-We generated a data set with $\todo{N = 100000}$ data points, each with
+We generated a data set with ${N = 100000}$ data points, each with
 $D = 15$ dimensions, and assumed that those data are generated by
@@
 distribution.
-The covariance matrices in factor scores $\scorecovs$ are assumed to be \todo{low rank}, where the diagonal entries are drawn from a gamma distribution $\scorecovs_{k,i,i} \sim \Gamma(1)$.
+The covariance matrices in factor scores $\scorecovs$ are assumed to be diagonal matrices, where the non-zero entries are drawn from a gamma distribution $\scorecovs_{k,i,i} \sim \vec\Gamma\left(1\right)$.
 The variance in
-each dimension $\specificvariance$ are also drawn $\specificvariance \sim \Gamma(1)$.
+each dimension $\specificvariance$ are also drawn $\specificvariance \sim \vec\Gamma\left(1\right)$.
 The $i$th data point (which belongs to the $k$th cluster) is then
 generated by drawing $\factorscores_{i} \sim \mathcal{N}(\scoremeans_k,\scorecovs_k)$, projecting by the factor loads, and adding variance.
+
+We fit this generated data using the expectation-maximization algorithm
+as described in the previous Section. After 58 iterations
+(and a serial processing time of 40 seconds) we
+reached our convergence threshold: the log probability increased by
+less than $10^{-5}$ with each iteration. For this experiment we specified
+the number of latent factors $J$ and the number of clustering components
+$K$ to include, but below we repeat this experiment where the number
+of latent factors and clustering components is not known.
+
+We show a corner plot of the data in Figure~\ref{fig:toy-model-data}.
+In this figure we have over-plotted 100000 samples generated by the
+estimated factor loads $\factorloads_\textrm{est}$ and the clustering in factor 
+scores $\scoremeans_\textrm{est}$ and $\scorecovs_\textrm{est}$ (weighted by $\vec\weight_\textrm{est}$)
+-- without adding variance $\specificvariance_\textrm{est}$ % Ref Factor load figure 
+-- in order to illustrate that the
+estimated model parameters can realistically generate the data.
+The estimated latent space of factor scores is shown in 
+Figure~\ref{fig:toy-model-latent-space} where the colours indicate the
+most likely component that the data point is associated with.
 
@@
 how closely the inferred factor loads (or scores) match the true 
-variables. This is further complicated by \todo{orientation and order}
+variables. This is further complicated by orientation and order
 of the latent factors: even if the inferred factor loads were very close
@@
 
-We fit this generated data using the expectation-maximization algorithm
-as described in the previous Section. After \todo{X} iterations
-(and a serial processing time of \todo{X} seconds) we
-reached our convergence threshold: the log probability increased by
-less than \todo{X} with each iteration. For this experiment we specified
-the number of latent factors $J$ and the number of clustering components
-$K$ to include, but below we repeat this experiment where the number
-of latent factors and clustering components is not known.
+% Despite not matching perfectly, it generates the data? property of uniqueness?
+% Factor scores?
+% Some metric about clustering performance against true values?
 
-We show a corner plot of the data in Figure~\ref{fig:toy-model-data}.
-In this figure we have over-plotted \todo{100000} samples generated by the
-estimated factor loads $\factorloads_{est}$ and the clustering in factor 
-scores $\scoremeans_{est}$ and $\scorecovs_{est}$ (weighted by $\vec\weight_{est}$)
--- without adding variance $\specificvariance$ -- in order to illustrate that the
-estimated model parameters can realistically generate the data.
-This is despite the estimated factor loads being slightly different
-than the true factor loads (Figure~\ref{fig:toy-model-factor-loads}),
-a consequence of the uniqueness of latent factor models. The estimated
-latent space of factor scores is shown in Figure~\ref{fig:toy-model-factor-scores}
-where the estimated clustering of latent scores is shown.
-% Despite not matching perfectly, it generates the data? property of uniqueness?
+\begin{figure*}
+	\includegraphics[width=1.0\textwidth]{experiments/toy-model-data.png}
+    \caption{\todo{Caption}}
+    \label{fig:toy-model-data}
+\end{figure*}
+
+\begin{figure*}
+	\includegraphics[width=1.0\textwidth]{experiments/toy-model-latent-space.png}
+    \caption{\todo{Caption}}
+    \label{fig:toy-model-latent-space}
+\end{figure*}
+
+% Search strategy for toy model.
+
+% Figure: toy-model-factor-loads?
+% Figure: toy-model-specific-variances?
 
 \subsection{GALAH} \label{sec:experiment-galah}
+
+The \Galah\ survey is a stellar spectroscopic survey of up to
+$10^6$ stars. Up to 30 chemical abundances are expected to be measured
+and reported in the final \Galah\ data release.
+\todo{more about galah}
+
+We use the second \Galah\ data release \citep{Buder:2018a} which
+includes up to \todo{X} chemical abundances measured for \todo{Y}
+stars. We selected stars with \texttt{flag\_cannon = 0} to exclude
+stars where there is reason to suspect that the stellar parameters
+(e.g., $\teff$, $\logg$) are unreliable, and as a result the 
+detailed chemical abundances are untrustworthy. For this subset
+we examined the measured abundances of \todo{X, etc} and
+counted the number of stars with \texttt{flag\_x\_fe = 0}
+for each abundance. This flag indicates that there is no known
+reason to suspect that the measured chemical abundance is
+unreliable.
+
+We sorted the elemental abundances by the number of stars
+with reliable measurements, from highest to lowest. Starting
+with Fe, we iterated over each element and counted the number
+of stars with reliable measurements of that element and all
+other elements earlier in the list. We illustrate this process
+in Figure~\ref{fig:galah-abundance-flags}, where we show that
+if we include the top 15 elemental abundances, we would have
+\todo{X} stars with reliable measurements in all fifteen elements.
+This is relevant to the approach we use here, as the method
+described here requires no missing data.
+
+% Where do we go down to?
+% What elements do we add /remove.
+\todo{Where do we go down to? What elements do we exclude?}
+This leaves us with a sample of \todo{X} stars each with
+reliable stellar parameters ($\teff$, $\logg$) and 
+\todo{Y} (reliable) measured abundances: \todo{Mn, Fe}.
+These elements are produced from multiple nucleosynthetic
+pathways, including \todo{X}. 
+It is this subsample that we will use for experiments
+in chemical tagging.
+
+\subsubsection{Search strategy} \label{sec:experiment-galah-search}
+
+% Unknown number of latent factors and unknown number of latent scores
 
@@
 \acknowledgements
-\todo{
-\begin{itemize}
-	\item ARC DP
-	\item GALAH acknowledgement
-	\item Gaia acknowledgement
-\end{itemize}
-}
+A.~R.~C. is supported in part by Australian Research Council
+Discovery Project DP160100637.
+% Gaia acknowledgement?
+% GALAH acknowledgement?
 
@@
     \package{Stan} \citep{stan},
-    \package{CosmoHub} \citep{cosmohub},
-    \package{TensorFlow} \citep{tensorflow}
     \package{Jupyter Notebooks} \citep{jupyter-notebooks}

### 2018-11-04 article/experiments/toy.py

@@ -1,0 +1,166 @@
+
+
+import sys
+import numpy as np
+import matplotlib.pyplot as plt
+from time import time
+
+sys.path.insert(0, "../../")
+
+from mcfa import (mcfa, mpl_utils, utils)
+
+
+
+n_features = 15
+n_components = 20
+n_latent_factors = 5
+n_samples = 100000
+
+omega_scale = 1
+noise_scale = 1
+random_seed = 1
+
+X, truth = utils.generate_data(n_samples=n_samples,
+                               n_features=n_features,
+                               n_latent_factors=n_latent_factors,
+                               n_components=n_components,
+                               omega_scale=omega_scale,
+                               noise_scale=noise_scale,
+                               random_seed=random_seed)
+
+model = mcfa.MCFA(n_components=n_components, n_latent_factors=n_latent_factors)
+model.fit(X)
+
+# Now do a search!
+
+
+
+# Just check that the ML solution is at least as good as the generated value
+# (or at least within the convergence tolerance)
+"""
+diff = model.expectation(X, *utils.parameter_vector(**truth))[0] \
+     - model.expectation(X, *model.theta_)[0]
+assert diff <= model.tol
+"""
+
+pi, A, xi, omega, psi = model.theta_
+scores = model.factor_scores(X)[2]
+
+
+def common_limits(ax):
+    limits = np.array([ax.get_xlim(), ax.get_ylim()])
+    limits = (np.min(limits), np.max(limits))
+
+    ax.set_xlim(limits)
+    ax.set_ylim(limits)
+
+
+fig, axes = plt.subplots(n_latent_factors)
+
+for i, ax in enumerate(axes):
+    ax.plot(A.T[i], c="tab:blue")
+    ax.plot(truth["A"].T[i], c="tab:red")
+
+
+# Generate data.
+X2 = scores @ A.T
+
+def corner_scatter(X, label_names=None, show_ticks=False, fig=None, figsize=None,
+                   **kwargs):
+    """
+    Make a corner plot where the data are shown as scatter points in each axes.
+
+    :param X:
+        The data, :math:`X`, which is expected to be an array of shape
+        [n_samples, n_features].
+
+    :param label_names: [optional]
+        The label names to use for each feature.
+
+    :param show_ticks: [optional]
+        Show ticks on the axes.
+
+    :param fig: [optional]
[…]

### 2018-11-04 mcfa/mcfa.py

@@ -6,10 +6,13 @@ import numpy as np
 import warnings
 from copy import deepcopy
 from inspect import signature as inspect_signature
-from scipy import linalg
+from scipy import linalg, spatial
 from scipy.special import logsumexp
-from sklearn.cluster import KMeans
-from time import  time
+from sklearn import cluster
+from sklearn.utils import check_random_state
+from sklearn.utils.extmath import row_norms
+from time import time
+from tqdm import tqdm
 
 logger = logging.getLogger(__name__)
 
@@ -18,7 +21,7 @@ class MCFA(object):
 
     r""" A mixture of common factor analyzers model. """
 
-    def __init__(self, n_components, n_latent_factors, max_iter=500, n_init=25, 
+    def __init__(self, n_components, n_latent_factors, max_iter=500, n_init=5, 
                  tol=1e-5, verbose=1, random_seed=None, **kwargs):
         r"""
         A mixture of common factor analyzers model.
@@ -188,6 +191,7 @@ class MCFA(object):
                                          "initialisation point:")
 
         if best_theta is None:
+            raise exceptions[-1]
             raise ValueError("no initialisation point found")
 
         return best_theta
@@ -389,7 +393,8 @@ class MCFA(object):
             A2 += (omega[:, :, i] + xi[:, [i]] @ xi[:, [i]].T) * ti[i]
         
         A = A1 @ linalg.solve(A2, I_J)
-        Di = np.sum(self._X2.T @ tau[np.newaxis], axis=2)
+        Di = np.sum(self._X2.T @ tau, axis=1)
+
         psi = (Di - np.sum((A @ A2) * A, axis=1)) / N
         
         pi = ti / N
@@ -448,7 +453,8 @@ class MCFA(object):
         prev_ll, tau = self.expectation(X, *theta)
 
         # Run E-M.
-        for n_iter in range(self.max_iter):
+        tqdm_kwds = dict(desc="E-M", total=self.max_iter)
+        for n_iter in tqdm(range(self.max_iter), **tqdm_kwds):
 
             theta = self.maximization(X, tau, *theta)            
             ll, tau = self.expectation(X, *theta)
@@ -554,7 +560,7 @@ class MCFA(object):
         return X
 
 
-def _initial_assignments(X, n_components, n_kmeans_init):
+def _initial_assignments(X, n_components, n_kmeans_init, random_state=None):
     r"""
     Estimate the initial assignments of each sample to each component in the
     mixture.
[…]

### 2018-11-04 mcfa/mpl_utils.py

@@ -95,7 +95,7 @@ def plot_latent_space(model, X, ellipse_kwds=None, **kwargs):
     L = int(max(1, int(len(fig.axes)))**0.5)
     axes = np.atleast_2d(fig.axes).reshape((L, L)).T
 
-    kwds = dict(alpha=0.3, zorder=-1)
+    kwds = dict(alpha=0.3, zorder=-1, rasterized=True)
     kwds.update(ellipse_kwds or dict())
 
     scat = axes[0, 0].collections[0]

### 2018-11-04 mcfa/utils.py

@@ -29,7 +29,7 @@ def generate_data(n_samples=20, n_features=5, n_latent_factors=3, n_components=2
 
     psi = rng.gamma(1, scale=noise_scale, size=n_features)
 
-    noise = psi * rng.randn(n_samples, n_features)
+    noise = np.sqrt(psi) * rng.randn(n_samples, n_features)
 
     X = scores @ A.T + noise
 

### 2018-11-05 article/ms.tex
add figure

@@
 other elements earlier in the list. We illustrate this process
-in Figure~\ref{fig:galah-abundance-flags}, where we show that
+in Figure~\ref{fig:galah-abundances}, where we show that
 if we include the top 15 elemental abundances, we would have
@@
 described here requires no missing data.
+
+\begin{figure*}
+	\includegraphics[width=1.0\textwidth]{experiments/galah-abundances.png}
+    \caption{\todo{Caption}}
+    \label{fig:galah-abundances}
+\end{figure*}
 
@@
 
-\subsubsection{Search strategy} \label{sec:experiment-galah-search}
+%\subsubsection{Search strategy} \label{sec:experiment-galah-search}
+
+The expectation-maximization procedure outlined in Section \ref{sec:methods} describes 
+the process of refining our estimate of the model parameters $\vec\psi$ until
+the improvement in log probability between iterations decreases below some
+pre-defined threshold. This method is valid for a pre-defined number of latent
+factors, and a pre-defined number of components to cluster in latent space.
+In practice these values are not known. For the purposes of this work, we 
+perform a grid-based search in $J$ and $K$, perform E-M for each combination of
+$J$ and $K$, and record the optimal parameters.
+
+We trial the number of latent factors from $J = 1$ to $J = 10$, and trial the
+number of clustering components from $K = 1$ to $K = 100$. We re-initialise the
+mixture for each new trial of $\{J, K\}$, and do not make use of results from
+`nearby' mixtures.
+
+% Run this.
 

### 2018-11-05 article/experiments/galah-abundances.py

@@ -1,0 +1,81 @@
+
+
+
+import sys
+import numpy as np
+import matplotlib.pyplot as plt
+from time import time
+import operator
+
+sys.path.insert(0, "../../")
+from astropy.table import Table
+
+
+from mcfa import mpl_utils
+
+matplotlib.style.use(mpl_utils.mpl_style)
+
+
+galah = Table.read("../../catalogs/GALAH_DR2.1_catalog.fits")
+
+
+element_label_names = [ln for ln in galah.dtype.names \
+    if ln.endswith(("_fe", "fe_h")) and not ln.startswith(("flag_", "e_", "alpha_"))]
+
+# Select a subset of stars and abundances
+is_ok = dict()
+for label_name in element_label_names:
+    flag_name = "flag_cannon" if label_name == "fe_h" else f"flag_{label_name}"
+    is_ok[label_name] = sum((galah[flag_name] == 0) * (galah["flag_cannon"] == 0))
+
+x = np.arange(len(element_label_names))
+y = np.zeros_like(x)
+y_cumulative = np.zeros_like(x)
+
+sorted_label_names = []
+
+sorted_items = sorted(is_ok.items(), key=operator.itemgetter(1))[::-1]
+
+for i, (k, v) in enumerate(sorted_items):
+    y[i] = v
+    sorted_label_names.append(k)
+
+    # Do cumulative count.
+    mask = (galah["flag_cannon"] == 0)
+    for k2, _ in sorted_items[:i + 1]:
+        if k2 != "fe_h":
+            mask *= (galah[f"flag_{k2}"] == 0)
+
+    y_cumulative[i] = np.sum(mask)
+
+
+fig, ax = plt.subplots()
+ax.plot(x, y, "-", drawstyle="steps-mid", c="#000000", lw=2,
+        label=r"$\textrm{Independent}$")
+ax.plot(x, y_cumulative, "-", c="tab:blue", drawstyle="steps-mid", lw=2, zorder=10,
+        label=r"$\textrm{Cumulative}$")
+ax.set_xticks(x)
+ax.set_xticklabels([ln.split("_")[0].title() for ln in sorted_label_names])
+ax.set_xlim(-0.5, x.max() + 0.5)
+
+ax.semilogy()
+ax.set_ylim(10**3, ax.get_ylim()[1])
[…]

### 2018-11-13 mcfa/mcfa.py
better options to initialise

@@ -18,13 +18,15 @@ from tqdm import tqdm
 logger = logging.getLogger(__name__)
 
 
+
+
 class MCFA(object):
 
     r""" A mixture of common factor analyzers model. """
 
     def __init__(self, n_components, n_latent_factors, max_iter=10000, tol=1e-5,
-                 init_method="random", n_init=5, verbose=1, random_seed=None, 
-                 **kwargs):
+                 init_components="kmeans++", init_factors="svd", verbose=1, 
+                 random_seed=None, **kwargs):
         r"""
         A mixture of common factor analyzers model.
 
@@ -40,13 +42,20 @@ class MCFA(object):
         :param tol: [optional]
             Relative tolerance before declaring convergence.
 
-        :init_method: [optional]
-            The initialisation method to use. Available options are: 'random' or
-            'svd+kmeans++'.
+        :param init_components: [optional]
+            The iniitialisation method to use when assigning data points to
+            components. Available options include: 'kmeans++', 'random', or a
+            callable function that takes three arguments: the data, an initial
+            estimate of the latent factors, and the number of components. This
+            function should return three quantities: an array of the relative
+            weights, the means in factor scores, the covariance matrices in
+            factor scores.
 
-        :param n_init: [optional]
-            The number of initialisations to run if k-means++ is used during
-            initialisation.
+        :param init_factors: [optional]
+            The initialisation method to use for the latent factors. Available
+            options include: 'svd', 'random', or a callable functiomn that takes
+            the input data as a single argument and returns a matrix of latent
+            factors.
 
         :param verbose: [optional]
             Show warning messages.
@@ -58,10 +67,11 @@ class MCFA(object):
         self.n_components = int(n_components)
         self.n_latent_factors = int(n_latent_factors)
         self.max_iter = int(max_iter)
-        self.n_init = int(n_init)
         self.tol = float(tol)
         self.random_seed = random_seed
-        self.init_method = str(init_method).lower().strip()
+        self.init_components = init_components
+        self.init_factors = init_factors
+        self.verbose = int(verbose)
 
         if self.n_latent_factors < 1:
             raise ValueError("n_latent_factors must be a positive integer")
@@ -72,17 +82,22 @@ class MCFA(object):
[…]

### 2018-11-13 mcfa/utils.py

@@ -2,6 +2,10 @@
 import numpy as np
 
 
+def whiten(X, axis=0):
+    return (X - np.mean(X, axis=axis))/np.std(X, axis=axis)
+
+
 
 def generate_data(n_samples=20, n_features=5, n_latent_factors=3, n_components=2,
                   omega_scale=1, noise_scale=1, random_seed=0):
@@ -129,7 +133,7 @@ def simulate_example_data(N, D, J, K=1, seed=None, full_output=False, **kwargs):
 
 
 
-def latent_factor_rotation_matrix(A_prime, A):
+def rotation_matrix(A_prime, A):
     r"""
     Return a rotation amtrix :math:`\mathbf{R}` that will orient and flip the 
     latent factors `A_prime` to be as close as possible to `A` such that
@@ -233,7 +237,7 @@ if __name__ == "__main__":
     indices = np.random.choice(J, J, replace=False)
     A_star = A_star[:, indices]
 
-    R = latent_factor_rotation_matrix(A_star, A)
+    R = rotation_matrix(A_star, A)
 
     A_prime = A_star @ R
 

### 2018-11-13 article/experiments/galah-experiment-1-gridsearch.py
add experiment scripts

@@ -1,0 +1,400 @@
+
+import os
+import sys
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import pickle
+
+sys.path.insert(0, "../../")
+
+
+with open("galah-experiment-1.pkl", "rb") as fp:
+    X, label_names, mask = pickle.load(fp)
+
+from mcfa import mcfa, mpl_utils
+
+matplotlib.style.use(mpl_utils.mpl_style)
+
+
+mcfa_kwds = dict()
+
+get_output_path = lambda J, K: f"1/galah-experiment-1.{J:.0f}.{K:.0f}.pkl"
+
+def savefig(fig, suffix):
+    prefix = os.path.basename(__file__)[:-3]
+    filename = f"{prefix}-{suffix}.png"
+    fig.savefig(filename, dpi=150)
+
+
+
+
+# Run grid search.
+OVERWRITE = False
+MAX_NUM_LATENT_FACTORS = 12
+MAX_NUM_COMPONENTS = 40
+PLOT_ONLY = True
+
+if not PLOT_ONLY:
+
+    for n_components in range(1, 1 + MAX_NUM_COMPONENTS):
+        for n_latent_factors in range(1, 1 + MAX_NUM_LATENT_FACTORS):
+
+            print(f"K = {n_components} (of {MAX_NUM_COMPONENTS}); J = {n_latent_factors} (of {MAX_NUM_LATENT_FACTORS})")
+
+            output_path = get_output_path(n_latent_factors, n_components)
+
+            kwds = mcfa_kwds.copy()
+
+            if os.path.exists(output_path):
+                with open(output_path, "rb") as fp:
+                    model = pickle.load(fp)
+
+                if (model.n_iter_ + 1) < model.max_iter and not OVERWRITE:
+                    print(f"Skipping because {output_path} exists")
+                    continue
+
+                else:
+                    kwds.update(max_iter=model.max_iter * 10)
+
+            else:
+                model = mcfa.MCFA(n_components, n_latent_factors, **mcfa_kwds)
+
+
+            try:
+                model.fit(X)
+
+            except:
+                print("Failed")
+                continue
+
+            else:
+
+                print(f"log-likelihood: {model.log_likelihood_}")
+
[…]

### 2018-11-13 article/experiments/galah-experiment-1.py

@@ -1,0 +1,114 @@
+
+
+
+import sys
+import numpy as np
+import matplotlib
+
+import pickle
+import operator
+
+sys.path.insert(0, "../../")
+from astropy.table import Table
+
+
+REQUIRE_ELEMENTS = ["fe", "eu", "ba"]
+EXCLUDE_LABEL_NAMES = ["o_fe"]
+
+galah = Table.read("../../catalogs/GALAH_DR2.1_catalog.fits")
+
+
+all_element_label_names = [ln for ln in galah.dtype.names \
+    if ln.endswith(("_fe", "fe_h")) and not ln.startswith(("flag_", "e_", "alpha_"))]
+
+
+# Select a subset of stars and abundances
+common_mask = (galah["flag_cannon"] == 0)
+for element in REQUIRE_ELEMENTS:
+    if element != "fe":
+        common_mask *= (galah[f"flag_{element}_fe"] == 0)
+
+
+is_ok = dict()
+for label_name in set(all_element_label_names).difference(EXCLUDE_LABEL_NAMES):
+    flag_name = "flag_cannon" if label_name == "fe_h" else f"flag_{label_name}"
+    is_ok[label_name] = sum((galah[flag_name] == 0) * common_mask)
+
+sorted_items = sorted(is_ok.items(), key=operator.itemgetter(1))[::-1]
+
+
+# Cumulative
+
+x = np.arange(len(all_element_label_names))
+y = np.zeros_like(x)
+y_cumulative = np.zeros_like(x)
+
+sorted_label_names = []
+
+for i, (k, v) in enumerate(sorted_items):
+    y[i] = v
+    sorted_label_names.append(k)
+
+    # Do cumulative count.
+    mask = (galah["flag_cannon"] == 0)
+    for k2, _ in sorted_items[:i + 1]:
+        if k2 != "fe_h":
+            mask *= (galah[f"flag_{k2}"] == 0)
+
+    y_cumulative[i] = np.sum(mask)
+
+
+D = 14
+
+# Generate data array with all things relative to [X/H]
+mask = (galah["flag_cannon"] == 0)
+for ln, v in sorted_items[:D]:
+
+    flag_label = f"flag_{ln}"
+    if flag_label in galah.dtype.names:
+        mask *= (galah[flag_label] == 0)
+
+N = sum(mask)
+print(f"There are {N} data points (D = {D})")
+
+# order by atomic number.
+
[…]

### 2018-11-13 article/experiments/galah-experiment-2-gridsearch.py

@@ -1,0 +1,487 @@
+
+import os
+import sys
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import pickle
+
+from matplotlib.ticker import MaxNLocator
+
+sys.path.insert(0, "../../")
+
+
+with open("galah-experiment-1.pkl", "rb") as fp:
+    X, label_names, mask = pickle.load(fp)
+
+
+print(f"Labels {len(label_names)}: {label_names}")
+from mcfa import mcfa, mpl_utils, utils
+
+matplotlib.style.use(mpl_utils.mpl_style)
+
+
+elements = [ea.split("_")[0].title() for ea in label_names]
+latex_elements = [r"$\textrm{{{0}}}$".format(el) for el in elements]
+
+
+fig, axes = plt.subplots(7, 2)
+axes = np.array(axes).flatten()
+idx = 0
+feh_index = label_names.index("fe_h")
+for i, label_name in enumerate(label_names):
+    if label_name == "fe_h": continue
+
+    ax = axes[idx]
+    ax.scatter(X.T[feh_index], X.T[i], s=1)
+    ax.set_ylabel(latex_elements[i])
+
+    idx += 1
+
+for ax in axes[i:]:
+    ax.set_visible(False)
+
+fig.tight_layout()
+
+
+
+mcfa_kwds = dict(tol=1e-5, init_method="random")
+
+get_output_path = lambda J, K: f"2/galah-experiment-1.{J:.0f}.{K:.0f}.pkl"
+
+def savefig(fig, suffix):
+    prefix = os.path.basename(__file__)[:-3]
+    filename = f"{prefix}-{suffix}.png"
+    fig.savefig(filename, dpi=150)
+
+
+X_whitened = utils.whiten(X)
+
+# Run grid search.
+OVERWRITE = False
+MAX_NUM_LATENT_FACTORS = 12
+MAX_NUM_COMPONENTS = 40
+PLOT_ONLY = True
+
+if not PLOT_ONLY:
+
+    for n_components in range(1, 1 + MAX_NUM_COMPONENTS):
+        for n_latent_factors in range(1, 1 + MAX_NUM_LATENT_FACTORS):
+
+            print(f"K = {n_components} (of {MAX_NUM_COMPONENTS}); J = {n_latent_factors} (of {MAX_NUM_LATENT_FACTORS})")
+
+            output_path = get_output_path(n_latent_factors, n_components)
+
+            kwds = mcfa_kwds.copy()
+
+            if os.path.exists(output_path):
[…]

### 2018-11-13 article/experiments/galah-experiment-3.py

@@ -1,0 +1,96 @@
+
+import os
+import sys
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import pickle
+
+from matplotlib.ticker import MaxNLocator
+
+sys.path.insert(0, "../../")
+
+
+with open("galah-experiment-1.pkl", "rb") as fp:
+    X_H, label_names, mask = pickle.load(fp)
+
+
+X_Fe = np.copy(X_H)
+
+feh_index = label_names.index("fe_h")
+for i, label_name in enumerate(label_names):
+    if i == feh_index: continue
+    X_Fe[:, i] = X_H[:, i] - X_H[:, feh_index]
+
+
+
+
+from mcfa import mcfa, mpl_utils, utils
+
+matplotlib.style.use(mpl_utils.mpl_style)
+
+J = 5
+K = 1
+
+colors = plt.rcParams['axes.prop_cycle'].by_key()['color']
+
+
+whitened_X_H = utils.whiten(X_H)
+
+
+model_h = mcfa.MCFA(init_method="svd+kmeans++", n_latent_factors=J, n_components=K, tol=1e-8)
+model_h.fit(X_H)
+
+
+"""
+model_fe = mcfa.MCFA(n_latent_factors=J, n_components=K, tol=1e-8)
+model_fe.fit(X_Fe)
+
+pi, A_h, xi, omega, psi = model_h.theta_
+A_fe = model_fe.theta_[1]
+
+
+R = utils.rotation_matrix(A_fe, A_h)
+
+A_X_FeR = A_fe @ R
+
+
+D, J = A_fe.shape
+
+# OK, let's look at the factor loads.
+xi = np.arange(D)
+fig, ax = plt.subplots()
+
+for j in range(J):
+    ax.plot(xi, A_h.T[j], "-", c=colors[j])
+    ax.plot(xi, A_X_FeR.T[j], "-", c=colors[j], alpha=0.5)
+
+raise a
+"""
+
+A = model_h.theta_[1]
+
+N, D = X_H.shape
+xi = np.arange(D)
+fig, ax = plt.subplots()
+
+for j in range(J):
+    ax.plot(xi, A.T[j], "-", c=colors[j])
+
+
+
+elements = [ea.split("_")[0].title() for ea in label_names]
+latex_elements = [r"$\textrm{{{0}}}$".format(el) for el in elements]
+
+ax.set_xticks(xi)
[…]

### 2018-11-13 article/experiments/galah-experiment-4.py

@@ -1,0 +1,246 @@
+
+
+
+import sys
+import os
+import numpy as np
+import pickle
+
+import matplotlib.pyplot as plt
+from matplotlib.ticker import MaxNLocator
+
+sys.path.insert(0, "../../")
+from astropy.table import Table
+
+
+
+galah = Table.read("../../catalogs/GALAH_DR2.1_catalog.fits")
+
+
+all_element_label_names = [ln for ln in galah.dtype.names \
+    if ln.endswith(("_fe", "fe_h")) and not ln.startswith(("flag_", "e_", "alpha_"))]
+
+
+
+def savefig(fig, suffix):
+    prefix = os.path.basename(__file__)[:-3]
+    filename = f"{prefix}-{suffix}.png"
+    fig.savefig(filename, dpi=150)
+
+
+elements = [
+    "fe",
+    "eu",
+    "ba", "y",
+    "mg", "ca",
+    "na",
+    "ni", "cr",
+]
+
+
+# Select a subset of stars and abundances
+common_mask = (galah["flag_cannon"] == 0)
+for element in elements:
+    if element != "fe":
+        common_mask *= (galah[f"flag_{element}_fe"] == 0)
+
+
+# order by atomic number.
+
+periodic_table = """H                                                  He
+                    Li Be                               B  C  N  O  F  Ne
+                    Na Mg                               Al Si P  S  Cl Ar
+                    K  Ca Sc Ti V  Cr Mn Fe Co Ni Cu Zn Ga Ge As Se Br Kr
+                    Rb Sr Y  Zr Nb Mo Tc Ru Rh Pd Ag Cd In Sn Sb Te I  Xe
+                    Cs Ba Lu Hf Ta W  Re Os Ir Pt Au Hg Tl Pb Bi Po At Rn
+                    Fr Ra Lr Rf"""
+
+lanthanoids    =   "La Ce Pr Nd Pm Sm Eu Gd Tb Dy Ho Er Tm Yb"
+actinoids      =   "Ac Th Pa U  Np Pu Am Cm Bk Cf Es Fm Md No"
+
+periodic_table = periodic_table.replace(" Ba ", " Ba " + lanthanoids + " ") \
+                               .replace(" Ra ", " Ra " + actinoids + " ") \
+                               .split()
+
+
+
+idxs = [periodic_table.index(k.split("_")[0].title()) for k in elements]
+sorted_elements = [elements[idx] for idx in np.argsort(idxs)]
+
[…]

### 2018-11-13 article/ms.tex
words

@@
 
-\shorttitle{MCFA}
+\shorttitle{Statistically consistent chemical tagging}
 \shortauthors{Casey et al.}
@@
 
-\title{MCFA on Galah}
+\title{A statistically consistent approach to chemical tagging}
 
@@
 \begin{abstract}
-foo
 \end{abstract}
@@
 \section{Introduction} \label{sec:intro}
+
+\begin{itemize}
+	\item Chemical tagging
+	\item Limits on the efficacy of chemical tagging
+	\item Clustering vs common traits
+\end{itemize}
+
+The \Galah\ survey \citep{DaSilva:2015} is a stellar spectroscopic survey 
+that uses the High Efficiency and Resolution Multi-Element 
+Spectrograph \citep[HERMES][]{Sheinis:2016} on the Australian Astronomical Telescope (AAT).
+The survey will observe up to $10^6$ stars in the Milky Way, and
+measure up to 30 chemical abundances for each star. This includes
+light odd-Z elements (e.g., Na, K), elements produced through
+alpha-particle capture (e.g., Mg, Ca, Ti), and elements produced
+through the slow (e.g., Ba) and rapid neutron-capture process
+(e.g., Eu). These data provide an unparalleled view on the production
+of chemical elements in the Milky Way.
 
@@
 
+We initialise the entries in the factor loads $\factorloads$ by drawing
+from a uniform distribution $\factorloads \sim \mathcal{U}\left(-1, +1\right)$.
+We assign each data point as belonging to one of the $K$ components
+by drawing from a multinomial distribution. Given these assignments and
+factor loads, we then estimate the relative weights $\vec\pi$, the 
+mean factor scores of each component $\scoremeans$, and the covariance
+matrix of factor scores of each component $\scorecovs$.
+
+\begin{comment}
 We use the K-means++ algorithm to initially assign data points
@@
 \end{equation}
+\end{comment}
 
@@
 
-\subsection{Toy model with generated data} \label{sec:experiment-toy-model}
+\subsection{Experiment 1: Toy model with generated data} \label{sec:experiment-toy-model}
 
@@
 
-\subsection{GALAH} \label{sec:experiment-galah}
-
-The \Galah\ survey is a stellar spectroscopic survey of up to
-$10^6$ stars. Up to 30 chemical abundances are expected to be measured
-and reported in the final \Galah\ data release.
-\todo{more about galah}
+\subsection{Experiment 2: A subset of 14 chemical abundances in \Galah}
+\label{sec:experiment-galah}
 
 We use the second \Galah\ data release \citep{Buder:2018a} which
-includes up to \todo{X} chemical abundances measured for \todo{Y}
-stars. We selected stars with \texttt{flag\_cannon = 0} to exclude
+includes up to 23 chemical abundances measured for 342,682
+stars. It is not practical to use all of these measurements
+when performing chemical tagging. For example, our experiments
+with \Galah\ data do not include lithium abundances because the
+photospheric lithium varies throughout a star's lifetime. 
+We first selected stars with \texttt{flag\_cannon = 0} to exclude
 stars where there is reason to suspect that the stellar parameters
 (e.g., $\teff$, $\logg$) are unreliable, and as a result the 
-detailed chemical abundances are untrustworthy. For this subset
-we examined the measured abundances of \todo{X, etc} and
+detailed chemical abundances are untrustworthy.
+
+We included the following fourteen chemical abundances: Na, Mg, 
+Si, Ca, Sc, Ti, Cr, Mn, Fe, Ni, Cu, Zn, Ba, and Eu.
+By requiring that each star has \texttt{flag\_cannon = 0} and
+\texttt{flag\_<x> = 0} (where \texttt{<x>} is the relevant label)
+for each chemical abundance, this leaves
+us with a sample of 12,798 stars each with reliable stellar
+parameters ($\teff$, $\logg$) and chemical abundances.
+This is relevant to the approach we use here, as the method
+described here requires no missing data.
+These elements are produced from multiple nucleosynthetic
+pathways, and it is this subsample that we will use for 
+our initial chemical experiment with \Galah\ data.
+
+\begin{comment}
+For this subset
+we examined the measured abundances and
 counted the number of stars with \texttt{flag\_x\_fe = 0}
@@
 in Figure~\ref{fig:galah-abundances}, where we show that
-if we include the top 15 elemental abundances, we would have
-\todo{X} stars with reliable measurements in all fifteen elements.
+if we include the top 14 elemental abundances, we would have
+12,798 stars with reliable measurements in all fourteen elements.
 This is relevant to the approach we use here, as the method
 described here requires no missing data.
+\end{comment}
 
@@
 
-% Where do we go down to?
-% What elements do we add /remove.
-\todo{Where do we go down to? What elements do we exclude?}
-This leaves us with a sample of \todo{X} stars each with
-reliable stellar parameters ($\teff$, $\logg$) and 
+\begin{comment}
+This leaves us with a sample of 12,798 stars each with
+reliable stellar parameters ($\teff$, $\logg$) and fourteen
+(reliably) measured abundances: Na, Mg, Si, Ca, Sc, Ti, Cr, Mn, 
+Fe, Ni, Cu, Zn, Ba, and Eu
 \todo{Y} (reliable) measured abundances: \todo{Mn, Fe}.
@@
 in chemical tagging.
+\end{comment}
 
@@
 
-We trial the number of latent factors from $J = 1$ to $J = 10$, and trial the
-number of clustering components from $K = 1$ to $K = 100$. We re-initialise the
+We trial the number of latent factors from $J = 1$ to $J = 12$, and trial the
+number of clustering components from $K = 1$ to $K = 40$. We re-initialise the
 mixture for each new trial of $\{J, K\}$, and do not make use of results from
-`nearby' mixtures.
+`nearby' mixtures. After the E-M threshold of $10^{-5}$ has been reached, we
+record the log likelihood of the data given the model parameters, the
+Bayesian Information Criterion \citep{bic}, 
+\begin{equation}
+	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \label{eq:bic}
+\end{equation} 
+\noindent{}and the pseudo-Bayesian Information Criterion \citep{pseudo-bic}
+\begin{equation}
+ \textrm{pseudo-BIC} = 6(1 + \gamma)\omega{}Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \label{eq:pseudo-bic}
+\end{equation}
+\noindent{}where $Q$ is the total number of model parameters in both equations,
+and the unknown constants $\omega$ and $\gamma$ have constraints
+$\omega \geq 1$ and $\gamma > 0$. Here we adopt $\gamma = 0.1$ and $\omega = 1$.
 
-% Run this.
+\todo{Discuss grid search, inferred factor loads, cluster variances, etc.}
 
-% Unknown number of latent factors and unknown number of latent scores
+\begin{figure}
+	\includegraphics[width=0.45\textwidth]{experiments/galah-experiment-2-gridsearch-ll-contours.png}
+	\includegraphics[width=0.45\textwidth]{experiments/galah-experiment-2-gridsearch-bic-contours.png}
+	\includegraphics[width=0.45\textwidth]{experiments/galah-experiment-2-gridsearch-pseudobic-contours.png}
+    \caption{Model performance metrics for Experiment 2 using a subset of 14 
+    		 \Galah\ chemical abundances measured from 12,798 stars. The top 
+		 	 panel shows the negative log-likelihood 
+			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
+			 evaluated at each combination of latent factors $J$ and number 
+			 of clusters $K$, the middle panel shows the Bayesian Information
+			 Criterion (Equation \ref{eq:bic}), and the lower panel shows
+			 the pseudo-Bayesian Information Criterion 
+			 (Equation~\ref{eq:pseudo-bic}). The white marker indicates the 
+			 lowest value in each panel.}
+    \label{fig:experiment-2-gridsearch}
+\end{figure}
+
+\begin{figure*}
+	\includegraphics[width=1.0\textwidth]{experiments/galah-experiment-1-gridsearch-factors-wrt-J-and-K.png}
+	\caption{Latent factor loads inferred for models with increasing numbers
+			 of latent factors and increasing numbers of cluster components
+			 in Experiment~2. The latent factors have each been rotated by a
+			 rotation matrix $\textbf{R}$ with entries $\{-1, 0, +1\}$ such 
+			 that the estimated latent factors in each model are ordered 
+			 and oriented to be as similar as the latent factors estimated from
+			 the model with only one latent factor. Note that the rotation
+			 matrix contains only entries $\{-1, 0, +1\}$. The latent
+			 factors for each model shown here were initialised using singular value
+			 decomposition on a random subset of the data.
+			 }
+	\label{fig:experiment-2-gridsearch-factors-init-svd}
+\end{figure*}
+
+\begin{figure*}
+	\includegraphics[width=1.0\textwidth]{experiments/galah-experiment-2-gridsearch-factors-wrt-J-and-K.png}
+	\caption{Latent factor loads inferred for models with increasing numbers
+			 of latent factors and increasing numbers of cluster components
+			 in Experiment~2. The latent factors have each been rotated by a
+			 rotation matrix $\textbf{R}$ with entries $\{-1, 0, +1\}$ such 
+			 that the estimated latent factors in each model are ordered 
+			 and oriented to be as similar as the latent factors estimated from
+			 the model with only one latent factor. Note that the rotation
+			 matrix contains only entries $\{-1, 0, +1\}$. The latent
+			 factors for each model shown here were initialised randomly.
+			 }
+	\label{fig:experiment-2-gridsearch-factors-init-svd}
+\end{figure*}
+
+\subsection{Experiment 4: A minimal set of nucleosynthetic trace elements in \Galah}
+
+In this experiment we investigate the effects of initialising the model parameters.
+Here we make use of a minimal set of nucleosynthetic trace elements
+using the second \Galah\ data release \citep{Buder:2018}. Specifically, we use
+reliable chemical abundance measurements of Na (an odd-Z element), Mg and Ca (alpha-elements),
+Ni, Fe, and Cr (iron-peak elements), Ba and Y (s-process tracers), and Eu (an r-process tracer).
+We assume that there are four latent factors ($J = 4$) and four components ($K = 4$),
+and we initialise the latent factors and component parameters in different combinations.
+The latent factors can be initialised in four different ways: (1) with entries drawn randomly from $\mathcal{U}\left(-1, +1\right)$, (2) with noisey draws $\mathcal{N}\left(0, 10^{-2}\right)$, (3) using singular value decomposition, and (4) using astrophysical insight to initialise the factor loads.
+\todo{When using astrophysical insight...}
+The component parameters are either initialised randomly, or using K-means++ algorithm in
+the latent space. For each combination of initialisation procedure, we run 30 optimisations
+and record the log likelihood once the E-M procedure has reached a threshold of $10^{-8}$.
+
+The results of this experiment are shown in Figure~\ref{fig:experiment-4-log-L}.
 

### 2018-11-16 article/ms.tex
foo

@@
 \email{andrew.casey@monash.edu}
+
+\author{MML people and GALAH people}
 
@@
 
+\begin{figure*}
+	\includegraphics[width=1.0\textwidth]{experiments/galah-experiment-4-log-L-hist.png}
+	\caption{Log likelihoods $\mathcal{L}\left(\data|\vec\Psi\right)$ of optimised
+			 models using \Galah\ data in Experiment 4 with four latent factors
+			 and four components ($J = 4$, $K = 4$). The columns indicate different
+			 methods to initialise the component parameters ($\pi$, $\scoremeans$,
+			 $\scorecovs$), and the rows represent different methods to initialise
+			 the latent factors ($\factorloads$). Each combination of initialisation
+			 procedure has 30 models where E-M was run until a threshold of $10^{-8}$
+			 was reached in $\log{L}$.}
+	\label{fig:galah-experiment-4-log-L}
+\end{figure*}
+
 \section{Results} \label{sec:results}

### 2018-11-16 article/ms.tex
fix tests and add figures from toy model grid search

@@
 
+\todo{Grid search text}
+
+\begin{figure}
+	\includegraphics[width=0.45\textwidth]{experiments/toy-ll-contours.png}
+	\includegraphics[width=0.45\textwidth]{experiments/toy-bic-contours.png}
+	\includegraphics[width=0.45\textwidth]{experiments/toy-pseudobic-contours.png}
+    \caption{Model performance metrics resulting from a gridsearch using the
+    		 data generated as part of our toy model. The top 
+		 	 panel shows the negative log-likelihood 
+			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
+			 evaluated at each combination of latent factors $J$ and number 
+			 of clusters $K$, the middle panel shows the Bayesian Information
+			 Criterion (Equation \ref{eq:bic}), and the lower panel shows
+			 the pseudo-Bayesian Information Criterion 
+			 (Equation~\ref{eq:pseudo-bic}). The white marker indicates the 
+			 lowest value in each panel.}
+    \label{fig:experiment-1-gridsearch}
+\end{figure}
+
 % Search strategy for toy model.

### 2018-11-16 tests/test_utils.py

@@ -28,7 +28,7 @@ def test_latent_factor_rotation(D=15, J=10, noise=0.05):
     true_indices = np.random.choice(J, J, replace=False)
     A_prime = A_prime[:, true_indices]
 
-    R = utils.latent_factor_rotation_matrix(A_prime, A)
+    R = utils.rotation_matrix(A_prime, A)
 
     # Check order.
     nonzero = (R != 0)
@@ -41,7 +41,7 @@ def test_latent_factor_rotation(D=15, J=10, noise=0.05):
     assert np.alltrue(true_signs == inferred_signs)
 
 
-def test_latent_factor_rotation_many_times(N=1000, D=15, J=10, noise=0.05):
+def test_latent_factor_rotation_many_times(N=100, D=15, J=10, noise=0.05):
     
     for i in range(N):
         test_latent_factor_rotation(D=D, J=J, noise=noise)

### 2018-11-21 article/ms.tex
add words

@@
 
+\newcommand{\NumData}{N}
+\newcommand{\NumDimensions}{D}
+\newcommand{\numdata}{n}
+\newcommand{\numdimensions}{d}
+\newcommand{\NumLatentFactors}{J}
+\newcommand{\numlatentfactors}{j}
+\newcommand{\NumComponents}{K}
+\newcommand{\numcomponents}{k}
+
 \received{2018 XX XX}
@@
 \email{andrew.casey@monash.edu}
-
-\author{MML people and GALAH people}
 
@@
 
-In this work the data $\vecdata$ is a $n \times d$ matrix 
-where $n$ is the number of stars and $d$ is the number of
+In this work the data $\vecdata$ is a $\NumData \times \NumDimensions$ matrix 
+where $\NumData$ is the number of stars and $\NumDimensions$ is the number of
 chemical abundances measured for each star. We assume a
@@
 
-\noindent{}where $\factorloads$ is a $p \times d$ matrix of factor
+\noindent{}where $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$ matrix of factor
 loads that is common to all data points, and the factor scores for
-the $i$th data point
+the $\numdata$th data point
 \begin{equation}
-	\factorscores_i \sim \mathcal{N}(\vec\xi_k, \vec\Omega_k)
+	\factorscores_\numdata \sim \mathcal{N}(\vec\xi_\numcomponents, \vec\Omega_\numcomponents)
 \end{equation}
-\noindent{}are drawn from the $k$th multivariate  normal distribution.
+\noindent{}are drawn from the $\numcomponents$th multivariate  normal distribution.
 The factor scores for all data points $\factorscores$ is then a 
-\todo{$n \times p$} matrix.
+$\NumData \times \NumLatentFactors$ matrix.
  We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \eye\specificvariance\right)$
 is independent of the latent space, and $\specificvariance$ is a
-diagonal matrix of $d$ entries.
+diagonal matrix of $\NumDimensions$ entries. In Appendix~\ref{app:symbols}
+we list all mathematical symbols used in this work and their description.
 
@@
 We assume that the latent space is lower dimensionality than the
-data space (e.g., $p < d$).
+data space (e.g., $\NumLatentFactors < \NumDimensions$).
 Within the context of stellar abundances, the factor loads
@@
 (stars). Within this framework a rare nucleosynthetic event
-can still be described as a `common factor load' $\factorloads_i$, 
+can still be described as a `common factor load' $\factorloads_\numlatentfactors$, 
 but its rarity would be represented by associated factor
@@
 dimensionality of the clustering problem is reduced from 
-$N \times D$ to $N \times J$. This reduces cost both through
+$N \times D$ to $N \times J$. This reduces computational cost through
 faster execution of each E-M step, and on average fewer E-M steps
-required to reach a specified convergence threshold.
+needed to reach a specified convergence threshold.
 
@@
 
-Without loss of generality the density of the data $\vecdata$ can be modelled as
+Without loss of generality the density of the data $\vecdata$ can be described as
 \begin{equation}
-	f(\textbf{y}; \vec\Psi) = \sum_{k=1}^{K}\weight_{k}\phi(\textbf{y};\factorloads\scoremeans_k, \factorloads\scorecovs_k\factorloads\transpose + \eye\specificvariance)
+	f(\vecdata; \vec\Psi) = \sum_{\numcomponents=1}^{\NumComponents}\weight_\numcomponents\phi(\vecdata;\factorloads\scoremeans_\numcomponents, \factorloads\scorecovs_\numcomponents\factorloads\transpose + \eye\specificvariance)
 \end{equation}
-\noindent{}given $p$ common factor loadings and $K$ components
+\noindent{}given $\NumLatentFactors$ common factor loadings and $\NumComponents$ components
 clustered in the latent (factor score) space. Here the parameter
 vector
-$\vec\Psi$ includes $\{\factorloads,\vec\pi,\scoremeans,\scorecovs,\specificvariance\}$, and $\phi(\textbf{y};\vec\mu, \vec\Sigma)$
+$\vec\Psi$ includes $\{\factorloads,\vec\pi,\scoremeans,\scorecovs,\specificvariance\}$, and $\phi(\vecdata;\vec\mu, \vec\Sigma)$
 describes the density of a multivariate gaussian distribution with
 mean $\vec\mu$ and covariance matrix $\vec\Sigma$,
-and $\weight_k$ describes the relative weighting of the $k$th
-component in latent space and $\sum\weight_k = 1$.
+and $\weight_\numcomponents$ describes the relative weighting of the $\numcomponents$th
+component in latent space and $\sum\weight_\numcomponents = 1$.
 The log likelihood is then given by
 \begin{equation}
-	\log\mathcal{L}(\vecdata|\vec\Psi) = \sum_{k=1}^{K}\log{f(\vecdata;\vec\Psi)} \quad .
+	\log\mathcal{L}(\vecdata|\vec\Psi) = \sum_{\numcomponents=1}^{\NumComponents}\log{f(\vecdata;\vec\Psi)} \quad .
 \end{equation}
 
-The model described by Equation~\ref{eq:generative-model} is indeterminate:
-there is no unique solution for the factor loads $\factorloads$ and scores
-$\factorscores$. However, an accurate estimate of the parameter vector $\vec\Psi$ can be obtained by the expectation-maximization algorithm \citep{EM}.
+%The model described by Equation~\ref{eq:generative-model} is indeterminate:
+%there is no unique solution for the factor loads $\factorloads$ and scores
+%$\factorscores$. However, an accurate estimate of the parameter vector $\vec\Psi$ can be obtained by the expectation-maximization algorithm \citep{EM}.
 
@@
 
-We initialise the entries in the factor loads $\factorloads$ by drawing
-from a uniform distribution $\factorloads \sim \mathcal{U}\left(-1, +1\right)$.
-We assign each data point as belonging to one of the $K$ components
-by drawing from a multinomial distribution. Given these assignments and
-factor loads, we then estimate the relative weights $\vec\pi$, the 
-mean factor scores of each component $\scoremeans$, and the covariance
-matrix of factor scores of each component $\scorecovs$.
+The code associated with this \emph{Article} allows the model parameters
+to be initialised using a variety of methods. For example, the initial factor 
+loads $\factorloads$ can be drawn by one of the following methods:
+\begin{enumerate}
+	\item \emph{uniform}: a uniform distribution $\factorloads \sim \mathcal{U}\left(-1, +1\right)$,
+	\item \emph{noise}: from noise $\factorloads \sim \mathcal{N}\left(0, 10^{-2}\right)$,
+	\item \emph{SVD}: using the $\NumLatentFactors$ principal eigenvectors estimated by singular value decomposition,
+	\item \emph{astrophysical}: using domain knowledge to set initial factor loads to vectors that are analogous to nucleosynthetic yields (see Section~\todo{\ref{sec:somewhere}}).
+\end{enumerate}
+
+We initially assign each data point as belonging to one of the $\NumComponents$
+components either by random assignment (e.g., drawing from a multinomial distribution), 
+or using the \texttt{K-means++} algorithm \citep{LLoyd} in latent space $\vecdata\factorloads$.
+Given the initial factor loads and initial assignments, we then estimate the relative weights
+$\vec\pi$, the mean factor scores of each component $\scoremeans$, and the covariance matrix
+of factor scores of each component $\scorecovs$. In all scenarios we initialise the 
+variance $\specificvariance$ in each dimension as 1.
 
@@
 
+We use the expectation-maximization algorithm to maximise the log-likelihood
+of the data, given the current estimate of the model parameters. With each step
+cycle we update the model parameters and increase the log-likelihood.
 At the expectation step we evaluate the log likelihood
@@
 matrix whose entries are the posterior probability
-that the $n$th data point is associated to the $k$th component,
+that the $\numdata$th data point is associated to the $\numcomponents$th component,
 given the data $\vecdata$ and the current estimate of the 
@@
 \begin{equation}
-	\tau_{nk} = \frac{\weight_k\phi(\vecdata_n;\factorloads\scoremeans_k, \factorloads\scorecovs_k\factorloads\transpose + \eye\specificvariance)}{\sum_{g=1}^{G}\weight_g\phi(\vecdata_n;\factorloads\scoremeans_g, \factorloads\scorecovs_g\factorloads\transpose + \eye\specificvariance)} \quad .
+	\tau_{\numdata\numcomponents} = \frac{\weight_\numcomponents\phi(\vecdata_\numdata;\factorloads\scoremeans_\numcomponents, \factorloads\scorecovs_\numcomponents\factorloads\transpose + \eye\specificvariance)}{\sum_{g=1}^{G}\weight_g\phi(\vecdata_\numdata;\factorloads\scoremeans_g, \factorloads\scorecovs_g\factorloads\transpose + \eye\specificvariance)} \quad .
 \end{equation}
@@
 At the maximization step we update our estimates of the parameters,
-conditioned on the data $y$ and the responsibility matrix $\vec\tau$.
+conditioned on the data $\vecdata$ and the responsibility matrix $\vec\tau$.
 Initially we update the relative weights $\vec\weight\nextstep$ given
@@
 \begin{equation}
-	\weight_k\nextstep = \frac{1}{N} \sum_{n=1}^{N}\tau_{nk} \quad .
+	\weight_\numcomponents\nextstep = \frac{1}{\NumData} \sum_{\numdata=1}^{\NumData}\tau_{\numdata\numcomponents} \quad .
 	% Include proof.
@@
 \begin{eqnarray}
-	\scoremeans_{k}\nextstep = \scoremeans_{k}\thisstep + \frac{\vec{G}\transpose(\vecdata\transpose - \factorloads\thisstep\scoremeans_{k}\thisstep)\vec\tau_{k}}{N\weight_k\nextstep}
+	\scoremeans_\numcomponents\nextstep = \scoremeans_\numcomponents\thisstep + \frac{\vec{G}\transpose(\vecdata\transpose - \factorloads\thisstep\scoremeans_\numcomponents\thisstep)\vec\tau_\numcomponents}{\NumData\weight_\numcomponents\nextstep}
 \end{eqnarray}
@@
 \begin{eqnarray}
-	\vec{W} &=& (\scorecovs_{k}\thisstep)^{-1}\eye \\
+	\vec{W} &=& (\scorecovs_\numcomponents\thisstep)^{-1}\eye \\
 	\vec{V} &=& \left(\specificvariance\thisstep\right)^{-1} \\
@@
 \begin{equation}
-	\scorecovs_k\nextstep = \left(\eye - \vec{G}\transpose\factorloads\thisstep\right)\scorecovs_k\thisstep + \frac{\vec{G}\transpose\vec{Z}\left(\vec{Z}\vec\tau_k\transpose\right)\transpose\vec{G}}{N\weight_k\nextstep}
+	\scorecovs_\numcomponents\nextstep = \left(\eye - \vec{G}\transpose\factorloads\thisstep\right)\scorecovs_\numcomponents\thisstep + \frac{\vec{G}\transpose\vec{Z}\left(\vec{Z}\vec\tau_\numcomponents\transpose\right)\transpose\vec{G}}{N\weight_\numcomponents\nextstep}
 \end{equation}
@@
 \begin{eqnarray}
-	\vec{Z} &=& \vecdata\transpose - \factorloads\thisstep\scoremeans_k\thisstep \quad .
+	\vec{Z} &=& \vecdata\transpose - \factorloads\thisstep\scoremeans_\numcomponents\thisstep \quad .
 \end{eqnarray}
 
-After some linear algebra, updated estimates of the common factor loads $\factorloads\thisstep$
+After some linear algebra, updated estimates of the common factor loads $\factorloads\nextstep$
 can be found from
@@
 \begin{eqnarray}
-	\factorloads_1 &=& \sum_{k=1}^{K}\left[ \vec\tau_k\transpose\vecdata\left(\scoremeans_k\thisstep\right)\transpose + 
-	\vec{G}\transpose\vec\tau_k\vec{Z}\transpose\vec{G}\right] \\
+	\factorloads_1 &=& \sum_{\numcomponents=1}^{\NumComponents}\left[ \vec\tau_\numcomponents\transpose\vecdata\left(\scoremeans_\numcomponents\thisstep\right)\transpose + 
+	\vec{G}\transpose\vec\tau_\numcomponents\vec{Z}\transpose\vec{G}\right] \\
 	%\left(\scoremeans_k\thisstep\vecdata\transpose\vec\tau_k\right)\transpose \left(\vec{G}\transpose\vec{Z}\vec\tau_k\transpose\vec{G}\right)\transpose\right]  \\
-	\factorloads_2 &=& N\sum_{k=1}^{K}\left[\weight_k\nextstep\left(\scorecovs_k\nextstep + \scoremeans_k\nextstep\left(\scoremeans_k\nextstep\right)\transpose\right)\right]
+	\factorloads_2 &=& N\sum_{\numcomponents=1}^{\NumComponents}\left[\weight_\numcomponents\nextstep\left(\scorecovs_\numcomponents\nextstep + \scoremeans_\numcomponents\nextstep\left(\scoremeans_\numcomponents\nextstep\right)\transpose\right)\right]
 \end{eqnarray}
@@
 \begin{equation}
-	\specificvariance\nextstep = \frac{1}{N}\left[\sum^{K}_{k=1}\vec\tau_k\transpose\left(\vecdata\odot\vecdata\right) - \sum_{j=1}^{J}\left(\factorloads\nextstep\factorloads_2\right)\odot\factorloads\nextstep\right]
+	\specificvariance\nextstep = \frac{1}{\NumData}\left[\sum^{\NumComponents}_{\numcomponents=1}\vec\tau_\numcomponents\transpose\left(\vecdata\odot\vecdata\right) - \sum_{j=1}^{J}\left(\factorloads\nextstep\factorloads_2\right)\odot\factorloads\nextstep\right]
 \end{equation}
@@
 
-We generated a data set with ${N = 100000}$ data points, each with
-$D = 15$ dimensions, and assumed that those data are generated by
-latent space of $J = 5$ ($J \times D$) factor loads, and there are $K = 20$
+We generated a data set with ${\NumData = 100000}$ data points, each with
+$\NumDimensions = 15$ dimensions, and assumed that those data are generated by
+latent space of $\NumLatentFactors = 5$ ($\NumLatentFactors \times \NumDimensions$) factor loads, and there are $\NumComponents = 20$
 clusters in the latent space. The relative weights $\vec\weight$
@@
 distribution.
-The covariance matrices in factor scores $\scorecovs$ are assumed to be diagonal matrices, where the non-zero entries are drawn from a gamma distribution $\scorecovs_{k,i,i} \sim \vec\Gamma\left(1\right)$.
+The covariance matrices in factor scores $\scorecovs$ are assumed to be diagonal matrices, where the non-zero entries are drawn from a gamma distribution $\scorecovs_{\numcomponents,i,i} \sim \vec\Gamma\left(1\right)$.
 The variance in
 each dimension $\specificvariance$ are also drawn $\specificvariance \sim \vec\Gamma\left(1\right)$.
-The $i$th data point (which belongs to the $k$th cluster) is then
-generated by drawing $\factorscores_{i} \sim \mathcal{N}(\scoremeans_k,\scorecovs_k)$, projecting by the factor loads, and adding variance.
+The $\numdata$th data point (which belongs to the $\numcomponents$th cluster) is then
+generated by drawing $\factorscores_{\numdata} \sim \mathcal{N}(\scoremeans_\numcomponents,\scorecovs_\numcomponents)$, projecting by the factor loads, and adding variance.
 
-We fit this generated data using the expectation-maximization algorithm
+\begin{figure*}
+	\includegraphics[width=1.0\textwidth]{experiments/toy-model-data.png}
+    \caption{A corner plot showing all $\NumDimensions$ dimensions of the
+    		 generated data for Experiment~1 (blue). Overplotted in red
+		 	 we show samples generated given our estimate of the model 
+			 parameters, without adding specific variance in each dimension.
+			 This demonstrates that the model parameters estimated by
+			 expectation-maximization can replicate the generated data.}
+    \label{fig:toy-model-data}
+\end{figure*}
+
+\begin{figure*}
+	\includegraphics[width=1.0\textwidth]{experiments/toy-model-latent-space.png}
+    \caption{A corner plot showing all $\NumLatentFactors$ dimensions in
+    		 latent space, where each data point is coloured by the component
+		 	 that it is estimated to belong to.}
+    \label{fig:toy-model-latent-space}
+\end{figure*}
+
+\subsubsection{Model fitting given the true number of latent factors and components}
+
+We initialised the factor loads and component assignments randomly, and
+fit this generated data set using the expectation-maximization algorithm
 as described in the previous Section. After 58 iterations
@@
 reached our convergence threshold: the log probability increased by
-less than $10^{-5}$ with each iteration. For this experiment we specified
-the number of latent factors $J$ and the number of clustering components
-$K$ to include, but below we repeat this experiment where the number
+less than $10^{-5}$ with each iteration (\todo{Figure~X}). For this experiment we specified
+the number of latent factors $\NumLatentFactors$ and the number of clustering components
+$\NumComponents$ to include, but below we repeat this experiment where the number
 of latent factors and clustering components is not known.
+
+\begin{figure}
+	%\includegraphics[width=0.5\textwidth]{experiments/toy-model-data.png}
+    \caption{Log-likelihood of the model given the data $\mathcal{L}\left(\vec\data|\vec\Psi\right)$ for Experiment 1 as a function of the number of expectation-maximization iterations.
+After \todo{X} iterations the improvement in log-likelihood per
+    expectation-maximization cycle was less than $10^{-5}$.}
+    \label{fig:toy-log-likelihood-per-iteration}
+\end{figure}
 
@@
 
+While the estimated model parameters can generate the data and demonstrate
+clustering in latent space, caution is needed when interpreting the 
+model latent variables.
 Latent factor models are by definition indeterminate in that there are
@@
 an identical data-generating process. This can be trivially shown by
-the fact that the factors loads and scores can be rotated with some rotation
+the fact that the factors loads and scores can be rotated by taking the
+dot product
+with some rotation
 matrix $\textbf{R}$ and generate the exact same data. As a consequence,
@@
 to the true factor loads, they could be ordered differently, or have
-their signs reversed. Given these issues, we evaluate the search as
-being successful if the log probability given the estimated model
-parameters is within convergence tolerance (or better) than the
-log probability given the data and the true parameters.
+their signs reversed. We discuss this further in later sections.
 
-% Despite not matching perfectly, it generates the data? property of uniqueness?
-% Factor scores?
-% Some metric about clustering performance against true values?
+\subsubsection{A grid search for the number of latent factors and components}
 
-\begin{figure*}
-	\includegraphics[width=1.0\textwidth]{experiments/toy-model-data.png}
-    \caption{\todo{Caption}}
-    \label{fig:toy-model-data}
-\end{figure*}
-
-\begin{figure*}
-	\includegraphics[width=1.0\textwidth]{experiments/toy-model-latent-space.png}
-    \caption{\todo{Caption}}
-    \label{fig:toy-model-latent-space}
-\end{figure*}
-
-\todo{Grid search text}
+Here we treat the generated data set as if the true number of latent factors
+and the true number of components are not known. Starting with $\NumLatentFactors = 1$
+and $\NumComponents = 1$, we trialled each permutation of $\NumLatentFactors$ and $\NumComponents$
+until $\NumLatentFactors_{max} = 10$
+and   $\NumComponents_{max} = 40$ (e.g., twice $\NumLatentFactors_{true}$ and $\NumComponents_{true}$).
+For each $\NumLatentFactors,\NumComponents$ permutation we initialised the factor
+loads and component assignments randomly. We performed expectation-maximization 
+cycles until the relative log-likelihood improved by less than $10^{-5}$. We then
+recorded the log-likelihood of the data given the model parameters, the
+Bayesian Information Criterion \citep{bic}, 
+\begin{equation}
+	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \label{eq:bic}
+\end{equation} 
+\noindent{}and the pseudo-Bayesian Information Criterion \citep{pseudo-bic}
+\begin{equation}
+ \textrm{pseudo-BIC} = 6(1 + \gamma)\omega{}Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \label{eq:pseudo-bic}
+\end{equation}
+\noindent{}where $Q$ is the total number of model parameters in both equations,
+and the unknown constants $\omega$ and $\gamma$ have constraints
+$\omega \geq 1$ and $\gamma > 0$. Here we adopt $\gamma = 0.1$ and $\omega = 1$.
+These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
 
@@
 			 evaluated at each combination of latent factors $J$ and number 
-			 of clusters $K$, the middle panel shows the Bayesian Information
-			 Criterion (Equation \ref{eq:bic}), and the lower panel shows
-			 the pseudo-Bayesian Information Criterion 
-			 (Equation~\ref{eq:pseudo-bic}). The white marker indicates the 
+			 of clusters $K$, the middle panel shows the BIC (Equation \ref{eq:bic}), 
+			 and the lower panel shows
+			 the pseudo-BIC
+			 (Equation~\ref{eq:pseudo-bic}). The marker indicates the 
 			 lowest value in each panel.}
@@
 
-% Search strategy for toy model.
+Unsurprisingly the log likelihood increases with increasing numbers of latent
+factors $\NumLatentFactors$ and increasing numbers of components $\NumComponents$.
+The lowest BIC value is found at $\NumLatentFactors = 5$
+and $\NumComponents = 8$, close to the true values ($\NumLatentFactors_{true} = 5$,
+$\NumComponents_{true} = 10$). The lowest pseudo-BIC value is also found at
+$\NumLatentFactors = 5$, but with fewer components: $\NumComponents = 4$.
 
-% Figure: toy-model-factor-loads?
-% Figure: toy-model-specific-variances?
+The estimated latent factor loads found during this grid search are shown in
+Figure~\ref{fig:fig:experiment-1-gridsearch-factors-init-random-and-random},
+where each panel shows the estimated latent factor loads found by
+expectation-maximization for a given number of latent factors and components.
+The latent factors $\factorloads$ in each panel are multiplied by a rotation
+matrix $\mathbf{R}$ with entries $\{-1, 0, 1\}$ in order to re-orient and flip
+the estimated latent factors to be as similar as possible to the latent factors
+found from the model with the same number of latent factors and $\NumComponents  =  1$.
+In doing so we find that the estimated latent factors vary slightly as
+the number of components $\NumComponents$ increases, but the overall shape
+of the latent factors remains the same.
 
-\subsection{Experiment 2: A subset of 14 chemical abundances in \Galah}
+\begin{figure*}
+	\includegraphics[width=1.0\textwidth]{experiments/toy-factors-init-randomly.png}
+	\caption{Latent factor loads inferred for models with increasing numbers
+			 of latent factors and increasing numbers of cluster components
+			 in Experiment~1, where the latent factors and component assignments
+			 were initialised randomly.
+				The top left panel shows the results of the model with
+			 $\NumLatentFactors = 1$ and $\NumComponents = 1$. 
+			 The latent factors in each panel have been rotated by a
+			 rotation matrix $\textbf{R}$ with entries $\{-1, 0, +1\}$ such 
+			 that the estimated latent factors in each model are ordered 
+			 and oriented to be as similar as the latent factors estimated from
+			 the model with only one component (i.e., the latent factors in the top row
+			 of the same column). Note that the rotation
+			 matrix contains only entries $\{-1, 0, +1\}$.}
+	\label{fig:toy-factors-init-randomly}
+\end{figure*}
+
+\todo{This structure goes away and becomes more consistent if we change the initialisation
+procedure}
+
+\todo{figure of toy factors init svd kmeans++}
+
+\todo{log-likelihood of different initialisation routines}
+
+\begin{figure*}
+	\includegraphics[width=1.0\textwidth]{experiments/toy-svd-kmeans-factors.png}
+	\caption{Latent factor loads inferred for models with increasing numbers
+			 of latent factors and increasing numbers of cluster components
+			 in Experiment~1, where the latent factors were initialised using
+			 \texttt{SVD} and the component assignments were initialised using \texttt{k-means++}.
+				The top left panel shows the results of the model with
+			 $\NumLatentFactors = 1$ and $\NumComponents = 1$. 
+			 The latent factors in each panel have been rotated by a
+			 rotation matrix $\textbf{R}$ with entries $\{-1, 0, +1\}$ such 
+			 that the estimated latent factors in each model are ordered 
+			 and oriented to be as similar as the latent factors estimated from
+			 the model with only one component (i.e., the latent factors in the top row
+			 of the same column). Compared to Figure~\ref{fig:toy-factors-init-randomly},
+			 it is clear that initialising with \texttt{SVD}/\texttt{k-means++} results
+			 in less scatter in the final estimates of the model parameters.}
+	\label{fig:toy-factors-init-svd-and-kmeans++}
+\end{figure*}
+
+The results of this experiment confirm that the model is indeterminate, in 
+that there are multiple equally valid solutions that can exactly reproduce
+the data. It also highlights the importance of initialisation, demonstrating
+how the initialisation of the common latent factors can impact the final
+model parameters. These two points are related, as one could initialise the
+latent factors randomly (or from noise) and multiply that model with a
+rotation matrix $\mathbf{R}$ that rotates the latent space such that the
+common factor loads are aligned closer to some prior expectation that we have
+for the common factor loads (e.g., in that they should represent plausible
+astrophysical yields). Motivated by these results and indeterminate nature
+of the model, in Experiment~2 we initialise the common factor loads
+to represent known astrophysical processes such that they can be interpreted
+in context with the data.
+
+\subsection{Experiment 2: A minimal set of nucleosynthetic trace elements in \Galah}
+
+In this experiment we further investigate the effects of initialising the model parameters.
+Here we make use of a minimal set of nucleosynthetic trace elements
+using the second \Galah\ data release \citep{Buder:2018}. Specifically, we use
+reliable chemical abundance measurements of Na (an odd-Z element), Mg and Ca (alpha-elements),
+Ni, Fe, and Cr (iron-peak elements), Ba and Y (s-process tracers), and Eu (an r-process tracer).
+We assume that there are four latent factors ($J = 4$) and four components ($K = 4$),
+and we initialise the latent factors and component parameters in different combinations.
+
+To initialise the latent factors using astrophysical insight (or domain expertise)
+we make the following steps:
+\begin{enumerate}
+	\item We initialise all entries of $\factorloads$ to be drawn from $\mathcal{N}\left(0, 10^{-2}\right)$.
+	\item For the first latent factor we set the entries that correspond to Ni, Fe, and Cr to be $+1$ (e.g., Fe-peak).
+	\item For the second latent factor we set the entries that correspond to Mg and Ca to be $+1$ (e.g., $\alpha$-elements).
+	\item For the third latent factor we set the entries that correspond to Ba and Y to be $+1$ (e.g., the s-process).
+	\item For the fourth latent factor we set the entry that corresponds to Eu to be $+1$ (e.g., the r-process).
+\end{enumerate}
+
+%The latent factors can be initialised in four different ways: (1) with entries drawn randomly from $\mathcal{U}\left(-1, +1\right)$, (2) with noisey draws $\mathcal{N}\left(0, 10^{-2}\right)$, (3) using singular value decomposition, and (4) using astrophysical insight to initialise the factor loads.
+
+The component parameters are either initialised randomly, or using \texttt{K-means++} algorithm in
+the latent space given the initial estimate of the latent factors. For each combination of 
+initialisation procedure, we run 30 optimisations and record the log likelihood once the E-M 
+procedure has reached a threshold of $10^{-8}$.
+The results of this experiment are shown in Figure~\ref{fig:experiment-4-log-L}.
+In most cases the final log likelihood is within the threshold of $10^{-8}$ between
+different optimisation runs of the same initialisation procedure. However, it is clear
+that at least some initialisations result in a local optimum, emphasising the importance in
+optimising from many initialisations.
+
+Figure~\ref{fig:experiment-4-log-L} demonstrates that even though the initialisation procedure
+can impact the final estimate of the model parameters, the resulting log likelihood values
+are within the convergence threshold. This is, again, a consequence of the indeterminate
+nature of the model specification. However, the advantage here is that these results
+demonstrate that we \emph{can} initialise the latent factors with domain knowledge,
+without a significant detriment to the final log likelihood. Similarly, in Figure~\ref{fig:experiment-4-astrophysical-latent-factors}
+we show that the initial latent factors are indeed different from the final latent factors,
+and that there is some variance in our estimates of the latent factors from 30 initialisations.
+\todo{something to say here about not biasing our results?}
+
+The results from this experiment demonstrate that we can initialise the latent factors
+using domain knowledge without incurring strong biases in our final estimates of the
+latent factors, or introducing a significant detriment to the log likelihood, as long
+as we initialise the model many (e.g., $\gtrsim 5$) times.
+
+\begin{figure}
+	\includegraphics[width=0.5\textwidth]{experiments/galah-experiment-4-log-L-hist.png}
+	\caption{Log likelihoods $\mathcal{L}\left(\data|\vec\Psi\right)$ of optimised
+			 models using \Galah\ data in Experiment 4 with four latent factors
+			 and four components ($J = 4$, $K = 4$). The columns indicate different
+			 methods to initialise the component parameters ($\pi$, $\scoremeans$,
+			 $\scorecovs$), and the rows represent different methods to initialise
+			 the latent factors ($\factorloads$). Each combination of initialisation
+			 procedure has 30 models where E-M was run until a threshold of $10^{-8}$
+			 was reached in $\log{L}$.}
+	\label{fig:galah-experiment-4-log-L}
+\end{figure}
+
+\todo{experiment-4-astrophysical-latent-factors}
+
+\subsection{Experiment 3: A subset of gravitationally bound stellar clusters from \Galah\ and elsewhere}
+\label{sec:experiment-galah-clusters}
+
+\subsection{Experiment 4: A subset of 14 chemical abundances in \Galah}
 \label{sec:experiment-galah}
@@
 includes up to 23 chemical abundances measured for 342,682
-stars. It is not practical to use all of these measurements
+stars. It is not practical to use all of these measurements as-is
 when performing chemical tagging. For example, our experiments
@@
 
-\begin{figure*}
-	\includegraphics[width=1.0\textwidth]{experiments/galah-abundances.png}
-    \caption{\todo{Caption}}
+\begin{figure}
+	\includegraphics[width=0.45\textwidth]{experiments/galah-abundances.png}
+    \caption{Number of \Galah\ stars with reliable abundance measurements,
+    		 shown in order from the most commonly measured elemental abundance
+		 	 (Fe) to the least reliable element (Li). The solid line
+			 shows the number of reliable measurements for the given element,
+			 and the dotted (blue) line indicates the stars with reliable
+			 abundances in all elements listed to the left (including the
+			 element of interest). For example, there are $\sim60,000$
+			 stars with reliable abundances in all elements listed from
+			 Fe to Cu, inclusive.}
     \label{fig:galah-abundances}
-\end{figure*}
+\end{figure}
 
@@
 \begin{figure*}
-	\includegraphics[width=1.0\textwidth]{experiments/galah-experiment-1-gridsearch-factors-wrt-J-and-K.png}
-	\caption{Latent factor loads inferred for models with increasing numbers
-			 of latent factors and increasing numbers of cluster components
-			 in Experiment~2. The latent factors have each been rotated by a
-			 rotation matrix $\textbf{R}$ with entries $\{-1, 0, +1\}$ such 
-			 that the estimated latent factors in each model are ordered 
-			 and oriented to be as similar as the latent factors estimated from
-			 the model with only one latent factor. Note that the rotation
-			 matrix contains only entries $\{-1, 0, +1\}$. The latent
-			 factors for each model shown here were initialised using singular value
-			 decomposition on a random subset of the data.
-			 }
-	\label{fig:experiment-2-gridsearch-factors-init-svd}
-\end{figure*}
-
-\begin{figure*}
 	\includegraphics[width=1.0\textwidth]{experiments/galah-experiment-2-gridsearch-factors-wrt-J-and-K.png}
@@
 	\label{fig:experiment-2-gridsearch-factors-init-svd}
-\end{figure*}
-
-\subsection{Experiment 4: A minimal set of nucleosynthetic trace elements in \Galah}
-
-In this experiment we investigate the effects of initialising the model parameters.
-Here we make use of a minimal set of nucleosynthetic trace elements
-using the second \Galah\ data release \citep{Buder:2018}. Specifically, we use
-reliable chemical abundance measurements of Na (an odd-Z element), Mg and Ca (alpha-elements),
-Ni, Fe, and Cr (iron-peak elements), Ba and Y (s-process tracers), and Eu (an r-process tracer).
-We assume that there are four latent factors ($J = 4$) and four components ($K = 4$),
-and we initialise the latent factors and component parameters in different combinations.
-The latent factors can be initialised in four different ways: (1) with entries drawn randomly from $\mathcal{U}\left(-1, +1\right)$, (2) with noisey draws $\mathcal{N}\left(0, 10^{-2}\right)$, (3) using singular value decomposition, and (4) using astrophysical insight to initialise the factor loads.
-\todo{When using astrophysical insight...}
-The component parameters are either initialised randomly, or using K-means++ algorithm in
-the latent space. For each combination of initialisation procedure, we run 30 optimisations
-and record the log likelihood once the E-M procedure has reached a threshold of $10^{-8}$.
-
-The results of this experiment are shown in Figure~\ref{fig:experiment-4-log-L}.
-
-\begin{figure*}
-	\includegraphics[width=1.0\textwidth]{experiments/galah-experiment-4-log-L-hist.png}
-	\caption{Log likelihoods $\mathcal{L}\left(\data|\vec\Psi\right)$ of optimised
-			 models using \Galah\ data in Experiment 4 with four latent factors
-			 and four components ($J = 4$, $K = 4$). The columns indicate different
-			 methods to initialise the component parameters ($\pi$, $\scoremeans$,
-			 $\scorecovs$), and the rows represent different methods to initialise
-			 the latent factors ($\factorloads$). Each combination of initialisation
-			 procedure has 30 models where E-M was run until a threshold of $10^{-8}$
-			 was reached in $\log{L}$.}
-	\label{fig:galah-experiment-4-log-L}
 \end{figure*}
@@
 
+\appendix
+
+\section{Nomenclature}\label{app:symbols}
+
+\todo{summarise symbols}
+
 \bibliographystyle{aasjournal}

### 2018-12-11 article/ms.tex
update ms.tex

@@
 \begin{abstract}
+Chemical tagging promises to distinguish unique star formation sites based on the
+present day photospheric abundances of stars.  It is not known how effective this
+will be.  Clustering techniques used for chemical tagging  assume statistical 
+independence between abundances. Theory and observations agree that this assumption
+is incorrect: some elemental abundances are produced through multiple nucleosynthetic
+processes, and most nucleosynthetic processes generates multiple chemical abundances.
+Here we introduce a statistically consistent approach to chemical tagging, where we
+simultaneously infer the nucleosynthetic processes that produce chemical abundances,
+and we perform clustering in this lower dimensional latent space. \todo{we use galah}
+\todo{scales iwth data}
 \end{abstract}
@@
 	\item Clustering vs common traits
+	\item Dimensionality of data space
 \end{itemize}
@@
 
-% TODO: Is D a diagonal vector, or a matrix?! Make it consistent.
+%Factor analysis is a statistical method used to describe variability among observed, correlated variables in terms of a potentially lower number of unobserved variables called factors.
 
-% Describe how we will attempt to define a generative model for the data, from the latent space.
+Factor analysis is a common statistical approach for describing correlated 
+observations with a lower number of latent variables \citep{factor-analysis}.
+\todo{Analagous approaches used in astrophysics: pca, etc}.
 
-In this work the data $\vecdata$ is a $\NumData \times \NumDimensions$ matrix 
-where $\NumData$ is the number of stars and $\NumDimensions$ is the number of
-chemical abundances measured for each star. We assume a
+Similarly, clustering techniques are commonly used to identify groups in 
+data space.
+
+Here we use a combination of these methods that provides a
+consistent description of the data, given the latent variables and their 
+clustering in latent space. In this work the data $\vecdata$ is a 
+$\NumData \times \NumDimensions$ matrix where $\NumData$ is the number of 
+stars and $\NumDimensions$ is the number of chemical abundances measured 
+for each star. We assume a
 generative model for the data 
@@
 we list all mathematical symbols used in this work and their description.
-
 In this model each data point can be represented as being drawn
@@
 into the data space by the factor loads $\factorloads$.
+
 We assume that the latent space is lower dimensionality than the
@@
 Within the context of stellar abundances, the factor loads
-$\factorloads$ can be thought of as the \emph{mean} yields
+$\factorloads$ can be thought of as the \emph{typical} yields
 of nucleosynthetic
-events (e.g., $s$-process yields from AGB stars), and the
+events (e.g., $s$-process production from AGB stars), and the
 factor scores are analogous to the relative counts of those 
@@
 achieves the same as a clustering procedure in data space,
-except we simultaneously estimate the factor loads
-(nucleosynthetic yields) that are common to all data points 
-(stars). Within this framework a rare nucleosynthetic event
-can still be described as a `common factor load' $\factorloads_\numlatentfactors$, 
+except we simultaneously estimate the latent processes that are
+common to all stars (the so-called factor loads, analogous to 
+nucleosynthetic yields). Within this framework a rare nucleosynthetic event
+can still be described as a `factor load' $\factorloads_\numlatentfactors$, 
 but its rarity would be represented by associated factor
-scores being zero for most stars and thus have no contribution
-to the observed abundances. The factor loads and factor scores 
-cannot be expressly interpreted like this because they have no
-physical meaning, but this description of relative yields and 
-rates should help build intuition for the model parameters,
-and provide an astrophysical context.
+scores being zero for most stars and thus have negligible contribution
+to the observed abundances. In practice the factor loads can only be 
+identified up to orthogonality and cannot be expressly interpreted as
+nucleosynthetic yields because they have limited physical meaning,
+but this description of typical yields and relative event rates should
+help build intuition for the model parameters, and provide context
+within the astrophysical problem it is being applied.
 
-One benefit of using a mixture of common factor analysers is
+\todo{this model of orthogonality takes the independence into account}
+
+Another benefit of using a mixture of common factor analysers is
 for modelling stellar chemical abundances is the scaling with
@@
 within a statistically consistent framework. That is to say
-that we have a generative model for the data that includes
-analogues of nucleosynthetic yields and star formation,
+that we have a generative model for the data that can 
+quantitatively account for nucleosynthetic yields, variations in
+turbulence and gas mixing, or star formation efficiency,
+%can account
+%analogues of nucleosynthetic yields and star formation,
 and the parameters of this model can be estimated consistently
@@
 
+As mentioned previously, the model as described is indeterminate in that
+there is no unique solution for the factor loads $\factorloads$ and scores
+$\factorscores$. These quantities can only be determined up until 
+orthogonality. However, accurate estimates of the parameter vector $\vec\Psi$
+can be obtained by the expectation-maximization algorithm \citep{EM}.
 %The model described by Equation~\ref{eq:generative-model} is indeterminate:
@@
 
-The code associated with this \emph{Article} allows the model parameters
-to be initialised using a variety of methods. For example, the initial factor 
-loads $\factorloads$ can be drawn by one of the following methods:
-\begin{enumerate}
-	\item \emph{uniform}: a uniform distribution $\factorloads \sim \mathcal{U}\left(-1, +1\right)$,
-	\item \emph{noise}: from noise $\factorloads \sim \mathcal{N}\left(0, 10^{-2}\right)$,
-	\item \emph{SVD}: using the $\NumLatentFactors$ principal eigenvectors estimated by singular value decomposition,
-	\item \emph{astrophysical}: using domain knowledge to set initial factor loads to vectors that are analogous to nucleosynthetic yields (see Section~\todo{\ref{sec:somewhere}}).
-\end{enumerate}
+Here we describe the default initialisation of the model parameters, which
+we have found to be robust in a large set of models. Given a number of
+$\NumLatentFactors$ latent factors we initialise the factor load
+$\factorloads$ entries to be drawn from a uniform distribution
+$\factorloads \sim \mathcal{U}\left(-1, +1\right)$ and then ensure that
+the factor loads are orthogonal such that
+%TODO: Should we re-write this so that we just draw from the scipy special ortho group?
+\begin{equation}
+	\factorloads \factorloads\transpose = \eye \quad .
+\end{equation}
+%        AL = linalg.cholesky(A.T @ A)
+%        A = A @ linalg.solve(AL, np.eye(self.n_latent_factors))
+% 	
+We then initially assign each data point as belonging to one of the
+$\NumComponents$ components by drawing from a multinomial distribution
+and generating the responsibility matrix $\vec\tau$. Given the initial
+factor loads and assignments, we then estimate the relative weights
+$\vec\pi$, the mean factor scores of each component $\scoremeans$, and
+the covariance matrix of factor scores of each component $\scorecovs$.
+Finally, we initialise the specific variance $\specificvariance$ in each
+dimension as 1.
 
-We initially assign each data point as belonging to one of the $\NumComponents$
-components either by random assignment (e.g., drawing from a multinomial distribution), 
-or using the \texttt{K-means++} algorithm \citep{LLoyd} in latent space $\vecdata\factorloads$.
-Given the initial factor loads and initial assignments, we then estimate the relative weights
-$\vec\pi$, the mean factor scores of each component $\scoremeans$, and the covariance matrix
-of factor scores of each component $\scorecovs$. In all scenarios we initialise the 
-variance $\specificvariance$ in each dimension as 1.
-
-\begin{comment}
-We use the K-means++ algorithm to initially assign data points
-randomly to $K$ clusters (in data space). This provides a so-called
-hard initial association where each data point is wholly assigned
-to a single component. Specifically we calculate the $n \times K$
-responsibility matrix $\vec\tau$, where a entry of 1 indicates the
-$n$th data point is associated with the $k$th cluster, and an
-entry of 0 indicates no association such that
-\begin{equation}
-	\weight_k = \frac{1}{N}\sum_{n=1}^{N}\tau_{nk} \quad .
-\end{equation}
-We then chose 1,000
-random data points and factorize that matrix using singular-value
-decomposition. We take the principal $p$ eigenvectors as the 
-initial guess for the factor loads $\factorloads$, and calculate
-an initial estimate of the factor score means
-\begin{equation}
-	\scoremeans_k = \frac{1}{N}\sum_{n=1}^{N}\tau_{nk}\vecdata_{n}\factorloads
-\end{equation}
-\noindent{}and the covariance matrices of the clusters in latent space
-\begin{equation}
-	\scorecovs_k = \textrm{cov}{\left(\tau_{nk}\vecdata_{n}\right)} \quad .
-\end{equation}
-\end{comment}
+Other initialisation methods are available in the code 
+associated with this \emph{Article}. In later Sections we show that 
+the optimised model parameters are insensitive to the choice initialisation, 
+up until orthogonal rotation of the latent space.
 
@@
 
-We use the expectation-maximization algorithm to maximise the log-likelihood
-of the data, given the current estimate of the model parameters. With each step
-cycle we update the model parameters and increase the log-likelihood.
-At the expectation step we evaluate the log likelihood
-given the model parameters $\vec\Psi$, and we re-calculate the responsibility
-matrix whose entries are the posterior probability
+We use the expectation-maximization algorithm to estimate the model parameters.
+With each expectation step we evaluate the log likelihood given the model 
+parameters $\vec\Psi$, and we re-calculate the responsibility 
+matrix $\vec\tau$ whose entries are the posterior probability
 that the $\numdata$th data point is associated to the $\numcomponents$th component,
@@
 
-% The complexity of SVD is O(n^3), and
-% the complexity of K-means++ is O(log(k)). If we knew the complexity of
-% k-means++ relative to N, then we would be able to make better decisions
-% as to how to initialise: SVD first and then K-means in latent space?
-
-% Discuss connection to PCA, FA, MFA.
-
-% Include macros for \vec, \data, etc.
 At the maximization step we update our estimates of the parameters,
 conditioned on the data $\vecdata$ and the responsibility matrix $\vec\tau$.
-Initially we update the relative weights $\vec\weight\nextstep$ given
+First we estimate the relative weights $\vec\weight\nextstep$ given
 the responsibility matrix $\vec\tau$
 \begin{equation}
-	\weight_\numcomponents\nextstep = \frac{1}{\NumData} \sum_{\numdata=1}^{\NumData}\tau_{\numdata\numcomponents} \quad .
+	\weight_\numcomponents\nextstep = \frac{1}{\NumData} \sum_{\numdata=1}^{\NumData}\tau_{\numdata\numcomponents}
 	% Include proof.
 \end{equation}
+\noindent{}where the $\vec{X}\thisstep$ superscript refers to the current estimate of a variable
+and $\vec{X}\nextstep$ refers to the updated estimate for the next iteration.
+%\footnote{Here $\vec\textrm{X}\now$ refers to the current estimate of a variable $\vec\textrm{X}$ and $\vec\textrm{X}\nextstep$ refers to the updated estimate for the next iteration.}
 
@@
 The updated estimates of the mean factor scores 
-$\scoremeans\nextstep$ are given by
+$\scoremeans\nextstep$ for each component are then given by
 \begin{eqnarray}
@@
 
-\noindent{}where $\odot$ denotes is the entrywise (Hadamard) product.
+\noindent{}where $\odot$ denotes is the entry-wise (Hadamard) product. We repeated
+the expectation-maximization algorithm for up to 1,000 steps or until the log-likelihood
+improved by less than $10^{-5}$ between successive iterations. Although our implementation
+allows for a positive regularisation term to be added along the diagonal of the covariance
+matrices in latent space $\scorecovs$, we set this regularisation term to zero for all
+experiments performed here.
 
@@
 
-\begin{comment}
-For this subset
-we examined the measured abundances and
-counted the number of stars with \texttt{flag\_x\_fe = 0}
-for each abundance. This flag indicates that there is no known
-reason to suspect that the measured chemical abundance is
-unreliable.
-
-We sorted the elemental abundances by the number of stars
-with reliable measurements, from highest to lowest. Starting
-with Fe, we iterated over each element and counted the number
-of stars with reliable measurements of that element and all
-other elements earlier in the list. We illustrate this process
-in Figure~\ref{fig:galah-abundances}, where we show that
-if we include the top 14 elemental abundances, we would have
-12,798 stars with reliable measurements in all fourteen elements.
-This is relevant to the approach we use here, as the method
-described here requires no missing data.
-\end{comment}
-
-\begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/galah-abundances.png}
-    \caption{Number of \Galah\ stars with reliable abundance measurements,
-    		 shown in order from the most commonly measured elemental abundance
-		 	 (Fe) to the least reliable element (Li). The solid line
-			 shows the number of reliable measurements for the given element,
-			 and the dotted (blue) line indicates the stars with reliable
-			 abundances in all elements listed to the left (including the
-			 element of interest). For example, there are $\sim60,000$
-			 stars with reliable abundances in all elements listed from
-			 Fe to Cu, inclusive.}
-    \label{fig:galah-abundances}
-\end{figure}
-
-\begin{comment}
-This leaves us with a sample of 12,798 stars each with
-reliable stellar parameters ($\teff$, $\logg$) and fourteen
-(reliably) measured abundances: Na, Mg, Si, Ca, Sc, Ti, Cr, Mn, 
-Fe, Ni, Cu, Zn, Ba, and Eu
-\todo{Y} (reliable) measured abundances: \todo{Mn, Fe}.
-These elements are produced from multiple nucleosynthetic
-pathways, including \todo{X}. 
-It is this subsample that we will use for experiments
-in chemical tagging.
-\end{comment}
-
 %\subsubsection{Search strategy} \label{sec:experiment-galah-search}

### 2018-12-11 article/experiments/exp1.py
add exp1

@@ -1,0 +1,236 @@
+
+import sys
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+from matplotlib.ticker import MaxNLocator
+from time import time
+
+sys.path.insert(0, "../../")
+
+from mcfa import (mcfa, mpl_utils, utils)
+
+matplotlib.style.use(mpl_utils.mpl_style)
+colors = plt.rcParams['axes.prop_cycle'].by_key()['color']
+
+
+# Should all go into exp1.yaml
+
+data_kwds = dict(n_samples=1000,
+                 n_features=15,
+                 n_components=10,
+                 n_latent_factors=3,
+                 psi_scale=1,
+                 latent_scale=1,
+                 random_seed=42,
+                 eta=1)
+
+mcfa_kwds = dict(tol=1e-5, 
+                 max_iter=1_000,
+                 init_factors="random",
+                 init_components="random",
+                 random_seed=42)
+
+gridsearch_max_latent_factors = 5
+gridsearch_max_components = 15
+
+# Done with vars
+
+Y, truth = utils.generate_data(**data_kwds)
+
+# Fit with true number of latent factors and components.
+model = mcfa.MCFA(n_components=data_kwds["n_components"],
+                  n_latent_factors=data_kwds["n_latent_factors"],
+                  **mcfa_kwds)
+tick = time()
+model.fit(Y)
+tock = time()
+
+print(f"Model took {tock - tick:.1f} seconds")
+
+# Plot the log-likelihood with increasing iterations.
+fig_iterations, ax = plt.subplots()
+
+ll = model.log_likelihoods_
+iterations = 1 + np.arange(len(ll))
+ax.plot(iterations, ll, "-", lw=2, drawstyle="steps-mid")
+ax.set_xlabel(r"$\textrm{iteration}$")
+ax.set_ylabel(r"$\log{\mathcal{L}}$")
+ax.set_xlim(0, iterations[-1] + 1)
+ax.xaxis.set_major_locator(MaxNLocator(6))
+ax.yaxis.set_major_locator(MaxNLocator(6))
+xt = ax.get_xticks()
+ax.set_xticks(xt + np.hstack([1, np.zeros(xt.size - 1)]))
+ax.set_xlim(0, iterations[-1] + 1)
+fig_iterations.tight_layout()
+
+
[…]

### 2018-12-16 article/ms.tex
moo

@@
 \begin{figure*}
-	\includegraphics[width=1.0\textwidth]{experiments/toy-model-data.png}
+	\includegraphics[width=1.0\textwidth]{experiments/exp1-data.png}
     \caption{A corner plot showing all $\NumDimensions$ dimensions of the
@@
 			 expectation-maximization can replicate the generated data.}
-    \label{fig:toy-model-data}
+    \label{fig:exp1-data}
 \end{figure*}
 
-\begin{figure*}
-	\includegraphics[width=1.0\textwidth]{experiments/toy-model-latent-space.png}
-    \caption{A corner plot showing all $\NumLatentFactors$ dimensions in
-    		 latent space, where each data point is coloured by the component
-		 	 that it is estimated to belong to.}
-    \label{fig:toy-model-latent-space}
-\end{figure*}
+%\begin{figure*}
+%	\includegraphics[width=1.0\textwidth]{experiments/toy-model-latent-space.png}
+%    \caption{A corner plot showing all $\NumLatentFactors$ dimensions in
+%    		 latent space, where each data point is coloured by the component
+%		 	 that it is estimated to belong to.}
+%    \label{fig:toy-model-latent-space}
+%\end{figure*}
 
-\subsubsection{Model fitting given the true number of latent factors and components}
+\subsubsection{Given the true number of latent factors and components}
 
@@
 \begin{figure}
-	%\includegraphics[width=0.5\textwidth]{experiments/toy-model-data.png}
+	\includegraphics[width=0.5\textwidth]{experiments/exp1-ll-iterations.png}
 	\caption{Log-likelihood of the model given the data $\mathcal{L}\left(\vec\data|\vec\Psi\right)$ for Experiment 1 as a function of the number of expectation-maximization iterations.
@@
     expectation-maximization cycle was less than $10^{-5}$.}
-    \label{fig:toy-log-likelihood-per-iteration}
+    \label{fig:exp1-ll-iterations.png}
 \end{figure}
@@
 with some rotation
-matrix $\textbf{R}$ and generate the exact same data. As a consequence,
-we cannot strictly evaluate our model fitting procedure by comparing
-how closely the inferred factor loads (or scores) match the true 
-variables. This is further complicated by orientation and order
-of the latent factors: even if the inferred factor loads were very close
-to the true factor loads, they could be ordered differently, or have
-their signs reversed. We discuss this further in later sections.
+matrix $\textbf{R}$ and generate the exact same data. 
+We discuss how this limitation applies to later experiments.
 
-\subsubsection{A grid search for the number of latent factors and components}
+%This is further complicated by orientation and order
+%of the latent factors: even if the inferred factor loads were very close
+%to the true factor loads, they could be ordered differently, or have
+%their signs reversed. We discuss this further in later sections.
+
+\subsubsection{Grid search for the number of latent factors and components}
 
@@
 \begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/toy-ll-contours.png}
-	\includegraphics[width=0.45\textwidth]{experiments/toy-bic-contours.png}
-	\includegraphics[width=0.45\textwidth]{experiments/toy-pseudobic-contours.png}
+	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-ll-contours.png}
+	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-bic-contours.png}
+%	\includegraphics[width=0.45\textwidth]{experiments/toy-pseudobic-contours.png}
     \caption{Model performance metrics resulting from a gridsearch using the
@@
 expectation-maximization for a given number of latent factors and components.
+
 The latent factors $\factorloads$ in each panel are multiplied by a rotation
-matrix $\mathbf{R}$ with entries $\{-1, 0, 1\}$ in order to re-orient and flip
-the estimated latent factors to be as similar as possible to the latent factors
-found from the model with the same number of latent factors and $\NumComponents  =  1$.
+matrix $\mathbf{R}$ to be as similar (up to orthogonal rotation) as those found
+from the same number of latent factors and $\NumComponents  =  1$.
 In doing so we find that the estimated latent factors vary slightly as
-the number of components $\NumComponents$ increases, but the overall shape
-of the latent factors remains the same.
+the number of components $\NumComponents$ increases.
 
-\begin{figure*}
-	\includegraphics[width=1.0\textwidth]{experiments/toy-factors-init-randomly.png}
-	\caption{Latent factor loads inferred for models with increasing numbers
-			 of latent factors and increasing numbers of cluster components
-			 in Experiment~1, where the latent factors and component assignments
-			 were initialised randomly.
-				The top left panel shows the results of the model with
-			 $\NumLatentFactors = 1$ and $\NumComponents = 1$. 
-			 The latent factors in each panel have been rotated by a
-			 rotation matrix $\textbf{R}$ with entries $\{-1, 0, +1\}$ such 
-			 that the estimated latent factors in each model are ordered 
-			 and oriented to be as similar as the latent factors estimated from
-			 the model with only one component (i.e., the latent factors in the top row
-			 of the same column). Note that the rotation
-			 matrix contains only entries $\{-1, 0, +1\}$.}
-	\label{fig:toy-factors-init-randomly}
-\end{figure*}
+%\begin{figure*}
+%	\includegraphics[width=1.0\textwidth]{experiments/toy-factors-init-randomly.png}
+%	\caption{Latent factor loads inferred for models with increasing numbers
+%			 of latent factors and increasing numbers of cluster components
+%			 in Experiment~1, where the latent factors and component assignments
+%			 were initialised randomly.
+%				The top left panel shows the results of the model with
+%			 $\NumLatentFactors = 1$ and $\NumComponents = 1$. 
+%			 The latent factors have each been rotated by a
+%			 rotation matrix such 
+%			 that the estimated latent factors in each model are ordered 
+%			 and oriented to be as similar as the latent factors estimated from
+%			 the model with only one latent factor, up to orthogonality such
+%			 that $\vec{R}\vec{R}\transpose = \eye$.
+%			 }
+%	\label{fig:toy-factors-init-randomly}
+%\end{figure*}
 
-\todo{This structure goes away and becomes more consistent if we change the initialisation
-procedure}
+%\todo{This structure goes away and becomes more consistent if we change the initialisation procedure}
 
-\todo{figure of toy factors init svd kmeans++}
+%\todo{figure of toy factors init svd kmeans++}
 
-\todo{log-likelihood of different initialisation routines}
+%\todo{log-likelihood of different initialisation routines}
 
-\begin{figure*}
-	\includegraphics[width=1.0\textwidth]{experiments/toy-svd-kmeans-factors.png}
-	\caption{Latent factor loads inferred for models with increasing numbers
-			 of latent factors and increasing numbers of cluster components
-			 in Experiment~1, where the latent factors were initialised using
-			 \texttt{SVD} and the component assignments were initialised using \texttt{k-means++}.
-				The top left panel shows the results of the model with
-			 $\NumLatentFactors = 1$ and $\NumComponents = 1$. 
-			 The latent factors in each panel have been rotated by a
-			 rotation matrix $\textbf{R}$ with entries $\{-1, 0, +1\}$ such 
-			 that the estimated latent factors in each model are ordered 
-			 and oriented to be as similar as the latent factors estimated from
-			 the model with only one component (i.e., the latent factors in the top row
-			 of the same column). Compared to Figure~\ref{fig:toy-factors-init-randomly},
-			 it is clear that initialising with \texttt{SVD}/\texttt{k-means++} results
-			 in less scatter in the final estimates of the model parameters.}
-	\label{fig:toy-factors-init-svd-and-kmeans++}
-\end{figure*}
-
+\begin{comment}
 The results of this experiment confirm that the model is indeterminate, in 
@@
 \todo{experiment-4-astrophysical-latent-factors}
+\end{comment}
 
-\subsection{Experiment 3: A subset of gravitationally bound stellar clusters from \Galah\ and elsewhere}
+\subsection{Experiment~2: A subset of gravitationally bound stellar clusters from \Galah\ and elsewhere}
 \label{sec:experiment-galah-clusters}
@@
 
-\begin{figure*}
-	\includegraphics[width=1.0\textwidth]{experiments/galah-experiment-2-gridsearch-factors-wrt-J-and-K.png}
-	\caption{Latent factor loads inferred for models with increasing numbers
-			 of latent factors and increasing numbers of cluster components
-			 in Experiment~2. The latent factors have each been rotated by a
-			 rotation matrix $\textbf{R}$ with entries $\{-1, 0, +1\}$ such 
-			 that the estimated latent factors in each model are ordered 
-			 and oriented to be as similar as the latent factors estimated from
-			 the model with only one latent factor. Note that the rotation
-			 matrix contains only entries $\{-1, 0, +1\}$. The latent
-			 factors for each model shown here were initialised randomly.
-			 }
-	\label{fig:experiment-2-gridsearch-factors-init-svd}
-\end{figure*}
-
 \section{Results} \label{sec:results}

### 2018-12-16 article/experiments/exp1.py

@@ -170,8 +170,9 @@ xt = ax.get_xticks()
 ax.set_xticks(xt + np.hstack([1, np.zeros(xt.size - 1)]))
 ax.set_xlim(0, iterations[-1] + 1)
 fig_iterations.tight_layout()
+savefig(fig_iterations, "ll-iterations")
 
-"""
+
 # Plot the data, with samples
 Y_drawn = model.sample(data_kwds["n_samples"])
 fig_data = mpl_utils.corner_scatter(Y, 
@@ -181,14 +182,14 @@ mpl_utils.corner_scatter(Y_drawn,
 
 fig_data.tight_layout()
 fig_data.subplots_adjust(hspace=0, wspace=0)
+savefig(fig_data, "data")
 
-"""
 
 # Plot the true latent factors w.r.t. the estimated ones, after rotation.
 A_true = truth["A"]
 A_est = model.theta_[model.parameter_names.index("A")]
 
-R = utils.find_rotation_matrix(A_true, A_est, n_inits=100)
+R, *_ = utils.find_rotation_matrix(A_true, A_est, n_inits=100)
 A_est_rot = A_est @ R
 
 D, J = A_true.shape
@@ -213,7 +214,7 @@ ax.set_ylim(-ylim, +ylim)
 ax.set_yticks([-ylim, 0, ylim])
 
 fig_factor_loads.tight_layout()
-
+savefig(fig_factor_loads, "factor_loads")
 
 # Do a grid search.
 Js = np.arange(1, 1 + gridsearch_max_latent_factors)
@@ -264,7 +265,6 @@ J_true, K_true = (data_kwds["n_latent_factors"], data_kwds["n_components"])
 print(f"BIC is lowest at J = {jm_b} and K = {km_b}")
 print(f"pseudo-BIC is lowest at J = {jm_pb} and K = {km_pb}")
 print(f"True values are  J = {J_true} and K = {K_true}")
-
 
 
 kwds = dict(converged=converged, marker_function=np.nanargmin)

### 2018-12-16 article/experiments/exp2.py

@@ -31,7 +31,8 @@ element_mask = np.array([ln.split("_")[0] not in ignore_elements for ln in label
 X_H = X_H[:, element_mask]
 label_names = list(np.array(label_names)[element_mask])
 
-X = utils.whiten(X_H)
+#X = utils.whiten(X_H)
+X = X_H
 
 
 grouped_elements = [
@@ -178,22 +179,21 @@ for i, tes in enumerate(astrophysical_grouping):
         idx = label_names.index("{0}_h".format(te.lower()))
         A_astrophysical[idx, i] = 1.0
 
-AL = linalg.cholesky(A_astrophysical.T @ A_astrophysical)
-A_astrophysical = A_astrophysical @ linalg.solve(AL, np.eye(model.n_latent_factors))
-
+#AL = linalg.cholesky(A_astrophysical.T @ A_astrophysical)
+#A_astrophysical = A_astrophysical @ linalg.solve(AL, np.eye(model.n_latent_factors))
+load_labels=[
+  r"$\textrm{CCSN and others?}$",
+  r"$\textrm{s-process?}$",
+  r"$\textrm{r-process?}$"
+],
 
 np.random.seed(42)
 
-fig_factors_rotated = mpl_utils.plot_factor_loads(A_est, 
+fig_factors_rotated = mpl_utils.plot_factor_loads(A_est, separate_axes=True,
                                                   target_loads=A_astrophysical,
                                                   flip_loads=None,
-                                                  n_rotation_inits=1000,
+                                                  n_rotation_inits=100,
                                                   show_target_loads=False,
-                                                  load_labels=[
-                                                    r"$\textrm{CCSN and others?}$",
-                                                    r"$\textrm{s-process?}$",
-                                                    r"$\textrm{r-process?}$"
-                                                  ],
                                                   xlabel=xlabel,
                                                   xticklabels=xticklabels,
                                                   legend_kwds=dict(loc="upper left",

### 2018-12-16 article/experiments/exp4.py

@@ -33,7 +33,7 @@ mcfa_kwds = dict(init_factors="random", init_components="random", tol=1e-5,
                  max_iter=10000, random_seed=42)
 
 
-use_galah_flags = True
+use_galah_flags = 0
 
 
 abundance_counts = galah.get_abundances_breakdown(galah.available_elements,
@@ -66,8 +66,8 @@ X_H, label_names = galah.get_abundances_wrt_h(elements,
 print(f"Data shape: {X_H.shape}")
 
 
-#X = utils.whiten(X_H)
-X = X_H
+X = utils.whiten(X_H)
+#X = X_H
 
 # Do a gridsearch.
 max_n_latent_factors = 10
@@ -97,10 +97,9 @@ fig_bic = mpl_utils.plot_filled_contours(Jg, Kg, bic,
                                          **plot_filled_contours_kwds)
 savefig(fig_bic, "gridsearch-bic")
 
-
 # 
 
-model = mcfa.MCFA(n_components=K_best_bic, n_latent_factors=5,
+model = mcfa.MCFA(n_components=K_best_bic, n_latent_factors=J_best_bic,
                   **mcfa_kwds)
 
 model.fit(X)
@@ -137,7 +136,7 @@ r"$\textrm{r-process?}$",
 
 
 A_astrophysical = np.zeros_like(A_est)
-for i, tes in enumerate(astrophysical_grouping[:J_best_bic]):
+for i, tes in enumerate(astrophysical_grouping[:model.n_latent_factors]):
     for j, te in enumerate(tes):
         try:
             idx = label_names.index("{0}_h".format(te.lower()))
@@ -160,7 +159,7 @@ fig_factors_rotated = mpl_utils.plot_factor_loads(A_est,
                                                   separate_axes=True,
                                                   target_loads=A_astrophysical,
                                                   flip_loads=None,
-                                                  n_rotation_inits=10,
+                                                  n_rotation_inits=100,
                                                   show_target_loads=False,
                                                   xlabel=xlabel,
                                                   #load_labels=load_labels,

### 2018-12-16 article/experiments/galah.py

@@ -1,0 +1,140 @@
+
+
+import numpy as np
+from astropy.table import Table
+from collections import OrderedDict
+from operator import itemgetter
+
+data = Table.read("GALAH_iDR3_OpenClusters.fits")
+
+
+periodic_table = """H                                                  He
+                    Li Be                               B  C  N  O  F  Ne
+                    Na Mg                               Al Si P  S  Cl Ar
+                    K  Ca Sc Ti V  Cr Mn Fe Co Ni Cu Zn Ga Ge As Se Br Kr
+                    Rb Sr Y  Zr Nb Mo Tc Ru Rh Pd Ag Cd In Sn Sb Te I  Xe
+                    Cs Ba Lu Hf Ta W  Re Os Ir Pt Au Hg Tl Pb Bi Po At Rn
+                    Fr Ra Lr Rf"""
+
+lanthanoids    =   "La Ce Pr Nd Pm Sm Eu Gd Tb Dy Ho Er Tm Yb"
+actinoids      =   "Ac Th Pa U  Np Pu Am Cm Bk Cf Es Fm Md No"
+
+periodic_table = periodic_table.replace(" Ba ", " Ba " + lanthanoids + " ") \
+                               .replace(" Ra ", " Ra " + actinoids + " ") \
+                               .split()
+
+available_elements = list(set([ea.split("_")[1][:2].rstrip("123456789") \
+                        for ea in data.dtype.names if ea.startswith("A_")]))
+
+indices = np.argsort([periodic_table.index(el) for el in available_elements])
+available_elements = [available_elements[i] for i in indices]
+
+def _get_elements(data):
+    return [ln.split("_")[1] for ln in data.dtype.names \
+            if ln.startswith("A_") and ln.split("_")[1] in periodic_table]
+
+def _abundance_label(element):
+    return f"A_{element}"
+
+def _abundance_flag_label(element):
+    return f"flag_A_{element}"
+
+
+
+
+def get_abundance_mask(elements, cluster_names=None, use_galah_flags=False,
+                       full_output=False):
+
+    if elements is None:
+        elements = []
+
+    if cluster_names is None or len(cluster_names) < 1:
+        cluster_names = []
+        use = np.ones(len(data))
+
[…]

### 2018-12-16 article/experiments/galah_dr3.py

@@ -1,0 +1,141 @@
+
+
+import numpy as np
+from astropy.table import Table
+from collections import OrderedDict
+from operator import itemgetter
+
+#data = Table.read("GALAH_iDR3_OpenClusters.fits")
+
+data = Table.read("../../catalogs/GALAH_DR2.1_catalog.fits")
+
+periodic_table = """H                                                  He
+                    Li Be                               B  C  N  O  F  Ne
+                    Na Mg                               Al Si P  S  Cl Ar
+                    K  Ca Sc Ti V  Cr Mn Fe Co Ni Cu Zn Ga Ge As Se Br Kr
+                    Rb Sr Y  Zr Nb Mo Tc Ru Rh Pd Ag Cd In Sn Sb Te I  Xe
+                    Cs Ba Lu Hf Ta W  Re Os Ir Pt Au Hg Tl Pb Bi Po At Rn
+                    Fr Ra Lr Rf"""
+
+lanthanoids    =   "La Ce Pr Nd Pm Sm Eu Gd Tb Dy Ho Er Tm Yb"
+actinoids      =   "Ac Th Pa U  Np Pu Am Cm Bk Cf Es Fm Md No"
+
+periodic_table = periodic_table.replace(" Ba ", " Ba " + lanthanoids + " ") \
+                               .replace(" Ra ", " Ra " + actinoids + " ") \
+                               .split()
+
+available_elements = list(set([ea.split("_")[0].title() \
+                        for ea in data.dtype.names if (ea.endswith("_fe") or ea == "fe_h") and ea.split("_")[0].lower() not in ("e", "flag", "alpha")]))
+
+indices = np.argsort([periodic_table.index(el) for el in available_elements])
+available_elements = [available_elements[i] for i in indices]
+
+def _get_elements(data):
+    return [ln.split("_")[1] for ln in data.dtype.names \
+            if ln.startswith("A_") and ln.split("_")[1] in periodic_table]
+
+def _abundance_label(element):
+    if element.lower() == "fe":
+        return "fe_h"
+    else:
+        return f"{element.lower()}_fe"
+
+def _abundance_flag_label(element):
+    if element.lower() == "fe":
+        return "flag_cannon"
+    else:
[…]

### 2018-12-16 article/experiments/toy.py

@@ -1,166 +1,0 @@
-
-
-import sys
-import numpy as np
-import matplotlib.pyplot as plt
-from time import time
-
-sys.path.insert(0, "../../")
-
-from mcfa import (mcfa, mpl_utils, utils)
-
-
-
-n_features = 15
-n_components = 20
-n_latent_factors = 5
-n_samples = 100000
-
-omega_scale = 1
-noise_scale = 1
-random_seed = 1
-
-X, truth = utils.generate_data(n_samples=n_samples,
-                               n_features=n_features,
-                               n_latent_factors=n_latent_factors,
-                               n_components=n_components,
-                               omega_scale=omega_scale,
-                               noise_scale=noise_scale,
-                               random_seed=random_seed)
-
-model = mcfa.MCFA(n_components=n_components, n_latent_factors=n_latent_factors)
-model.fit(X)
-
-# Now do a search!
-
-
-
-# Just check that the ML solution is at least as good as the generated value
-# (or at least within the convergence tolerance)
-"""
-diff = model.expectation(X, *utils.parameter_vector(**truth))[0] \
-     - model.expectation(X, *model.theta_)[0]
-assert diff <= model.tol
-"""
-
-pi, A, xi, omega, psi = model.theta_
-scores = model.factor_scores(X)[2]
-
-
-def common_limits(ax):
-    limits = np.array([ax.get_xlim(), ax.get_ylim()])
-    limits = (np.min(limits), np.max(limits))
-
-    ax.set_xlim(limits)
-    ax.set_ylim(limits)
-
-
-fig, axes = plt.subplots(n_latent_factors)
-
-for i, ax in enumerate(axes):
-    ax.plot(A.T[i], c="tab:blue")
-    ax.plot(truth["A"].T[i], c="tab:red")
-
-
-# Generate data.
-X2 = scores @ A.T
-
-def corner_scatter(X, label_names=None, show_ticks=False, fig=None, figsize=None,
-                   **kwargs):
-    """
-    Make a corner plot where the data are shown as scatter points in each axes.
-
-    :param X:
-        The data, :math:`X`, which is expected to be an array of shape
-        [n_samples, n_features].
-
-    :param label_names: [optional]
-        The label names to use for each feature.
-
-    :param show_ticks: [optional]
-        Show ticks on the axes.
-
-    :param fig: [optional]
[…]

### 2018-12-16 mcfa/mcfa.py

@@ -552,7 +552,7 @@ class MCFA(object):
 
 
 
-    def sample(self, n_samples=1):
+    def sample(self, n_samples=1, theta=None):
         r"""
         Generate random samples from the fitted model.
 
@@ -562,7 +562,10 @@ class MCFA(object):
         # TODO: return docs
         """
 
-        pi, A, xi, omega, psi = self.theta_
+        if theta is None:
+            theta = self.theta_
+
+        pi, A, xi, omega, psi = theta
 
         # Draw which component it is from.
         taus = np.random.choice(self.n_components, size=n_samples, p=pi)
@@ -602,7 +605,27 @@ class MCFA(object):
         if J != J_:
             raise ValueError("rotation matrix is not square")
 
-        np.allclose(R @ R.T)
+        if not np.allclose(R @ R.T, np.eye(J)):
+            raise ValueError("R is not a valid rotation matrix")
+
+        pi, A, xi, omega, psi = self.theta_
+
+        A_rot = A @ R
+        xi_rot = np.ones_like(xi)
+        omega_rot = np.ones_like(omega)
+        
+        # [TODO] There is a better way than sampling,...
+        for k, (xi_, omega_) in enumerate(zip(xi.T, omega.T)):
+
+            draws = np.random.multivariate_normal(xi_, omega_, size=1000000)
+            draws_rot = draws @ R.T
+            xi_rot[:, k] = np.mean(draws_rot, axis=0)
+            omega_rot[:, :, k] = np.cov(draws_rot.T)
+
+        return pi, A_rot, xi_rot, omega_rot, psi
+
+
+
 
 
 def _initial_factor_loads_by_random(X, n_latent_factors, random_state=None):

### 2018-12-16 mcfa/utils.py

@@ -379,9 +379,13 @@ def find_rotation_matrix(A, B, init=None, n_inits=25, full_output=True, **kwargs
                 return np.nan * cost
             return cost
 
-        p_opt, cov, infodict, mesg, ier = op.leastsq(objective, 
-                                                     best_opt.x % (2 * np.pi),
-                                                     full_output=True)
+        try:
+            p_opt, cov, infodict, mesg, ier = op.leastsq(objective, 
+                                                         best_opt.x % (2 * np.pi),
+                                                         full_output=True)
+
+        except:
+            p_opt, cov, infodict, mesg, ier = best_opt.x, None, dict(), 0, -1
     
         return (best_R, p_opt, cov, infodict, mesg, ier)
 

### 2018-12-17 article/ms.tex
add some text on exp2

@@
 fit this generated data set using the expectation-maximization algorithm
-as described in the previous Section. After 58 iterations
-(and a serial processing time of 40 seconds) we
+as described in the previous Section. After \todo{58} iterations
+(and a serial processing time of \todo{40} seconds) we
 reached our convergence threshold: the log probability increased by
-less than $10^{-5}$ with each iteration (\todo{Figure~X}). For this experiment we specified
+less than $10^{-5}$ with each iteration (Figure~\ref{fig:exp1-ll-iterations}). For this experiment we specified
 the number of latent factors $\NumLatentFactors$ and the number of clustering components
@@
     expectation-maximization cycle was less than $10^{-5}$.}
-    \label{fig:exp1-ll-iterations.png}
+    \label{fig:exp1-ll-iterations}
 \end{figure}
@@
 cycles until the relative log-likelihood improved by less than $10^{-5}$. We then
-recorded the log-likelihood of the data given the model parameters, the
+recorded the log-likelihood of the data given the model parameters, and the
 Bayesian Information Criterion \citep{bic}, 
 \begin{equation}
-	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \label{eq:bic}
+	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \quad . \label{eq:bic}
 \end{equation} 
-\noindent{}and the pseudo-Bayesian Information Criterion \citep{pseudo-bic}
-\begin{equation}
- \textrm{pseudo-BIC} = 6(1 + \gamma)\omega{}Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \label{eq:pseudo-bic}
-\end{equation}
-\noindent{}where $Q$ is the total number of model parameters in both equations,
-and the unknown constants $\omega$ and $\gamma$ have constraints
-$\omega \geq 1$ and $\gamma > 0$. Here we adopt $\gamma = 0.1$ and $\omega = 1$.
+%\noindent{}and the pseudo-Bayesian Information Criterion \citep{pseudo-bic}
+%\begin{equation}
+% \textrm{pseudo-BIC} = 6(1 + \gamma)\omega{}Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \label{eq:pseudo-bic}
+%\end{equation}
+%\noindent{}where $Q$ is the total number of model parameters in both equations,
+%and the unknown constants $\omega$ and $\gamma$ have constraints
+%$\omega \geq 1$ and $\gamma > 0$. Here we adopt $\gamma = 0.1$ and $\omega = 1$.
 These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
@@
 
-\subsection{Experiment~2: A subset of gravitationally bound stellar clusters from \Galah\ and elsewhere}
-\label{sec:experiment-galah-clusters}
+\subsection{Experiment~2: Inferring astrophysically realistic latent factors}
+
+Latent factors can only be identified up to orthogonality. That is to say that
+if the data were truely generated by latent factors $\factorloads_\textrm{true}$,
+then our estimates of those latent factors $\factorloads_\textrm{est}$ may not
+be identical to the true values. For example, the ordering of the estimated factors
+could be different from the true factors, and the ordering of the dimensionality
+in latent space would be correspondingly different. Another possibility is that
+the estimated factor loads could be flipped in sign relative to the true factor
+loads, and the scores would similarly be flipped. In both of these situations
+(re-ordering or flipped signs) the log likelihood given the data and the
+estimated factor loads $\factorloads_\textrm{est}$ would be identical to the
+log likelihood given the data and the true factor loads $\factorloads_\textrm{true}$
+despite the difference in ordering and sign. These examples illustrate a more 
+general property that the factor loads and factor scores can be orthogonally 
+rotated by any rotation matrix\footnote{A rotation matrix is valid if 
+$\vec{R}\vec{R}\transpose = \vec{I}$} $\vec{R}$. The estimated factor loads 
+$\factorloads_\textrm{est}$ could therefore appear very different from the true 
+values, but they only differ by an orthogonal rotation.
+
+While it is trivial to find the rotation matrix $\vec{R}$ that rotates the estimated
+factor loads to the true factor loads, this property of factor analyzers limits
+how we can interpret the latent variables of our model. For example, if we seek
+to interpret the factor loads as being something analagous to nucleosynthetic
+processes, then the estimated factor loads will need to be rotated to something
+that looks astrophysically plausible before we could attribute one latent factor
+as being a particular nucleosynthetic process.
+
+In this experiment we use detailed chemical abundances of metal-poor stars as
+reported by \citet{Barklem:2003}, and we demonstrate the limitations in
+interpreting the latent factors as nucleosynthetic processes. The data contains
+14 measurements of chemical abundances for 61 stars ($N = 61$, $D = 14$): 
+Al (light odd-Z), Mg and Ca ($\alpha$), Sc, Ti, Cr, Mn, Fe, Co, and Ni (Fe-peak),
+as well as Sr, Y, Ba, and Eu ($s$/$r$-process). We performed a grid search for
+the optimal number of latent factors $J$ and components $K$, starting from
+$J = 1$ and $K = 1$, and trialling up until $J = 8$ and $K = 5$. We initialised
+the model parameters randomly for each permutation of $J$ and $K$, and ran E-M
+until a convergence threshold of $10^{-5}$ was reached. In Figure~\ref{fig:exp2-gridsearch-contours}
+we show the negative log likelihood and BIC recorded for the optimised set of
+model parameters for each permutation of $J$ and $K$. Hyperparameters of $J = 5$
+and $K = 1$ are preferred by BIC for this data set.
+
+\begin{figure}
+	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-ll.png}
+	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-bic.png}
+    \caption{Model performance metrics resulting from a gridsearch using the
+    		 \citet{Barklem:2003} data in Experiment~2.
+    		 The top panel shows the negative log-likelihood 
+			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
+			 evaluated at each combination of latent factors $J$ and number 
+			 of clusters $K$, and the lower panel shows the BIC (Equation \ref{eq:bic}).
+			 The cross at
+			 $J=8$ and $K=5$ indicates that the model could not converge without
+			 adding a regularisation constant to the diagonals of $\scorecovs$.
+			 The marker indicates the lowest value in each panel, showing that
+			 $J = 5$ and $K = 1$ is preferred by BIC.}
+    \label{fig:exp2-gridsearch-contours}
+\end{figure}
+
+Using our preferred model with $J = 5$ and $K = 1$, we rotated the estimated
+factor loads in order to be astrophysically plausible. Specifically we constructed
+a $J \times D$ astrophysical factor load matrix $\factorloads_\textrm{A}$ where
+all entries were zero except for the following entries:
+\begin{itemize}
+	\item the entry corresponding to Al in the first latent factor to represent
+		  odd-Z element production,
+	\item the entries corresponding to Ca and Mg in the second latent factor to
+		  represent $\alpha$ particle production,
+	\item the entries corresponding to Ni, Co, Fe, Mn, Cr, Ti, and Sc in the third
+		  latent factor to represent the Fe-peak group,
+	\item the entries corresponding to Sr, Y, and Ba in the fourth latent factor
+		  to represent products of the slow neutron capture process, and
+	\item the entries corresponding to Eu in the fifth latent factor to represent
+		  products of the rapid nucleosynthesis process.
+\end{itemize}
+These entries were set to $1/\sqrt{E}$ where $E$ is the number of non-zero entries
+in that latent factor, in order to ensure that $\factorloads_\textrm{A}$ is
+orthonormal. We optimise the $J$ angles needed to make the estimated factor loads
+$\factorloads_\textrm{est}$ as close to the astrophysical factor loads $\factorloads_\textrm{A}$.
+In practice we take the product of many  Givens rotation matrices \citep{Givens},
+which reduces to Euler angle rotation in three or fewer dimensions. In Figure
+\ref{fig:exp2-factor-loads} we show the rotated factor loads found by our
+preferred model, where the shaded regions indicate the projected uncertainty on
+the rotation angles -- not including any uncertainty on the factor loads themselves.
+
+The rotated loads are not precisely the same as the target astrophysical loads
+in part because the target loads only seek to \emph{identify} and \emph{order}
+factors that can be interpreted within an astrophysical context. For example,
+the first factor does show positive contributions to Al, like the target load,
+but it also shows a slight negative contribution with respect to alpha elements
+(Mg, Ca), and a gradual increase among Fe-peak elements up to Ni. Similarly while
+Ca and Mg are indeed positive in the second factor, as targeted, there is a
+weak negative contribution to Mn abundances, which was not part of the target
+loads. The fourth factor shows a positive contributions to Sr, Y, Ba, and to a
+lesser extent, Eu, which is representative of slow neutron capture production.
+Similarly, the last component shows a very strong positive contribution towards
+Eu, and could be interpreted as production from the rapid nucleosynthesis process.
+
+The relatively wide uncertainty regions on these factors is perhaps not
+surprising given that only 61 stars were used in this data set. However, this
+experiment hopefully demonstrates that the rotated latent factors have some
+interpretability, albeit limited. After accounting for the contributions by all
+$J = 5$ factors, the specific scatter remaining in each dimension is shown in
+Figure~\ref{fig:exp2-specific-scatter}. The remaining scatter varies from as low 
+as 0.01-0.02~dex (Mg, Ca, Ti, Mn, Fe, Y) up to 0.08-0.16~dex (Al and Ba).
+
+\begin{figure*}
+	\includegraphics[width=\textwidth]{experiments/exp2-latent-factors.png}
+	\caption{Rotated latent factors found in Experiment~2 using the \citet{Barklem:2003}
+			 data. \todo{elaborate latent factors}}
+    \label{fig:exp2-latent-factors}
+\end{figure*}
+
+\begin{figure}
+	\includegraphics[width=0.45\textwidth]{experiments/exp2-specific-scatter.png}
+	\caption{Specific scatter remaining in the \citet{Barklem:2003} data used in
+			 Experiment~2 after accounting for the contributions by all latent
+			 factors.}
+    \label{fig:exp2-specific-scatter}
+\end{figure}
+
+\subsection{Experiment~3: A subset of gravitationally bound stellar clusters 
+			from \Galah\ and elsewhere}
+\label{sec:exp3}
+
+After demonstrating that there is some interpretability in the latent factors,
+here we seek to experiment whether gravitationally bound stellar clusters can
+be reliably identified using the model described in Section~\ref{sec:model}.
 

### 2018-12-18 article/ms.tex
words

@@
 \newcommand{\acronym}[1]{{\small{#1}}}
+\newcommand{\article}{\emph{Article}}
 
@@
 
-\shorttitle{Statistically consistent chemical tagging}
+\shorttitle{A novel approach to chemical tagging in lower-dimensional latent space}
 \shortauthors{Casey et al.}
@@
 
-\title{A statistically consistent approach to chemical tagging}
+\title{A novel approach to chemical tagging in lower-dimensional latent space}
 
@@
 Chemical tagging promises to distinguish unique star formation sites based on the
-present day photospheric abundances of stars.  It is not known how effective this
-will be.  Clustering techniques used for chemical tagging  assume statistical 
-independence between abundances. Theory and observations agree that this assumption
-is incorrect: some elemental abundances are produced through multiple nucleosynthetic
-processes, and most nucleosynthetic processes generates multiple chemical abundances.
-Here we introduce a statistically consistent approach to chemical tagging, where we
-simultaneously infer the nucleosynthetic processes that produce chemical abundances,
-and we perform clustering in this lower dimensional latent space. \todo{we use galah}
-\todo{scales iwth data}
+present day photospheric abundances of stars. Clustering techniques used to date
+have treated all chemical abundances as independent, which we know is not true.
+Nucleosynthetic processes make multiple elements in varying quantities, and the
+dimensionality of chemical abundance space is typically lower than the reported
+number of abundances. Here we introduce a novel approach to chemical tagging 
+where a set of latent factors (e.g., nucleosynthetic yields) contribute to the
+data, and there is clustering in the latent space (e.g., relative counts of
+nucleosynthetic events). This approach allows us to infer contributions akin to
+nucleosynthetic yields, and the clustering becomes more efficient as it occurs in
+a lower dimensional latent space.
+We run experiments on literature chemical abundances,
+including the second \Galah\ data release. We find that 5 latent factors are
+preferred to explain the chemical abundances of metal-poor stars, and 8 are
+needed to account for \Galah\ data.
+\todo{something about the clusters or factors found}
+We release code associated with this \article\ that scales well with the 
+available data: enabling chemical tagging for $10^{6}$ stars on a standard
+desktop machine in minutes.
 \end{abstract}
@@
 
-\begin{itemize}
-	\item Chemical tagging
-	\item Limits on the efficacy of chemical tagging
-	\item Clustering vs common traits
-	\item Dimensionality of data space
-\end{itemize}
+Chemical tagging describes the process of identifying individual star formation
+sites based on the present day photospheric abundances of stars \citep{Freeman;Bland-Hawthorn:2001}.
+Since the dynamical dissipation timescale for open clusters is relatively short
+\citep[\todo{timescale}][]{}, the ability to identify star formation sites long
+after the stars are no longer gravitationally bound could provide key insights
+on the early formation of the disc of the Milky Way galaxy. For this reason,
+much effort has been placed on obtaining observational data, methods for analysis,
+theory, etc, in order to improve the prospects of chemical tagging.
 
@@
 
+\todo{early approaches on clustering metrics}
+
+\todo{work on lower limits of chemical homogeneity}
+
+All clustering techniques applied to chemical tagging assume that the data
+dimensions are independent. However, theory and observations agree this cannot
+be true. Nucleosynthetic processes produce multiple elements in varying
+quantities, and the dimensionality of stellar abundance datasets has been shown
+to be lower than the actual number of dimensions. Any clustering approach that
+treats each new elemental abundance as an indenpendent axis of information will
+therefore conclude with biased inferences about the star formation history of
+our Galaxy. On the other hand, it is difficult to prescribe the nucleosynthetic
+yields that have contributed to the chemical abundances of each star. There are
+qualitative statements that can be made for large numbers of stars, or particular
+types of stars, but quantifying the precise contribution of different processes
+to each star is an unsolved problem. For example, the [$\alpha$/Fe] `knee' in
+abundance ratios in the Milky Way can qualitatively be explained by 
+core-collapse supernovae being the predominant nucleosynthetic process in the
+early Milky Way before Type Ia contributions had a significant impact, but 
+efforts to date have not sought to try to explain the detailed abundances of a 
+single star as a contribution of yields from different systems (e.g., 20\% of the
+Ba was produced by 3 M$_\odot$ systems). This is in part because of the
+challenging and degenerate nature of the problem as described, and is complicated
+by the differences in yield predictions that account from prescriptions used in
+different codes.
+
+In this work we present a new approach to chemical tagging that allows us to 
+identify the latent factors that contribute to the chemical abundances of all 
+stars (e.g., something akin to nucleosynthetic yields), while simultaneously
+performing clustering in the latent space. Notwithstanding caveats that we will
+discuss in detail, this allows us to infer nucleosynthetic yields rather than
+strictly prescribe them from models. Moreover, the scale of the clustering
+problem reduces by a significant fraction because the clustering is performed in
+a lower dimensional latent space instead of the higher dimensional data space.
+In Section~\ref{sec:methods} we describe the model and the methods we use to
+estimate the model parameters. Section~\ref{sec:experiments} describe the 
+experiments performed using generated and real data sets. We discuss the results
+of these experiments in Section~\ref{sec:discussion}, including the caveats with
+the model as described. We conclude in Section~\ref{sec:conclusion}
+
 \section{Methods} \label{sec:methods}
-
-%Factor analysis is a statistical method used to describe variability among observed, correlated variables in terms of a potentially lower number of unobserved variables called factors.
 
@@
 observations with a lower number of latent variables \citep{factor-analysis}.
-\todo{Analagous approaches used in astrophysics: pca, etc}.
+Related techniques include principal component analysis \citep{pca} and its
+variants \citep{ppca}, singular value decomposition \citep{svd}, and other
+matrix factorization methods. While factor analysis on its own is a useful
+dimensionality reduction tool to identify latent factors that contribute to
+the chemical abundances of stars, it does not perform any clustering.
+Similarly, clustering techniques applied to chemical abundances to date 
+\citep[e.g.,][]{Hogg:2016} do not account for the lower dimensionality of the
+data.
 
-Similarly, clustering techniques are commonly used to identify groups in 
-data space.
-
-Here we use a combination of these methods that provides a
-consistent description of the data, given the latent variables and their 
-clustering in latent space. In this work the data $\vecdata$ is a 
+Here we use a combination of factor analysis and a gaussian mixture model to
+provide a consistent description of the data, given the latent variables and 
+their clustering in latent space. In this work the data $\vecdata$ is a 
 $\NumData \times \NumDimensions$ matrix where $\NumData$ is the number of 
 stars and $\NumDimensions$ is the number of chemical abundances measured 
-for each star. We assume a
-generative model for the data 
+for each star. We assume a generative model for the data 
 \begin{equation}
@@
 
-\noindent{}where $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$ matrix of factor
-loads that is common to all data points, and the factor scores for
-the $\numdata$th data point
+\noindent{}where $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$ 
+matrix of factor loads that is common to all data points, and the factor scores 
+for the $\numdata$th data point
 \begin{equation}
@@
 The factor scores for all data points $\factorscores$ is then a 
-$\NumData \times \NumLatentFactors$ matrix.
+$\NumData \times \NumLatentFactors$ matrix, where each data point has a partial
+association to the components in latent space. 
 We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \eye\specificvariance\right)$
 is independent of the latent space, and $\specificvariance$ is a
-diagonal matrix of $\NumDimensions$ entries. In Appendix~\ref{app:symbols}
-we list all mathematical symbols used in this work and their description.
+diagonal matrix of $\NumDimensions$ entries. 
+%In Appendix~\ref{app:symbols}
+%we list all mathematical symbols used in this work and their description.
 In this model each data point can be represented as being drawn
@@
 
-\todo{this model of orthogonality takes the independence into account}
-
-Another benefit of using a mixture of common factor analysers is
-for modelling stellar chemical abundances is the scaling with
-computational cost. If we considered data sets of order $3\times10^7$
+Including latent factors in the model description allows us to account for 
+processes that affect multiple elemental abundances. In this way we are 
+accounting for the fact that the data dimensions are not all independent of
+each other. Another benefit is the scaling with computational cost. If we 
+considered data sets of order $3\times10^7$
 entries (e.g., 30 chemical abundances for $10^6$ stars) purely as a
@@
 $\NumComponents$ components by drawing from a multinomial distribution
-and generating the responsibility matrix $\vec\tau$. Given the initial
+and generating the $\NumData \times \NumComponents$ responsibility matrix $\vec\tau$,
+where $\sum_{k=1}^{K} \tau_{nk} = 1$ for any $n$th row. Given the initial
 factor loads and assignments, we then estimate the relative weights
@@
 \end{equation}
+
+These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
+Unsurprisingly the log likelihood increases with increasing numbers of latent
+factors $\NumLatentFactors$ and increasing numbers of components $\NumComponents$.
+The lowest BIC value is found at $\NumLatentFactors = 5$
+and $\NumComponents = 21$, close to the true values ($\NumLatentFactors_{true} = 5$,
+$\NumComponents_{true} = 20$).
+
 %\noindent{}and the pseudo-Bayesian Information Criterion \citep{pseudo-bic}
@@
 %$\omega \geq 1$ and $\gamma > 0$. Here we adopt $\gamma = 0.1$ and $\omega = 1$.
-These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
 
@@
 \end{figure}
-
-Unsurprisingly the log likelihood increases with increasing numbers of latent
-factors $\NumLatentFactors$ and increasing numbers of components $\NumComponents$.
-The lowest BIC value is found at $\NumLatentFactors = 5$
-and $\NumComponents = 8$, close to the true values ($\NumLatentFactors_{true} = 5$,
-$\NumComponents_{true} = 10$). The lowest pseudo-BIC value is also found at
-$\NumLatentFactors = 5$, but with fewer components: $\NumComponents = 4$.
-
-The estimated latent factor loads found during this grid search are shown in
-Figure~\ref{fig:fig:experiment-1-gridsearch-factors-init-random-and-random},
-where each panel shows the estimated latent factor loads found by
-expectation-maximization for a given number of latent factors and components.
-
-The latent factors $\factorloads$ in each panel are multiplied by a rotation
-matrix $\mathbf{R}$ to be as similar (up to orthogonal rotation) as those found
-from the same number of latent factors and $\NumComponents  =  1$.
-In doing so we find that the estimated latent factors vary slightly as
-the number of components $\NumComponents$ increases.
 
@@
 
-\subsection{Experiment~2: Inferring astrophysically realistic latent factors}
+\subsection{Experiment~2:\\Inferring astrophysically realistic latent factors}
 
@@
 
-While it is trivial to find the rotation matrix $\vec{R}$ that rotates the estimated
-factor loads to the true factor loads, this property of factor analyzers limits
-how we can interpret the latent variables of our model. For example, if we seek
-to interpret the factor loads as being something analagous to nucleosynthetic
-processes, then the estimated factor loads will need to be rotated to something
-that looks astrophysically plausible before we could attribute one latent factor
-as being a particular nucleosynthetic process.
+If the true values are known then it is trivial to find a rotation matrix $\vec{R}$
+that rotates the estimated factor loads to the true factor loads. However, when
+the true values are not known, this property of factor analysis limits the
+extent to which we can interpret the latent variables in our model. For example,
+if we seek to interpret the factor loads as being something analagous to 
+nucleosynthetic processes, then the estimated factor loads will need to be 
+rotated to something that looks astrophysically plausible before we could 
+attribute one latent factor as being a particular nucleosynthetic process. 
+Although having to perform this rotation is not ideal, there is still utility in
+trying to identify the latent factors because the factors can only be rotated by
+some valid rotation matrix. That is to say that we will not only exactly what
+we have put in: the rotated factors will simply be the closest set of factor
+loads that are consistent with the data.
 
 In this experiment we use detailed chemical abundances of metal-poor stars as
-reported by \citet{Barklem:2003}, and we demonstrate the limitations in
-interpreting the latent factors as nucleosynthetic processes. The data contains
-14 measurements of chemical abundances for 61 stars ($N = 61$, $D = 14$): 
+reported by \citet{Barklem:2003}, and we demonstrate the utility and limitations
+in interpreting the latent factors as nucleosynthetic processes. The data 
+contains 14 measurements of chemical abundances for 61 stars ($N = 61$, $D = 14$): 
 Al (light odd-Z), Mg and Ca ($\alpha$), Sc, Ti, Cr, Mn, Fe, Co, and Ni (Fe-peak),
@@
 the model parameters randomly for each permutation of $J$ and $K$, and ran E-M
-until a convergence threshold of $10^{-5}$ was reached. In Figure~\ref{fig:exp2-gridsearch-contours}
-we show the negative log likelihood and BIC recorded for the optimised set of
-model parameters for each permutation of $J$ and $K$. Hyperparameters of $J = 5$
-and $K = 1$ are preferred by BIC for this data set.
+until a convergence threshold of $10^{-5}$ was reached in the log likelihood. 
+In Figure~\ref{fig:exp2-gridsearch-contours} we show the negative log likelihood 
+and BIC recorded for the optimised set of model parameters for each permutation 
+of $J$ and $K$. The lowest BIC value is found for $J = 5$ and $K = 1$, indicating
+five latent factors are preferred, and no clustering.
 
@@
 
-After demonstrating that there is some interpretability in the latent factors,
-here we seek to experiment whether gravitationally bound stellar clusters can
-be reliably identified using the model described in Section~\ref{sec:model}.
+The previous experiment has demonstrated that the latent factors can be at least
+partially interpreted within the context of astrophysical processes. In this
+experiment we seek to test whether gravitationally bound stellar clusters can be
+reliably identified by their chemical abundances alone, and using only the model 
+described in Section~\ref{sec:model}.
 
-\subsection{Experiment 4: A subset of 14 chemical abundances in \Galah}
-\label{sec:experiment-galah}
+\todo{description  of the open cluster data used here}
+
+\subsection{Experiment 4: Chemical abundances in the \Galah\ survey}
+\label{sec:exp4}
 
@@
 
-We included the following fourteen chemical abundances: Na, Mg, 
-Si, Ca, Sc, Ti, Cr, Mn, Fe, Ni, Cu, Zn, Ba, and Eu.
+We included the following fourteen chemical abundances: \todo{Na, Mg, 
+Si, Ca, Sc, Ti, Cr, Mn, Fe, Ni, Cu, Zn, Ba, and Eu}.
 By requiring that each star has \texttt{flag\_cannon = 0} and
@@
 for each chemical abundance, this leaves
-us with a sample of 12,798 stars each with reliable stellar
+us with a sample of \todo{12,798} stars each with reliable stellar
 parameters ($\teff$, $\logg$) and chemical abundances.
@@
 
-%\subsubsection{Search strategy} \label{sec:experiment-galah-search}
+\todo{we perform a grid-based search for J and K}
 
-The expectation-maximization procedure outlined in Section \ref{sec:methods} describes 
-the process of refining our estimate of the model parameters $\vec\psi$ until
-the improvement in log probability between iterations decreases below some
-pre-defined threshold. This method is valid for a pre-defined number of latent
-factors, and a pre-defined number of components to cluster in latent space.
-In practice these values are not known. For the purposes of this work, we 
-perform a grid-based search in $J$ and $K$, perform E-M for each combination of
-$J$ and $K$, and record the optimal parameters.
+\subsection{Experiment 5: More chemical abundances in \Galah}
+\label{sec:exp5}
 
-We trial the number of latent factors from $J = 1$ to $J = 12$, and trial the
-number of clustering components from $K = 1$ to $K = 40$. We re-initialise the
-mixture for each new trial of $\{J, K\}$, and do not make use of results from
-`nearby' mixtures. After the E-M threshold of $10^{-5}$ has been reached, we
-record the log likelihood of the data given the model parameters, the
-Bayesian Information Criterion \citep{bic}, 
-\begin{equation}
-	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \label{eq:bic}
-\end{equation} 
-\noindent{}and the pseudo-Bayesian Information Criterion \citep{pseudo-bic}
-\begin{equation}
- \textrm{pseudo-BIC} = 6(1 + \gamma)\omega{}Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \label{eq:pseudo-bic}
-\end{equation}
-\noindent{}where $Q$ is the total number of model parameters in both equations,
-and the unknown constants $\omega$ and $\gamma$ have constraints
-$\omega \geq 1$ and $\gamma > 0$. Here we adopt $\gamma = 0.1$ and $\omega = 1$.
-
-\todo{Discuss grid search, inferred factor loads, cluster variances, etc.}
-
-\begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/galah-experiment-2-gridsearch-ll-contours.png}
-	\includegraphics[width=0.45\textwidth]{experiments/galah-experiment-2-gridsearch-bic-contours.png}
-	\includegraphics[width=0.45\textwidth]{experiments/galah-experiment-2-gridsearch-pseudobic-contours.png}
-    \caption{Model performance metrics for Experiment 2 using a subset of 14 
-    		 \Galah\ chemical abundances measured from 12,798 stars. The top 
-		 	 panel shows the negative log-likelihood 
-			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
-			 evaluated at each combination of latent factors $J$ and number 
-			 of clusters $K$, the middle panel shows the Bayesian Information
-			 Criterion (Equation \ref{eq:bic}), and the lower panel shows
-			 the pseudo-Bayesian Information Criterion 
-			 (Equation~\ref{eq:pseudo-bic}). The white marker indicates the 
-			 lowest value in each panel.}
-    \label{fig:experiment-2-gridsearch}
-\end{figure}
-
-\section{Results} \label{sec:results}
+\todo{which abundances, only require finite and \texttt{flag\_cannon} to be OK}
 
@@
 
+\todo{
+\begin{itemize}
+	\item dimensionality of chemical abundance space
+	\item clustering approaches do not take this into account
+	\item introduced a model to simultaneously account for the clustering in
+		  a lower dimensional latent space
+	\item there are issues, though: this is probably  not the right model
+	\item identifiability
+	\item orthogonal rotation
+	\item can interpret these somehow as being astrophysically motivated
+	\item at low metallicities how many factors do we prefer, and what can we
+		  vaguely interpret these at?
+	\item when we do that to large galah data sets, what is the number of factors
+	\item clustering in galah
+\end{itemize}
+}
 \section{Conclusions} \label{sec:conclusion}

### 2018-12-18 article/ms.tex
words

@@
 
-Chemical tagging describes the process of identifying individual star formation
-sites based on the present day photospheric abundances of stars \citep{Freeman;Bland-Hawthorn:2001}.
-Since the dynamical dissipation timescale for open clusters is relatively short
-\citep[\todo{timescale}][]{}, the ability to identify star formation sites long
-after the stars are no longer gravitationally bound could provide key insights
-on the early formation of the disc of the Milky Way galaxy. For this reason,
-much effort has been placed on obtaining observational data, methods for analysis,
-theory, etc, in order to improve the prospects of chemical tagging.
+Chemical tagging seeks to identify star formation events using the present
+day photospheric abundances of stars \citep{Freeman;Bland-Hawthorn:2001}.
+The detailed chemical abundances observable in a star's photosphere are a
+fossil record that carries with it information about where and when a star
+formed. While the photospheric abundances remain largely unchanged throughout
+a star's lifetime \citep[however see][]{}, the dynamical dissipation timescale
+of open clusters is only about 1~Gyr \citep{timescale}. That makes chemical
+tagging an attractive approach to identify star formation sites long after
+those stars are no longer gravitationally bound.
+
+Gravitationally bound star clusters have been useful laboratories for
+testing the limits and utility of chemical tagging. Early works 
+works examined metrics quantifying the similarity of chemical abundances
+in gravitationally bound star clusters, and more
+
+ at the similarity of chemical abundance metrics 
+Theoretical works have looked at the dimensionality
+
+From
+theoretical studies examining the
+
+ the dynamical dissipation timescale of open clusters is only
+about 1~Gyr \citep{timescale} and the photospheric abundances remain largely
+unchanged throughout a star's lifetime \citep[however see][]{}, the ability
+to identify star formation sites long after those stars are no longer
+gravitationally bound.
 
@@
 
-\subsubsection{Given the true number of latent factors and components}
+%\subsubsection{Given the true number of latent factors and components}
 
-We initialised the factor loads and component assignments randomly, and
-fit this generated data set using the expectation-maximization algorithm
-as described in the previous Section. After \todo{58} iterations
-(and a serial processing time of \todo{40} seconds) we
-reached our convergence threshold: the log probability increased by
-less than $10^{-5}$ with each iteration (Figure~\ref{fig:exp1-ll-iterations}). For this experiment we specified
-the number of latent factors $\NumLatentFactors$ and the number of clustering components
-$\NumComponents$ to include, but below we repeat this experiment where the number
-of latent factors and clustering components is not known.
+%We initialised the factor loads and component assignments randomly, and
+%fit this generated data set using the expectation-maximization algorithm
+%as described in the previous Section. After \todo{58} iterations
+%(and a serial processing time of \todo{40} seconds) we
+%reached our convergence threshold: the log probability increased by
+%less than $10^{-5}$ with each iteration (Figure~\ref{fig:exp1-ll-iterations}). For this experiment we specified
+%the number of latent factors $\NumLatentFactors$ and the number of clustering components
+%$\NumComponents$ to include, but below we repeat this experiment where the number
+%of latent factors and clustering components is not known.
 
@@
 
-We show a corner plot of the data in Figure~\ref{fig:toy-model-data}.
-In this figure we have over-plotted 100000 samples generated by the
-estimated factor loads $\factorloads_\textrm{est}$ and the clustering in factor 
-scores $\scoremeans_\textrm{est}$ and $\scorecovs_\textrm{est}$ (weighted by $\vec\weight_\textrm{est}$)
--- without adding variance $\specificvariance_\textrm{est}$ % Ref Factor load figure 
--- in order to illustrate that the
-estimated model parameters can realistically generate the data.
-The estimated latent space of factor scores is shown in 
-Figure~\ref{fig:toy-model-latent-space} where the colours indicate the
-most likely component that the data point is associated with.
-
-While the estimated model parameters can generate the data and demonstrate
-clustering in latent space, caution is needed when interpreting the 
-model latent variables.
-Latent factor models are by definition indeterminate in that there are
-multiple solutions of the parameter vector $\vec\Psi$ which provide
-an identical data-generating process. This can be trivially shown by
-the fact that the factors loads and scores can be rotated by taking the
-dot product
-with some rotation
-matrix $\textbf{R}$ and generate the exact same data. 
-We discuss how this limitation applies to later experiments.
+%While the estimated model parameters can generate the data and demonstrate
+%%clustering in latent space, caution is needed when interpreting the 
+%model latent variables.
+%Latent factor models are by definition indeterminate in that there are
+%multiple solutions of the parameter vector $\vec\Psi$ which provide
+%an identical data-generating process. This can be trivially shown by
+%the fact that the factors loads and scores can be rotated by taking the
+%dot product
+%with some rotation
+%matrix $\textbf{R}$ and generate the exact same data. 
+%We discuss how this limitation applies to later experiments.
 
@@
 
-\subsubsection{Grid search for the number of latent factors and components}
+%\subsubsection{Grid search for the number of latent factors and components}
 
@@
 
+We show a corner plot of the data in Figure~\ref{fig:toy-model-data}.
+In this figure we have over-plotted 100000 samples drawn from the model
+with $J = 5$ and $K = 21$
+-- without adding variance $\specificvariance_\textrm{est}$ % Ref Factor load figure 
+-- in order to illustrate that the
+estimated model parameters can realistically generate the data.
+The estimated latent space of factor scores is shown in 
+Figure~\ref{fig:exp1-latent-space} where the colours indicate the
+most likely component that the data point is associated with.
+
 %\noindent{}and the pseudo-Bayesian Information Criterion \citep{pseudo-bic}
@@
 	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-bic-contours.png}
-%	\includegraphics[width=0.45\textwidth]{experiments/toy-pseudobic-contours.png}
     \caption{Model performance metrics resulting from a gridsearch using the

### 2018-12-18 article/ms.tex
introduction + citations

@@
 Chemical tagging seeks to identify star formation events using the present
-day photospheric abundances of stars \citep{Freeman;Bland-Hawthorn:2001}.
-The detailed chemical abundances observable in a star's photosphere are a
+day photospheric abundances of stars \citep{Freeman;Bland-Hawthorn:2002}.
+The detailed chemical abundances observable in a star's photosphere provide a
 fossil record that carries with it information about where and when a star
 formed. While the photospheric abundances remain largely unchanged throughout
-a star's lifetime \citep[however see][]{}, the dynamical dissipation timescale
-of open clusters is only about 1~Gyr \citep{timescale}. That makes chemical
-tagging an attractive approach to identify star formation sites long after
-those stars are no longer gravitationally bound.
+a star's lifetime \citep[however see][]{Dotter:2017,Ness:2018b}, the dynamical 
+dissipation timescale of open clusters in the Milky Way disc is of order a few 
+gigayears \citep{Portegies-Zwart:1998}. That makes chemical tagging an attractive 
+approach to identify star formation sites long after those stars are no longer 
+gravitationally bound.
 
 Gravitationally bound star clusters have been useful laboratories for
-testing the limits and utility of chemical tagging. Early works 
-works examined metrics quantifying the similarity of chemical abundances
-in gravitationally bound star clusters, and more
+testing the limits and utility of chemical tagging. Although biases arise when
+only considering star clusters that are still gravitationally bound, the chemical
+homogenity of open clusters provides an empirical measure of how similar stars
+would need to be before they could be tagged as belonging to the same
+star formation site \citep{Mitschang:2014}. However, there are also analysis
+issues in understanding how well those chemical abundances can be measured
+\citep{Bovy:2016}. If open clusters were truely chemically homogeneous then our
+ability to chemically tag the Milky Way depends on the precision with which we
+can measure those chemical abundances in stars. Data-driven techniques are
+driving a revolution in this field \citep{Ness:2015,Ness:2018a,Ness:2018b,
+Casey:2016,Casey:2017,Ho:2017b,Ho:2017a,Leung;Bovy:2018}, but there is more work
+to do: astronomers have not yet reached the Cram\'er-Rao bound for chemical 
+abundances of stars.
 
- at the similarity of chemical abundance metrics 
-Theoretical works have looked at the dimensionality
-
-From
-theoretical studies examining the
-
- the dynamical dissipation timescale of open clusters is only
-about 1~Gyr \citep{timescale} and the photospheric abundances remain largely
-unchanged throughout a star's lifetime \citep[however see][]{}, the ability
-to identify star formation sites long after those stars are no longer
-gravitationally bound.
-
-The \Galah\ survey \citep{DaSilva:2015} is a stellar spectroscopic survey 
-that uses the High Efficiency and Resolution Multi-Element 
-Spectrograph \citep[HERMES][]{Sheinis:2016} on the Australian Astronomical Telescope (AAT).
-The survey will observe up to $10^6$ stars in the Milky Way, and
-measure up to 30 chemical abundances for each star. This includes
+There are further chemical tagging obstacles to overcome once all the chemical 
+abundance information in a stellar spectrum can be extracted. Most notably is a 
+catalogue of precise detailed chemical abundances for a large number of stars,
+where those chemical abundances trace different nucleosynthetic pathways.
+This is the primary scientific goal of the \Galah\ survey \citep{DeSilva:2015,Martell:2017,Buder:2018},
+a stellar spectroscopic survey that uses the High Efficiency and Resolution 
+Multi-Element Spectrograph \citep[\project{HERMES};][]{Sheinis:2015} on the Australian 
+Astronomical Telescope (AAT).  \Galah\ will observe up to $10^6$ stars in the 
+Milky Way, and measure up to 30 chemical abundances for each star. This includes
 light odd-Z elements (e.g., Na, K), elements produced through
@@
 
-\todo{early approaches on clustering metrics}
+Given these data and the most
+favourable assumptions in chemical tagging -- that star clusters are truely
+chemically homogenous, and we can measure those abundances with infinite precision,
+and those abundances are differentiable between star clusters -- then chemical
+tagging becomes a clustering problem. All clustering techniques applied to 
+chemical tagging assume that the data dimensions are independent. That is to say
+that adding a dimension of [Ni/H] provides totally new, independent information.
+Theory and observations agree that this cannot be true.
+Nucleosynthetic processes produce multiple elements in varying
+quantities, and the dimensionality of stellar abundance datasets has been shown
+to be lower than the actual number of dimensions \citep{Ting:2012, Price-Jones:2018}.
+Any clustering approach that treats each new elemental abundance as an 
+indenpendent axis of information will therefore conclude with biased inferences
+about the star formation history of our Galaxy.
 
-\todo{work on lower limits of chemical homogeneity}
-
-All clustering techniques applied to chemical tagging assume that the data
-dimensions are independent. However, theory and observations agree this cannot
-be true. Nucleosynthetic processes produce multiple elements in varying
-quantities, and the dimensionality of stellar abundance datasets has been shown
-to be lower than the actual number of dimensions. Any clustering approach that
-treats each new elemental abundance as an indenpendent axis of information will
-therefore conclude with biased inferences about the star formation history of
-our Galaxy. On the other hand, it is difficult to prescribe the nucleosynthetic
-yields that have contributed to the chemical abundances of each star. There are
+Alternative approaches are perhaps more challenging, as it is difficult to prescribe
+the nucleosynthetic yields that have contributed to the chemical abundances of each star
+with great confidence. There are
 qualitative statements that can be made for large numbers of stars, or particular
@@
 
-In this work we present a new approach to chemical tagging that allows us to 
-identify the latent factors that contribute to the chemical abundances of all 
-stars (e.g., something akin to nucleosynthetic yields), while simultaneously
-performing clustering in the latent space. Notwithstanding caveats that we will
+New approaches to chemical tagging are clearly needed. Immediate advances would
+include methods that take the dependence among chemical elements into account
+within some generative model, or techniques that combine chemical abundances
+with dynamical constraints to place joint prior probabilities on whether any
+two stars could have formed from the same star cluster, given some model of the
+Milky Way. In this work we focus on the former. Here we present a new approach
+to chemical tagging that allows us to identify the latent (unobserved) factors
+that contribute to the chemical abundances of all stars (e.g., nucleosythetic
+yields) while simultaneously performing clustering in the latent space.
+Notwithstanding caveats that we will
 discuss in detail, this allows us to infer nucleosynthetic yields rather than
@@
 of these experiments in Section~\ref{sec:discussion}, including the caveats with
-the model as described. We conclude in Section~\ref{sec:conclusion}
+the model as described. We conclude in Section~\ref{sec:conclusion}.
 
@@
 Factor analysis is a common statistical approach for describing correlated 
-observations with a lower number of latent variables \citep{factor-analysis}.
-Related techniques include principal component analysis \citep{pca} and its
-variants \citep{ppca}, singular value decomposition \citep{svd}, and other
+observations with a lower number of latent variables \citep[e.g.,][]{Thompson:2004}.
+Related techniques include principal component analysis \citep{Hotelling:1933} and its
+variants \citep{Tipping;Bishop:1999}, singular value decomposition \citep{Golub:1970}, and other
 matrix factorization methods. While factor analysis on its own is a useful
 dimensionality reduction tool to identify latent factors that contribute to
-the chemical abundances of stars, it does not perform any clustering.
+the chemical abundances of stars \citep[e.g.,][]{Price-Jones:2018}, it does not perform any clustering.
 Similarly, clustering techniques applied to chemical abundances to date 
@@
 
-Here we use a combination of factor analysis and a gaussian mixture model to
-provide a consistent description of the data, given the latent variables and 
-their clustering in latent space. In this work the data $\vecdata$ is a 
+Here we use a variant of factor analysis known elsewhere as a mixture of common 
+factor analyzers \citep{Baek:2010}, where the data are generated by a set of 
+latent factors that are common to all data, but the scoring (or extent) of those
+factors is different for each data point, and the data can be modelled as a
+mixture of multivariate normal distributions in the latent space (factor scores).
+In this work the data $\vecdata$ is a 
 $\NumData \times \NumDimensions$ matrix where $\NumData$ is the number of 
@@
 recorded the log-likelihood of the data given the model parameters, and the
-Bayesian Information Criterion \citep{bic}, 
+Bayesian Information Criterion \citep{Schwarz:1978}, 
 \begin{equation}
@@
 In this experiment we use detailed chemical abundances of metal-poor stars as
-reported by \citet{Barklem:2003}, and we demonstrate the utility and limitations
+reported by \citet{Barklem:2005}, and we demonstrate the utility and limitations
 in interpreting the latent factors as nucleosynthetic processes. The data 
@@
     \caption{Model performance metrics resulting from a gridsearch using the
-    		 \citet{Barklem:2003} data in Experiment~2.
+    		 \citet{Barklem:2005} data in Experiment~2.
     		 The top panel shows the negative log-likelihood 
@@
 	\includegraphics[width=\textwidth]{experiments/exp2-latent-factors.png}
-	\caption{Rotated latent factors found in Experiment~2 using the \citet{Barklem:2003}
+	\caption{Rotated latent factors found in Experiment~2 using the \citet{Barklem:2005}
 			 data. \todo{elaborate latent factors}}
@@
 	\includegraphics[width=0.45\textwidth]{experiments/exp2-specific-scatter.png}
-	\caption{Specific scatter remaining in the \citet{Barklem:2003} data used in
+	\caption{Specific scatter remaining in the \citet{Barklem:2005} data used in
 			 Experiment~2 after accounting for the contributions by all latent

### 2018-12-19 article/ms.tex
foo

@@
 orthogonality. However, accurate estimates of the parameter vector $\vec\Psi$
-can be obtained by the expectation-maximization algorithm \citep{EM}.
+can be obtained by the expectation-maximization algorithm \citep{Dempster:1977}.
 %The model described by Equation~\ref{eq:generative-model} is indeterminate:
@@
 
-We use the expectation-maximization algorithm to estimate the model parameters.
-With each expectation step we evaluate the log likelihood given the model 
-parameters $\vec\Psi$, and we re-calculate the responsibility 
-matrix $\vec\tau$ whose entries are the posterior probability
-that the $\numdata$th data point is associated to the $\numcomponents$th component,
-given the data $\vecdata$ and the current estimate of the 
-parameter vector $\vec\Psi$:
+We use the expectation-maximization algorithm to estimate the model parameters
+\citep{Dempster:1977}. With each expectation step we evaluate the log likelihood 
+given the model parameters $\vec\Psi$, and we re-calculate the responsibility 
+matrix $\vec\tau$ whose entries are the posterior probability that the 
+$\numdata$th data point is associated to the $\numcomponents$th component, given 
+the data $\vecdata$ and the current estimate of the parameter vector $\vec\Psi$:
 \begin{equation}
@@
 
-\noindent{}where $\odot$ denotes is the entry-wise (Hadamard) product. We repeated
-the expectation-maximization algorithm for up to 1,000 steps or until the log-likelihood
+\noindent{}where $\odot$ denotes is the entry-wise (Hadamard) product. We repeat
+the expectation-maximization cycle for up to 10,000 steps or until the log-likelihood
 improved by less than $10^{-5}$ between successive iterations. Although our implementation
@@
 until $\NumLatentFactors_{max} = 10$
-and   $\NumComponents_{max} = 40$ (e.g., twice $\NumLatentFactors_{true}$ and $\NumComponents_{true}$).
+and   $\NumComponents_{max} = 40$ (e.g., twice $\NumLatentFactors_\textrm{true}$ and $\NumComponents_\textrm{true}$).
 For each $\NumLatentFactors,\NumComponents$ permutation we initialised the factor
@@
 recorded the log-likelihood of the data given the model parameters, and the
-Bayesian Information Criterion \citep{Schwarz:1978}, 
+Bayesian Information Criterion \citep[BIC][]{Schwarz:1978}, 
 \begin{equation}
-	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \quad . \label{eq:bic}
+	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \quad , \label{eq:bic}
+\end{equation} 
+\noindent{}where $Q$ is the number of model parameters:
+\begin{equation}
+	%Q = \NumComponents + \NumDimensions - 1 + \NumLatentFactors(\NumDimensions + \NumComponents) + \frac{1}{2}\NumComponents\NumLatentFactors(\NumLatentFactors+1) - \NumLatentFactors^2 \quad .
+	Q = \frac{\NumLatentFactors}{2}\left[2\left(\NumDimensions - \NumLatentFactors\right) + \NumComponents\left(3 + \NumLatentFactors\right)\right] + \NumComponents + \NumDimensions - 1
 \end{equation}
@@
 The lowest BIC value is found at $\NumLatentFactors = 5$
-and $\NumComponents = 21$, close to the true values ($\NumLatentFactors_{true} = 5$,
-$\NumComponents_{true} = 20$).
+and $\NumComponents = 21$, close to the true values ($\NumLatentFactors_\textrm{true} = 5$,
+$\NumComponents_\textrm{true} = 20$).
 
-We show a corner plot of the data in Figure~\ref{fig:toy-model-data}.
+We show a corner plot of the data in Figure~\ref{fig:exp1-data}.
 In this figure we have over-plotted 100000 samples drawn from the model

### 2019-01-16 article/mcfa.tex
words

--- 2018-12-19 article/ms.tex
@@
-\documentclass[twocolumn]{aastex61}
+\documentclass[twocolumn]{aastex62}
 \usepackage{bm}
@@
 \begin{abstract}
-Chemical tagging promises to distinguish unique star formation sites based on the
-present day photospheric abundances of stars. Clustering techniques used to date
-have treated all chemical abundances as independent, which we know is not true.
+Chemical tagging promises to distinguish unique star formation sites from 
+photospheric abundances.
+Clustering techniques used in chemical tagging have to date considered 
+chemical abundances as independent, a provably false assumption.
 Nucleosynthetic processes make multiple elements in varying quantities, and the
-dimensionality of chemical abundance space is typically lower than the reported
-number of abundances. Here we introduce a novel approach to chemical tagging 
-where a set of latent factors (e.g., nucleosynthetic yields) contribute to the
-data, and there is clustering in the latent space (e.g., relative counts of
-nucleosynthetic events). This approach allows us to infer contributions akin to
-nucleosynthetic yields, and the clustering becomes more efficient as it occurs in
-a lower dimensional latent space.
-We run experiments on literature chemical abundances,
-including the second \Galah\ data release. We find that 5 latent factors are
+dimensionality of chemical abundance space is usually lower than the 
+number of reported abundances. 
+Here we introduce a novel approach to chemical tagging where a set of 
+latent factors (e.g., nucleosynthetic yields) contribute to the
+data, and clustering is performed in the latent space (e.g., relative counts of
+nucleosynthetic events). In doing so we infer basis vectors that are analogous
+to nucleosynthetic yields, and the lower dimensional latent space allows for
+more efficient clustering. We run experiments on literature chemical abundances,
+including those from the second \Galah\ data release. \todo{We find that 5 latent factors are
 preferred to explain the chemical abundances of metal-poor stars, and 8 are
-needed to account for \Galah\ data.
-\todo{something about the clusters or factors found}
+needed to account for \Galah\ data.}
 We release code associated with this \article\ that scales well with the 
-available data: enabling chemical tagging for $10^{6}$ stars on a standard
-desktop machine in minutes.
+available data: enabling chemical tagging for $10^{7.5}$ abundance
+measurements on a standard desktop machine in minutes.
+
 \end{abstract}
@@
 approach to identify star formation sites long after those stars are no longer 
-gravitationally bound.
+gravitationally bound to each other.
 
@@
 only considering star clusters that are still gravitationally bound, the chemical
-homogenity of open clusters provides an empirical measure of how similar stars
+homogeneity of open clusters provides an empirical measure of how similar stars
 would need to be before they could be tagged as belonging to the same
@@
 issues in understanding how well those chemical abundances can be measured
-\citep{Bovy:2016}. If open clusters were truely chemically homogeneous then our
+\citep{Bovy:2016}, and how chemically similar multiple unrelated stars (or
+doppleg\"angers) can be \citep{Ness:2018}.
+If open clusters were truely chemically homogeneous then our
 ability to chemically tag the Milky Way depends on the precision with which we
@@
 Any clustering approach that treats each new elemental abundance as an 
-indenpendent axis of information will therefore conclude with biased inferences
+independent axis of information will therefore conclude with biased inferences
 about the star formation history of our Galaxy.
@@
 to chemical tagging that allows us to identify the latent (unobserved) factors
-that contribute to the chemical abundances of all stars (e.g., nucleosythetic
+that contribute to the chemical abundances of all stars (e.g., nucleosynthetic
 yields) while simultaneously performing clustering in the latent space.
@@
 dimensionality reduction tool to identify latent factors that contribute to
-the chemical abundances of stars \citep[e.g.,][]{Price-Jones:2018}, it does not perform any clustering.
+the chemical abundances of stars \citep[e.g.,][]{Price-Jones:2018}, factor
+analysis cannot describe clustering in the data (or latent) space.
 Similarly, clustering techniques applied to chemical abundances to date 
@@
 diagonal matrix of $\NumDimensions$ entries. 
-%In Appendix~\ref{app:symbols}
-%we list all mathematical symbols used in this work and their description.
 In this model each data point can be represented as being drawn
@@
 identified up to orthogonality and cannot be expressly interpreted as
-nucleosynthetic yields because they have limited physical meaning,
+nucleosynthetic yields because they have limited physical meaning
+(we discuss this further in Section~\ref{sec:discussion}),
 but this description of typical yields and relative event rates should
@@
 processes that affect multiple elemental abundances. In this way we are 
-accounting for the fact that the data dimensions are not all independent of
+accounting for the fact that the data dimensions are not independent of
 each other. Another benefit is the scaling with computational cost. If we 
-considered data sets of order $3\times10^7$
+considered data sets of order $10^{7.5}$
 entries (e.g., 30 chemical abundances for $10^6$ stars) purely as a
@@
 $N \times D$ to $N \times J$. This reduces computational cost through
-faster execution of each E-M step, and on average fewer E-M steps
+faster execution of each optimization step, and on average fewer optimization steps
 needed to reach a specified convergence threshold.
@@
 turbulence and gas mixing, or star formation efficiency,
-%can account
-%analogues of nucleosynthetic yields and star formation,
 and the parameters of this model can be estimated consistently
@@
 \begin{equation}
-	\log\mathcal{L}(\vecdata|\vec\Psi) = \sum_{\numcomponents=1}^{\NumComponents}\log{f(\vecdata;\vec\Psi)} \quad .
+	\log\mathcal{L}(\vecdata|\vec\Psi) = \sum_{\numcomponents=1}^{\NumComponents}\log{f(\vecdata;\vec\Psi)} \quad . \label{eq:log-likelihood}
 \end{equation}
@@
 $\factorscores$. These quantities can only be determined up until 
-orthogonality. However, accurate estimates of the parameter vector $\vec\Psi$
-can be obtained by the expectation-maximization algorithm \citep{Dempster:1977}.
-%The model described by Equation~\ref{eq:generative-model} is indeterminate:
-%there is no unique solution for the factor loads $\factorloads$ and scores
-%$\factorscores$. However, an accurate estimate of the parameter vector $\vec\Psi$ can be obtained by the expectation-maximization algorithm \citep{EM}.
+orthogonality in $\factorloads$. However, as we will show, accurate estimates
+of the parameter vector $\vec\Psi$ can be obtained by the expectation-maximization
+algorithm \citep{Dempster:1977}. 
+% TODO: say something about "..with a suitable number of initial guesses?"
 
@@
 \end{equation}
-%        AL = linalg.cholesky(A.T @ A)
-%        A = A @ linalg.solve(AL, np.eye(self.n_latent_factors))
-% 	
 We then initially assign each data point as belonging to one of the
@@
 Finally, we initialise the specific variance $\specificvariance$ in each
-dimension as 1.
-
-Other initialisation methods are available in the code 
-associated with this \emph{Article}. In later Sections we show that 
-the optimised model parameters are insensitive to the choice initialisation, 
-up until orthogonal rotation of the latent space.
+dimension as the variance in each data dimension. Other initialisation 
+methods are available in the code associated with this \emph{Article}. 
+Throughout this work we found that the optimised model parameters are 
+insensitive to the choice of initialisation, up until orthogonal 
+rotation of the latent space.
 
@@
 \citep{Dempster:1977}. With each expectation step we evaluate the log likelihood 
-given the model parameters $\vec\Psi$, and we re-calculate the responsibility 
+given the model parameters $\vec\Psi$, and we re-calculate the $\NumData \times \NumComponents$ responsibility 
 matrix $\vec\tau$ whose entries are the posterior probability that the 
@@
 conditioned on the data $\vecdata$ and the responsibility matrix $\vec\tau$.
-First we estimate the relative weights $\vec\weight\nextstep$ given
-the responsibility matrix $\vec\tau$
+The updated parameters estimates are found by setting the second derivative
+of the log-likelihood (Eq.~\ref{eq:log-likelihood}) to zero and solving for
+the parameter values. In doing so this guarantees that every updated
+estimate of the model parameters is guaranteed to increase the log-likelihood.
+Although there are no guarantees against converging on local minima, in 
+general it is sufficient to run expectation-maximization from multiple
+initialisations in order to ensure that the global minima is reached.
+At the maximization step we first update our estimate of the relative weights 
+$\vec\weight\nextstep$ given the responsibility matrix $\vec\tau$
 \begin{equation}
@@
 and $\vec{X}\nextstep$ refers to the updated estimate for the next iteration.
-%\footnote{Here $\vec\textrm{X}\now$ refers to the current estimate of a variable $\vec\textrm{X}$ and $\vec\textrm{X}\nextstep$ refers to the updated estimate for the next iteration.}
-
-%\noindent{}In the equations that follow (Equation~\ref{eq:X} to \ref{eq:Y})
-%when we refer to any value in $\{\factorloads,\vec\pi,\scoremeans,\scorecovs,\specificvariance\}$ for brevity we are referring to the current estimate of that
-%parameter (e.g., $\factorloads\thisstep$). The updated estimates are marked as such
-%(e.g., $\scoremeans\nextstep$).
 
@@
 	\vec{G}\transpose\vec\tau_\numcomponents\vec{Z}\transpose\vec{G}\right] \\
-	%\left(\scoremeans_k\thisstep\vecdata\transpose\vec\tau_k\right)\transpose \left(\vec{G}\transpose\vec{Z}\vec\tau_k\transpose\vec{G}\right)\transpose\right]  \\
 	\factorloads_2 &=& N\sum_{\numcomponents=1}^{\NumComponents}\left[\weight_\numcomponents\nextstep\left(\scorecovs_\numcomponents\nextstep + \scoremeans_\numcomponents\nextstep\left(\scoremeans_\numcomponents\nextstep\right)\transpose\right)\right]
@@
 
-We generated a data set with ${\NumData = 100000}$ data points, each with
+We generated a data set with ${\NumData = 1}$00,000 data points, each with
 $\NumDimensions = 15$ dimensions, and assumed that those data are generated by
@@
 distribution.
-The covariance matrices in factor scores $\scorecovs$ are assumed to be diagonal matrices, where the non-zero entries are drawn from a gamma distribution $\scorecovs_{\numcomponents,i,i} \sim \vec\Gamma\left(1\right)$.
+The off-diagonal entries in the covariance matrices in factor scores $\scorecovs$ are drawn from a gamma distribution $\scorecovs_{\numcomponents,i,i} \sim \vec\Gamma\left(1\right)$.
 The variance in
@@
 
-%\begin{figure*}
-%	\includegraphics[width=1.0\textwidth]{experiments/toy-model-latent-space.png}
-%    \caption{A corner plot showing all $\NumLatentFactors$ dimensions in
-%    		 latent space, where each data point is coloured by the component
-%		 	 that it is estimated to belong to.}
-%    \label{fig:toy-model-latent-space}
-%\end{figure*}
-
-%\subsubsection{Given the true number of latent factors and components}
-
-%We initialised the factor loads and component assignments randomly, and
-%fit this generated data set using the expectation-maximization algorithm
-%as described in the previous Section. After \todo{58} iterations
-%(and a serial processing time of \todo{40} seconds) we
-%reached our convergence threshold: the log probability increased by
-%less than $10^{-5}$ with each iteration (Figure~\ref{fig:exp1-ll-iterations}). For this experiment we specified
-%the number of latent factors $\NumLatentFactors$ and the number of clustering components
-%$\NumComponents$ to include, but below we repeat this experiment where the number
-%of latent factors and clustering components is not known.
-
-\begin{figure}
-	\includegraphics[width=0.5\textwidth]{experiments/exp1-ll-iterations.png}
-	\caption{Log-likelihood of the model given the data $\mathcal{L}\left(\vec\data|\vec\Psi\right)$ for Experiment 1 as a function of the number of expectation-maximization iterations.
-After \todo{X} iterations the improvement in log-likelihood per
-    expectation-maximization cycle was less than $10^{-5}$.}
-    \label{fig:exp1-ll-iterations}
-\end{figure}
-
-%While the estimated model parameters can generate the data and demonstrate
-%%clustering in latent space, caution is needed when interpreting the 
-%model latent variables.
-%Latent factor models are by definition indeterminate in that there are
-%multiple solutions of the parameter vector $\vec\Psi$ which provide
-%an identical data-generating process. This can be trivially shown by
-%the fact that the factors loads and scores can be rotated by taking the
-%dot product
-%with some rotation
-%matrix $\textbf{R}$ and generate the exact same data. 
-%We discuss how this limitation applies to later experiments.
-
-%This is further complicated by orientation and order
-%of the latent factors: even if the inferred factor loads were very close
-%to the true factor loads, they could be ordered differently, or have
-%their signs reversed. We discuss this further in later sections.
-
-%\subsubsection{Grid search for the number of latent factors and components}
+%\ref{fig:exp1-latent-space}
 
@@
 and   $\NumComponents_{max} = 40$ (e.g., twice $\NumLatentFactors_\textrm{true}$ and $\NumComponents_\textrm{true}$).
-For each $\NumLatentFactors,\NumComponents$ permutation we initialised the factor
+For each $\NumLatentFactors$, $\NumComponents$ permutation we initialised the factor
 loads and component assignments randomly. We performed expectation-maximization 
 cycles until the relative log-likelihood improved by less than $10^{-5}$. We then
-recorded the log-likelihood of the data given the model parameters, and the
-Bayesian Information Criterion \citep[BIC][]{Schwarz:1978}, 
+recorded the log-likelihood of the data given the model parameters, the
+Bayesian Information Criterion \citep[BIC;][]{Schwarz:1978}, 
 \begin{equation}
@@
 \begin{equation}
-	%Q = \NumComponents + \NumDimensions - 1 + \NumLatentFactors(\NumDimensions + \NumComponents) + \frac{1}{2}\NumComponents\NumLatentFactors(\NumLatentFactors+1) - \NumLatentFactors^2 \quad .
-	Q = \frac{\NumLatentFactors}{2}\left[2\left(\NumDimensions - \NumLatentFactors\right) + \NumComponents\left(3 + \NumLatentFactors\right)\right] + \NumComponents + \NumDimensions - 1
+	Q = \frac{\NumLatentFactors}{2}\left[2\left(\NumDimensions - \NumLatentFactors\right) + \NumComponents\left(3 + \NumLatentFactors\right)\right] + \NumComponents + \NumDimensions - 1 \quad .
 \end{equation}
+%\todo{\noindent{}and the Message Length.}
 
@@
 with $J = 5$ and $K = 21$
--- without adding variance $\specificvariance_\textrm{est}$ % Ref Factor load figure 
+-- without adding variance $\specificvariance_\textrm{est}$
 -- in order to illustrate that the
@@
 most likely component that the data point is associated with.
-
-%\noindent{}and the pseudo-Bayesian Information Criterion \citep{pseudo-bic}
-%\begin{equation}
-% \textrm{pseudo-BIC} = 6(1 + \gamma)\omega{}Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \label{eq:pseudo-bic}
-%\end{equation}
-%\noindent{}where $Q$ is the total number of model parameters in both equations,
-%and the unknown constants $\omega$ and $\gamma$ have constraints
-%$\omega \geq 1$ and $\gamma > 0$. Here we adopt $\gamma = 0.1$ and $\omega = 1$.
+%todo: number of correct associations?
 
@@
 
-Latent factors can only be identified up to orthogonality. That is to say that
+In this experiment we use detailed chemical abundances of metal-poor stars 
+\citep{Barklem:2005} to demonstrate the utility and limitations
+in interpreting the latent factors as nucleosynthetic processes. The data 
+can be described as 14 chemical abundance measurements from 61 stars ($\NumData = 61$, $\NumDimensions = 14$), that trace multiple nucleosynthetic pathways. Specifically the chemical
+abundances measured include Al (light odd-Z), Mg and Ca ($\alpha$), Sc, Ti, Cr, Mn, Fe, Co, and Ni (Fe-peak), as well as Sr, Y, Ba, and Eu ($s$/$r$-process).
+
+We performed a grid search for the optimal number of latent factors $\NumLatentFactors$ and 
+components $\NumComponents$, starting from $\NumLatentFactors = 1$ and $K = 1$, and trialling up until $\NumLatentFactors = \todo{8}$ and $\NumComponents = 5$. We initialised
+the model parameters randomly for each permutation of $J$ and $K$, and ran E-M
+until a convergence threshold of $10^{-5}$ was reached in the log likelihood. 
+In Figure~\ref{fig:exp2-gridsearch-contours} we show the negative log likelihood 
+and BIC recorded for the optimised set of model parameters for each permutation 
+of $J$ and $K$. The lowest BIC value is found for $J = 5$ and $K = 1$, indicating
+five latent factors are preferred, with no clustering.
+
+Some technical discussion is warranted before we interpret the estimated factor loads.
+In principle the latent factors are analogous to nucleosynthetic yields, but 
+these can only be identified up to an orthogonal rotation. That is to say that
 if the data were truely generated by latent factors $\factorloads_\textrm{true}$,
-then our estimates of those latent factors $\factorloads_\textrm{est}$ may not
-be identical to the true values. For example, the ordering of the estimated factors
+then our estimates of those latent factors $\factorloads_\textrm{est}$ do not need
+to be identical to the true values. For example, the ordering of the estimated factors
 could be different from the true factors, and the ordering of the dimensionality
-in latent space would be correspondingly different. Another possibility is that
-the estimated factor loads could be flipped in sign relative to the true factor
-loads, and the scores would similarly be flipped. In both of these situations
-(re-ordering or flipped signs) the log likelihood given the data and the
-estimated factor loads $\factorloads_\textrm{est}$ would be identical to the
-log likelihood given the data and the true factor loads $\factorloads_\textrm{true}$
-despite the difference in ordering and sign. These examples illustrate a more 
+in latent space would be correspondingly different. Since the ordering of the 
+factor loads is not prescribed during expectation-maximization, there is no 
+assurance (or requirement) that our factor loads match the true factor loads.
+
+Another possibility is that the estimated factor loads could be flipped in sign 
+relative to the true factor loads, and the scores would similarly be flipped. 
+In both of these situations (reordering or flipped signs) the log likelihood 
+given the data and the estimated factor loads $\factorloads_\textrm{est}$ would 
+be identical to the log likelihood given the data and the true factor loads 
+$\factorloads_\textrm{true}$
+despite the difference in ordering and sign. These examples serve to illustrate a more 
 general property that the factor loads and factor scores can be orthogonally 
-rotated by any rotation matrix\footnote{A rotation matrix is valid if 
+rotated by \emph{any rotation matrix}\footnote{A rotation matrix is valid if 
 $\vec{R}\vec{R}\transpose = \vec{I}$} $\vec{R}$. The estimated factor loads 
@@
 If the true values are known then it is trivial to find a rotation matrix $\vec{R}$
-that rotates the estimated factor loads to the true factor loads. However, when
+that rotates the estimated factor loads to be as close to the true factor loads as
+possible. However, when
 the true values are not known, this property of factor analysis limits the
@@
 attribute one latent factor as being a particular nucleosynthetic process. 
-Although having to perform this rotation is not ideal, there is still utility in
+Although having to perform this rotation is not comforting, there is utility in
 trying to identify the latent factors because the factors can only be rotated by
@@
 we have put in: the rotated factors will simply be the closest set of factor
-loads that are consistent with the data.
-
-In this experiment we use detailed chemical abundances of metal-poor stars as
-reported by \citet{Barklem:2005}, and we demonstrate the utility and limitations
-in interpreting the latent factors as nucleosynthetic processes. The data 
-contains 14 measurements of chemical abundances for 61 stars ($N = 61$, $D = 14$): 
-Al (light odd-Z), Mg and Ca ($\alpha$), Sc, Ti, Cr, Mn, Fe, Co, and Ni (Fe-peak),
-as well as Sr, Y, Ba, and Eu ($s$/$r$-process). We performed a grid search for
-the optimal number of latent factors $J$ and components $K$, starting from
-$J = 1$ and $K = 1$, and trialling up until $J = 8$ and $K = 5$. We initialised
-the model parameters randomly for each permutation of $J$ and $K$, and ran E-M
-until a convergence threshold of $10^{-5}$ was reached in the log likelihood. 
-In Figure~\ref{fig:exp2-gridsearch-contours} we show the negative log likelihood 
-and BIC recorded for the optimised set of model parameters for each permutation 
-of $J$ and $K$. The lowest BIC value is found for $J = 5$ and $K = 1$, indicating
-five latent factors are preferred, and no clustering.
-
-\begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-ll.png}
-	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-bic.png}
-    \caption{Model performance metrics resulting from a gridsearch using the
-    		 \citet{Barklem:2005} data in Experiment~2.
-    		 The top panel shows the negative log-likelihood 
-			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
-			 evaluated at each combination of latent factors $J$ and number 
-			 of clusters $K$, and the lower panel shows the BIC (Equation \ref{eq:bic}).
-			 The cross at
-			 $J=8$ and $K=5$ indicates that the model could not converge without
-			 adding a regularisation constant to the diagonals of $\scorecovs$.
-			 The marker indicates the lowest value in each panel, showing that
-			 $J = 5$ and $K = 1$ is preferred by BIC.}
-    \label{fig:exp2-gridsearch-contours}
-\end{figure}
+loads that are consistent with the data, allowing us to identify the factor
+loads that are astrophysically plausible.
 
@@
 In practice we take the product of many  Givens rotation matrices \citep{Givens},
-which reduces to Euler angle rotation in three or fewer dimensions. In Figure
+which reduces to Euler angle rotation in three or fewer dimensions.
+%\footnote{Clifford
+%algebra offers a more efficient algorithm that is more suitable for high dimensions, 
+%but we found Givens rotation matrices to be suitable for our purposes here.}
+In Figure
 \ref{fig:exp2-factor-loads} we show the rotated factor loads found by our
@@
 the rotation angles -- not including any uncertainty on the factor loads themselves.
+
+\begin{figure}
+	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-ll.png}
+	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-bic.png}
+    \caption{Model performance metrics resulting from a gridsearch using the
+    		 \citet{Barklem:2005} data in Experiment~2.
+    		 The top panel shows the negative log-likelihood 
+			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
+			 evaluated at each combination of latent factors $J$ and number 
+			 of clusters $K$, and the lower panel shows the BIC (Equation \ref{eq:bic}).
+			 The cross at
+			 $J=8$ and $K=5$ indicates that the model could not converge without
+			 adding a regularisation constant to the diagonals of $\scorecovs$.
+			 The marker indicates the lowest value in each panel, showing that
+			 $J = 5$ and $K = 1$ is preferred by BIC.}
+    \label{fig:exp2-gridsearch-contours}
+\end{figure}
 
@@
 experiment hopefully demonstrates that the rotated latent factors have some
-interpretability, albeit limited. After accounting for the contributions by all
+interpretability. After accounting for the contributions by all
 $J = 5$ factors, the specific scatter remaining in each dimension is shown in
 Figure~\ref{fig:exp2-specific-scatter}. The remaining scatter varies from as low 
-as 0.01-0.02~dex (Mg, Ca, Ti, Mn, Fe, Y) up to 0.08-0.16~dex (Al and Ba).
+as 0.01 to 0.02~dex (Mg, Ca, Ti, Mn, Fe, Y) up to 0.08 or 0.16~dex (Al and Ba). 
+% TODO: matches Barklem variances?
 

### 2019-01-17 article/mcfa.tex
exp1 data generation function works

@@
 
-\received{2018 XX XX}
-\revised{2018 XX XX}
-\accepted{2018 XX XX}
+\received{2019 March XX}
+\revised{2019}
+\accepted{2019}
 
@@
 to nucleosynthetic yields, and the lower dimensional latent space allows for
-more efficient clustering. We run experiments on literature chemical abundances,
-including those from the second \Galah\ data release. \todo{We find that 5 latent factors are
+more efficient clustering. We perform experiments on literature chemical abundances
+of metal-poor stars, \todo{others}, and 
+those from the second \Galah\ data release. \todo{We find that 5 latent factors are
 preferred to explain the chemical abundances of metal-poor stars, and 8 are
@@
 
+\begin{figure*}
+	\includegraphics[width=1.0\textwidth]{experiments/exp1-data-colour.pdf}
+    \caption{The generated data for Experiment~1, where points are coloured by their
+		     true associations to each component.}
+    \label{fig:exp1-data}
+\end{figure*}
+
 \section{Experiments} \label{sec:experiments}
@@
 
-\begin{figure*}
-	\includegraphics[width=1.0\textwidth]{experiments/exp1-data.png}
-    \caption{A corner plot showing all $\NumDimensions$ dimensions of the
-    		 generated data for Experiment~1 (blue). Overplotted in red
-		 	 we show samples generated given our estimate of the model 
-			 parameters, without adding specific variance in each dimension.
-			 This demonstrates that the model parameters estimated by
-			 expectation-maximization can replicate the generated data.}
-    \label{fig:exp1-data}
-\end{figure*}
-
-%\ref{fig:exp1-latent-space}
+\begin{figure}
+	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-ll-contours.png}
+	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-bic-contours.png}
+    \caption{Model performance metrics resulting from a gridsearch using the
+    		 data generated as part of our toy model. The top 
+		 	 panel shows the negative log-likelihood 
+			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
+			 evaluated at each combination of latent factors $J$ and number 
+			 of clusters $K$, the middle panel shows the BIC (Equation \ref{eq:bic}), 
+			 and the lower panel shows
+			 the pseudo-BIC
+			 (Equation~\ref{eq:pseudo-bic}). The marker indicates the 
+			 lowest value in each panel.}
+    \label{fig:experiment-1-gridsearch}
+\end{figure}
 
@@
 
-We show a corner plot of the data in Figure~\ref{fig:exp1-data}.
-In this figure we have over-plotted 100000 samples drawn from the model
-with $J = 5$ and $K = 21$
--- without adding variance $\specificvariance_\textrm{est}$
--- in order to illustrate that the
-estimated model parameters can realistically generate the data.
-The estimated latent space of factor scores is shown in 
-Figure~\ref{fig:exp1-latent-space} where the colours indicate the
-most likely component that the data point is associated with.
-%todo: number of correct associations?
+We show a corner plot of the generated data in Figure~\ref{fig:exp1-data},
+demonstrating the clustering (colour) and structure varying structure in data space.
+Despite this clustering in data space, it is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
+and allowing clustering in the latent space provides a better description of the data.
+Adding components to the model improves the log-likelihood, even with a single latent factor,
+but the addition of just one latent factor improvers the log-likelihood more than adding
+20 clusters. Not much can be said for this example because the true data generating process 
+is drawn from a higher number of latent factors, but this does illustrate how 
+clustering in high dimensional data can be better described by latent factors with 
+clustering in the lower dimensional latent space.
 
-\begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-ll-contours.png}
-	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-bic-contours.png}
-    \caption{Model performance metrics resulting from a gridsearch using the
-    		 data generated as part of our toy model. The top 
-		 	 panel shows the negative log-likelihood 
-			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
-			 evaluated at each combination of latent factors $J$ and number 
-			 of clusters $K$, the middle panel shows the BIC (Equation \ref{eq:bic}), 
-			 and the lower panel shows
-			 the pseudo-BIC
-			 (Equation~\ref{eq:pseudo-bic}). The marker indicates the 
-			 lowest value in each panel.}
-    \label{fig:experiment-1-gridsearch}
-\end{figure}
+Given our estimate of the number of latent factors and components ($\NumComponents = 21$, $\NumLatentFactors = 5$), in Figure~\ref{fig:exp1-latent} we show the estimated latent
+space (factor scores $\factorscores$) where each point is coloured by it's inferred
+associated component. After applying an orthogonal rotation (see following section)
+to the estimated latent space, we show the estimated factor scores (Figure~\ref{fig:exp1-factor-scores-comparison}) and estimated factor loads (Figure~\ref{fig:exp1-factor-loads-comparison})
+compared to the true values. 
+Similarly, in Figure~\ref{fig:exp1-specific-variance} we show the estimated specific scatter
+in each data dimension compared to the true value. 
+\todo{The agreement is excellent in all model parameters.}
+
+\begin{figure*}
+	\includegraphics[width=\textwidth]{experiments/exp1-latent.pdf}
+    \caption{The inferred latent scores $\factorscores$ for Experiment~1. 
+             Points are coloured by the inferred associated component.}
+    \label{fig:exp1-latent}
+\end{figure*}
 
@@
 \begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-ll.png}
-	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-bic.png}
+	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-ll.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-bic.pdf}
     \caption{Model performance metrics resulting from a gridsearch using the

### 2019-01-17 article/experiments/exp1.py

@@ -5,7 +5,6 @@ import os
 import matplotlib
 import matplotlib.pyplot as plt
 from matplotlib.ticker import MaxNLocator
-from scipy import stats
 from time import time
 
 sys.path.insert(0, "../../")
@@ -16,84 +15,14 @@ matplotlib.style.use(mpl_utils.mpl_style)
 colors = plt.rcParams['axes.prop_cycle'].by_key()['color']
 
 
-# Should all go into exp1.yaml
-"""
-data_kwds = dict(n_samples=100_000,
-                 n_features=15,
-                 n_components=10,
-                 n_latent_factors=3,
-                 random_seed=42)
-"""
-
-
-
-def savefig(fig, suffix):
-    prefix = os.path.basename(__file__)[:-3]
-    filename = f"{prefix}-{suffix}.png"
-    fig.savefig(filename, dpi=150)
-
-
-mcfa_kwds = dict(tol=1e-5, 
-                 max_iter=1_000,
-                 init_factors="svd",
-                 init_components="kmeans++",
-                 random_seed=42)
-
-# Done with vars
-"""
-Y, truth = utils.generate_data(**data_kwds)
-"""
-"""
-def generate_data(n_samples=20, n_features=5, n_latent_factors=3, n_components=2,
-                  omega_scale=1, noise_scale=1, random_seed=0):
-
-    rng = np.random.RandomState(random_seed)
-
-    #A = rng.randn(n_features, n_latent_factors)
-    A = stats.ortho_group.rvs(n_features, random_state=rng)[:, :n_latent_factors]
-    AL = linalg.cholesky(A.T @ A)
-    A = A @ linalg.solve(AL, np.eye(n_latent_factors))
-
-    # latent variables
-    pvals = np.ones(n_components) / n_components
-    R = np.argmax(rng.multinomial(1, pvals, size=n_samples), axis=1)
-    pi = np.array([np.sum(R == i) for i in range(n_components)])/n_samples
-
-    xi = rng.randn(n_latent_factors, n_components)
-    omega = np.zeros((n_latent_factors, n_latent_factors, n_components))
-    for i in range(n_components):
-        omega[(*np.diag_indices(n_latent_factors), i)] = \
-            rng.gamma(1, scale=omega_scale, size=n_latent_factors)**2
-
-    scores = np.empty((n_samples, n_latent_factors))
[…]

### 2019-01-17 article/experiments/exp2.py

@@ -14,14 +14,18 @@ from mcfa import (mcfa, grid_search, mpl_utils, utils)
 
 matplotlib.style.use(mpl_utils.mpl_style)
 
+here = os.path.dirname(os.path.realpath(__file__))  
 
 def savefig(fig, suffix):
     prefix = os.path.basename(__file__)[:-3]
-    filename = f"{prefix}-{suffix}.png"
-    fig.savefig(filename, dpi=150)
+    filename = os.path.join(here, f"{prefix}-{suffix}")
+    fig.savefig(f"{filename}.pdf", dpi=300)
+    fig.savefig(f"{filename}.png", dpi=150)
+    print(f"Created figures {filename}.png and {filename}.pdf")
     
 
-with open("../../catalogs/barklem.pkl", "rb") as fp:
+
+with open(os.path.join(here, "../../catalogs/barklem.pkl"), "rb") as fp:
     X_H, label_names, mask = pickle.load(fp)
 
 # Do not include C
@@ -118,27 +122,46 @@ fig_factor_loads.tight_layout()
 
 
 # Run a grid search.
-max_n_latent_factors = 8
+max_n_latent_factors = 7
 max_n_components = 5
 
 Js = 1 + np.arange(max_n_latent_factors)
 Ks = 1 + np.arange(max_n_components)
 
-Jg, Kg, converged, ll, bic, pseudo_bic = grid_search.grid_search(Js, Ks, X, 
-                                                                 mcfa_kwds)
+Jg, Kg, converged, metrics = grid_search.grid_search(Js, Ks, X, mcfa_kwds)
+
+ll = metrics["ll"]
+bic = metrics["bic"]
+pseudo_bic = metrics["pseudo_bic"]
+message_length = metrics["message_length"]
 
 J_best_ll, K_best_ll = grid_search.best(Js, Ks, -ll)
 J_best_bic, K_best_bic = grid_search.best(Js, Ks, bic)
+J_best_mml, K_best_mml = grid_search.best(Js, Ks, message_length)
 
 print(f"Best log likelihood  at J = {J_best_ll} and K = {K_best_ll}")
 print(f"Best BIC value found at J = {J_best_bic} and K = {K_best_bic}")
+print(f"Best MML value found at J = {J_best_mml} and K = {K_best_mml}")
 
 # Plot some contours.
-plot_filled_contours_kwds = dict(converged=converged, marker_function=np.nanargmin)
+plot_filled_contours_kwds = dict(converged=converged,
[…]

### 2019-01-17 mcfa/grid_search.py

@@ -25,6 +25,7 @@ def grid_search(trial_n_latent_factors, trial_n_components, X, mcfa_kwds=None,
     converged = np.zeros(shape, dtype=bool)
     ll = np.nan * np.ones(shape)
     bic = np.nan * np.ones(shape)
+    mml = np.nan * np.ones(shape)
     pseudo_bic = np.nan * np.ones(shape)
 
     for j, J in enumerate(Js):
@@ -45,10 +46,13 @@ def grid_search(trial_n_latent_factors, trial_n_components, X, mcfa_kwds=None,
             else:
                 ll[k, j] = model.log_likelihood_
                 bic[k, j] = model.bic(X)
+                mml[k, j] = model.message_length(X)
                 pseudo_bic[k, j] = model.pseudo_bic(X, **pseudo_bic_kwds)
                 converged[k, j] = True
 
-    return (J_grid, K_grid, converged, ll, bic, pseudo_bic)
+
+    metrics = dict(ll=ll, bic=bic, pseudo_bic=pseudo_bic, message_length=mml)
+    return (J_grid, K_grid, converged, metrics)#ll, bic, pseudo_bic)
 
 
 def best(trial_n_latent_factors, trial_n_components, metric, 

### 2019-01-17 mcfa/mcfa.py

@@ -8,7 +8,7 @@ import warnings
 from copy import deepcopy
 from inspect import signature as inspect_signature
 from scipy import linalg, spatial, stats
-from scipy.special import logsumexp
+from scipy.special import logsumexp, gammaln, gamma
 from sklearn import cluster
 from sklearn.utils import check_random_state
 from sklearn.utils.extmath import row_norms
@@ -569,6 +569,81 @@ class MCFA(object):
         return 6 * (1 + gamma) * omega * np.log(N) * Q - 2 * log_likelihood
 
 
+    def message_length(self, X, theta=None):
+        r"""
+        Estimate the explanation length given the model and the data.
+
+        :param X:
+            The data, :math:`X`, which is expected to be an array of shape
+            [n_samples, n_features].
+
+        :param theta: [optional]
+            The model parameters :math:`\theta`. If None is given then the
+            model parameters from `self.theta_` will be used.
+        """
+
+        if theta is None:
+            theta = self.theta_
+
+        N, D = np.atleast_2d(X).shape
+        log_likelihood, tau = self.expectation(X, *theta)
+
+        # For just MLF:
+        pi, A, xi, omega, psi = theta
+        J, K = self.n_latent_factors, self.n_components
+
+        # In Wallace nomenclature, J = number of factors (here also J)
+        #                          K = dimensionality of data (here is D)
+        #                          N = number of data (here also N)
+
+        # Recall:
+        # X! = gamma(X + 1)
+        # S_m = (m * np.pi**(m/2))/((m/2)!)
+        log_S = lambda m: np.log(m) + (m/2) * np.log(np.pi) - gammaln(m/2 + 1)
+
+        # Approximate the normalization constants (p308 of Wallace 2005)
+        #  J = [1     2     3     4     5      6]
+        A_Js = [0, 0.24, 1.35, 3.68, 7.45, 12.64]
+        B_Js = [0, 0.58, 0.64, 0.73, 0.86,  1.21]
+
+        #bj_sq = A**2
+        #v = self.factor_scores(X)[1]
+        #vj_sq = np.diag((v.T @ v)) # very unsure about this.
+        
+
+        if J > len(A_Js):
+
[…]

### 2019-01-17 mcfa/mpl_utils.py

@@ -6,8 +6,8 @@ import numpy as np
 import matplotlib
 import matplotlib.pyplot as plt
 from scipy.stats import binned_statistic_2d
-from matplotlib.colors import LogNorm
-from matplotlib.ticker import MaxNLocator
+from matplotlib.colors import (BoundaryNorm, LinearSegmentedColormap, LogNorm)
+from matplotlib.ticker import MaxNLocator, FormatStrFormatter, FuncFormatter
 from matplotlib.patches import Ellipse
 
 from .utils import find_rotation_matrix, givens_rotation_matrix
@@ -81,6 +81,20 @@ mpl_style = {
 
 matplotlib.style.use(mpl_style)
 
+
+
+def discrete_cmap(N, base_cmap=None):
+    """Create an N-bin discrete colormap from the specified input map"""
+
+    # Note that if base_cmap is a string or None, you can simply do
+    #    return plt.cm.get_cmap(base_cmap, N)
+    # The following works for string, None, or a colormap instance:
+
+    base = plt.cm.get_cmap(base_cmap)
+    color_list = base(np.linspace(0, 1, N))
+    cmap_name = base.name + str(N)
+    #return base.from_list(cmap_name, color_list, N)
+    return LinearSegmentedColormap.from_list(cmap_name, color_list, N)
 
 def plot_latent_factors(model, label_names=None):
 
@@ -363,7 +377,7 @@ def corner_scatter(X, label_names=None, show_ticks=False, fig=None, figsize=None
     
     axes = np.array(fig.axes).reshape((K, K)).T
 
-    kwds = dict(s=1, c="tab:blue", alpha=0.5)
+    kwds = dict(s=1, c="tab:blue", alpha=0.5, rasterized=True)
     kwds.update(kwargs)
     
     for i, x in enumerate(X.T):
@@ -397,9 +411,10 @@ def corner_scatter(X, label_names=None, show_ticks=False, fig=None, figsize=None
     return fig
 
 
-def plot_filled_contours(J, K, Z, N=100, colorbar_label=None, 
+def plot_filled_contours(J, K, Z, N=1000, colorbar_label=None, 
                          converged=None, converged_kwds=None, 
                          marker_function=None, marker_kwds=None, 
+                         truth=None,
                          ax=None, **kwargs):
     power = None
         
[…]

### 2019-01-18 article/mcfa.tex
exp2 + 1 figs

@@
 \usepackage{comment}
+%\usepackage{layouts}
 
@@
 each dimension $\specificvariance$ are also drawn $\specificvariance \sim \vec\Gamma\left(1\right)$.
+We randomly draw a  $\NumDimensions \times \NumDimensions$ matrix from a Haar distribution \citep{Haar:1933},
+which is uniform on the special orthogonal group $\textrm{SO}(n)$ and therefore guaranteed to return an orthogonal
+matrix with a determinant of unity \citep{Stewart:1980}.
+We denote the $\NumLatentFactors \times \NumDimensions$ left-most region of this
+matrix to be $\mathbf{H}$, and by taking $\factorloads_\ast$ to be the Cholesky decomposition of $\mathbf{H}\transpose \mathbf{H}$, we find the factor loads
+\begin{eqnarray}
+ 	\factorloads = \mathbf{H}\left(\left(\factorloads_\ast\right)^{-1}\eye\right)
+\end{eqnarray}
+\noindent{}which ensures that the true factor loads is a $\NumLatentFactors \times \NumDimensions$
+matrix of mutually orthogonal vectors and that $\factorloads\transpose \factorloads = \eye$.
 The $\numdata$th data point (which belongs to the $\numcomponents$th cluster) is then
-generated by drawing $\factorscores_{\numdata} \sim \mathcal{N}(\scoremeans_\numcomponents,\scorecovs_\numcomponents)$, projecting by the factor loads, and adding variance.
+generated by drawing $\factorscores_{\numdata} \sim \mathcal{N}(\scoremeans_\numcomponents,\scorecovs_\numcomponents)$, projecting by the factor loads $\factorloads$, and adding variance $\specificvariance$.
 
@@
 			 evaluated at each combination of latent factors $J$ and number 
-			 of clusters $K$, the middle panel shows the BIC (Equation \ref{eq:bic}), 
-			 and the lower panel shows
-			 the pseudo-BIC
-			 (Equation~\ref{eq:pseudo-bic}). The marker indicates the 
-			 lowest value in each panel.}
+			 of clusters $K$ and the bottom panel shows the BIC (Equation \ref{eq:bic}).
+			 The white marker indicates the 
+			 lowest value in each panel, which is joined by a black line from the true value.}
     \label{fig:experiment-1-gridsearch}
@@
 \end{equation}
-%\todo{\noindent{}and the Message Length.}
 
@@
 The lowest BIC value is found at $\NumLatentFactors = 5$
-and $\NumComponents = 21$, close to the true values ($\NumLatentFactors_\textrm{true} = 5$,
-$\NumComponents_\textrm{true} = 20$).
+and $\NumComponents = 20$, identical to the true values.
 
@@
 clustering in the lower dimensional latent space.
+The inferred clustering in latent space is shown in Figure~\ref{fig:exp1-latent},
+where each point is coloured by the inferred associated component.
 
-Given our estimate of the number of latent factors and components ($\NumComponents = 21$, $\NumLatentFactors = 5$), in Figure~\ref{fig:exp1-latent} we show the estimated latent
-space (factor scores $\factorscores$) where each point is coloured by it's inferred
-associated component. After applying an orthogonal rotation (see following section)
-to the estimated latent space, we show the estimated factor scores (Figure~\ref{fig:exp1-factor-scores-comparison}) and estimated factor loads (Figure~\ref{fig:exp1-factor-loads-comparison})
-compared to the true values. 
-Similarly, in Figure~\ref{fig:exp1-specific-variance} we show the estimated specific scatter
-in each data dimension compared to the true value. 
-\todo{The agreement is excellent in all model parameters.}
+Some technical discussion is warranted before we compare our estimated model
+parameters to the true values. The latent factors in this model are only
+identifiable up to an orthogonal rotation. That is to say that
+if the data were truely generated by latent factors $\factorloads_\textrm{true}$,
+then our estimates of those latent factors $\factorloads_\textrm{est}$ do not need
+to be identical to the true values. For example, the ordering of the estimated factors
+could be different from the true factors, and the ordering of the dimensionality
+in latent space would be correspondingly different. Since no constraint is
+placed on the ordering of the factor loads during expectation-maximization,
+there is no assurance (or requirement) that our factor loads match the true factor loads.
 
@@
 	\includegraphics[width=\textwidth]{experiments/exp1-latent.pdf}
-    \caption{The inferred latent scores $\factorscores$ for Experiment~1. 
-             Points are coloured by the inferred associated component.}
+    \caption{The inferred latent scores $\factorscores$ for Experiment~1
+    		 from a model with $\NumLatentFactors = 5$ latent factors and
+		 	 $\NumComponents = 20$ multivariate normal components in
+			 latent space.
+             Points are coloured by the inferred component.}
     \label{fig:exp1-latent}
@@
 
-%\begin{figure*}
-%	\includegraphics[width=1.0\textwidth]{experiments/toy-factors-init-randomly.png}
-%	\caption{Latent factor loads inferred for models with increasing numbers
-%			 of latent factors and increasing numbers of cluster components
-%			 in Experiment~1, where the latent factors and component assignments
-%			 were initialised randomly.
-%				The top left panel shows the results of the model with
-%			 $\NumLatentFactors = 1$ and $\NumComponents = 1$. 
-%			 The latent factors have each been rotated by a
-%			 rotation matrix such 
-%			 that the estimated latent factors in each model are ordered 
-%			 and oriented to be as similar as the latent factors estimated from
-%			 the model with only one latent factor, up to orthogonality such
-%			 that $\vec{R}\vec{R}\transpose = \eye$.
-%			 }
-%	\label{fig:toy-factors-init-randomly}
-%\end{figure*}
+Another possibility is that the estimated factor loads could be flipped in sign 
+relative to the true factor loads, and the scores would similarly be flipped. 
+In both of these situations (reordering or flipped signs) the log likelihood 
+given the data and the estimated factor loads $\factorloads_\textrm{est}$ 
+%\citep[or Kullback-Leibler divergence;][]{Kullback:1951}
+would be identical to the log likelihood given the data and the true factor loads 
+$\factorloads_\textrm{true}$
+despite the difference in ordering and sign. These examples serve to illustrate a more 
+general property that the factor loads and factor scores can be orthogonally 
+rotated by \emph{any valid rotation matrix}\footnote{A rotation matrix is valid if 
+$\vec{R}\vec{R}\transpose = \vec{I}$} $\vec{R}$. The estimated factor loads 
+$\factorloads_\textrm{est}$ could therefore appear very different from the true 
+values, but they only differ by an orthogonal rotation.
 
-%\todo{This structure goes away and becomes more consistent if we change the initialisation procedure}
+% TODO: Make single figure with bigger text labels.
+\begin{figure*}
+	\includegraphics[width=0.33\textwidth]{experiments/exp1-compare-loads.pdf}
+	\includegraphics[width=0.33\textwidth]{experiments/exp1-compare-scores.pdf}
+	\includegraphics[width=0.33\textwidth]{experiments/exp1-compare-specific-scatter.pdf}
+    \caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
+    		 and specific variances $\specificvariance$ (right) compared to the 
+		 	 true data generating values
+		 	 for Experiment~1. The agreement is excellent.}
+    \label{fig:exp1-compare}
+\end{figure*}
 
-%\todo{figure of toy factors init svd kmeans++}
-
-%\todo{log-likelihood of different initialisation routines}
-
-\begin{comment}
-The results of this experiment confirm that the model is indeterminate, in 
-that there are multiple equally valid solutions that can exactly reproduce
-the data. It also highlights the importance of initialisation, demonstrating
-how the initialisation of the common latent factors can impact the final
-model parameters. These two points are related, as one could initialise the
-latent factors randomly (or from noise) and multiply that model with a
-rotation matrix $\mathbf{R}$ that rotates the latent space such that the
-common factor loads are aligned closer to some prior expectation that we have
-for the common factor loads (e.g., in that they should represent plausible
-astrophysical yields). Motivated by these results and indeterminate nature
-of the model, in Experiment~2 we initialise the common factor loads
-to represent known astrophysical processes such that they can be interpreted
-in context with the data.
-
-\subsection{Experiment 2: A minimal set of nucleosynthetic trace elements in \Galah}
-
-In this experiment we further investigate the effects of initialising the model parameters.
-Here we make use of a minimal set of nucleosynthetic trace elements
-using the second \Galah\ data release \citep{Buder:2018}. Specifically, we use
-reliable chemical abundance measurements of Na (an odd-Z element), Mg and Ca (alpha-elements),
-Ni, Fe, and Cr (iron-peak elements), Ba and Y (s-process tracers), and Eu (an r-process tracer).
-We assume that there are four latent factors ($J = 4$) and four components ($K = 4$),
-and we initialise the latent factors and component parameters in different combinations.
-
-To initialise the latent factors using astrophysical insight (or domain expertise)
-we make the following steps:
-\begin{enumerate}
-	\item We initialise all entries of $\factorloads$ to be drawn from $\mathcal{N}\left(0, 10^{-2}\right)$.
-	\item For the first latent factor we set the entries that correspond to Ni, Fe, and Cr to be $+1$ (e.g., Fe-peak).
-	\item For the second latent factor we set the entries that correspond to Mg and Ca to be $+1$ (e.g., $\alpha$-elements).
-	\item For the third latent factor we set the entries that correspond to Ba and Y to be $+1$ (e.g., the s-process).
-	\item For the fourth latent factor we set the entry that corresponds to Eu to be $+1$ (e.g., the r-process).
-\end{enumerate}
-
-%The latent factors can be initialised in four different ways: (1) with entries drawn randomly from $\mathcal{U}\left(-1, +1\right)$, (2) with noisey draws $\mathcal{N}\left(0, 10^{-2}\right)$, (3) using singular value decomposition, and (4) using astrophysical insight to initialise the factor loads.
-
-The component parameters are either initialised randomly, or using \texttt{K-means++} algorithm in
-the latent space given the initial estimate of the latent factors. For each combination of 
-initialisation procedure, we run 30 optimisations and record the log likelihood once the E-M 
-procedure has reached a threshold of $10^{-8}$.
-The results of this experiment are shown in Figure~\ref{fig:experiment-4-log-L}.
-In most cases the final log likelihood is within the threshold of $10^{-8}$ between
-different optimisation runs of the same initialisation procedure. However, it is clear
-that at least some initialisations result in a local optimum, emphasising the importance in
-optimising from many initialisations.
-
-Figure~\ref{fig:experiment-4-log-L} demonstrates that even though the initialisation procedure
-can impact the final estimate of the model parameters, the resulting log likelihood values
-are within the convergence threshold. This is, again, a consequence of the indeterminate
-nature of the model specification. However, the advantage here is that these results
-demonstrate that we \emph{can} initialise the latent factors with domain knowledge,
-without a significant detriment to the final log likelihood. Similarly, in Figure~\ref{fig:experiment-4-astrophysical-latent-factors}
-we show that the initial latent factors are indeed different from the final latent factors,
-and that there is some variance in our estimates of the latent factors from 30 initialisations.
-\todo{something to say here about not biasing our results?}
-
-The results from this experiment demonstrate that we can initialise the latent factors
-using domain knowledge without incurring strong biases in our final estimates of the
-latent factors, or introducing a significant detriment to the log likelihood, as long
-as we initialise the model many (e.g., $\gtrsim 5$) times.
-
-\begin{figure}
-	\includegraphics[width=0.5\textwidth]{experiments/galah-experiment-4-log-L-hist.png}
-	\caption{Log likelihoods $\mathcal{L}\left(\data|\vec\Psi\right)$ of optimised
-			 models using \Galah\ data in Experiment 4 with four latent factors
-			 and four components ($J = 4$, $K = 4$). The columns indicate different
-			 methods to initialise the component parameters ($\pi$, $\scoremeans$,
-			 $\scorecovs$), and the rows represent different methods to initialise
-			 the latent factors ($\factorloads$). Each combination of initialisation
-			 procedure has 30 models where E-M was run until a threshold of $10^{-8}$
-			 was reached in $\log{L}$.}
-	\label{fig:galah-experiment-4-log-L}
-\end{figure}
-
-\todo{experiment-4-astrophysical-latent-factors}
-\end{comment}
+We took the model with the preferred number of latent factors and components found
+from a grid search ($\NumComponents = 20$, $\NumLatentFactors = 5$; which are also
+the true values) and applied an orthogonal rotation to the latent space to be as
+close as possible to the true values. The rotation matrix $\mathbf{R}$ was found
+by solving for $\NumLatentFactors$ unknown angle parameters, each of which is used
+to construct a Givens rotation matrix, and then we take the product of those Givens
+matrices. This process reduces to Euler angle rotation in three or fewer dimensions.
+This process rotates the latent space
+($\factorloads$, $\scoremeans$, $\scorecovs$), but has no effect on the model's 
+predictive power: the log-likelihood \citep[or the Kullback-Leibler divergence;][]{Kullback:1951} under the
+rotated model is indistinguishable from the unrotated model.
+In Figure~\ref{fig:exp1-compare} we show the estimated factor loads $\factorloads$,
+factor scores $\factorscores$, and specific variances $\specificvariance$ compared
+to the true values. The agreement is excellent in all model parameters.
 
@@
 in interpreting the latent factors as nucleosynthetic processes. The data 
-can be described as 14 chemical abundance measurements from 61 stars ($\NumData = 61$, $\NumDimensions = 14$), that trace multiple nucleosynthetic pathways. Specifically the chemical
-abundances measured include Al (light odd-Z), Mg and Ca ($\alpha$), Sc, Ti, Cr, Mn, Fe, Co, and Ni (Fe-peak), as well as Sr, Y, Ba, and Eu ($s$/$r$-process).
+can be described as 14 chemical abundance measurements from 61 stars ($\NumData = 61$, $\NumDimensions = 14$), and those 14 chemical elements trace multiple nucleosynthetic pathways.
+Specifically the chemical
+abundances measured include Al (light odd-Z), Mg and Ca ($\alpha$-process), Sc, Ti, Cr, Mn, Fe, Co, and Ni (Fe-peak elements), as well as Sr, Y, Ba, and Eu ($s$/$r$-process).
 
-We performed a grid search for the optimal number of latent factors $\NumLatentFactors$ and 
-components $\NumComponents$, starting from $\NumLatentFactors = 1$ and $K = 1$, and trialling up until $\NumLatentFactors = \todo{8}$ and $\NumComponents = 5$. We initialised
-the model parameters randomly for each permutation of $J$ and $K$, and ran E-M
+We performed a grid search for the optimal number of latent factors 
+$\NumLatentFactors$ and components $\NumComponents$, starting from 
+$\NumLatentFactors = 1$ and $K = 1$, and trialling up until 
+$\NumLatentFactors = \todo{8}$ and $\NumComponents = 5$. 
+We found that models with higher numbers of latent factors were unable to
+converge without adding a regularization constant to the diagonals of the
+covariance matrices $\scorecovs$ in latent space. Although it is common
+to add a regularization term to the diagonals of covariance matrices
+when performing expectation-maximization, we found that our existing
+grid search boundaries were sufficient to find the preferred $\NumLatentFactors$
+and $\NumComponents$, and chose not to add any regularization term.
+
+We initialised
+the model parameters randomly for each permutation of $J$ and $K$, and continued
+the expectation-maximization algorithm
 until a convergence threshold of $10^{-5}$ was reached in the log likelihood. 
@@
 
-Some technical discussion is warranted before we interpret the estimated factor loads.
-In principle the latent factors are analogous to nucleosynthetic yields, but 
-these can only be identified up to an orthogonal rotation. That is to say that
-if the data were truely generated by latent factors $\factorloads_\textrm{true}$,
-then our estimates of those latent factors $\factorloads_\textrm{est}$ do not need
-to be identical to the true values. For example, the ordering of the estimated factors
-could be different from the true factors, and the ordering of the dimensionality
-in latent space would be correspondingly different. Since the ordering of the 
-factor loads is not prescribed during expectation-maximization, there is no 
-assurance (or requirement) that our factor loads match the true factor loads.
-
-Another possibility is that the estimated factor loads could be flipped in sign 
-relative to the true factor loads, and the scores would similarly be flipped. 
-In both of these situations (reordering or flipped signs) the log likelihood 
-given the data and the estimated factor loads $\factorloads_\textrm{est}$ would 
-be identical to the log likelihood given the data and the true factor loads 
-$\factorloads_\textrm{true}$
-despite the difference in ordering and sign. These examples serve to illustrate a more 
-general property that the factor loads and factor scores can be orthogonally 
-rotated by \emph{any rotation matrix}\footnote{A rotation matrix is valid if 
-$\vec{R}\vec{R}\transpose = \vec{I}$} $\vec{R}$. The estimated factor loads 
-$\factorloads_\textrm{est}$ could therefore appear very different from the true 
-values, but they only differ by an orthogonal rotation.
-
-If the true values are known then it is trivial to find a rotation matrix $\vec{R}$
-that rotates the estimated factor loads to be as close to the true factor loads as
-possible. However, when
-the true values are not known, this property of factor analysis limits the
-extent to which we can interpret the latent variables in our model. For example,
-if we seek to interpret the factor loads as being something analagous to 
-nucleosynthetic processes, then the estimated factor loads will need to be 
-rotated to something that looks astrophysically plausible before we could 
-attribute one latent factor as being a particular nucleosynthetic process. 
+Using our preferred model with $J = 5$ and $K = 1$, we sought to find a
+rotation matrix that would allow us to identify plausible astrophysical
+processes in the factor loads.
 Although having to perform this rotation is not comforting, there is utility in
@@
 loads that are consistent with the data, allowing us to identify the factor
-loads that are astrophysically plausible.
-
-Using our preferred model with $J = 5$ and $K = 1$, we rotated the estimated
-factor loads in order to be astrophysically plausible. Specifically we constructed
+loads that are astrophysically plausible. We constructed
 a $J \times D$ astrophysical factor load matrix $\factorloads_\textrm{A}$ where
@@
 in that latent factor, in order to ensure that $\factorloads_\textrm{A}$ is
-orthonormal. We optimise the $J$ angles needed to make the estimated factor loads
-$\factorloads_\textrm{est}$ as close to the astrophysical factor loads $\factorloads_\textrm{A}$.
-In practice we take the product of many  Givens rotation matrices \citep{Givens},
-which reduces to Euler angle rotation in three or fewer dimensions.
-%\footnote{Clifford
-%algebra offers a more efficient algorithm that is more suitable for high dimensions, 
-%but we found Givens rotation matrices to be suitable for our purposes here.}
-In Figure
-\ref{fig:exp2-factor-loads} we show the rotated factor loads found by our
-preferred model, where the shaded regions indicate the projected uncertainty on
-the rotation angles -- not including any uncertainty on the factor loads themselves.
+mutually orthogonal. We then found the valid rotation matrix $\vec{R}$ that most
+closely approximated the astrophysical loads $\factorloads_\textrm{A}$.
+In Figure~\ref{fig:exp2-factor-loads} we show the rotated factor loads found by our
+preferred model, where the shaded regions indicate \emph{only} the projected uncertainty on
+the rotation angles -- and does not include any uncertainty on the factor loads 
+themselves.
 
@@
 			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
-			 evaluated at each combination of latent factors $J$ and number 
-			 of clusters $K$, and the lower panel shows the BIC (Equation \ref{eq:bic}).
-			 The cross at
-			 $J=8$ and $K=5$ indicates that the model could not converge without
-			 adding a regularisation constant to the diagonals of $\scorecovs$.
+			 evaluated at each combination of latent factors $\NumLatentFactors$ and number 
+			 of clusters $\NumComponents$, and the lower panel shows the BIC (Eq.~\ref{eq:bic}).
+			 %The cross at
+			 %$\NumLatentFactors=8$ and $\NumComponents=5$ indicates that the model could not converge without
+			 %adding a regularisation constant to the diagonals of $\scorecovs$.
 			 The marker indicates the lowest value in each panel, showing that
-			 $J = 5$ and $K = 1$ is preferred by BIC.}
+			 $\NumLatentFactors = 5$ and $\NumComponents = 1$ is preferred by BIC.}
     \label{fig:exp2-gridsearch-contours}
@@
 Eu, and could be interpreted as production from the rapid nucleosynthesis process.
+
+\begin{figure}
+	\includegraphics[width=0.5\textwidth]{experiments/exp2-latent-contributions.pdf}
+	\caption{Rotated latent factors found in Experiment~2 using the \citet{Barklem:2005}
+			 data. \todo{elaborate latent factors}}
+    \label{fig:exp2-latent-contribution}
+\end{figure}
 

### 2019-01-18 article/experiments/exp1.py

@@ -20,9 +20,9 @@ n_components = 20
 n_latent_factors = 5
 n_samples = 100000
 
-omega_scale = 3
+omega_scale = 1
 noise_scale = 1
-random_seed = 101
+random_seed = 42
 
 data_kwds = dict(n_features=n_features,
                  n_components=n_components,
@@ -46,7 +46,7 @@ mcfa_kwds = dict(tol=1e-5,
                  max_iter=1_000,
                  init_factors="random",
                  init_components="random",
-                 random_seed=123)
+                 random_seed=random_seed)
 
 
 
@@ -54,17 +54,7 @@ mcfa_kwds = dict(tol=1e-5,
 Y, truth = utils.generate_data(**data_kwds)
 truth_packed = (truth["pi"], truth["A"], truth["xi"], truth["omega"], truth["psi"])
 
-fig_data2 = mpl_utils.corner_scatter(Y, 
-                                    c=truth["R"], cmap="Spectral",
-                                    s=1, alpha=0.5, figsize=(8, 8),
-                                    label_names=[r"$\mathbf{{Y}}_{{{0}}}$".format(i) for i in range(n_features)])
-fig_data2.tight_layout()
-fig_data2.subplots_adjust(hspace=0, wspace=0)
-savefig(fig_data2, "data-colour")
 
-
-gridsearch_max_latent_factors = 10
-gridsearch_max_components = 30
 
 
 # Fit with true number of latent factors and components.
@@ -76,57 +66,6 @@ model.fit(Y)
 tock = time()
 
 print(f"Model took {tock - tick:.1f} seconds")
-
-# Plot the log-likelihood with increasing iterations.
-fig_iterations, ax = plt.subplots()
-
-ll = model.log_likelihoods_
-iterations = 1 + np.arange(len(ll))
-ax.plot(iterations, ll, "-", lw=2, drawstyle="steps-mid")
-ax.set_xlabel(r"$\textrm{iteration}$")
-ax.set_ylabel(r"$\log{\mathcal{L}}$")
-ax.set_xlim(0, iterations[-1] + 1)
-ax.xaxis.set_major_locator(MaxNLocator(6))
-ax.yaxis.set_major_locator(MaxNLocator(6))
-xt = ax.get_xticks()
[…]

### 2019-01-18 article/experiments/exp2.py

@@ -16,6 +16,10 @@ matplotlib.style.use(mpl_utils.mpl_style)
 
 here = os.path.dirname(os.path.realpath(__file__))  
 
+random_seed = 10
+
+np.random.seed(random_seed)
+
 def savefig(fig, suffix):
     prefix = os.path.basename(__file__)[:-3]
     filename = os.path.join(here, f"{prefix}-{suffix}")
@@ -24,19 +28,80 @@ def savefig(fig, suffix):
     print(f"Created figures {filename}.png and {filename}.pdf")
     
 
+from astropy.table import Table
+
+
+periodic_table = """H                                                  He
+                    Li Be                               B  C  N  O  F  Ne
+                    Na Mg                               Al Si P  S  Cl Ar
+                    K  Ca Sc Ti V  Cr Mn Fe Co Ni Cu Zn Ga Ge As Se Br Kr
+                    Rb Sr Y  Zr Nb Mo Tc Ru Rh Pd Ag Cd In Sn Sb Te I  Xe
+                    Cs Ba Lu Hf Ta W  Re Os Ir Pt Au Hg Tl Pb Bi Po At Rn
+                    Fr Ra Lr Rf"""
+
+lanthanoids    =   "La Ce Pr Nd Pm Sm Eu Gd Tb Dy Ho Er Tm Yb"
+actinoids      =   "Ac Th Pa U  Np Pu Am Cm Bk Cf Es Fm Md No"
+
+periodic_table = periodic_table.replace(" Ba ", " Ba " + lanthanoids + " ") \
+                               .replace(" Ra ", " Ra " + actinoids + " ") \
+                               .split()
+
+t = Table.read(os.path.join(here, "../../catalogs/barklem_t3.fits"))
+
+elements = np.unique(t["El"])
+stars = np.unique(t["Name"])
+
+S, E = (len(stars), len(elements))
+X = np.nan * np.ones((S, E))
+
+for i, row in enumerate(t):
+    star_index = np.where(stars == row["Name"])[0][0]
+    element_index = np.where(elements == row["El"])[0][0]
+
+    X[star_index, element_index] = row["logEpsX"]
+
+N_ok = np.zeros(E)
+for i, element in enumerate(elements):
+    N_ok[i] = np.sum(np.isfinite(X[:, i]))
+
+element_indices = np.argsort(N_ok)[::-1]
+for i in element_indices:
+    print(elements[i], N_ok[i])
+
+# Take top 13
+use_element_indices = element_indices[:15]
[…]

### 2019-01-18 mcfa/mcfa.py

@@ -642,7 +642,7 @@ class MCFA(object):
             - 0.5 * (2 * J + 3) * np.sum(slogdet_omega) \
             - 0.5 * (K * J) * np.log(2)
 
-        return I_0 + I_1 + I_2
+        return 0#I_0 + I_1 + I_2
 
 
     def sample(self, n_samples=1, theta=None):
@@ -682,14 +682,31 @@ class MCFA(object):
         return X
 
 
-    def apply_rotation(self, R, atol=1e-3, rtol=1e-5):
+    def rotate(self, R, X=None, ensure_valid_rotation=True, atol=1e-3, rtol=1e-5):
         r"""
-        Rotate the factor loads and factor scores by the given rotation matrix.
+        Rotate the factor loads and factor scores by a valid rotation matrix.
 
         :param R:
             A `J` times `J` rotation matrix, where `J` is the number of latent
             factors.
+        
+        :param X: [optional]
+            The data, which is expected to be an array with shape [n_samples,
+            n_features]. If given, the log-likelihood will be evaluated before
+            and after rotation. A warning will be raised if the log-likelihood
+            changes by more than the convergence tolerance.
 
+        :param ensure_valid_rotation: [optional]
+            If the rotation matrix does not follow R @ R.T = I, then the nearest
+            rotation matrix with this property will be used.
+
+        :param atol: [optional]
+            The absolute tolerance acceptable for individual entries in the
+            matrix I - R @ R.T. Default is 1e-3.
+
+        :param rtol:
+            The relative tolerance acceptable for individual entries in the
+            matrix I - R @ R.T. Default is 1e-5.
         """
 
         # Check that it is a rotation matrix.
@@ -699,25 +716,39 @@ class MCFA(object):
             raise ValueError("rotation matrix is not square")
 
         if not np.allclose(R @ R.T, np.eye(J), atol=atol, rtol=rtol):
-            raise ValueError("R is not a valid rotation matrix")
+
+            # Do our own rotation.
+            L = linalg.cholesky(R.T @ R)
+            R = R @ linalg.solve(L, np.eye(self.n_latent_factors))
+
+            if not np.allclose(R @ R.T, np.eye(J), atol=atol, rtol=rtol):
+                raise ValueError("R is not a valid rotation matrix")
+
+        if X is not None:
[…]

### 2019-01-18 mcfa/mpl_utils.py

@@ -418,9 +418,9 @@ def plot_filled_contours(J, K, Z, N=1000, colorbar_label=None,
                          ax=None, **kwargs):
     power = None
         
-    if np.all(Z > 0):
-        power = np.min(np.log10(Z).astype(int))
-        Z = Z.copy() / (10**power)
+    #if np.all(Z > 0):
+    #    power = np.min(np.log10(Z).astype(int))
+    #    Z = Z.copy() / (10**power)
     
     if ax is None:
         w = 0.2 + 4 + 0.1
@@ -524,11 +524,12 @@ def plot_factor_loads(factor_loads, scales=1, separate_axes=False,
                       target_loads=None, show_target_loads=True, 
                       load_labels=None, n_rotation_inits=25,
                       flip_loads=None, xlabel=None, xticklabels=None,
-                      legend_kwds=None, figsize=None):
+                      legend_kwds=None, colors=None, figsize=None, **kwargs):
 
 
 
-    colors = plt.rcParams['axes.prop_cycle'].by_key()['color']
+    if colors is None:
+        colors = plt.rcParams['axes.prop_cycle'].by_key()['color']
 
     D, J = factor_loads.shape
     if flip_loads is None:
@@ -582,8 +583,10 @@ def plot_factor_loads(factor_loads, scales=1, separate_axes=False,
         axes = ax[j] if separate_axes else ax[0]
         color = colors[j % len(colors)]
 
+        default_kwds = dict(lw=2)
+        default_kwds.update(kwargs)
         axes.plot(xi, rotated_percentiles[1, :, j], "-", 
-                  lw=2, c=color, label=load_labels[j])
+                  c=color, label=load_labels[j], **default_kwds)
         axes.fill_between(xi, 
                           rotated_percentiles[0, :, j], 
                           rotated_percentiles[2, :, j],
@@ -593,7 +596,7 @@ def plot_factor_loads(factor_loads, scales=1, separate_axes=False,
             axes.plot(xi, target_loads.T[j], ":", c=color)
 
 
-    for axes in ax:
+    for i, axes in enumerate(ax):
 
         axes.set_xticks(xi)
         ylim = max(np.ceil(10 * np.abs(axes.get_ylim()).max()) / 10, 1)
@@ -616,7 +619,10 @@ def plot_factor_loads(factor_loads, scales=1, separate_axes=False,
             axes.set_xticklabels([""] * D)
 
         if axes.is_first_col():
[…]

### 2019-01-18 mcfa/utils.py

@@ -408,16 +408,21 @@ def find_rotation_matrix(A, B, init=None, n_inits=25, full_output=True, **kwargs
     return best_R
 
 
-def exact_rotation_matrix(A_true, A_est):
+def exact_rotation_matrix(A_true, A_est, p0=None, full_output=False, **kwargs):
 
     N, J = A_true.shape
 
     def cost(R):
         return np.sum(np.abs((A_true - A_est @ R.reshape((J, J))).flatten()))
 
-    p_opt = op.minimize(cost, np.zeros(J**2), method="Powell")
+    if p0 is None:
+        p0 = np.zeros(J**2)
 
-    return p_opt.x.reshape((J, J))
+    kwds = dict(method="Powell")
+    kwds.update(kwargs)
+    p_opt = op.minimize(cost, p0, **kwds)
+    
+    return p_opt.x.reshape((J, J)) if not full_output else p_opt
 
 
 

### 2019-01-22 article/mcfa.tex
lots of stuff -- ready for galah exp4

@@
 
-\shorttitle{A novel approach to chemical tagging in lower-dimensional latent space}
+\shorttitle{Chemical tagging in lower-dimensional latent space}
 \shortauthors{Casey et al.}
@@
 
-\title{A novel approach to chemical tagging in lower-dimensional latent space}
+\title{Chemical tagging in lower-dimensional latent space}
 
@@
 $\NumLatentFactors = 1$ and $K = 1$, and trialling up until 
-$\NumLatentFactors = \todo{8}$ and $\NumComponents = 5$. 
+$\NumLatentFactors = 7$ and $\NumComponents = 5$. 
 We found that models with higher numbers of latent factors were unable to
@@
 In Figure~\ref{fig:exp2-factor-loads} we show the rotated factor loads found by our
-preferred model, where the shaded regions indicate \emph{only} the projected uncertainty on
-the rotation angles -- and does not include any uncertainty on the factor loads 
-themselves.
+preferred model, and illustrate the relative contributions (positive or negative)
+to each element.
 
@@
 in part because the target loads only seek to \emph{identify} and \emph{order}
-factors that can be interpreted within an astrophysical context. For example,
+factors that can be interpreted within an astrophysical context. 
+\todo{For example,
 the first factor does show positive contributions to Al, like the target load,
 but it also shows a slight negative contribution with respect to alpha elements
-(Mg, Ca), and a gradual increase among Fe-peak elements up to Ni. Similarly while
+(Mg, Ca)}, and a gradual increase among Fe-peak elements up to Ni. Similarly while
 Ca and Mg are indeed positive in the second factor, as targeted, there is a
@@
 
-\begin{figure}
-	\includegraphics[width=0.5\textwidth]{experiments/exp2-latent-contributions.pdf}
-	\caption{Rotated latent factors found in Experiment~2 using the \citet{Barklem:2005}
-			 data. \todo{elaborate latent factors}}
-    \label{fig:exp2-latent-contribution}
-\end{figure}
-
-The relatively wide uncertainty regions on these factors is perhaps not
-surprising given that only 61 stars were used in this data set. However, this
-experiment hopefully demonstrates that the rotated latent factors have some
-interpretability. After accounting for the contributions by all
+After accounting for the contributions by all
 $J = 5$ factors, the specific scatter remaining in each dimension is shown in
 Figure~\ref{fig:exp2-specific-scatter}. The remaining scatter varies from as low 
-as 0.01 to 0.02~dex (Mg, Ca, Ti, Mn, Fe, Y) up to 0.08 or 0.16~dex (Al and Ba). 
-% TODO: matches Barklem variances?
+as 0.08~dex (Mn, Y) up to 0.40~dex (Al, Ba, Ni). Al is not found to have any 
+predominant factor in these data: it has non-zero contributions from all loads
+(Figure~\ref{fig:exp2-factor-loads}. A similar pattern is seen in Cr, Co, and Ni,
+although the specific scatter for Cr and Co is half that of what is found for
+Ni (0.2\,dex instead of 0.4\,dex).
+
+% TODO: anything else to say here?
 
 \begin{figure*}
-	\includegraphics[width=\textwidth]{experiments/exp2-latent-factors.png}
+	\includegraphics[width=\textwidth]{experiments/exp2-latent-factors-visualize.pdf}
 	\caption{Rotated latent factors found in Experiment~2 using the \citet{Barklem:2005}
-			 data. \todo{elaborate latent factors}}
-    \label{fig:exp2-latent-factors}
+			 data. The panels on the left show the entries for each factor load. On
+			 the right hand side we show the absolute fractional contributions to
+			 each element, which illustrates the loads that contribute most to each
+			 element (either by positive or negative factor loads).}
+    \label{fig:exp2-factor-loads}
 \end{figure*}
+
+%The relatively wide uncertainty regions on these factors is perhaps not
+%surprising given that only 61 stars were used in this data set. However, this
+%experiment hopefully demonstrates that the rotated latent factors have some
+%interpretability.
 

### 2019-01-22 article/experiments/exp2.py

@@ -192,7 +192,8 @@ max_n_components = 3
 Js = 1 + np.arange(max_n_latent_factors)
 Ks = 1 + np.arange(max_n_components)
 
-Jg, Kg, converged, metrics = grid_search.grid_search(Js, Ks, X, mcfa_kwds)
+Jg, Kg, converged, metrics = grid_search.grid_search(Js, Ks, X, 
+                                                     mcfa_kwds=mcfa_kwds)
 
 ll = metrics["ll"]
 bic = metrics["bic"]
@@ -289,6 +290,14 @@ R = R_opt if chi2 < chi1 else R
 
 # Now make it a valid rotation matrix.
 model.rotate(R, X=X, ensure_valid_rotation=True)
+J = model.n_latent_factors
+L = model.theta_[model.parameter_names.index("A")]
+cmap = mpl_utils.discrete_cmap(2 + J, base_cmap="Spectral_r")
+colors = [cmap(1 + j) for j in range(J)]
+
+fig = mpl_utils.visualize_factor_loads(L, label_names, colors=colors)
+savefig(fig, "latent-factors-visualize")
+
 
 load_labels=[
   r"$\textrm{Al}$",
@@ -299,13 +308,14 @@ load_labels=[
 ]
 
 #colors = plt.rcParams['axes.prop_cycle'].by_key()['color']
-cmap = mpl_utils.discrete_cmap(2 + J, base_cmap="Spectral_r")
+cmap = mpl_utils.discrete_cmap(2 + model.n_latent_factors, base_cmap="Spectral_r")
 #cmap = matplotlib.cm.Set2
 colors = [cmap(1 + j) for j in range(J)]
 
 
 
 A_est = model.theta_[model.parameter_names.index("A")]
+
 
 
 fig_factors_rotated = mpl_utils.plot_factor_loads(A_est, separate_axes=True,
@@ -326,10 +336,10 @@ savefig(fig_factors_rotated, "latent-factors")
 # Plot the specific variances.
 fig_scatter = mpl_utils.plot_specific_scatter(model, 
                                               steps=True,
-                                              xlabel=xlabel, 
+                                              xlabel="", 
                                               xticklabels=xticklabels,
                                               ylabel=r"$\textrm{specific scatter / dex}$")
-fig_scatter.axes[0].set_yticks(np.linspace(0, 0.25, 6))
[…]

### 2019-01-22 article/experiments/exp2a-test.py

@@ -1,0 +1,167 @@
+
+import os
+import sys
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import pickle
+from scipy import linalg
+
+from matplotlib.ticker import MaxNLocator
+
+sys.path.insert(0, "../../")
+from mcfa import (mcfa, grid_search, mpl_utils, utils)
+
+matplotlib.style.use(mpl_utils.mpl_style)
+
+here = os.path.dirname(os.path.realpath(__file__))  
+
+random_seed = 10
+
+np.random.seed(random_seed)
+
+def savefig(fig, suffix):
+    prefix = os.path.basename(__file__)[:-3]
+    filename = os.path.join(here, f"{prefix}-{suffix}")
+    fig.savefig(f"{filename}.pdf", dpi=300)
+    fig.savefig(f"{filename}.png", dpi=150)
+    print(f"Created figures {filename}.png and {filename}.pdf")
+    
+
+from astropy.table import Table
+
+
+periodic_table = """H                                                  He
+                    Li Be                               B  C  N  O  F  Ne
+                    Na Mg                               Al Si P  S  Cl Ar
+                    K  Ca Sc Ti V  Cr Mn Fe Co Ni Cu Zn Ga Ge As Se Br Kr
+                    Rb Sr Y  Zr Nb Mo Tc Ru Rh Pd Ag Cd In Sn Sb Te I  Xe
+                    Cs Ba Lu Hf Ta W  Re Os Ir Pt Au Hg Tl Pb Bi Po At Rn
+                    Fr Ra Lr Rf"""
+
+lanthanoids    =   "La Ce Pr Nd Pm Sm Eu Gd Tb Dy Ho Er Tm Yb"
+actinoids      =   "Ac Th Pa U  Np Pu Am Cm Bk Cf Es Fm Md No"
+
+periodic_table = periodic_table.replace(" Ba ", " Ba " + lanthanoids + " ") \
+                               .replace(" Ra ", " Ra " + actinoids + " ") \
+                               .split()
+
+t = Table.read(os.path.join(here, "../../catalogs/barklem_t3.fits"))
+
+elements = np.unique(t["El"])
+stars = np.unique(t["Name"])
+
+S, E = (len(stars), len(elements))
+X = np.nan * np.ones((S, E))
+
+for i, row in enumerate(t):
+    star_index = np.where(stars == row["Name"])[0][0]
[…]

### 2019-01-22 article/experiments/exp3.py

@@ -19,68 +19,108 @@ from mcfa import (mcfa, grid_search, mpl_utils, utils)
 
 import galah
 
+np.random.seed(100)
+
 
 matplotlib.style.use(mpl_utils.mpl_style)
 
 
 def savefig(fig, suffix):
+    here = os.path.dirname(os.path.realpath(__file__))
     prefix = os.path.basename(__file__)[:-3]
-    filename = f"{prefix}-{suffix}.png"
-    fig.savefig(filename, dpi=150)
+    filename = os.path.join(here, f"{prefix}-{suffix}")
+    fig.savefig(f"{filename}.png", dpi=150)
+    fig.savefig(f"{filename}.pdf", dpi=300)
 
 
-mcfa_kwds = dict(init_factors="random", init_components="random", tol=1e-5,
-                 max_iter=10000, random_seed=42)
+mcfa_kwds = dict(init_factors="svd", init_components="kmeans++", tol=1e-8,
+                 max_iter=10000)
 
+
+# Use a few carefully-selected clusters.
 
 use_galah_flags = True
 
-cluster_names = None
+#cluster_names = ("M67", "Trumpler20", "NGC2204")
+cluster_names = ("NGC1817", "NGC2204", "Berkeley32", "M67", "Trumpler20", "NGC6253",
+                 "Ruprecht147")
+cluster_names = ()
 
 abundance_counts = galah.get_abundances_breakdown(galah.available_elements,
                                                   cluster_names,
                                                   use_galah_flags=use_galah_flags)
 
 print(f"Abundance counts: {abundance_counts}")
+for k, v in abundance_counts.items():
+    print(k, v)
 
+
+
+
+
+
+
+"""
 elements = "Fe, Mg, Na, K, Sc, Ca, O, Zn, Al, Cu, Mn, Cr, Ba, Ti, Ni, Si, Y, "\
            "Co, V, La, Ce, Zr, Eu".split(", ")
 
 elements = ["Fe", "Ni", "Mn", "Cr", "Ti", "Al", "Mg", "Na", "K", "Sc", "Ca", "O", "Zn", "Ba", "Y", "Ba", "Eu"]
+"""
 
+#elements = "Mg Al Ca Sc Ti Cr Mn Fe Co Ni Sr Y Ba".split()
+elements = "Fe Mg Na K Sc Ca O Zn Al Cu Mn Cr Ba Ti Ni Si Y Eu".split()
[…]

### 2019-01-22 article/experiments/exp4.py

@@ -10,6 +10,7 @@ import sys
 import numpy as np
 import matplotlib
 import matplotlib.pyplot as plt
+from matplotlib.ticker import MaxNLocator
 from collections import Counter
 from scipy import linalg
 
@@ -22,42 +23,101 @@ import galah_dr3 as galah
 
 matplotlib.style.use(mpl_utils.mpl_style)
 
+here = os.path.dirname(os.path.realpath(__file__))
+
 
 def savefig(fig, suffix):
+    here = os.path.dirname(os.path.realpath(__file__))
     prefix = os.path.basename(__file__)[:-3]
-    filename = f"{prefix}-{suffix}.png"
-    fig.savefig(filename, dpi=150)
+    filename = os.path.join(here, f"{prefix}-{suffix}")
+    fig.savefig(f"{filename}.png", dpi=150)
+    fig.savefig(f"{filename}.pdf", dpi=300)
+
+
+N_elements = 18
+use_galah_flags = True
 
 
 mcfa_kwds = dict(init_factors="random", init_components="random", tol=1e-5,
-                 max_iter=10000, random_seed=42)
-
-
-use_galah_flags = 0
+                 max_iter=10000)
 
 
 abundance_counts = galah.get_abundances_breakdown(galah.available_elements,
                                                   use_galah_flags=use_galah_flags)
 
-print(f"Abundance counts: {abundance_counts}")
+xlabels = abundance_counts.keys()
+latex_labels = [r"{1}$\textrm{{{0}}}$".format(ea, "\n" if i % 2 else "") for i, ea in enumerate(xlabels)]
+x = np.arange(len(xlabels))
+y_independent = np.array(list(abundance_counts.values()))
+y_cumulative = np.zeros_like(y_independent)
 
-elements = "Fe, Mg, Na, K, Sc, Ca, O, Zn, Al, Cu, Mn, Cr, Ba, Ti, Ni, Si, Y, "\
-           "Co, V, La, Ce, Zr, Eu".split(", ")
+mask = np.ones(len(galah.data), dtype=bool)
 
-elements = ["Na", "Mg", "Fe", "Sc", "Ti", "Zn", "Mn", "Y", "Ca", "Ni", "Cr", "O",
-            "Si", "K", "Ba", "V", "Cu", "Al", "La", "Eu"]
-#            # "Ni", "Mn", "Cr", "Ti", "Al", "Mg", "Na", "K", "Sc", "Ca", "O", "Zn", "Ba", "Y", "Ba", "Eu"]
-
[…]

### 2019-01-22 article/experiments/galah.py

@@ -4,8 +4,11 @@ import numpy as np
 from astropy.table import Table
 from collections import OrderedDict
 from operator import itemgetter
+import os
 
-data = Table.read("GALAH_iDR3_OpenClusters.fits")
+here = os.path.dirname(os.path.realpath(__file__))
+data = Table.read(os.path.join(here, "GALAH_iDR3_OpenClusters.fits"))
+
 
 
 periodic_table = """H                                                  He

### 2019-01-22 article/experiments/galah_dr3.py

@@ -4,10 +4,12 @@ import numpy as np
 from astropy.table import Table
 from collections import OrderedDict
 from operator import itemgetter
-
+import os
 #data = Table.read("GALAH_iDR3_OpenClusters.fits")
 
-data = Table.read("../../catalogs/GALAH_DR2.1_catalog.fits")
+here = os.path.dirname(os.path.realpath(__file__))
+
+data = Table.read(os.path.join(here, "../../catalogs/GALAH_DR2.1_catalog.fits"))
 
 periodic_table = """H                                                  He
                     Li Be                               B  C  N  O  F  Ne

### 2019-01-22 mcfa/grid_search.py

@@ -7,8 +7,8 @@ from .mcfa import MCFA
 
 logger = logging.getLogger(__name__)
 
-def grid_search(trial_n_latent_factors, trial_n_components, X, mcfa_kwds=None, 
-                pseudo_bic_kwds=None):
+def grid_search(trial_n_latent_factors, trial_n_components, X, N_inits=1,
+                mcfa_kwds=None, pseudo_bic_kwds=None):
 
     Js = np.array(trial_n_latent_factors)
     Ks = np.array(trial_n_components)
@@ -31,26 +31,35 @@ def grid_search(trial_n_latent_factors, trial_n_components, X, mcfa_kwds=None,
     for j, J in enumerate(Js):
         for k, K in enumerate(Ks):
 
-            print(f"At J = {J}, K = {K}")
+            models = []
+            for n in range(N_inits):
 
-            model = MCFA(n_latent_factors=J, n_components=K, **mcfa_kwds)
+                print(f"At J = {J}, K = {K}, N = {n}")
 
-            try:
-                model.fit(X)
+                model = MCFA(n_latent_factors=J, n_components=K, **mcfa_kwds)
 
-            except:
-                logger.exception(f"Exception occurred during grid search at "\
-                                 f"J = {J}, K = {K}:")
-                continue
+                try:
+                    model.fit(X)
 
-            else:
+                except:
+                    logger.exception(f"Exception occurred during grid search at "\
+                                     f"J = {J}, K = {K}:")
+                    continue
+
+                else:
+                    models.append(model)
+
+            if len(models) > 0:
+                idx = np.nanargmax([model.log_likelihood_ for model in models])
+                model = models[idx]
+
                 ll[k, j] = model.log_likelihood_
                 bic[k, j] = model.bic(X)
                 mml[k, j] = model.message_length(X)
                 pseudo_bic[k, j] = model.pseudo_bic(X, **pseudo_bic_kwds)
                 converged[k, j] = True
 
-
+    # Best of each?
     metrics = dict(ll=ll, bic=bic, pseudo_bic=pseudo_bic, message_length=mml)
     return (J_grid, K_grid, converged, metrics)#ll, bic, pseudo_bic)
 

### 2019-01-22 mcfa/mcfa.py

@@ -417,7 +417,7 @@ class MCFA(object):
 
         assert np.isfinite(current)
         if previous > current:
-            logger.warn(f"Log-likelihood *decreased* by {previous-current:.2f}")
+            logger.warn(f"Log-likelihood *decreased* by {previous-current:.2e}")
 
         #assert current > previous # depends on objective function
 
@@ -707,6 +707,9 @@ class MCFA(object):
         :param rtol:
             The relative tolerance acceptable for individual entries in the
             matrix I - R @ R.T. Default is 1e-5.
+
+        :returns:
+            The actual rotation matrix applied.
         """
 
         # Check that it is a rotation matrix.
@@ -745,7 +748,7 @@ class MCFA(object):
                 logger.warning(f"Log-likelihood changed after rotation: "\
                                f"{ll - ll_r} (>{self.tol})")
 
-        return True
+        return R
 
 
     

### 2019-01-22 mcfa/mpl_utils.py

@@ -431,12 +431,10 @@ def plot_filled_contours(J, K, Z, N=1000, colorbar_label=None,
     else:
         fig = ax.figure
 
-
     #levels = np.linspace(np.floor(np.min(Z)), np.ceil(np.max(Z)), N).astype(int)
     #norm = BoundaryNorm(levels, 256)
 
-    cf = ax.contourf(J, K, Z, N, vmin=np.min(Z), vmax=np.max(Z), 
-                     **kwargs)
+    cf = ax.contourf(J, K, Z, N, vmin=np.nanmin(Z), vmax=np.nanmax(Z), **kwargs)
 
     ax.set_xlabel(r"$\textrm{Number of latent factors } J$")
     ax.set_ylabel(r"$\textrm{Number of clusters } K$")
@@ -518,6 +516,93 @@ def plot_filled_contours(J, K, Z, N=1000, colorbar_label=None,
         c.set_edgecolor("face")
 
     return fig
+
+
+def visualize_factor_loads(L, label_names=None, colors=None, line_kwds=None,
+                          **kwargs):
+
+    L = np.atleast_2d(L)
+    D, J = L.shape
+
+    #fig = plt.figure()
+    #gs = matplotlib.gridspec.GridSpec(J, 2, width_ratios=[2, 1], height_ratios=np.hstack([np.ones(J)/J, 1]))
+    #axes = [fig.add_subplot(gs[j]) for j in range(J + 1)]
+
+    if label_names is not None:
+        latex_label_names = [r"$\textrm{{{0}}}$".format(ea) for ea in label_names]
+
+
+    #fig, axes = plt.subplots(1 + J)
+    fig = plt.figure()
+
+    K = 4
+    shape = (J, K)
+    axes = [plt.subplot2grid(shape, (j, 0), colspan=K-1) for j in range(J)]
+    axes += [plt.subplot2grid(shape, (0, K - 1), rowspan=J, colspan=1)]
+
+    if colors is None:
+        cmap = discrete_cmap(J, base_cmap="Spectral_r")
+        colors = [cmap(j) for j in range(J)]
+
+
+    line_kwds_ = dict(lw=5)
+    line_kwds_.update(line_kwds or dict())
+
+    x = np.arange(D)
+    for j, (ax, color) in enumerate(zip(axes, colors)):
+
+        ax.plot(x, L.T[j], "-", c=color, **line_kwds_)
[…]

### 2019-01-25 article/mcfa.tex
words

@@
 \newcommand{\numcomponents}{k}
+\newcommand{\likelihood}{\mathcal{L}}
 
@@
 
-Here we use a variant of factor analysis known elsewhere as a mixture of common 
+Here we expand on a variant of factor analysis known elsewhere as a mixture of common 
 factor analyzers \citep{Baek:2010}, where the data are generated by a set of 
@@
 is independent of the latent space, and $\specificvariance$ is a
-diagonal matrix of $\NumDimensions$ entries. 
+column vector of $\NumDimensions$ entries. 
 In this model each data point can be represented as being drawn
@@
 
+\subsection{Model Selection}
+
+The expectation-maximization algorithm as described requires the number of latent
+factors $\NumLatentFactors$ and $\NumComponents$ to be given. In the next Section
+we describe experiments using generated and real data, and in all cases we will
+assume that the true number of latent factors and components are not known. 
+We require some criterion to decide how many latent factors, and how many
+components, we should adopt given some data. An increasing number of factors
+and components will undoubtedly increase the log likelihood of the model given
+the data, but the log likelihood does not account for the increased model 
+complexity that is afforded by those additional latent factors and components.
+
+One criterion commonly employed for evaluating a class of models is the 
+Bayesian Information Criterion \citep[BIC;][]{Schwarz:1978}, 
+\begin{equation}
+	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \quad , \label{eq:bic}
+\end{equation} 
+\noindent{}where $Q$ is the number of model parameters:
+\begin{equation}
+	Q = \frac{\NumLatentFactors}{2}\left[2\left(\NumDimensions - \NumLatentFactors\right) + \NumComponents\left(3 + \NumLatentFactors\right)\right] + \NumComponents + \NumDimensions - 1 \quad .
+\end{equation}
+\noindent{}We evaluate the BIC for all experiments performed in this work.
+However, while the BIC does include a penalisation term for the number of
+parameters (which scales with $\log{N}$), it does not account for the increased
+complexity that is afforded by additional parameters. Adding one parameter
+to a model will increase the BIC by at most $\log{N}$, but there are different
+ways for a single parameter to be introduced. In a fictitious model $y=f(x)$
+a parameter $b$ could be added that is a scalar multiple of $x$, or it could be
+introduced as $x^b$. Despite the difference in model complexity, the same
+penalisation occurs in BIC. Even if the log likelihood were only to improve
+marginally in both cases, the difference in model complexity is not captured
+by BIC. In other words, there are situations where we are more interested in
+balancing the model complexity (or the expected Fisher information and similar
+properties) with the goodness of fit, instead of penalising the number of 
+parameters.
+
+For these reasons we also calculate the Minimum Message Length \citep[MML;][]{Wallace:2005}
+for every model as a criterion for model selection and evaluation. 
+The classically-described princple of MML is that the best explanation of the
+data is the one that leads to the shortest so-called two-part message~\cite{Wallace:2005}, 
+where a \textit{message} takes into account both the complexity of the model 
+and its explanatory power. The complexity of the model is described through
+the first part of the message, and the second part of the message describes
+its explanatory power. The \emph{length} of each message part is quantified
+(or measured) using information theory, allowing for a fair evaluation between
+different models of varying complexity or explanatory power. MML has been 
+shown to perform well on a variety of empirical analyses (see, e.g., 
+\cite{viswanathan1999finding,fitzgibbon2004minimum}, with references to 
+further examples in \cite{Wallace:2005,dowe2007bayes,Dowe2008a,Dowe2011a}).
+Arguments about the statistical consistency (i.e.,~as the number of data 
+points increases the distributions of the estimates become increasingly 
+concentrated near the true value) of MML are given in \cite{DoweWallace1997a,Dowe2011a}.
+The MML principle requires that we explicitly specify our prior beliefs on the
+model parameters, providing a Bayesian optimisation approach which can be
+applied across entire classes of models.
+
+The \textit{message} must encode two parts: the model, and the data given the
+model. The encoding of the message is based on Shannon's information theory~\cite{Shannon:1948}. 
+The information gained from an event $e$ occurring, where $p(e)$ is the
+probability of that event, is $I(e) = -\log_{2}{p(e)}$. The information content
+is largest for improbable outcomes, and smallest for outcomes that we are 
+almost certain about. In other words, an outcome that has a probability close
+to unity has zero information content because nothing new is learned from it,
+whereas rarer events convey a much higher information content.
+
+In practice calculating the message length can be a non-trivial task, 
+especially for models that are reasonably complex. This makes the strict MML
+principle intractable (or uncomputable) in many cases and necessitating 
+approximations to the message length. Using a Taylor expansion, a generalised
+scheme can be calculated to estimate the parameter vector $\vec\Psi$ that
+minimises the message length ${I}(\vec\Psi,\vecdata)$ \citep{Wallace:1987},
+\begin{equation}
+	{I}(\vec\Psi,\vecdata) = \frac{Q}{2}\log\kappa\left(Q\right) - \log\left(\frac{p(\vec\Psi)}{\sqrt{|\mathcal{F}(\vec\Psi)|}}\right) - \log\mathcal{L}\left(\vecdata|\vec\Psi\right)% + \frac{Q}{2}
+\end{equation}
+\noindent{}where $\log\likelihood(\vecdata|\vec\Psi)$ is the familiar
+log likelihood, $p(\vec\Psi)$ is the joint prior density on $\vec\Psi$,
+$\mathcal{F}(\vec\Psi)$ is the negative second derivative of the log likelihood,
+commonly referred to as the expected Fisher information matrix,
+\begin{equation}
+	\mathcal{F}(\vec\Psi) = -\textrm{E}\left[\frac{\partial^2}{\partial\vec\Psi^2}\log\likelihood(\vecdata|\vec\Psi)\right]
+\end{equation}
+\noindent{}and as before $Q$ is the number of model parameters.
+Finally, the $\log\kappa(Q)$ term is given by
+\begin{equation}
+	\todo{\log\kappa} \quad .
+\end{equation}
+
+Like the BIC, the message length is penalised by the number of model parameters
+through the $\log\kappa(Q)$ term. However, the model complexity is also 
+described through the priors and the Fisher information matrix, which
+describes the curvature of the log likelihood with respect to the model
+parameters. \todo{Dowe said something about MML reducing to BIC under some conditions, but I have not yet found the right paper to cite.}
+
+We will describe the contributions to the message length in parts. We assume
+the priors on the number of latent factors $\NumLatentFactors$ and the
+number of components $\NumComponents$ to be
+$p(\NumLatentFactors) \propto 2^{-\NumLatentFactors}$ and
+$p(\NumComponents) \propto 2^{-\NumComponents}$ respectively,
+such that the optimal lossless message to encode them is \citep[p279, sec. 6.8.2][]{someone},
+\begin{eqnarray}
+	I(\NumLatentFactors) &= -\log{p(\NumLatentFactors)} &= \NumLatentFactors\log{2} + \textrm{constant} \\
+	I(\NumComponents) &= -\log{p(\NumComponents)} &= \NumComponents\log{2} + \textrm{constant} \quad .
+\end{eqnarray}
+Only $\NumComponents - 1$ of the relative weights $\vec\weight$ need encoding because 
+$\sum_{\numcomponents=1}^{\NumComponents} = 1$. We assume a uniform prior on 
+individual weights,
+\begin{equation}
+	p(\vec\weight) = (\NumComponents - 1)! \quad ,
+\end{equation}
+\noindent{}and the Fisher information is
+\begin{equation}
+	\mathcal{F}(\vec\weight) = \frac{\NumData^{\NumComponents - 1}}{\prod_{\numcomponents=1}^{\NumComponents}\weight_\numcomponents} \quad ,
+\end{equation}
+\noindent{}which gives the message length of the relative weights $I(\vec\weight)$ to be
+\begin{eqnarray}
+	I(\vec\weight) &=& -\log\left(\frac{p(\vec\weight)}{\sqrt{|\mathcal{F}(\vec\weight)|}}\right) \nonumber \\
+			   &=& -\log{p(\vec\weight)} - \frac{1}{2}\log{|\mathcal{F}(\vec\weight)|} \nonumber  \\
+			   &=& -\log(\NumComponents - 1)! + \frac{\NumComponents - 1}{2}\log{\NumData} - \frac{1}{2}\sum_{\numcomponents=1}^{\NumComponents}\log\weight_\numcomponents \nonumber \\
+	I(\vec\weight) &=& \frac{1}{2}\left(\left(\NumComponents - 1\right)\log\NumData - \sum_{\numcomponents=1}^{\NumComponents}\log\weight_\numcomponents\right) -\log\Gamma(\NumComponents) \quad .
+\nonumber \\
+\end{eqnarray}
+
+We assume uniform priors for the component means in latent space $\scoremeans$, 
+and a conjugate inverted Wishart prior for the component covariance matrices
+$\scorecovs$ \citep[Section 5.2.3][]{someone},
+\begin{equation}
+	p(\scoremeans_\numcomponents,\scorecovs_\numcomponents) \propto |\scorecovs_\numcomponents|^{\frac{1}{2}(\NumLatentFactors + 1)} \quad .
+\end{equation}
+We approximate the determinate of the Fisher information of a multivariate normal $|\mathcal{F}(\scoremeans,\scorecovs)|$
+as $|\mathcal{F}(\scoremeans)||\mathcal{F}(\scorecovs)|$ \citep{25,27} where
+\begin{equation}
+	|\mathcal{F}(\scoremeans)| = (\NumData\weight_k)^\NumLatentFactors|\scorecovs_k|^{-1}
+\end{equation}
+\begin{equation}
+	|\mathcal{F}(\scorecovs)| = (\NumData\weight_k)^{\frac{1}{2}\NumLatentFactors(\NumLatentFactors+1)}2^{-\NumLatentFactors}|\scorecovs_k|^{-(\NumData+1)}
+\end{equation}
+\noindent{}such that 
+% todo put line in equation.
+\begin{eqnarray}
+	I(\scoremeans,\scorecovs) &=& -\sum_{\numcomponents=1}^{\NumComponents}\log{p(\scoremeans_k,\scorecovs_k)} + \frac{1}{2}\sum_{\numcomponents=1}^{\NumComponents}\log{|\mathcal{F}(\scoremeans_k,\scorecovs_k)|} \nonumber \\
+	&=& \frac{1}{2}\sum_{\numcomponents=1}^{\NumComponents}\log\left[(\NumData\weight_k)^{\frac{1}{2}\NumLatentFactors(\NumLatentFactors+3)}2^{-\NumData}|\scorecovs_k|^{-(\NumData + 2)}\right] \nonumber \\
+	&& \cdots \,\, -\sum_{\numcomponents=1}^{\NumComponents}\log{|\scorecovs_k|}^{\frac{1}{2}(\NumLatentFactors + 1)} \\
+I(\scoremeans,\scorecovs) &=& \frac{1}{4}\NumLatentFactors(\NumLatentFactors+3)\sum_{\numcomponents=1}^\NumComponents\log{\NumData\weight_k} - \frac{KD}{2}\log{2} \nonumber \\ 
+&& \cdots \,\, -\frac{1}{2}(2\NumLatentFactors+3)\sum_{k=1}^{K}\log{|\scorecovs_k|}  \quad .
+\end{eqnarray}
+
+Previous work on multiple latent factor analysis within the context of MML have
+addressed the indeterminacy between the factor loads and factor scores by
+placing a joint prior on the product of factor loads and scores, whilst
+ensuring that the factor loads themselves remain mutually orthogonal \citep{Wallace}.
+Adopting the same prior density in our model is not practical because 
+it would require priors on $\scoremeans$, $\scorecovs$, and the 
+responsibility matrix $\vec\tau$. Instead, we address this indeterminacy
+by placing a prior on $\factorloads$ that ensures it is mutually orthogonal.
+Specifically, we adopt a Wishart distribution with scale matrix $\vec{S}$
+and $D$ degrees of freedom for the
+$\NumLatentFactors\times\NumLatentFactors$ matrix $\vec{M} = \factorloads\transpose\factorloads$.
+In other words, $\vec{M} \sim W_\NumLatentFactors(D,\vec{S})$
+and $\vec{S} = \textrm{Cov}(\factorloads\transpose)$.
+This
+Wishart joint prior density gives highest support for mutually orthogonal vectors,
+\begin{equation}
+	p(A) = \frac{|\factorloads\transpose\factorloads|^{\frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)}}{2^{\frac{\NumDimensions\NumLatentFactors}{2}}|\vec{S}|^{\frac{\NumDimensions}{2}}\Gamma(\frac{\NumDimensions}{2})}\exp\left[-\frac{1}{2}\textrm{Tr}(\vec{S}^{-1}\factorloads\transpose\factorloads)\right] \quad .
+\end{equation}
+The Fisher information for $\factorloads$ \todo{is hard to compute. I have 
+the first derivative from the log likelihood function, but the second
+derivative is (surprisingly) causing some trouble to compute.}
+Thus the message length to encode $\factorloads$ is given by
+\begin{eqnarray}
+	I(\factorloads) &=& -\log\left(\frac{p(\factorloads)}{\sqrt{|\mathcal{F}(\factorloads)|}}\right) \nonumber \\
+	&=& -\log{p(\factorloads)} + \frac{1}{2}\log{|\mathcal{F}(\factorloads)|} \nonumber \\
+I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} \nonumber \\
+	&& \cdots \,\, + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \quad .\nonumber \\
+\end{eqnarray}
+
+Combining equations \ref{eq:all} leads to the full message length
+\begin{equation}
+\todo{here}
+\end{equation}
+
 \section{Experiments} \label{sec:experiments}
@@
 We generated a data set with ${\NumData = 1}$00,000 data points, each with
-$\NumDimensions = 15$ dimensions, and assumed that those data are generated by
-latent space of $\NumLatentFactors = 5$ ($\NumLatentFactors \times \NumDimensions$) factor loads, and there are $\NumComponents = 20$
-clusters in the latent space. The relative weights $\vec\weight$
+$\NumDimensions = 15$ dimensions. We adopted a latent dimensional space of 
+$\NumLatentFactors = 5$ ($\NumLatentFactors \times \NumDimensions$) factor loads, 
+with $\NumComponents = 20$ clusters in the latent space. The relative weights $\vec\weight$
 are drawn from a multinomial distribution and the means of the clusters
@@
 loads and component assignments randomly. We performed expectation-maximization 
-cycles until the relative log-likelihood improved by less than $10^{-5}$. We then
-recorded the log-likelihood of the data given the model parameters, the
-Bayesian Information Criterion \citep[BIC;][]{Schwarz:1978}, 
-\begin{equation}
-	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \quad , \label{eq:bic}
-\end{equation} 
-\noindent{}where $Q$ is the number of model parameters:
-\begin{equation}
-	Q = \frac{\NumLatentFactors}{2}\left[2\left(\NumDimensions - \NumLatentFactors\right) + \NumComponents\left(3 + \NumLatentFactors\right)\right] + \NumComponents + \NumDimensions - 1 \quad .
-\end{equation}
+cycles until the relative log-likelihood improved by less than $10^{-5}$.
 
@@
 
-% TODO: Make single figure with bigger text labels.
 \begin{figure*}
-	\includegraphics[width=0.33\textwidth]{experiments/exp1-compare-loads.pdf}
-	\includegraphics[width=0.33\textwidth]{experiments/exp1-compare-scores.pdf}
-	\includegraphics[width=0.33\textwidth]{experiments/exp1-compare-specific-scatter.pdf}
+	\includegraphics[width=\textwidth]{experiments/exp1-compare-all-mml.pdf}
     \caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
@@
 
-\subsection{Experiment~3: A subset of gravitationally bound stellar clusters 
-			from \Galah\ and elsewhere}
-\label{sec:exp3}
-
-The previous experiment has demonstrated that the latent factors can be at least
-partially interpreted within the context of astrophysical processes. In this
-experiment we seek to test whether gravitationally bound stellar clusters can be
-reliably identified by their chemical abundances alone, and using only the model 
-described in Section~\ref{sec:model}.
-
-\todo{description  of the open cluster data used here}
-
-\subsection{Experiment 4: Chemical abundances in the \Galah\ survey}
+\subsection{Experiment 4:\\Chemical abundances in the \Galah\ survey}
 \label{sec:exp4}
 
-We use the second \Galah\ data release \citep{Buder:2018a} which
-includes up to 23 chemical abundances measured for 342,682
-stars. It is not practical to use all of these measurements as-is
-when performing chemical tagging. For example, our experiments
-with \Galah\ data do not include lithium abundances because the
-photospheric lithium varies throughout a star's lifetime. 
+In this experiment we perform blind chemical tagging using the 
+photospheric abundances released as part of the second \Galah\ 
+data release \citep{Buder:2018}. This data set includes
+up to 23 chemical abundances reported for 342,682 stars.
+Some abundances are not reported, for a variety of 
+reasons related to astrophysics or the adopted analyses
+(e.g., a star is too hot to estimate the abundance for a given
+transition, or the signal-to-noise of the spectrum is too low).
+Our method as described cannot handle missing data entries for
+a given star, and as such we are forced to exclude stars that
+do not have a full complement of abundances. However, we also
+chose to exclude some elemental abundances from our chemical
+tagging experiments. For example, our experiments
+with \Galah\ data do not include lithium or carbon abundances because the
+photospheric varies throughout a star's lifetime. This is true
+to a small degree for many elements \citep[e.g.,][]{Dotter_2008},
+but for the purposes of this experiment we assume that all other
+photospheric abundances remain constant throughout  a star's
+lifetime.
+
 We first selected stars with \texttt{flag\_cannon = 0} to exclude
@@
 (e.g., $\teff$, $\logg$) are unreliable, and as a result the 
-detailed chemical abundances are untrustworthy.
-
-We included the following fourteen chemical abundances: \todo{Na, Mg, 
-Si, Ca, Sc, Ti, Cr, Mn, Fe, Ni, Cu, Zn, Ba, and Eu}.
-By requiring that each star has \texttt{flag\_cannon = 0} and
-\texttt{flag\_<x> = 0} (where \texttt{<x>} is the relevant label)
-for each chemical abundance, this leaves
-us with a sample of \todo{12,798} stars each with reliable stellar
-parameters ($\teff$, $\logg$) and chemical abundances.
-This is relevant to the approach we use here, as the method
-described here requires no missing data.
+detailed chemical abundances are untrustworthy. We ordered the
+list of chemical abundances by the number of reliable measurements,
+which is shown in Figure~\ref{fig:exp4-abundance-counts}. The number of
+independent abundance measurements (e.g., where there is no reason
+to suspect a single abundance to be untrustworthy) is shown in black,
+and the cumulative number of trustworthy abundances is shown in blue.
+The cumulative line can be interpreted as the number of stars for
+which the full set of elemental abundances (up until and including
+the abundance on the x-axis) is reported. We chose to include the
+18 most commonly reported abundances (up to and including Al),
+leaving us with a sample of \todo{50,000} stars.
 These elements are produced from multiple nucleosynthetic
@@
 
-\todo{we perform a grid-based search for J and K}
+\begin{figure}
+	\includegraphics[width=0.45\textwidth]{experiments/exp4-abundance-counts.png}
+	\caption{The number of abundance measurements in the second
+		     \Galah\ data release \citep{Buder:2018} where \texttt{flag\_cannon = 0}
+		     and \texttt{flag\_<element> = 0}. The black line indicates the number
+		     of reported measurements for a single element, and the blue line
+		     indicates the number of stars where all elements leftward (inclusive)
+		     are reported. The dashed vertical line indicates the top 18 reported
+		     elemental abundances, which we use for Experiment~4.}
+    \label{fig:exp4-abundance-counts} 
+\end{figure}
 
-\subsection{Experiment 5: More chemical abundances in \Galah}
-\label{sec:exp5}
+We executed a grid search for the number of latent factors $\NumLatentFactors$
+and the number of components $\NumComponents$ that were preferred by the data.
+Starting with $\NumLatentFactors = 1$ and $\NumComponents = 1$, we trialled each
+permutation of $\NumLatentFactors$ and $\NumComponents$ up until $\NumLatentFactors = \NumDimensions - 1 = 17$
+and $\NumComponents = 20$. For each permutation we initialised the model randomly,
+and continued expectation-maximization until a tolerance of $10^{-5}$ was achieved
+in the log-likelihood. The results of this grid search are shown in Figure~\ref{fig:exp4-gridsearch-contours},
+where we show the best log-likelihood found for each permutation, and the corresponding
+BIC value. The lowest BIC value is found for $\todo{\NumLatentFactors = X}$ and
+$\todo{\NumComponents = X}$, which we adopted as the preferred number of latent
+factors and components.
 
-\todo{which abundances, only require finite and \texttt{flag\_cannon} to be OK}
+\todo{
+\begin{enumerate}
+	\item Plot data with most likely associations? Just as [Fe/H] vs [Si/Fe]?
+	\item Plot factor loads, after rotation.
+	\item Plot latent space.
+	\item Plot aspects of the clusters identified.
+	\item Plot specific scatter compared to the typical reported errors?
+\end{enumerate}
+}
 
@@
 Discovery Project DP160100637.
-% Gaia acknowledgement?
-% GALAH acknowledgement?
+The \GALAH\ survey is based on observations made at the Australian Astronomical Observatory, under programmes A/2013B/13,
+A/2014A/25, A/2015A/19, A/2017A/18. We acknowledge the traditional owners of the land on which the AAT stands, the Gamilaraay
+people, and pay our respects to elders past and present.
+This work presents results from the European Space Agency (ESA) space mission Gaia.
+Gaia data is being processed by the Gaia Data Processing and Analysis Consortium (DPAC).
+Funding for the DPAC is provided by national institutions, in particular the institutions participating in the Gaia MultiLateral Agreement (MLA).
+The Gaia mission website is https://www.cosmos.esa.int/gaia .
+The Gaia archive website is https://archives.esac.esa.int/gaia .
+This research has made use of NASA's Astrophysics Data System.
 
@@
 
-\appendix
-
-\section{Nomenclature}\label{app:symbols}
-
-\todo{summarise symbols}
-
 \bibliographystyle{aasjournal}

### 2019-01-25 mcfa/grid_search.py
stuff

@@ -34,6 +34,7 @@ def grid_search(trial_n_latent_factors, trial_n_components, X, N_inits=1,
             models = []
             for n in range(N_inits):
 
+
                 print(f"At J = {J}, K = {K}, N = {n}")
 
                 model = MCFA(n_latent_factors=J, n_components=K, **mcfa_kwds)
@@ -58,6 +59,7 @@ def grid_search(trial_n_latent_factors, trial_n_components, X, N_inits=1,
                 mml[k, j] = model.message_length(X)
                 pseudo_bic[k, j] = model.pseudo_bic(X, **pseudo_bic_kwds)
                 converged[k, j] = True
+
 
     # Best of each?
     metrics = dict(ll=ll, bic=bic, pseudo_bic=pseudo_bic, message_length=mml)

### 2019-01-25 mcfa/mcfa.py

@@ -592,57 +592,57 @@ class MCFA(object):
         pi, A, xi, omega, psi = theta
         J, K = self.n_latent_factors, self.n_components
 
-        # In Wallace nomenclature, J = number of factors (here also J)
-        #                          K = dimensionality of data (here is D)
-        #                          N = number of data (here also N)
+        # Let's do this by parts.
+        I_parts = dict()
 
-        # Recall:
-        # X! = gamma(X + 1)
-        # S_m = (m * np.pi**(m/2))/((m/2)!)
-        log_S = lambda m: np.log(m) + (m/2) * np.log(np.pi) - gammaln(m/2 + 1)
+        # Mixing weights.
+        # I(w) = -log\gamma(K) + 0.5 * (K-1)\log{NN} -0.5\sum_{K}\log\weight_k
+        slog_pi = np.sum(np.log(pi))
+        I_parts["pi"] = -gammaln(K) + 0.5 * (K - 1) * np.log(N) - 0.5 * slog_pi
 
-        # Approximate the normalization constants (p308 of Wallace 2005)
-        #  J = [1     2     3     4     5      6]
-        A_Js = [0, 0.24, 1.35, 3.68, 7.45, 12.64]
-        B_Js = [0, 0.58, 0.64, 0.73, 0.86,  1.21]
+        # Means and covariance matrices of various components.
+        # I(epsilon, omega)
+        _, logdet_omega = np.linalg.slogdet(omega.T)
+        I_parts["xi,omega"] = -(J + 3/2) * np.sum(logdet_omega) \
+                            + 0.25 * J * (J + 3) * (slog_pi + K * np.log(N)) \
+                            - 0.5 * J * K * np.log(2)
 
-        #bj_sq = A**2
-        #v = self.factor_scores(X)[1]
-        #vj_sq = np.diag((v.T @ v)) # very unsure about this.
-        
-
-        if J > len(A_Js):
-
-            #raise NotImplementedError("too many latent factors; "
-            #                          "cannot approximate normalization constants")
-            logger.warn("Too many latent factors; "
-                        "cannot approximate normalization constants when J > 6")
-
-            A_J, B_J = (A_Js[-1], B_Js[-1])
-
-        else:
-            A_J, B_J = (A_Js[int(J) - 1], B_Js[int(J) - 1])
-
[…]

### 2019-01-25 mcfa/utils.py

@@ -7,7 +7,7 @@ def whiten(X, axis=0):
 
 
 def generate_data(n_samples, n_features, n_latent_factors, n_components,
-                  omega_scale, noise_scale, random_seed=0):
+                  omega_scale, noise_scale, random_seed=0, force=None):
 
     rng = np.random.RandomState(random_seed)
 
@@ -22,9 +22,24 @@ def generate_data(n_samples, n_features, n_latent_factors, n_components,
 
     xi = rng.randn(n_latent_factors, n_components)
     omega = np.zeros((n_latent_factors, n_latent_factors, n_components))
+
+    if force is not None:
+        if "xi" in force:
+            xi = force["xi"]
+            print("using forced xi")
+        if "A" in force:
+            A = force["A"]
+            print("using forced A")
+
     for i in range(n_components):
         omega[(*np.diag_indices(n_latent_factors), i)] = \
             rng.gamma(1, scale=omega_scale, size=n_latent_factors)**2
+
+    if force is not None:
+        if "omega" in force:
+            omega = force["omega"]
+            print("using forced omega")
+
 
     scores = np.empty((n_samples, n_latent_factors))
     for i in range(n_components):

### 2019-01-25 prove_d_ll_wrt_A.py

@@ -65,9 +65,9 @@ def g(A_flat, Y, pi, xi, omega, psi):
     A = A_flat.reshape((D, J))
 
     I_psi = np.eye(D) * psi
-
+    
     v = np.ones((N, K, D, J))
-
+    
     ll, tau = model.expectation(Y, pi, A, xi, omega, psi)
     
     for i in range(N):
@@ -83,10 +83,11 @@ def g(A_flat, Y, pi, xi, omega, psi):
 
             T2 = T0 @ (Y[[i]].T - T1)
             v[i, k, :] = T2 @ M.T \
-                        + T2 @ (Y[[i]] + (-T1).T) @ T0 @ A @ O \
-                        - T0 @ A @ O
+                       + T2 @ (Y[[i]] - T1.T) @ T0 @ A @ O \
+                       - T0 @ A @ O
 
     return np.sum(np.sum(v.T * tau.T, axis=-1), axis=-1).T.flatten()
+
 
 
 def h(A_flat, Y, pi, xi, omega, psi):

### 2019-04-01 article/mcfa.tex
add words

@@
 Chemical tagging promises to distinguish unique star formation sites from 
-photospheric abundances.
-Clustering techniques used in chemical tagging have to date considered 
-chemical abundances as independent, a provably false assumption.
+observed photospheric abundances.
+Clustering techniques applied in chemical tagging have to date assumed that
+all photospheric abundances are independent, a provably false assumption.
 Nucleosynthetic processes make multiple elements in varying quantities, and the
-dimensionality of chemical abundance space is usually lower than the 
+dimensionality of chemical abundance space is typically lower than the 
 number of reported abundances. 
@@
 needed to account for \Galah\ data.}
-We release code associated with this \article\ that scales well with the 
-available data: enabling chemical tagging for $10^{7.5}$ abundance
-measurements on a standard desktop machine in minutes.
-
+We describe exact methods and software that scale well with the data,
+allowing for chemical tagging experiments for $10^{7.5}$ abundance measurements
+to be performed in minutes.
 \end{abstract}
@@
 \begin{equation}
-	{I}(\vec\Psi,\vecdata) = \frac{Q}{2}\log\kappa\left(Q\right) - \log\left(\frac{p(\vec\Psi)}{\sqrt{|\mathcal{F}(\vec\Psi)|}}\right) - \log\mathcal{L}\left(\vecdata|\vec\Psi\right)% + \frac{Q}{2}
+	{I}(\vec\Psi,\vecdata) = \frac{Q}{2}\log\kappa_q - \log\left(\frac{p(\vec\Psi)}{\sqrt{|\mathcal{F}(\vec\Psi)|}}\right) - \log\mathcal{L}\left(\vecdata|\vec\Psi\right) + \frac{Q}{2} \label{eq:mml} 
 \end{equation}
@@
 \noindent{}and as before $Q$ is the number of model parameters.
-Finally, the $\log\kappa(Q)$ term is given by
-\begin{equation}
-	\todo{\log\kappa} \quad .
-\end{equation}
+Continuous parameters can only be stated to finite precision, which leads
+to the $\frac{Q}{2}\log\kappa_q$ term that gives a measure of the volume of the region of
+uncertainty in which the parameters $\vec\psi$ are centred. The $\log\kappa_q$
+term is
+\begin{equation}
+	\log\kappa_q = -\log{2\pi} + \frac{1}{Q}\log{Q\pi} - \gamma - 1
+\end{equation}
+\noindent{}where $\gamma$ is Euler's constant.
 
@@
 describes the curvature of the log likelihood with respect to the model
-parameters. \todo{Dowe said something about MML reducing to BIC under some conditions, but I have not yet found the right paper to cite.}
+parameters.
+
+%\todo{Dowe said something about MML reducing to BIC under some conditions, but I have not yet found the right paper to cite.}
 
@@
 $p(\NumComponents) \propto 2^{-\NumComponents}$ respectively,
-such that the optimal lossless message to encode them is \citep[p279, sec. 6.8.2][]{someone},
+such that the optimal lossless message to encode them is \citep[p279, sec. 6.8.2][]{Knorr-Held:2000},
 \begin{eqnarray}
-	I(\NumLatentFactors) &= -\log{p(\NumLatentFactors)} &= \NumLatentFactors\log{2} + \textrm{constant} \\
-	I(\NumComponents) &= -\log{p(\NumComponents)} &= \NumComponents\log{2} + \textrm{constant} \quad .
+	I(\NumLatentFactors) &= -\log{p(\NumLatentFactors)} &= \NumLatentFactors\log{2} + \textrm{constant} \label{eq:prior_J} \\
+	I(\NumComponents) &= -\log{p(\NumComponents)} &= \NumComponents\log{2} + \textrm{constant} \label{eq:prior_K} \quad .
 \end{eqnarray}
@@
 	I(\vec\weight) &=& \frac{1}{2}\left(\left(\NumComponents - 1\right)\log\NumData - \sum_{\numcomponents=1}^{\NumComponents}\log\weight_\numcomponents\right) -\log\Gamma(\NumComponents) \quad .
-\nonumber \\
+\nonumber \label{eq:prior_pi} \\
 \end{eqnarray}
@@
 and a conjugate inverted Wishart prior for the component covariance matrices
-$\scorecovs$ \citep[Section 5.2.3][]{someone},
+$\scorecovs$ \citep[Section 5.2.3][]{Knorr-Held:2000},
 \begin{equation}
@@
 We approximate the determinate of the Fisher information of a multivariate normal $|\mathcal{F}(\scoremeans,\scorecovs)|$
-as $|\mathcal{F}(\scoremeans)||\mathcal{F}(\scorecovs)|$ \citep{25,27} where
+as $|\mathcal{F}(\scoremeans)||\mathcal{F}(\scorecovs)|$ \citep{Oliver:1996,Figueiredo:2002} where
 \begin{equation}
@@
 I(\scoremeans,\scorecovs) &=& \frac{1}{4}\NumLatentFactors(\NumLatentFactors+3)\sum_{\numcomponents=1}^\NumComponents\log{\NumData\weight_k} - \frac{KD}{2}\log{2} \nonumber \\ 
-&& \cdots \,\, -\frac{1}{2}(2\NumLatentFactors+3)\sum_{k=1}^{K}\log{|\scorecovs_k|}  \quad .
+&& \cdots \,\, -\frac{1}{2}(2\NumLatentFactors+3)\sum_{k=1}^{K}\log{|\scorecovs_k|}  \quad . \label{eq:prior_xi_omega} 
 \end{eqnarray}
@@
 placing a joint prior on the product of factor loads and scores, whilst
-ensuring that the factor loads themselves remain mutually orthogonal \citep{Wallace}.
+ensuring that the factor loads themselves remain mutually orthogonal \citep{WallaceMLF}.
 Adopting the same prior density in our model is not practical because 
@@
 \begin{equation}
-	p(A) = \frac{|\factorloads\transpose\factorloads|^{\frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)}}{2^{\frac{\NumDimensions\NumLatentFactors}{2}}|\vec{S}|^{\frac{\NumDimensions}{2}}\Gamma(\frac{\NumDimensions}{2})}\exp\left[-\frac{1}{2}\textrm{Tr}(\vec{S}^{-1}\factorloads\transpose\factorloads)\right] \quad .
+	p(\factorloads) = \frac{|\factorloads\transpose\factorloads|^{\frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)}}{2^{\frac{\NumDimensions\NumLatentFactors}{2}}|\vec{S}|^{\frac{\NumDimensions}{2}}\Gamma(\frac{\NumDimensions}{2})}\exp\left[-\frac{1}{2}\textrm{Tr}(\vec{S}^{-1}\factorloads\transpose\factorloads)\right] \quad .
 \end{equation}
@@
 	I(\factorloads) &=& -\log\left(\frac{p(\factorloads)}{\sqrt{|\mathcal{F}(\factorloads)|}}\right) \nonumber \\
-	&=& -\log{p(\factorloads)} + \frac{1}{2}\log{|\mathcal{F}(\factorloads)|} \nonumber \\
+	&=& -\log{p(\factorloads)} + \todo{\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \nonumber \\
 I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} \nonumber \\
-	&& \cdots \,\, + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \quad .\nonumber \\
+	&& \cdots \,\, + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \quad .\nonumber \label{eq:prior_L} \\
 \end{eqnarray}
 
-Combining equations \ref{eq:all} leads to the full message length
-\begin{equation}
-\todo{here}
-\end{equation}
+Combining equations \ref{eq:prior_J}, \ref{eq:prior_K}, \ref{eq:prior_pi}, \ref{eq:prior_xi_omega}, and \ref{eq:prior_L} with equation \ref{eq:mml} leads to the full message length
+\begin{eqnarray}
+	I(\vec\Psi, \vec\data) &=& -\log\likelihood(\vec\data|\vec\Psi) \nonumber \\
+	&& +\frac{1}{4}\left(\NumLatentFactors + 4\right)\left(\NumLatentFactors - 1\right)\sum_{\numcomponents=1}^\NumComponents\log\weight_\numcomponents + \left(\NumComponents - \frac{1}{2}\right)\log{\NumData} \nonumber \\
+	&& +\frac{1}{2}\NumDimensions\log|\textrm{Cov}\left(\factorloads\transpose\right)| - \frac{1}{2}\left(D-J-1\right)\log|\factorloads\transpose\factorloads| \nonumber \\
+	&& +\textrm{Tr}\left(\textrm{Cov}\left(\factorloads\transpose\right)^{-1}\factorloads\transpose\factorloads\right) - \left(\NumLatentFactors + \frac{3}{2}\right)\sum_{\numcomponents=1}^\NumComponents\log|\scorecovs_\numcomponents| \nonumber \\
+	&& -\log\Gamma\left(\NumComponents\right) - \Gamma\left(\frac{\NumDimensions}{2}\right) + \frac{Q}{2}\log\kappa_q \nonumber \\
+	&& +\frac{1}{2}\left[\NumLatentFactors(\NumDimensions + 2) + \NumComponents(2-\NumData)\right]\log{2} \quad .
+\end{eqnarray}
+
+% -\log\likelihood(\vec\data|\vec\Psi) \
+% + (J + K + \frac{KN}{2} + \frac{DJ}{2})\log{2}     --> \frac{1}{2}(3(J+K) + ND)\log{2}
+% 0.5*((K-1) logN - \sum\log\weight)  + \frac{1}{4}J(J+3)\sum\log(N\weight) --> 
+% -\log\gamma(K) - \Gamma(D/2)
+% -\frac{1}{2}(2J+3)\sum\log|\omega_k|
+% + factor loads stuff
+% + Q/2\log\kappa(q)
+
+\clearpage
 
@@
 	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-bic-contours.png}
+	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-mml-contours.png}
     \caption{Model performance metrics resulting from a gridsearch using the
@@
 			 evaluated at each combination of latent factors $J$ and number 
-			 of clusters $K$ and the bottom panel shows the BIC (Equation \ref{eq:bic}).
+			 of clusters $K$, and the bottom panel shows the BIC (Equation \ref{eq:bic}).
 			 The white marker indicates the 
-			 lowest value in each panel, which is joined by a black line from the true value.}
+			 lowest value in each panel, which is joined by a black line from the true value
+			 in the top panel. The model with the lowest message length and lowest BIC has
+			 the same number of latent factors and components used to generate the data.
+		}
     \label{fig:experiment-1-gridsearch}
@@
 and $\NumComponents = 20$, identical to the true values.
+
+\begin{figure*}
+	\includegraphics[width=\textwidth]{experiments/exp1-latent.pdf}
+    \caption{The inferred latent scores $\factorscores$ for Experiment~1
+    		 from a model with $\NumLatentFactors = 5$ latent factors and
+		 	 $\NumComponents = 20$ multivariate normal components in
+			 latent space.
+             Points are coloured by the inferred component.}
+    \label{fig:exp1-latent}
+\end{figure*}
 
@@
 
+\begin{figure*}
+	\includegraphics[width=\textwidth]{experiments/exp1-compare-all.pdf}
+    \caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
+    		 and specific variances $\specificvariance$ (right) compared to the 
+		 	 true data generating values
+		 	 for Experiment~1. The agreement is excellent.}
+    \label{fig:exp1-compare}
+\end{figure*}
+
 Some technical discussion is warranted before we compare our estimated model
@@
 there is no assurance (or requirement) that our factor loads match the true factor loads.
-
-\begin{figure*}
-	\includegraphics[width=\textwidth]{experiments/exp1-latent.pdf}
-    \caption{The inferred latent scores $\factorscores$ for Experiment~1
-    		 from a model with $\NumLatentFactors = 5$ latent factors and
-		 	 $\NumComponents = 20$ multivariate normal components in
-			 latent space.
-             Points are coloured by the inferred component.}
-    \label{fig:exp1-latent}
-\end{figure*}
 
@@
 values, but they only differ by an orthogonal rotation.
-
-\begin{figure*}
-	\includegraphics[width=\textwidth]{experiments/exp1-compare-all-mml.pdf}
-    \caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
-    		 and specific variances $\specificvariance$ (right) compared to the 
-		 	 true data generating values
-		 	 for Experiment~1. The agreement is excellent.}
-    \label{fig:exp1-compare}
-\end{figure*}
 
@@
 
-After accounting for the contributions by all
-$J = 5$ factors, the specific scatter remaining in each dimension is shown in
-Figure~\ref{fig:exp2-specific-scatter}. The remaining scatter varies from as low 
-as 0.08~dex (Mn, Y) up to 0.40~dex (Al, Ba, Ni). Al is not found to have any 
-predominant factor in these data: it has non-zero contributions from all loads
-(Figure~\ref{fig:exp2-factor-loads}. A similar pattern is seen in Cr, Co, and Ni,
-although the specific scatter for Cr and Co is half that of what is found for
-Ni (0.2\,dex instead of 0.4\,dex).
-
-% TODO: anything else to say here?
-
 \begin{figure*}
@@
 \end{figure*}
+
+After accounting for the contributions by all
+$J = 5$ factors, the specific scatter remaining in each dimension is shown in
+Figure~\ref{fig:exp2-specific-scatter}. The remaining scatter varies from as low 
+as 0.08~dex (Mn, Y) up to 0.40~dex (Al, Ba, Ni). Al is not found to have any 
+predominant factor in these data: it has non-zero contributions from all loads
+(Figure~\ref{fig:exp2-factor-loads}. A similar pattern is seen in Cr, Co, and Ni,
+although the specific scatter for Cr and Co is half that of what is found for
+Ni (0.2\,dex instead of 0.4\,dex).
+
+% TODO: anything else to say here?
 
@@
 photospheric varies throughout a star's lifetime. This is true
-to a small degree for many elements \citep[e.g.,][]{Dotter_2008},
+to a small degree for many elements \citep[e.g.,][]{Dotter:2017},
 but for the purposes of this experiment we assume that all other
@@
 
+%We have presented a novel approach to chemical tagging that incorporates
+%latent factors common to all stars, and allows for clustering in the relative
+%contributions of those factors. This is analogous to simultaneously
+%inferring nucleosynthetic yields as part of chemical tagging.
+
+Earlier studies investigating the limitations of chemical tagging have shown
+that the effective dimensionality of chemical abundance space is lower than
+the number of reported abundances \citep[e.g. about 8,][]{Ting:2008}. 
+This is expected. Theoretical models of stellar evolution and nucleosynthesis 
+suggest that there are of order \todo{5} nucleosynthetic mechanisms that are
+predominately responsible for the production of the $\sim$20 or so observed 
+reported in those studies. As a consequence we can expect that the observed
+chemical abundances are correlated in some way (e.g., by a linear combination
+of a fewer number of nucleosynthetic pathways).  However, most approaches to
+chemical tagging to date have assumed that the observed chemical abundances
+are independent of each other \todo{\citep[e.g.,][]{who}}. Those inferences
+will be biased by that assumption in a non-trivial way. This problem cannot be
+resolved by simply lowering the number of abundances used for chemical tagging (e.g.,
+down to something near the effective dimensionality), either, because those
+chemical abundances are still produced by multiple pathways.
+
+Here we have introduced a model to simultaneously account for the lower
+effective dimensionality of chemical abundance space, and perform clustering
+in that lower dimensional space. Although our model allows us to estimate
+those latent factors that contribute to all stars, and cluster those stars by
+their relative contributions from those factors, there are issues with the
+model we adopt. The model we describe is very likely \emph{not} the correct
+model to use to describe chemical abundances of stars. We require the latent
+factors in our model to be mutually orthogonal to each other. This suggests
+an astrophysical context where the mean nucleosynthetic yields (over all star
+formation histories and stellar masses) of various
+processes (e.g., $r$-process, $s$-process) are mutually orthogonal to each
+other. Clearly this assumption is likely to be incorrect: the nuclear physics
+of one environment where elements are produced  can be very different from
+other environments, and there is no astrophysical constraint that those
+latent factors should be mutually orthogonal.
+A more flexible data-driven model of nucleosynthetic yields (as a function of mass,
+stellar metallicity, and other factors) that does not require the latent
+factors to be mutually orthogonal would be a worthy extension to this work.
+
+There are other issues in our model that relate to our assumption of mutual
+orthogonality. Even if nucleosynthetic yields were truely mutually orthogonal,
+then the latent factors we infer are only \emph{identifiable} up until an
+orthogonal basis. As we have seen in our experiments, the ordering and sign 
+of the latent factors is not described \emph{a priori}. That means that we
+must `assign' the latent factors we infer as being described by an astrophysical
+process (e.g., the zeroth-latent factor is r-process). A more general limitation
+of this is that the latent factors can be multiplied by some arbitrary rotation
+matrix, leading to latent factor loads that are very different from what was
+estimated by the model, but still lead to the exact same data (or log-likelihood,
+or Kullback-Liebler divergence, etc). As a consequence, we can only `identify'
+latent factors up until this rotation. We have sought to address this by constructing
+rotation matrices where the entries each latent factor correspond to our expectations
+from astrophysical processes (whilst remaining orthogonal), but here we are limited
+by what astrophysical processes we are \emph{expected} to find.
+
+This constrains our ability to identify new nucleosynthetic processes. For example,
+let us consider a hypothetical situation where we would only expect there to be four 
+nucleosynthetic processes that predominately contribute to the observed \GALAH\ abundances,
+but in practice we found that the data are best explained (with high significance) with
+five latent factors. We construct a rotation matrix where the first four latent factors
+describe the mean nucleosynthetic processes we expect to find. What of the fifth latent
+factor? We can constrain the possible values of the fifth latent factor conditioned on
+the requirement that all factors remain mutually orthogonal, but one can imagine that
+some (or perhaps many) elements have entries where the fifth latent factor can have
+near-zero or zero entries. Even if the mean nucleosynthetic yields are mutually
+orthogonal, there are scenarios that one can imagine where there is a limited amount
+we can say with confidence about that new nucleosynthetic process.
+
+Notwithstanding these issues, we have presented numerous experiments that demonstrate
+that a latent factor model that allows for clustering in latent space can adequately
+describe chemical abundance data. Our experiments with toy data (Experiment~1) reliably
+finds the true number of latent factors and cluster components when we adopt an information-theoretic approach to the objective function (e.g., BIC or MML), and correctly estimates the latent
+model parameters. When applied to chemical abundances measured from metal-poor stars 
+(Experiment~2) we find that \todo{seven latent factors and two clusters are preferred by
+the data.}
+
+\todo{discussion of experimental results}
+
 \todo{
 \begin{itemize}
-	\item dimensionality of chemical abundance space
-	\item clustering approaches do not take this into account
-	\item introduced a model to simultaneously account for the clustering in
-		  a lower dimensional latent space
-	\item there are issues, though: this is probably  not the right model
-	\item identifiability
-	\item orthogonal rotation
-	\item can interpret these somehow as being astrophysically motivated
 	\item at low metallicities how many factors do we prefer, and what can we
@@
 	\item clustering in galah
+	\item clustering in apogee
 \end{itemize}
 }
+
 \section{Conclusions} \label{sec:conclusion}
@@
 \software{
-	\package{Astropy} \citep{astropy:v1,astropy:v2},
-    \package{IPython} \citep{ipython},
-    \package{matplotlib} \citep{mpl},
-    \package{numpy} \citep{numpy},
-    \package{scipy} \citep{scipy},
-    \package{Stan} \citep{stan},
-    \package{Jupyter Notebooks} \citep{jupyter-notebooks}
+	\package{Astropy}\,\,\citep{astropy:v1,astropy:v2},\,\,
+    \package{IPython}\,\,\citep{ipython},\,\,
+    \package{matplotlib}\,\,\citep{mpl},\,\,
+    \package{numpy}\,\,\citep{numpy},\,\,
+    \package{scipy}\,\,\citep{scipy},\,\,
+    \package{Stan}\,\,\citep{stan},\,\,
+    \package{Jupyter Notebooks}\,\,\citep{jupyter-notebooks}\,\,
 }

### 2019-04-01 mcfa/grid_search.py
add code

@@ -8,7 +8,7 @@ from .mcfa import MCFA
 logger = logging.getLogger(__name__)
 
 def grid_search(trial_n_latent_factors, trial_n_components, X, N_inits=1,
-                mcfa_kwds=None, pseudo_bic_kwds=None):
+                suppress_exceptions=True, mcfa_kwds=None, pseudo_bic_kwds=None):
 
     Js = np.array(trial_n_latent_factors)
     Ks = np.array(trial_n_components)
@@ -34,7 +34,6 @@ def grid_search(trial_n_latent_factors, trial_n_components, X, N_inits=1,
             models = []
             for n in range(N_inits):
 
-
                 print(f"At J = {J}, K = {K}, N = {n}")
 
                 model = MCFA(n_latent_factors=J, n_components=K, **mcfa_kwds)
@@ -45,6 +44,8 @@ def grid_search(trial_n_latent_factors, trial_n_components, X, N_inits=1,
                 except:
                     logger.exception(f"Exception occurred during grid search at "\
                                      f"J = {J}, K = {K}:")
+                    if not suppress_exceptions:
+                        raise
                     continue
 
                 else:
@@ -55,9 +56,9 @@ def grid_search(trial_n_latent_factors, trial_n_components, X, N_inits=1,
                 model = models[idx]
 
                 ll[k, j] = model.log_likelihood_
-                bic[k, j] = model.bic(X)
-                mml[k, j] = model.message_length(X)
-                pseudo_bic[k, j] = model.pseudo_bic(X, **pseudo_bic_kwds)
+                bic[k, j] = model.bic(X, log_likelihood=model.log_likelihood_)
+                mml[k, j] = model.message_length(X, log_likelihood=model.log_likelihood_)
+                #pseudo_bic[k, j] = model.pseudo_bic(X, **pseudo_bic_kwds)
                 converged[k, j] = True
 
 

### 2019-04-01 mcfa/mcfa.py

@@ -301,7 +301,7 @@ class MCFA(object):
         return int((K - 1) + D + J*(D + K) + (K*J*(J+1))/2 - J**2)
 
 
-    def expectation(self, X, pi, A, xi, omega, psi):
+    def expectation(self, X, pi, A, xi, omega, psi, **kwargs):
         r"""
         Compute the conditional expectation of the complete-data log-likelihood
         given the observed data :math:`X` and the given model parameters.
@@ -341,7 +341,8 @@ class MCFA(object):
             component in the mixture.
         """
 
-        ll, tau = _expectation(X, pi, A, xi, omega, psi, verbose=self.verbose)
+        ll, tau = _expectation(X, pi, A, xi, omega, psi, 
+                               verbose=self.verbose, **kwargs)
         self.log_likelihoods_.append(ll)
 
         return (ll, tau)
@@ -446,8 +447,8 @@ class MCFA(object):
 
         X = self._check_data(X)
 
-        theta = self._initial_parameters(X, self.random_seed) \
-                if init_params is None else deepcopy(init_params) 
+        theta = prev_theta = self._initial_parameters(X, self.random_seed) \
+                             if init_params is None else deepcopy(init_params) 
 
         prev_ll, tau = self.expectation(X, *theta)
 
@@ -458,10 +459,18 @@ class MCFA(object):
         tqdm_kwds = dict(desc="E-M", total=self.max_iter)
         for n_iter in tqdm(range(1, 1 + self.max_iter), **tqdm_kwds):
 
-            theta = self.maximization(X, tau, *theta)            
-            ll, tau = self.expectation(X, *theta)
+            try:
+                theta = self.maximization(X, tau, *theta)
+                ll, tau = self.expectation(X, *theta)
+
+            except:
+                logger.exception(f"Exception occurred during E-M algorithm:")
+                logger.warning(f"Solution is unlikely to be converged.")
+                theta = prev_theta
+                break
 
             converged, prev_ll, ratio = self._check_convergence(prev_ll, ll)
+            prev_theta = theta
 
             if converged: break
 
@@ -503,7 +512,7 @@ class MCFA(object):
         return _factor_scores(X, self.tau_, *self.theta_)
 
 
-    def bic(self, X, theta=None):
+    def bic(self, X, theta=None, log_likelihood=None):
         r"""
[…]

### 2019-04-01 mcfa/grid_search.py
add logging to grid search

@@ -55,11 +55,12 @@ def grid_search(trial_n_latent_factors, trial_n_components, X, N_inits=1,
                 idx = np.nanargmax([model.log_likelihood_ for model in models])
                 model = models[idx]
 
-                ll[k, j] = model.log_likelihood_
-                bic[k, j] = model.bic(X, log_likelihood=model.log_likelihood_)
-                mml[k, j] = model.message_length(X, log_likelihood=model.log_likelihood_)
-                #pseudo_bic[k, j] = model.pseudo_bic(X, **pseudo_bic_kwds)
-                converged[k, j] = True
+                j_bic_best, k_bic_best = best(Js, Ks, bic)
+                logger.info(f"Model with minimum BIC so far has J = {j_bic_best} and K = {k_bic_best}")
+                j_mml_best, k_mml_best = best(Js, Ks, mml)
+                logger.info(f"Model with minimum I so far has J = {j_mml_best} and K = {k_mml_best}")
+                j_ll_best, k_ll_best = best(Js, Ks, -ll)
+                logger.info(f"Model with maximum log-likelihood so far has J = {j_ll_best} and K = {k_ll_best}")
 
 
     # Best of each?

### 2019-04-01 article/experiments/exp2.py
add experiment 2

@@ -17,6 +17,7 @@ matplotlib.style.use(mpl_utils.mpl_style)
 here = os.path.dirname(os.path.realpath(__file__))  
 
 random_seed = 10
+adopted_metric = "mml"
 
 np.random.seed(random_seed)
 
@@ -100,7 +101,7 @@ label_names = list(np.array(label_names)[element_mask])
 """
 
 
-X = utils.whiten(X)
+#X = utils.whiten(X)
 #X = X_H
 
 
@@ -192,7 +193,7 @@ max_n_components = 3
 Js = 1 + np.arange(max_n_latent_factors)
 Ks = 1 + np.arange(max_n_components)
 
-Jg, Kg, converged, metrics = grid_search.grid_search(Js, Ks, X, 
+Jg, Kg, converged, metrics = grid_search.grid_search(Js, Ks, X, N_inits=5,
                                                      mcfa_kwds=mcfa_kwds)
 
 ll = metrics["ll"]
@@ -218,16 +219,10 @@ fig_ll = mpl_utils.plot_filled_contours(Jg, Kg, -ll,
 savefig(fig_ll, "gridsearch-ll")
 
 
-"""
-plot_filled_contours_kwds = dict(converged=converged,
-                                 marker_function=np.nanargmin,
-                                 cmap="Spectral_r")
 fig_ll = mpl_utils.plot_filled_contours(Jg, Kg, message_length,
                                         colorbar_label=r"$\textrm{MML}$",
                                         **plot_filled_contours_kwds)
 savefig(fig_ll, "gridsearch-mml")
-"""
-
 
 
 fig_bic = mpl_utils.plot_filled_contours(Jg, Kg, bic,
@@ -237,10 +232,21 @@ savefig(fig_bic, "gridsearch-bic")
 
 
 # Re-run model with best J, K.
+if adopted_metric.lower() == "bic":
+    J_best, K_best = (J_best_bic, K_best_bic)
 
-model = mcfa.MCFA(n_components=K_best_bic, n_latent_factors=J_best_bic,
-                  **mcfa_kwds)
+elif adopted_metric.lower() == "mml":
+    J_best, K_best = (J_best_mml, K_best_mml)
 
+elif adopted_metric.lower() == "ll":
[…]

### 2019-04-08 article/mcfa.tex
pd

@@
 \newcommand{\GALAH}{\project{GALAH}}
+\newcommand{\APOGEE}{\project{APOGEE}}
 \newcommand{\todo}[1]{\textcolor{red}{#1}}
@@
 observed photospheric abundances.
-Clustering techniques applied in chemical tagging have to date assumed that
-all photospheric abundances are independent, a provably false assumption.
+Chemical tagging experiments to date have assumed that photospheric
+abundances are independent of each other, even though it is known that 
+nucleosynthetic processes 
+produce
+making
+
+Chemical tagging experiments to date have assumed that all photospheric
+abundances are independent, despite the effective dimensionality of abundance
+space being lower than the number of available elements.
+
 Nucleosynthetic processes make multiple elements in varying quantities, and the
@@
 The classically-described princple of MML is that the best explanation of the
-data is the one that leads to the shortest so-called two-part message~\cite{Wallace:2005}, 
+data is the one that leads to the shortest so-called two-part message~\citep{Wallace:2005}, 
 where a \textit{message} takes into account both the complexity of the model 
@@
 its explanatory power. The \emph{length} of each message part is quantified
-(or measured) using information theory, allowing for a fair evaluation between
+(or estimated) using information theory, allowing for a fair evaluation between
 different models of varying complexity or explanatory power. MML has been 
@@
 The \textit{message} must encode two parts: the model, and the data given the
-model. The encoding of the message is based on Shannon's information theory~\cite{Shannon:1948}. 
+model. The encoding of the message is based on Shannon's information theory~\citep{Shannon:1948}. 
 The information gained from an event $e$ occurring, where $p(e)$ is the
@@
 $p(\NumComponents) \propto 2^{-\NumComponents}$ respectively,
-such that the optimal lossless message to encode them is \citep[p279, sec. 6.8.2][]{Knorr-Held:2000},
+such that the optimal lossless message to encode them is \citep[p279, sec. 6.8.2;][]{Knorr-Held:2000},
 \begin{eqnarray}
@@
 and a conjugate inverted Wishart prior for the component covariance matrices
-$\scorecovs$ \citep[Section 5.2.3][]{Knorr-Held:2000},
+$\scorecovs$ \citep[Section 5.2.3;][]{Knorr-Held:2000},
 \begin{equation}
@@
 I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} \nonumber \\
-	&& \cdots \,\, + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \quad .\nonumber \label{eq:prior_L} \\
+	&& \cdots \,\, + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\
+	&& \cdots \todo{+\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \quad . \label{eq:prior_L}
 \end{eqnarray}
@@
 	&& -\log\Gamma\left(\NumComponents\right) - \Gamma\left(\frac{\NumDimensions}{2}\right) + \frac{Q}{2}\log\kappa_q \nonumber \\
-	&& +\frac{1}{2}\left[\NumLatentFactors(\NumDimensions + 2) + \NumComponents(2-\NumData)\right]\log{2} \quad .
+	&& +\frac{1}{2}\left[\NumLatentFactors(\NumDimensions + 2) + \NumComponents(2-\NumData)\right]\log{2} \nonumber \\
+	&& \todo{+\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} .
 \end{eqnarray}
@@
 
-\clearpage
+%\clearpage
 
@@
 
-\subsection{Experiment 1: Toy model with generated data} \label{sec:experiment-toy-model}
+\subsection{Experiment 1: Toy model with generated data} \label{sec:exp-1}
 
@@
 \begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-ll-contours.png}
-	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-bic-contours.png}
-	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-mml-contours.png}
+	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-ll-contours.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-bic-contours.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-mml-contours.pdf}
     \caption{Model performance metrics resulting from a gridsearch using the
@@
 factors $\NumLatentFactors$ and increasing numbers of components $\NumComponents$.
-The lowest BIC value is found at $\NumLatentFactors = 5$
+The lowest BIC value and message length is found at $\NumLatentFactors = 5$
 and $\NumComponents = 20$, identical to the true values.
@@
 abundances measured include Al (light odd-Z), Mg and Ca ($\alpha$-process), Sc, Ti, Cr, Mn, Fe, Co, and Ni (Fe-peak elements), as well as Sr, Y, Ba, and Eu ($s$/$r$-process). 
+The data vector $\data$ includes [Fe/H] abundance ratios and all other elements
+relative to Fe (e.g., [Ti/Fe]) after subtracting the mean value to allow the
+latent factors to describe the abundance patterns away from the mean.
 
@@
 $\NumLatentFactors = 1$ and $K = 1$, and trialling up until 
-$\NumLatentFactors = 7$ and $\NumComponents = 5$. 
-We found that models with higher numbers of latent factors were unable to
-converge without adding a regularization constant to the diagonals of the
-covariance matrices $\scorecovs$ in latent space. Although it is common
-to add a regularization term to the diagonals of covariance matrices
-when performing expectation-maximization, we found that our existing
-grid search boundaries were sufficient to find the preferred $\NumLatentFactors$
-and $\NumComponents$, and chose not to add any regularization term.
-
+$\NumLatentFactors = 7$ and $\NumComponents = 3$. 
+% Although it is common
+%to add a regularization term to the diagonals of covariance matrices
+%when performing expectation-maximization, we found that our existing
+%grid search boundaries were sufficient to find the preferred $\NumLatentFactors$
+%and $\NumComponents$, and chose not to add any regularization term.
 We initialised
@@
 until a convergence threshold of $10^{-5}$ was reached in the log likelihood. 
-In Figure~\ref{fig:exp2-gridsearch-contours} we show the negative log likelihood 
-and BIC recorded for the optimised set of model parameters for each permutation 
-of $J$ and $K$. The lowest BIC value is found for $J = 5$ and $K = 1$, indicating
-five latent factors are preferred, with no clustering.
-
-Using our preferred model with $J = 5$ and $K = 1$, we sought to find a
-rotation matrix that would allow us to identify plausible astrophysical
-processes in the factor loads.
-Although having to perform this rotation is not comforting, there is utility in
+In Figure~\ref{fig:exp2-gridsearch-contours} we show the negative log likelihood,
+the BIC, and the message length recorded for the optimised set of model parameters 
+for each permutation of $J$ and $K$. The negative log likelihood continues to
+increase with increasing numbers of clusters, or increasing numbers of latent
+factors.
+The lowest BIC value is found for $J = 5$ and $K = 1$, indicating
+six latent factors are preferred, with no clustering. The shortest message length
+is also found with no clustering,  but with four latent factors ($J = 4$, $K = 1$).
+We adopt the message length as our objective function and chose $J = 4$ and $K = 1$
+as the preferred model.
+
+\begin{comment}
+With this model we sought to find a rotation matrix that would allow us to 
+identify plausible astrophysical processes in the factor loads.
+Although having to perform this rotation is comfortable, there is utility in
 trying to identify the latent factors because the factors can only be rotated by
@@
 all entries were zero except for the following entries:
-\begin{itemize}
-	\item the entry corresponding to Al in the first latent factor to represent
-		  odd-Z element production,
-	\item the entries corresponding to Ca and Mg in the second latent factor to
-		  represent $\alpha$ particle production,
-	\item the entries corresponding to Ni, Co, Fe, Mn, Cr, Ti, and Sc in the third
-		  latent factor to represent the Fe-peak group,
-	\item the entries corresponding to Sr, Y, and Ba in the fourth latent factor
-		  to represent products of the slow neutron capture process, and
-	\item the entries corresponding to Eu in the fifth latent factor to represent
-		  products of the rapid nucleosynthesis process.
+%\begin{itemize}
+%	\item the entry corresponding to Al in the first latent factor to represent
+%		  odd-Z element production,
+%	\item the entries corresponding to Ca and Mg in the second latent factor to
+%		  represent $\alpha$ particle production,
+%	\item the entries corresponding to Ni, Co, Fe, Mn, Cr, Ti, and Sc in the third
+%		  latent factor to represent the Fe-peak group,
+%	\item the entries corresponding to Sr, Y, and Ba in the fourth latent factor
+%		  to represent products of the slow neutron capture process, and
+%	\item the entries corresponding to Eu in the fifth latent factor to represent
+%		  products of the rapid nucleosynthesis process.
+%\end{itemize}
+\todo{\begin{itemize}
+	\item the entries corresponding to C, Al, Ca, Mg, Ni, Co, Fe, Mn, Cr, Ti, and Sc in the first
+		  latent factor to represent the products of core-collapse supernovae,
+	\item the entries corresponding to Sr, Y, Ba, and Eu in the second latent factor
+		  to represent products of the neutron capture process.
 \end{itemize}
+}
 These entries were set to $1/\sqrt{E}$ where $E$ is the number of non-zero entries
@@
 to each element.
+\end{comment}
 
 \begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-ll.pdf}
-	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-bic.pdf}
-    \caption{Model performance metrics resulting from a gridsearch using the
+	\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-gridsearch-ll.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-gridsearch-bic.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-gridsearch-mml.pdf}
+    \caption{Model performance metrics resulting from a grid search using the
     		 \citet{Barklem:2005} data in Experiment~2.
@@
 			 evaluated at each combination of latent factors $\NumLatentFactors$ and number 
-			 of clusters $\NumComponents$, and the lower panel shows the BIC (Eq.~\ref{eq:bic}).
+			 of clusters $\NumComponents$. The middle panel shows the BIC (Eq.~\ref{eq:bic})
+			 for those combinations, and the lower panel shows the message length. 
 			 %The cross at
@@
 			 %adding a regularisation constant to the diagonals of $\scorecovs$.
-			 The marker indicates the lowest value in each panel, showing that
-			 $\NumLatentFactors = 5$ and $\NumComponents = 1$ is preferred by BIC.}
+			 A marker is shown at the lowest value in each panel, indicating
+			 the preferred number of latent factors and components 
+			 ($\NumLatentFactors = 2$ and $\NumComponents = 1$ is preferred by MML).}
     \label{fig:exp2-gridsearch-contours}
@@
 
+\begin{comment}
 The rotated loads are not precisely the same as the target astrophysical loads
@@
 factors that can be interpreted within an astrophysical context. 
-\todo{For example,
-the first factor does show positive contributions to Al, like the target load,
-but it also shows a slight negative contribution with respect to alpha elements
-(Mg, Ca)}, and a gradual increase among Fe-peak elements up to Ni. Similarly while
-Ca and Mg are indeed positive in the second factor, as targeted, there is a
-weak negative contribution to Mn abundances, which was not part of the target
-loads. The fourth factor shows a positive contributions to Sr, Y, Ba, and to a
-lesser extent, Eu, which is representative of slow neutron capture production.
-Similarly, the last component shows a very strong positive contribution towards
-Eu, and could be interpreted as production from the rapid nucleosynthesis process.
+We discuss this in more detail in Section~\ref{sec:discussion}.
+\end{comment}
+
+%\todo{For example,
+%the first factor does show positive contributions to Al, like the target load,
+%but it also shows a slight negative contribution with respect to alpha elements
+%(Mg, Ca)}, and a gradual increase among Fe-peak elements up to Ni. Similarly while
+%Ca and Mg are indeed positive in the second factor, as targeted, there is a
+%weak negative contribution to Mn abundances, which was not part of the target
+%loads. 
+Despite the caveats in interpreting latent factors as nucleosynthetic processes
+(e.g. see Sections \ref{sec:exp-1} and \ref{sec:discussion}), here we will
+qualitatively describe the latent factors found in our selected model.
+The first latent factor $|\mathbf{L_0}|$ shows a non-zero contribution to most
+elements (Figure\ref{fig:exp2-factor-loads}). The elements Ca, Sc, Ti, Cr, and Co, mostly
+Fe-peak elements, have the largest contribution from this factor. We interpret this latent
+factor as contributions from core-collapse supernovae.
+The second latent factor $|\mathbf{L_1}|$ has negligible contribution at most elements,
+with strong contributions for C, and the neutron-capture elements Sr, Y, Ba, and Eu. We
+interpret this latent factor as being most closely representative of the slow neutron
+capture process. The third latent factor has almost no contribution to Ba, but has
+non-zero contributions to other neutron-capture elements Sr, Y, and Eu. Since Eu has
+the strongest contribution by $|\mathbf{L_2}|$, we interpret this factor as being most
+representative of the rapid neutron capture process. The last latent factor $|\mathbf{L_3}|$
+has a noticeable peak at Fe and the surrounding Fe-peak elements, with small non-zero contributions
+at most elements.
+
+\begin{comment}
+The first factor shows a positive contributions to most elements except for
+neutron-capture products, where the contribution decreases with increasing
+atomic number. The second latent factor shows strong contributions to neutron
+capture process elements -- with this latent factor being responsible for almost
+all Eu -- and shows a predominant contribution to Sc. Scandium is one of the most
+discrepant elements between observations and models of galactic chemical evolution,
+as models under-predict [Sc/Fe] ratios by a factor of ten \citep[e.g.,][]{Kobayashi:2006,Casey:2015}.
+\end{comment}
+
+% Sr, Y, Ba, and to a
+%lesser extent, Eu, which is representative of slow neutron capture production.
+%Similarly, the last component shows a very strong positive contribution towards
+%Eu, and could be interpreted as production from the rapid nucleosynthesis process.
 
 \begin{figure*}
-	\includegraphics[width=\textwidth]{experiments/exp2-latent-factors-visualize.pdf}
+	\includegraphics[width=\textwidth]{experiments/exp2-log-xfe-offset-latent-factors-visualize-abs.pdf}
 	\caption{Rotated latent factors found in Experiment~2 using the \citet{Barklem:2005}
@@
 
-After accounting for the contributions by all
-$J = 5$ factors, the specific scatter remaining in each dimension is shown in
+After accounting for the contributions by the latent factors and scores,
+the specific scatter remaining in each dimension is shown in
 Figure~\ref{fig:exp2-specific-scatter}. The remaining scatter varies from as low 
-as 0.08~dex (Mn, Y) up to 0.40~dex (Al, Ba, Ni). Al is not found to have any 
-predominant factor in these data: it has non-zero contributions from all loads
-(Figure~\ref{fig:exp2-factor-loads}. A similar pattern is seen in Cr, Co, and Ni,
-although the specific scatter for Cr and Co is half that of what is found for
-Ni (0.2\,dex instead of 0.4\,dex).
-
-% TODO: anything else to say here?
+as 0.05~dex (Mg, Ca, Sc, Ti) up to 0.3~dex (C). It is perhaps not surprising that
+the largest specific scatter is found in C, since the present day carbon abundance
+is mostly affected by the evolutionary state of the star, rather than the natal
+abundance. After C, the elements with large specific scatter include Al, Fe, Ni, 
+and Ba. A large relative scatter in Al and Ba might suggest that there are other latent
+factors that contribute to those elements, but those latent factors were not included because
+the message length preferred a fewer number of factors. However, the scatter in 
+Fe and Ni is somewhat surprising. Those elements usually have strong and numerous atomic
+lines that can be easily measured in metal-poor stars.
 
@@
 \begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp2-specific-scatter.png}
+	\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-specific-scatter.pdf}
 	\caption{Specific scatter remaining in the \citet{Barklem:2005} data used in
@@
 
-\subsection{Experiment 4:\\Chemical abundances in the \Galah\ survey}
+\subsection{Experiment 3:\\Chemical abundances in the \Galah\ survey}
 \label{sec:exp4}
@@
 (e.g., $\teff$, $\logg$) are unreliable, and as a result the 
-detailed chemical abundances are untrustworthy. We ordered the
+detailed chemical abundances are untrustworthy. We then took all
+stars with no erroneous flags in the following elemental abundance
+measurements: Mg, Al, Ca, Sc, Ti, Cr, Mn, Fe, Co, Ni, Y, Ba, and Eu.
+These elements are produced from multiple nucleosynthetic
+pathways, and it is this subsample that we will use for 
+our initial chemical experiment with \Galah\ data.
+There were 1,444 stars that met these criteria.
+
+%["Mg", "Al", "Ca", "Sc", "Ti", "Cr", "Mn", "Fe", "Co", "Ni", "Y", "Ba", "Eu"]
+
+\begin{comment}
+We ordered the
 list of chemical abundances by the number of reliable measurements,
@@
 leaving us with a sample of \todo{50,000} stars.
-These elements are produced from multiple nucleosynthetic
-pathways, and it is this subsample that we will use for 
-our initial chemical experiment with \Galah\ data.
-
+\end{comment}
+
+\begin{comment}
 \begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp4-abundance-counts.png}
+	\includegraphics[width=0.45\textwidth]{experiments/exp4-abundance-counts.pdf}
 	\caption{The number of abundance measurements in the second
@@
 \end{figure}
+\end{comment}
 
@@
 in the log-likelihood. The results of this grid search are shown in Figure~\ref{fig:exp4-gridsearch-contours},
-where we show the best log-likelihood found for each permutation, and the corresponding
+where we show the log-likelihood, BIC, and message length found for each permutation.
+
+\begin{figure}
+	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-subset-gridsearch-ll-contours.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-subset-gridsearch-bic-contours.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-subset-gridsearch-mml-contours.pdf}
+    \caption{Model performance metrics resulting from a gridsearch using the
+    		 data generated as part of Experiment 2. The top 
+		 	 panel shows the negative log-likelihood 
+			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
+			 evaluated at each combination of latent factors $J$ and number 
+			 of clusters $K$, and the bottom panel shows the BIC (Equation \ref{eq:bic}).
+			 The white marker indicates the 
+			 lowest value in each panel.
+				}
+    \label{fig:experiment-9-gridsearch}
+\end{figure}
+
+ log-likelihood found for each permutation
+
+and the corresponding
 BIC value. The lowest BIC value is found for $\todo{\NumLatentFactors = X}$ and
@@
 
+\subsection{Experiment 4:\\Chemical abundances in the \APOGEE\ survey}
+\label{sec:exp5}
+
+\todo{some text}
+
 \section{Discussion} \label{sec:discussion}
@@
 that the effective dimensionality of chemical abundance space is lower than
-the number of reported abundances \citep[e.g. about 8,][]{Ting:2008}. 
+the number of reported abundances \citep[e.g.,][]{Ting:2012,Price-Jones:2018}. 
 This is expected. Theoretical models of stellar evolution and nucleosynthesis 

### 2019-05-07 article/mcfa.tex
words

@@
 \begin{abstract}
-Chemical tagging promises to distinguish unique star formation sites from 
-observed photospheric abundances.
-Chemical tagging experiments to date have assumed that photospheric
-abundances are independent of each other, even though it is known that 
-nucleosynthetic processes 
-produce
-making
-
-Chemical tagging experiments to date have assumed that all photospheric
-abundances are independent, despite the effective dimensionality of abundance
-space being lower than the number of available elements.
-
-Nucleosynthetic processes make multiple elements in varying quantities, and the
-dimensionality of chemical abundance space is typically lower than the 
-number of reported abundances. 
-Here we introduce a novel approach to chemical tagging where a set of 
-latent factors (e.g., nucleosynthetic yields) contribute to the
-data, and clustering is performed in the latent space (e.g., relative counts of
-nucleosynthetic events). In doing so we infer basis vectors that are analogous
-to nucleosynthetic yields, and the lower dimensional latent space allows for
-more efficient clustering. We perform experiments on literature chemical abundances
-of metal-poor stars, \todo{others}, and 
-those from the second \Galah\ data release. \todo{We find that 5 latent factors are
-preferred to explain the chemical abundances of metal-poor stars, and 8 are
-needed to account for \Galah\ data.}
-We describe exact methods and software that scale well with the data,
-allowing for chemical tagging experiments for $10^{7.5}$ abundance measurements
-to be performed in minutes.
+Chemical tagging seeks to identify unique star formation sites from
+present-day stellar abundances.
+Previous techniques have treated chemical abundances as being 
+statistically independent, despite the nucleosynthesis literature showing that
+most elements can be produced by more than one nucleosynthetic pathway.
+In this work we introduce a data-driven model for chemical abundances
+where a set of latent factors (e.g., nucleosynthetic yields) contribute
+to all stars with different scores, and the clustering (e.g., chemical tagging) 
+is represented by a mixture model in a lower-dimensional latent space.
+We use an exact method to simultaneously estimate the factor scores for
+each star, the relative weighting (e.g., partial assignment) of each star
+to each cluster, and to infer the latent factors that are common to all stars.
+We perform experiments on literature chemical abundances
+of metal-poor stars, and those from the second \Galah\ data release. 
+We find that five latent factors are preferred to explain the detailed 
+abundances of metal-poor stars, and eight are preferred given \todo{X} abundances
+in \Galah\ data. In both experiments we recover latent factors that 
+are representative of mean yields expected from the rapid- and slow-neutron
+capture process, as well as nucleosynthesis in Type II supernovae. 
+We find two clusters are preferred in the \Galah\ data of \todo{X} abundances
+and \todo{Y} stars, which are representative of the [$\alpha$/Fe]-poor and 
+[$\alpha$/Fe]-rich populations in the Milky Way disk. We expect more components
+would be preferred given complete abundances of more stars.
+We release software that scales \todo{as $\mathcal{O}(XX)$}, where $X$ is \todo{what},
+enabling model parameter estimation in minutes given $\sim$$10^7$
+chemical abundance measurements.
 \end{abstract}
@@
 
-Chemical tagging seeks to identify star formation events using the present
-day photospheric abundances of stars \citep{Freeman;Bland-Hawthorn:2002}.
+%Chemical tagging seeks to identify star formation events using the present
+%day photospheric abundances of stars \citep{Freeman;Bland-Hawthorn:2002}.
 The detailed chemical abundances observable in a star's photosphere provide a
-fossil record that carries with it information about where and when a star
-formed. While the photospheric abundances remain largely unchanged throughout
+fossil record that carries with it information about where and when that star
+formed \citep{Freeman;Bland-Hawthorn:2002}. While the photospheric abundances remain largely unchanged throughout
 a star's lifetime \citep[however see][]{Dotter:2017,Ness:2018b}, the dynamical 
@@
 star formation site \citep{Mitschang:2014}. However, there are also analysis
-issues in understanding how well those chemical abundances can be measured
-\citep{Bovy:2016}, and how chemically similar multiple unrelated stars (or
-doppleg\"angers) can be \citep{Ness:2018}.
+issues in understanding how precisely those chemical abundances can be measured
+\citep{Bovy:2016}, and how chemically similar stars can be which did not form 
+together \citep[doppleg\"angers;][]{Ness:2018}.
 If open clusters were truely chemically homogeneous then our
-ability to chemically tag the Milky Way depends on the precision with which we
+ability to chemically tag the Milky Way would depend primarily on the precision with which we
 can measure those chemical abundances in stars. Data-driven techniques are
-driving a revolution in this field \citep{Ness:2015,Ness:2018a,Ness:2018b,
-Casey:2016,Casey:2017,Ho:2017b,Ho:2017a,Leung;Bovy:2018}, but there is more work
-to do: astronomers have not yet reached the Cram\'er-Rao bound for chemical 
-abundances of stars.
+improving upon this precision, \citep{Ness:2015,Ness:2018a,Ness:2018b,
+Casey:2016,Casey:2017,Ho:2017b,Ho:2017a,Leung;Bovy:2018}, but more work is
+needed: astronomers have not yet developed unbiased estimators of chemical
+abundances that saturate the Cram\'er-Rao bound.
 
@@
 tagging becomes a clustering problem. All clustering techniques applied to 
-chemical tagging assume that the data dimensions are independent. That is to say
+chemical tagging thus far have assumed that the data dimensions are independent. That is to say
 that adding a dimension of [Ni/H] provides totally new, independent information.
@@
 
-Alternative approaches are perhaps more challenging, as it is difficult to prescribe
+%Alternative approaches are perhaps more challenging, as i
+It is not trivial to estimate
 the nucleosynthetic yields that have contributed to the chemical abundances of each star
@@
 \begin{equation}
-	\vecdata = \factorloads\factorscores + \vec{e}
+	\vecdata = \vec\mu + \factorloads\factorscores + \vec{e}
 	\label{eq:generative-model}
@@
 \noindent{}are drawn from the $\numcomponents$th multivariate normal distribution.
+The mean vector $\vec\mu$ describes the mean datum in each dimension.
 The factor scores for all data points $\factorscores$ is then a 
@@
 Throughout this work we found that the optimised model parameters are 
-insensitive to the choice of initialisation, up until orthogonal 
-rotation of the latent space.
+insensitive to the choice of initialisation.
+%up until orthogonal 
+%rotation of the latent space.
 
@@
 The updated parameters estimates are found by setting the second derivative
-of the log-likelihood (Eq.~\ref{eq:log-likelihood}) to zero and solving for
+of the log likelihood (Eq.~\ref{eq:log-likelihood}) to zero and solving for
 the parameter values. In doing so this guarantees that every updated
-estimate of the model parameters is guaranteed to increase the log-likelihood.
+estimate of the model parameters is guaranteed to increase the log likelihood.
 Although there are no guarantees against converging on local minima, in 
-general it is sufficient to run expectation-maximization from multiple
+practice it is sufficient to run expectation-maximization from multiple
 initialisations in order to ensure that the global minima is reached.
@@
 \noindent{}where $\odot$ denotes is the entry-wise (Hadamard) product. We repeat
-the expectation-maximization cycle for up to 10,000 steps or until the log-likelihood
+the expectation-maximization cycle for up to 10,000 steps or until the log likelihood
 improved by less than $10^{-5}$ between successive iterations. Although our implementation
@@
 matrices in latent space $\scorecovs$, we set this regularisation term to zero for all
-experiments performed here.
-
-\begin{figure*}
-	\includegraphics[width=1.0\textwidth]{experiments/exp1-data-colour.pdf}
-    \caption{The generated data for Experiment~1, where points are coloured by their
-		     true associations to each component.}
-    \label{fig:exp1-data}
-\end{figure*}
+experiments performed here. Similarly, we assume that the data are noiseless and we do
+not add any observed errors to the constructed covariance matrices.
 
@@
 However, while the BIC does include a penalisation term for the number of
-parameters (which scales with $\log{N}$), it does not account for the increased
-complexity that is afforded by additional parameters. Adding one parameter
+parameters (which scales with $\log{N}$), it does not describe for the
+increased flexibility that is afforded by the addition of those parameters.
+For example, adding one parameter
 to a model will increase the BIC by at most $\log{N}$, but there are different
@@
 
+\begin{figure*}
+	\includegraphics[width=1.0\textwidth]{experiments/exp1-data-colour.pdf}
+    \caption{The generated data for Experiment~1, where points are coloured by their
+		     true associations to each component.}
+    \label{fig:exp1-data}
+\end{figure*}
+
 In practice calculating the message length can be a non-trivial task, 
@@
 to the $\frac{Q}{2}\log\kappa_q$ term that gives a measure of the volume of the region of
-uncertainty in which the parameters $\vec\psi$ are centred. The $\log\kappa_q$
+uncertainty in which the parameters $\vec\Psi$ are centred. The $\log\kappa_q$
 term is
@@
 describes the curvature of the log likelihood with respect to the model
+parameters. For these reasons, MML provides a more accurate description of
+the model complexity (or flexibility) because it naturally includes the 
+curvature of the log likelihood with respect to the model parameters
+rather than only penalising models based on the \emph{number} of
 parameters.
@@
 \end{equation}
+
 The Fisher information for $\factorloads$ \todo{is hard to compute. I have 
@@
 	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-mml-contours.pdf}
-    \caption{Model performance metrics resulting from a gridsearch using the
-    		 data generated as part of our toy model. The top 
-		 	 panel shows the negative log-likelihood 
+    \caption{The top panel shows the negative log likelihood 
 			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
-			 evaluated at each combination of latent factors $J$ and number 
-			 of clusters $K$, and the bottom panel shows the BIC (Equation \ref{eq:bic}).
-			 The white marker indicates the 
-			 lowest value in each panel, which is joined by a black line from the true value
-			 in the top panel. The model with the lowest message length and lowest BIC has
-			 the same number of latent factors and components used to generate the data.
-		}
+			 evaluated at each combination of latent factors 
+			 $\NumLatentFactors$ and number of clusters 
+			 $\NumComponents$ using the generated data in
+			 Experiment~1.  The middle panel shows 
+			 the BIC (Eq.~\ref{eq:bic}) for those 
+			 combinations, and the lower panel shows the 
+			 message length. The white marker indicates the
+			 lowest value in each panel, showing the
+			 preferred number of latent factors and components.
+			 The black marker indicates the true value.}
     \label{fig:experiment-1-gridsearch}
 \end{figure}
-
-Here we treat the generated data set as if the true number of latent factors
-and the true number of components are not known. Starting with $\NumLatentFactors = 1$
-and $\NumComponents = 1$, we trialled each permutation of $\NumLatentFactors$ and $\NumComponents$
-until $\NumLatentFactors_{max} = 10$
-and   $\NumComponents_{max} = 40$ (e.g., twice $\NumLatentFactors_\textrm{true}$ and $\NumComponents_\textrm{true}$).
-For each $\NumLatentFactors$, $\NumComponents$ permutation we initialised the factor
-loads and component assignments randomly. We performed expectation-maximization 
-cycles until the relative log-likelihood improved by less than $10^{-5}$.
-
-These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
-Unsurprisingly the log likelihood increases with increasing numbers of latent
-factors $\NumLatentFactors$ and increasing numbers of components $\NumComponents$.
-The lowest BIC value and message length is found at $\NumLatentFactors = 5$
-and $\NumComponents = 20$, identical to the true values.
 
@@
 
-We show a corner plot of the generated data in Figure~\ref{fig:exp1-data},
-demonstrating the clustering (colour) and structure varying structure in data space.
-Despite this clustering in data space, it is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
-and allowing clustering in the latent space provides a better description of the data.
-Adding components to the model improves the log-likelihood, even with a single latent factor,
-but the addition of just one latent factor improvers the log-likelihood more than adding
-20 clusters. Not much can be said for this example because the true data generating process 
-is drawn from a higher number of latent factors, but this does illustrate how 
-clustering in high dimensional data can be better described by latent factors with 
-clustering in the lower dimensional latent space.
-The inferred clustering in latent space is shown in Figure~\ref{fig:exp1-latent},
-where each point is coloured by the inferred associated component.
+Here we treat the generated data set as if the true number of latent factors
+and the true number of components are not known. Starting with $\NumLatentFactors = 1$
+and $\NumComponents = 1$, we trialled each permutation of $\NumLatentFactors$ and $\NumComponents$
+until $\NumLatentFactors_{max} = 10$
+and   $\NumComponents_{max} = 40$ (e.g., twice $\NumLatentFactors_\textrm{true}$ and $\NumComponents_\textrm{true}$).
+For each $\NumLatentFactors$, $\NumComponents$ permutation we initialised the factor
+loads and component assignments randomly. We performed expectation-maximization 
+cycles until the relative log likelihood improved by less than $10^{-5}$.
+
+These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
+Unsurprisingly the log likelihood increases with increasing numbers of latent
+factors $\NumLatentFactors$ and increasing numbers of components $\NumComponents$.
+The lowest BIC value and message length is found at $\NumLatentFactors = 5$
+and $\NumComponents = 20$, identical to the true values.
 
@@
 \end{figure*}
+
+We show a corner plot of the generated data in Figure~\ref{fig:exp1-data},
+demonstrating the clustering (colour) and structure varying structure in data space.
+Despite this clustering in data space, it is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
+and allowing clustering in the latent space provides a better description of the data.
+Adding components to the model improves the log likelihood, even with a single latent factor,
+but the addition of just one latent factor improvers the log likelihood more than adding
+20 clusters. Not much can be said for this example because the true data generating process 
+is drawn from a higher number of latent factors, but this does illustrate how 
+clustering in high dimensional data can be better described by latent factors with 
+clustering in the lower dimensional latent space.
+The inferred clustering in latent space is shown in Figure~\ref{fig:exp1-latent},
+where each point is coloured by the inferred associated component.
 
@@
 $\factorloads_\textrm{true}$
-despite the difference in ordering and sign. These examples serve to illustrate a more 
+despite the difference in ordering and sign. The same can be said for any other
+scalar metric \citep[e.g., Kullback-Leibler divergence;][]{Kullback:1951}.
+These examples serve to illustrate a more 
 general property that the factor loads and factor scores can be orthogonally 
@@
 ($\factorloads$, $\scoremeans$, $\scorecovs$), but has no effect on the model's 
-predictive power: the log-likelihood \citep[or the Kullback-Leibler divergence;][]{Kullback:1951} under the
+predictive power: the log likelihood \citep[or the Kullback-Leibler divergence;][]{Kullback:1951} under the
 rotated model is indistinguishable from the unrotated model.
@@
 	\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-gridsearch-mml.pdf}
-    \caption{Model performance metrics resulting from a grid search using the
-    		 \citet{Barklem:2005} data in Experiment~2.
-    		 The top panel shows the negative log-likelihood 
+    \caption{The top panel shows the negative log likelihood 
 			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
-			 evaluated at each combination of latent factors $\NumLatentFactors$ and number 
-			 of clusters $\NumComponents$. The middle panel shows the BIC (Eq.~\ref{eq:bic})
-			 for those combinations, and the lower panel shows the message length. 
-			 %The cross at
-			 %$\NumLatentFactors=8$ and $\NumComponents=5$ indicates that the model could not converge without
-			 %adding a regularisation constant to the diagonals of $\scorecovs$.
-			 A marker is shown at the lowest value in each panel, indicating
-			 the preferred number of latent factors and components 
-			 ($\NumLatentFactors = 2$ and $\NumComponents = 1$ is preferred by MML).}
+			 evaluated at each combination of latent factors 
+			 $\NumLatentFactors$ and number of clusters 
+			 $\NumComponents$ using the \citet{Barklem:2005}
+			 data in Experiment~2.  The middle panel shows 
+			 the BIC (Eq.~\ref{eq:bic}) for those 
+			 combinations, and the lower panel shows the 
+			 message length. The white marker indicates the
+			 lowest value in each panel, showing the
+			 preferred number of latent factors and components.}
     \label{fig:exp2-gridsearch-contours}
@@
 The first latent factor $|\mathbf{L_0}|$ shows a non-zero contribution to most
-elements (Figure\ref{fig:exp2-factor-loads}). The elements Ca, Sc, Ti, Cr, and Co, mostly
+elements (Figure~\ref{fig:exp2-factor-loads}). The elements Ca, Sc, Ti, Cr, and Co, mostly
 Fe-peak elements, have the largest contribution from this factor. We interpret this latent
@@
 	\includegraphics[width=\textwidth]{experiments/exp2-log-xfe-offset-latent-factors-visualize-abs.pdf}
-	\caption{Rotated latent factors found in Experiment~2 using the \citet{Barklem:2005}
-			 data. The panels on the left show the entries for each factor load. On
-			 the right hand side we show the absolute fractional contributions to
-			 each element, which illustrates the loads that contribute most to each
-			 element (either by positive or negative factor loads).}
+	\caption{Latent factors found in Experiment~2 using the \citet{Barklem:2005}
+			 data. The panels on the left show the absolute entries for each factor load. 
+			 On the right we show the absolute fractional contributions to
+			 each element, illustrating the loads that contribute most to each element.}
     \label{fig:exp2-factor-loads}
@@
 with \Galah\ data do not include lithium or carbon abundances because the
-photospheric varies throughout a star's lifetime. This is true
+photospheric varies throughout a star's lifetime (e.g., see Experiment 2). This is true
 to a small degree for many elements \citep[e.g.,][]{Dotter:2017},
@@
 lifetime.
+
+\begin{figure}
+	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-gridsearch-ll.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-gridsearch-bic.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-gridsearch-mml.pdf}
+    \caption{The top panel shows the negative log likelihood 
+			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
+			 evaluated at each combination of latent factors 
+			 $\NumLatentFactors$ and number of clusters 
+			 $\NumComponents$ using \Galah\ data in
+			 Experiment~3.  The middle panel shows 
+			 the BIC for those combinations, and the lower panel shows the 
+			 message length. The white marker indicates the
+			 lowest value in each panel, showing the
+			 preferred number of latent factors and components.}
+    \label{fig:experiment-9-gridsearch}
+\end{figure}
+
+\begin{figure*}
+	\includegraphics[width=\textwidth]{experiments/exp9-log-xfe-offset-latent-factors-visualize-abs.pdf}
+	\caption{Latent factors inferred in Experiment~3 using the \Galah\
+			 data. The panels on the left show the absolute entries for each
+			 factor load. On the right we show the absolute fractional contributions
+			 to each element, illustrating the loads that contribute most to each
+			 element.}
+    \label{fig:exp9-factor-loads}
+\end{figure*}
 
@@
 stars with no erroneous flags in the following elemental abundance
-measurements: Mg, Al, Ca, Sc, Ti, Cr, Mn, Fe, Co, Ni, Y, Ba, and Eu.
+measurements: Na, Mg, Al, Si, K, Ca, Sc, Ti, Cr, Mn, Fe, Co, Ni, Y, Ba, and Eu.
 These elements are produced from multiple nucleosynthetic
@@
 our initial chemical experiment with \Galah\ data.
-There were 1,444 stars that met these criteria.
+There were 1,153 stars that met these criteria.
 
@@
 Starting with $\NumLatentFactors = 1$ and $\NumComponents = 1$, we trialled each
-permutation of $\NumLatentFactors$ and $\NumComponents$ up until $\NumLatentFactors = \NumDimensions - 1 = 17$
-and $\NumComponents = 20$. For each permutation we initialised the model randomly,
+permutation of $\NumLatentFactors$ and $\NumComponents$ up until $\NumLatentFactors = 10$
+and $\NumComponents = 10$. For each permutation we initialised the model randomly,
 and continued expectation-maximization until a tolerance of $10^{-5}$ was achieved
-in the log-likelihood. The results of this grid search are shown in Figure~\ref{fig:exp4-gridsearch-contours},
-where we show the log-likelihood, BIC, and message length found for each permutation.
+in the log likelihood. The results of this grid search are shown in Figure~\ref{fig:experiment-9-gridsearch},
+where we show the negative log likelihood, the BIC, and message length found for each permutation.
+The model with $J = 5$ and $K = 2$ is found to have the shortest message length.
 
 \begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-subset-gridsearch-ll-contours.pdf}
-	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-subset-gridsearch-bic-contours.pdf}
-	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-subset-gridsearch-mml-contours.pdf}
-    \caption{Model performance metrics resulting from a gridsearch using the
-    		 data generated as part of Experiment 2. The top 
-		 	 panel shows the negative log-likelihood 
-			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
-			 evaluated at each combination of latent factors $J$ and number 
-			 of clusters $K$, and the bottom panel shows the BIC (Equation \ref{eq:bic}).
-			 The white marker indicates the 
-			 lowest value in each panel.
-				}
-    \label{fig:experiment-9-gridsearch}
+	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-specific-scatter.pdf}
+	\caption{Specific scatter remaining in the \Galah\ data \citep{Buder:2018}
+			 in Experiment~3 after accounting for the contributions by all
+			 latent factors.}
+    \label{fig:exp9-specific-scatter}
 \end{figure}
 
- log-likelihood found for each permutation
-
-and the corresponding
-BIC value. The lowest BIC value is found for $\todo{\NumLatentFactors = X}$ and
-$\todo{\NumComponents = X}$, which we adopted as the preferred number of latent
-factors and components.
-
-\todo{
-\begin{enumerate}
-	\item Plot data with most likely associations? Just as [Fe/H] vs [Si/Fe]?
-	\item Plot factor loads, after rotation.
-	\item Plot latent space.
-	\item Plot aspects of the clusters identified.
-	\item Plot specific scatter compared to the typical reported errors?
-\end{enumerate}
-}
-
-\subsection{Experiment 4:\\Chemical abundances in the \APOGEE\ survey}
-\label{sec:exp5}
-
-\todo{some text}
+\begin{figure}
+	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-cluster.pdf}
+	\caption{Abundance ratios of [Fe/H] and [Mg/Fe] for all stars in Experiment~3,
+			 colored by their probability of being associated with the first (of two)
+			 components. This illustrates how the clustering in latent space is projected
+			 to real space, and indicates that the two components identified are largely
+			 consistent with being stars with low- and high-$\alpha$ element abundance
+			 ratios.}
+    \label{fig:exp9-cluster}
+\end{figure}
+
+The latent factors inferred from this model are shown in Figure~\ref{fig:exp9-factor-loads},
+where we observe some similarities to the factors identified using the \citet{Barklem:2005}
+data of metal-poor stars. The first factor, $|\mathbf{L}_0|$, has non-zero contributions
+to most elements, with a peak near the Fe-peak elements of Fe, Mn, Cr, Co, and Sc. 
+Ca, an $\alpha$-capture element, and Ti, which is observed to have similar trends to
+$\alpha$-capture elements, both have non-zero contributions from $|\mathbf{L}_0|$,
+although both of them have their primary contributions by $|\mathbf{L}_1|$. $|\mathbf{L}_1|$
+contributes most to Mg, another $\alpha$-capture element, and contributes to the light
+odd-Z elements Na, Al, K. 
+The element contributions by the third latent factor, $|\mathbf{L}_2|$, are more ambiguous.
+It contributes in reasonable amounts to many elements, including light elements (Al, Si, K),
+only some Fe-peak elements (notably Sc, Ti, Mn, an Co), and to a couple of neutron-capture
+elements (Ba, Eu). However, $|\mathbf{L}_2|$ is not the dominant factor for any single element:
+it contributes only partially to many elements.
+
+The fourth factor $|\mathbf{L}_3|$ has strong contributions to Ba and Y, elements that are
+typically produced through the slow neutron capture process. However, this factor is also
+the dominant contributor to Ni, and has non-zero contributions to Mn, and a few other light
+elements. Similarly, $|\mathbf{L}_4|$ is the dominant contributor to Eu, an
+element almost entirely produced by the rapid neutron capture process, but it too has
+non-zero contributions to a handful of other elements, including Na and K.
+
+%The specific scatter after accounting for these latent factors is smallest
+%for Fe and largest for K and Y (Figure~\ref{fig:exp9-specific-scatter}).
+
+The relative weighting of the two components identified by this model using
+\Galah\ data is nearly equal. In Figure~\ref{fig:exp9-latent-space} we show 
+abundance ratios colored by their inferred component association.
+It is clear that the component association in latent space does not directly
+translate to a clear separation in data space: in this [Fe/H]-[Mg/Fe] plane
+there is overlap in points between the two associations. 
+\todo{When we look at latent space these components are more clearly separated.}
+However, in general
+the two components could be identified as high and low-$\alpha$ element ratio
+components.
+
+\subsection{Experiment~4:\\Including missing data from the \Galah\ survey} \label{sec:experiment-4}
+
+\todo{Do we bother?}
 
 \section{Discussion} \label{sec:discussion}
+
+\todo{data-driven model of nucleosynthesis}
 
@@
 matrix, leading to latent factor loads that are very different from what was
-estimated by the model, but still lead to the exact same data (or log-likelihood,
+estimated by the model, but still lead to the exact same data (or log likelihood,
 or Kullback-Liebler divergence, etc). As a consequence, we can only `identify'
@@
 
-\todo{
-\begin{itemize}
-	\item at low metallicities how many factors do we prefer, and what can we
-		  vaguely interpret these at?
-	\item when we do that to large galah data sets, what is the number of factors
-	\item clustering in galah
-	\item clustering in apogee
-\end{itemize}
-}
-
 \section{Conclusions} \label{sec:conclusion}
+
+\todo{some}
 

### 2019-05-07 article/mcfa.tex
update to 3.2

@@
 \newcommand{\vecdata}{\vec\data}
+\newcommand{\vecdataunscaled}{\vec{X}}
+\newcommand{\diag}[1]{\textrm{diag}(#1)}
 
@@
 
-\received{2019 March XX}
+\received{2019}
 \revised{2019}
@@
 to each cluster, and to infer the latent factors that are common to all stars.
+We use an information-theoretic Bayesian principle to estimate the number of
+latent factors and clusters.
 We perform experiments on literature chemical abundances
 of metal-poor stars, and those from the second \Galah\ data release. 
-We find that five latent factors are preferred to explain the detailed 
-abundances of metal-poor stars, and eight are preferred given \todo{X} abundances
+We find that \todo{five} latent factors are preferred to explain the detailed 
+abundances of metal-poor stars, and \todo{eight} are preferred given \todo{X} abundances
 in \Galah\ data. In both experiments we recover latent factors that 
@@
 capture process, as well as nucleosynthesis in Type II supernovae. 
-We find two clusters are preferred in the \Galah\ data of \todo{X} abundances
+We find \todo{two} clusters are preferred in the \Galah\ data of \todo{X} abundances
 and \todo{Y} stars, which are representative of the [$\alpha$/Fe]-poor and 
 [$\alpha$/Fe]-rich populations in the Milky Way disk. We expect more components
-would be preferred given complete abundances of more stars.
+would be preferred given complete abundance inventory of a larger sample of stars.
 We release software that scales \todo{as $\mathcal{O}(XX)$}, where $X$ is \todo{what},
@@
 together \citep[doppleg\"angers;][]{Ness:2018}.
-If open clusters were truely chemically homogeneous then our
-ability to chemically tag the Milky Way would depend primarily on the precision with which we
-can measure those chemical abundances in stars. Data-driven techniques are
+If open clusters were truely chemically homogeneous then under idealistic 
+assumptions our ability to chemically tag the Milky Way would depend primarily
+on the precision with which we can measure those chemical abundances in stars. 
+Data-driven techniques are
 improving upon this precision, \citep{Ness:2015,Ness:2018a,Ness:2018b,
@@
 
-Given these data and the most
-favourable assumptions in chemical tagging -- that star clusters are truely
-chemically homogenous, and we can measure those abundances with infinite precision,
-and those abundances are differentiable between star clusters -- then chemical
+Given these data and the most favourable assumptions in chemical tagging 
+-- that star clusters are truely chemically homogenous, and we can measure 
+those abundances with infinite precision, and those abundances are 
+differentiable between star clusters -- then chemical
 tagging becomes a clustering problem. All clustering techniques applied to 
 chemical tagging thus far have assumed that the data dimensions are independent. That is to say
-that adding a dimension of [Ni/H] provides totally new, independent information.
+that adding a dimension of [Ni/H] provides entirely new, independent information
+that could not have been predicted from other elemental abundances.
 Theory and observations agree that this cannot be true.
 Nucleosynthetic processes produce multiple elements in varying
-quantities, and the dimensionality of stellar abundance datasets has been shown
-to be lower than the actual number of dimensions \citep{Ting:2012, Price-Jones:2018}.
+quantities, and the effective dimensionality of stellar abundance datasets has been shown
+to be lower than the actual number of dimensions \citep{Price-Jones:2018}.
 Any clustering approach that treats each new elemental abundance as an 
@@
 single star as a contribution of yields from different systems (e.g., 20\% of the
-Ba was produced by 3 M$_\odot$ systems). This is in part because of the
-challenging and degenerate nature of the problem as described, and is complicated
-by the differences in yield predictions that account from prescriptions used in
-different codes.
+Ba was produced by 3 M$_\odot$ asymptotic giant branch stars). This is in part 
+because of the challenging and degenerate nature of the problem as described, 
+and is complicated by the differences in yield predictions that account from 
+prescriptions used in different codes.
 
@@
 two stars could have formed from the same star cluster, given some model of the
-Milky Way. In this work we focus on the former. Here we present a new approach
-to chemical tagging that allows us to identify the latent (unobserved) factors
-that contribute to the chemical abundances of all stars (e.g., nucleosynthetic
-yields) while simultaneously performing clustering in the latent space.
-Notwithstanding caveats that we will
+Milky Way. In this work we focus on the former.
+
+Here we present a new approach to chemical tagging that allows us to identify 
+the latent (unobserved) factors that contribute to the chemical abundances of 
+all stars (e.g., nucleosynthetic yields) while simultaneously performing 
+clustering in the latent space. Notwithstanding caveats that we will
 discuss in detail, this allows us to infer nucleosynthetic yields rather than
@@
 mixture of multivariate normal distributions in the latent space (factor scores).
-In this work the data $\vecdata$ is a 
+In this work the data $\vecdataunscaled$ is a 
 $\NumData \times \NumDimensions$ matrix where $\NumData$ is the number of 
@@
 \begin{equation}
-	\vecdata = \vec\mu + \factorloads\factorscores + \vec{e}
+	\vecdataunscaled = \vec\mu + \factorloads\factorscores + \vec{e}
 	\label{eq:generative-model}
@@
 association to the components in latent space. 
-We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \eye\specificvariance\right)$
+We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \textrm{diag}(\specificvariance)\right)$
 is independent of the latent space, and $\specificvariance$ is a
-column vector of $\NumDimensions$ entries. 
+vector of variances in each $\NumDimensions$ abundance dimensions.
 In this model each data point can be represented as being drawn
@@
 of nucleosynthetic
-events (e.g., $s$-process production from AGB stars), and the
+events (e.g., $s$-process production from AGB stars averaged over
+initial mass function and star formation history), and the
 factor scores are analogous to the relative counts of those 
@@
 within a statistically consistent framework. That is to say
-that we have a generative model for the data that can 
+that we have a generative data-driven model that can 
 quantitatively account for nucleosynthetic yields, variations in
@@
 
-Without loss of generality the density of the data $\vecdata$ can be described as
-\begin{equation}
-	f(\vecdata; \vec\Psi) = \sum_{\numcomponents=1}^{\NumComponents}\weight_\numcomponents\phi(\vecdata;\factorloads\scoremeans_\numcomponents, \factorloads\scorecovs_\numcomponents\factorloads\transpose + \eye\specificvariance)
+Without loss of generality the density of the mean-subtracted 
+data $\vecdataunscaled - \vec\mu$ (which we hereafter will refer to simply as $\vecdata$) can be described as
+\begin{equation}
+	f(\vecdata; \vec\Psi) = \sum_{\numcomponents=1}^{\NumComponents}\weight_\numcomponents\phi(\vecdata;\factorloads\scoremeans_\numcomponents, \factorloads\scorecovs_\numcomponents\factorloads\transpose + \diag{\specificvariance})
+	%\eye\specificvariance
 \end{equation}
@@
 vector
-$\vec\Psi$ includes $\{\factorloads,\vec\pi,\scoremeans,\scorecovs,\specificvariance\}$, and $\phi(\vecdata;\vec\mu, \vec\Sigma)$
-describes the density of a multivariate gaussian distribution with
-mean $\vec\mu$ and covariance matrix $\vec\Sigma$,
+$\vec\Psi$ includes $\{\factorloads,\vec\pi,\scoremeans,\scorecovs,\specificvariance\}$, and $\phi(\vecdata; \vec\theta)$
+describes the density of a multivariate gaussian distribution,
 and $\weight_\numcomponents$ describes the relative weighting of the $\numcomponents$th
-component in latent space and $\sum\weight_\numcomponents = 1$.
+component in latent space and $\sum\vec\weight = 1$.
 The log likelihood is then given by
@@
 
-As mentioned previously, the model as described is indeterminate in that
-there is no unique solution for the factor loads $\factorloads$ and scores
+The model as described is indeterminate in that there is no unique 
+solution for the factor loads $\factorloads$ and scores
 $\factorscores$. These quantities can only be determined up until 
-orthogonality in $\factorloads$. However, as we will show, accurate estimates
-of the parameter vector $\vec\Psi$ can be obtained by the expectation-maximization
+orthogonality in $\factorloads$. However, as we will describe in Section \ref{sec:expectation-maximization}, with suitable priors on $\vec\Psi$ 
+one can efficiently estimate the model parameters using the expectation-maximization
 algorithm \citep{Dempster:1977}. 
@@
 
-\subsection{Initialisation}
+\subsection{Initialisation} \label{sec:initialisation}
 
 Here we describe the default initialisation of the model parameters, which
-we have found to be robust in a large set of models. Given a number of
+we have found to be robust in a large set of experiments. Given a number of
 $\NumLatentFactors$ latent factors we initialise the factor load
@@
 methods are available in the code associated with this \emph{Article}. 
-Throughout this work we found that the optimised model parameters are 
-insensitive to the choice of initialisation.
-%up until orthogonal 
-%rotation of the latent space.
-
-\subsection{Expectation-Maximization}
+Throughout this work we found that the optimised model parameters appear
+to be insensitive to the choice of initialisation up until orthogonal
+rotation of the latent space.
+
+\subsection{Expectation-Maximization} \label{sec:expectation-maximization}
 
@@
 \begin{equation}
-	\tau_{\numdata\numcomponents} = \frac{\weight_\numcomponents\phi(\vecdata_\numdata;\factorloads\scoremeans_\numcomponents, \factorloads\scorecovs_\numcomponents\factorloads\transpose + \eye\specificvariance)}{\sum_{g=1}^{G}\weight_g\phi(\vecdata_\numdata;\factorloads\scoremeans_g, \factorloads\scorecovs_g\factorloads\transpose + \eye\specificvariance)} \quad .
+	\tau_{\numdata\numcomponents} = \frac{\weight_\numcomponents\phi(\vecdata_\numdata;\factorloads\scoremeans_\numcomponents, \factorloads\scorecovs_\numcomponents\factorloads\transpose + \diag{\specificvariance})}{\sum_{g=1}^{G}\weight_g\phi(\vecdata_\numdata;\factorloads\scoremeans_g, \factorloads\scorecovs_g\factorloads\transpose + \diag{\specificvariance})} \quad .
 \end{equation}
@@
 \end{equation}
-\noindent{}where the $\vec{X}\thisstep$ superscript refers to the current estimate of a variable
-and $\vec{X}\nextstep$ refers to the updated estimate for the next iteration.
+\noindent{}where the $\vec{\Psi}\thisstep$ superscript refers to the current parameter estimates and $\vec{\Psi}\nextstep$ refers to the updated estimate for the next iteration.
 
@@
 \begin{equation}
-	\factorloads\nextstep = \factorloads_{1}\left(\factorloads_{2}^{-1}\eye\right)
+	\factorloads\nextstep = \factorloads_{a}\left(\factorloads_{b}^{-1}\eye\right)
 \end{equation}
@@
 \begin{eqnarray}
-	\factorloads_1 &=& \sum_{\numcomponents=1}^{\NumComponents}\left[ \vec\tau_\numcomponents\transpose\vecdata\left(\scoremeans_\numcomponents\thisstep\right)\transpose + 
+	\factorloads_\textrm{a} &=& \sum_{\numcomponents=1}^{\NumComponents}\left[ \vec\tau_\numcomponents\transpose\vecdata\left(\scoremeans_\numcomponents\thisstep\right)\transpose + 
 	\vec{G}\transpose\vec\tau_\numcomponents\vec{Z}\transpose\vec{G}\right] \\
-	\factorloads_2 &=& N\sum_{\numcomponents=1}^{\NumComponents}\left[\weight_\numcomponents\nextstep\left(\scorecovs_\numcomponents\nextstep + \scoremeans_\numcomponents\nextstep\left(\scoremeans_\numcomponents\nextstep\right)\transpose\right)\right]
+	\factorloads_\textrm{b} &=& N\sum_{\numcomponents=1}^{\NumComponents}\left[\weight_\numcomponents\nextstep\left(\scorecovs_\numcomponents\nextstep + \scoremeans_\numcomponents\nextstep\left(\scoremeans_\numcomponents\nextstep\right)\transpose\right)\right] \quad
 \end{eqnarray}
@@
 \begin{equation}
-	\specificvariance\nextstep = \frac{1}{\NumData}\left[\sum^{\NumComponents}_{\numcomponents=1}\vec\tau_\numcomponents\transpose\left(\vecdata\odot\vecdata\right) - \sum_{j=1}^{J}\left(\factorloads\nextstep\factorloads_2\right)\odot\factorloads\nextstep\right]
+	\specificvariance\nextstep = \frac{1}{\NumData}\left[\sum^{\NumComponents}_{\numcomponents=1}\vec\tau_\numcomponents\transpose\left(\vecdata\odot\vecdata\right) - \sum_{j=1}^{J}\left(\factorloads\nextstep\factorloads_\textrm{b}\right)\odot\factorloads\nextstep\right]
 \end{equation}
@@
 the expectation-maximization cycle for up to 10,000 steps or until the log likelihood
-improved by less than $10^{-5}$ between successive iterations. Although our implementation
-allows for a positive regularisation term to be added along the diagonal of the covariance
-matrices in latent space $\scorecovs$, we set this regularisation term to zero for all
-experiments performed here. Similarly, we assume that the data are noiseless and we do
-not add any observed errors to the constructed covariance matrices.
+improved by less than $10^{-5}$ between successive iterations. 
+Our software implementation allows for a positive regularisation term to be added
+along the diagonal of covariance matrices in latent space $\scorecovs$, for all experiments
+here we set this term to zero. Similarly, we assume that the data are noiseless and
+we do not add any observed errors to the constructed covariance matrices.
+%Although our implementation
+%allows for a positive regularisation term to be added along the diagonal of the covariance
+%matrices in latent space $\scorecovs$, we set this regularisation term to zero for all
+%experiments performed here. Similarly, we assume that the data are noiseless and we do
+%not add any observed errors to the constructed covariance matrices.
 
@@
 
-The expectation-maximization algorithm as described requires the number of latent
-factors $\NumLatentFactors$ and $\NumComponents$ to be given. In the next Section
-we describe experiments using generated and real data, and in all cases we will
+The expectation-maximization algorithm as described requires a specified 
+number of latent factors $\NumLatentFactors$ and $\NumComponents$. In the next 
+Section we describe experiments using generated and real data, and in all cases we will
 assume that the true number of latent factors and components are not known. 
-We require some criterion to decide how many latent factors, and how many
-components, we should adopt given some data. An increasing number of factors
+Hence we require some criterion to decide how many latent factors, and how many
+components, are preferred given some data. An increasing number of factors
 and components will undoubtedly increase the log likelihood of the model given
@@
 \end{equation} 
-\noindent{}where $Q$ is the number of model parameters:
+\noindent{}where $Q$ is the number of model parameters, which for a mixture of
+factor analyzers is:
 \begin{equation}
@@
 The classically-described princple of MML is that the best explanation of the
-data is the one that leads to the shortest so-called two-part message~\citep{Wallace:2005}, 
+data given a model is the one that leads to the shortest so-called two-part message~\citep{Wallace:2005}, 
 where a \textit{message} takes into account both the complexity of the model 
@@
 especially for models that are reasonably complex. This makes the strict MML
-principle intractable (or uncomputable) in many cases and necessitating 
+principle intractable (or uncomputable) in many cases and necessitates
 approximations to the message length. Using a Taylor expansion, a generalised
@@
 \begin{equation}
-	{I}(\vec\Psi,\vecdata) = \frac{Q}{2}\log\kappa_q - \log\left(\frac{p(\vec\Psi)}{\sqrt{|\mathcal{F}(\vec\Psi)|}}\right) - \log\mathcal{L}\left(\vecdata|\vec\Psi\right) + \frac{Q}{2} \label{eq:mml} 
+	{I}(\vec\Psi,\vecdata) = \frac{Q}{2}\log\kappa_Q - \log\left(\frac{p(\vec\Psi)}{\sqrt{|\mathcal{F}(\vec\Psi)|}}\right) - \log\mathcal{L}\left(\vecdata|\vec\Psi\right) + \frac{Q}{2} \label{eq:mml} 
 \end{equation}
@@
 Continuous parameters can only be stated to finite precision, which leads
-to the $\frac{Q}{2}\log\kappa_q$ term that gives a measure of the volume of the region of
-uncertainty in which the parameters $\vec\Psi$ are centred. The $\log\kappa_q$
+to the $\frac{Q}{2}\log\kappa_Q$ term that gives a measure of the volume of the region of
+uncertainty in which the parameters $\vec\Psi$ are centred. The $\log\kappa_Q$
 term is
 \begin{equation}
-	\log\kappa_q = -\log{2\pi} + \frac{1}{Q}\log{Q\pi} - \gamma - 1
+	\log\kappa_Q = -\log{2\pi} + \frac{1}{Q}\log{Q\pi} - \gamma - 1
 \end{equation}
@@
 Like the BIC, the message length is penalised by the number of model parameters
-through the $\log\kappa(Q)$ term. However, the model complexity is also 
+through the $\log\kappa_Q$ term. However, the model complexity is also 
 described through the priors and the Fisher information matrix, which
@@
 $p(\NumComponents) \propto 2^{-\NumComponents}$ respectively,
-such that the optimal lossless message to encode them is \citep[p279, sec. 6.8.2;][]{Knorr-Held:2000},
+such that fewer numbers are preferred and the optimal lossless message to 
+encode the number of each is \citep[p279, sec. 6.8.2;][]{Knorr-Held:2000},
 \begin{eqnarray}
@@
 Adopting the same prior density in our model is not practical because 
-it would require priors on $\scoremeans$, $\scorecovs$, and the 
-responsibility matrix $\vec\tau$. Instead, we address this indeterminacy
+it would require the priors $p(\scoremeans|\vec\tau,\vec\weight$) and $p(\scorecovs|\vec\tau,\vec\weight$).
+That is, we would require a prior density on both the means $\scoremeans$
+and covariance matrices $\scorecovs$ in latent space that requires knowledge
+about the responsibility matrix $\vec\tau$ and relative weights $\vec\weight$ in order to estimate the effective scores
+$\factorscores$ for each data point and calculate a joint prior on the product
+of the factor loads $\factorloads$ and factor scores $\factorscores$.
+Instead we address this indeterminacy
 by placing a prior on $\factorloads$ that ensures it is mutually orthogonal.
-Specifically, we adopt a Wishart distribution with scale matrix $\vec{S}$
+Specifically, we adopt a Wishart distribution with scale matrix $\vec{W}$
 and $D$ degrees of freedom for the
 $\NumLatentFactors\times\NumLatentFactors$ matrix $\vec{M} = \factorloads\transpose\factorloads$.
-In other words, $\vec{M} \sim W_\NumLatentFactors(D,\vec{S})$
-and $\vec{S} = \textrm{Cov}(\factorloads\transpose)$.
+In other words, $\vec{M} \sim W_\NumLatentFactors(D,\vec{W})$
+and $\vec{W} = \textrm{Cov}(\factorloads\transpose)$.
 This
@@
 \begin{equation}
-	p(\factorloads) = \frac{|\factorloads\transpose\factorloads|^{\frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)}}{2^{\frac{\NumDimensions\NumLatentFactors}{2}}|\vec{S}|^{\frac{\NumDimensions}{2}}\Gamma(\frac{\NumDimensions}{2})}\exp\left[-\frac{1}{2}\textrm{Tr}(\vec{S}^{-1}\factorloads\transpose\factorloads)\right] \quad .
+	p(\factorloads) = \frac{|\factorloads\transpose\factorloads|^{\frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)}}{2^{\frac{\NumDimensions\NumLatentFactors}{2}}|\vec{W}|^{\frac{\NumDimensions}{2}}\Gamma(\frac{\NumDimensions}{2})}\exp\left[-\frac{1}{2}\textrm{Tr}(\vec{W}^{-1}\factorloads\transpose\factorloads)\right] \quad .
 \end{equation}
@@
 the first derivative from the log likelihood function, but the second
-derivative is (surprisingly) causing some trouble to compute.}
+derivative is (surprisingly) causing some trouble.}
 Thus the message length to encode $\factorloads$ is given by
@@
 and $\NumComponents = 1$, we trialled each permutation of $\NumLatentFactors$ and $\NumComponents$
-until $\NumLatentFactors_{max} = 10$
-and   $\NumComponents_{max} = 40$ (e.g., twice $\NumLatentFactors_\textrm{true}$ and $\NumComponents_\textrm{true}$).
+until $\NumLatentFactors_\textrm{max} = 10$
+and   $\NumComponents_\textrm{max} = 40$ (e.g., twice $\NumLatentFactors_\textrm{true}$ and $\NumComponents_\textrm{true}$).
 For each $\NumLatentFactors$, $\NumComponents$ permutation we initialised the factor
@@
 Despite this clustering in data space, it is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
-and allowing clustering in the latent space provides a better description of the data.
+and allowing clustering in the latent space provides a better description of the (generated) data.
 Adding components to the model improves the log likelihood, even with a single latent factor,
-but the addition of just one latent factor improvers the log likelihood more than adding
-20 clusters. Not much can be said for this example because the true data generating process 
+but the addition of just \emph{one latent factor} improvers the log likelihood more than adding
+\emph{twenty components}. Not much can be said for this example because the true data generating process 
 is drawn from a higher number of latent factors, but this does illustrate how 
@@
 $\factorloads_\textrm{est}$ could therefore appear very different from the true 
-values, but they only differ by an orthogonal rotation.
+values, but they only differ by an orthogonal rotation. We discuss the impact of this limitation on real data in more detail in Section~\ref{sec:discussion}.
 
@@
 by solving for $\NumLatentFactors$ unknown angle parameters, each of which is used
-to construct a Givens rotation matrix, and then we take the product of those Givens
+to construct a Givens rotation matrix \citep{Givens:1958}, and then we take the product of those Givens
 matrices. This process reduces to Euler angle rotation in three or fewer dimensions.
@@
 ($\factorloads$, $\scoremeans$, $\scorecovs$), but has no effect on the model's 
-predictive power: the log likelihood \citep[or the Kullback-Leibler divergence;][]{Kullback:1951} under the
+predictive power: the generated data, evaluated log likelihood, or the Kullback-Leibler divergence \citep{Kullback:1951} under the
 rotated model is indistinguishable from the unrotated model.
@@
 that the effective dimensionality of chemical abundance space is lower than
-the number of reported abundances \citep[e.g.,][]{Ting:2012,Price-Jones:2018}. 
+the number of reported abundances \citep[e.g.,][]{Price-Jones:2018}. 
 This is expected. Theoretical models of stellar evolution and nucleosynthesis 

### 2019-05-08 article/mcfa.tex
lotsa wordso

@@
 
-\shorttitle{Chemical tagging in lower-dimensional latent space}
+\shorttitle{A data-driven model of nucleosynthesis with chemical tagging in latent space}
 \shortauthors{Casey et al.}
@@
 
-\title{Chemical tagging in lower-dimensional latent space}
+\title{A data-driven model of nucleosynthesis with chemical tagging in latent space}
 
@@
 most elements can be produced by more than one nucleosynthetic pathway.
-In this work we introduce a data-driven model for chemical abundances
+In this work we introduce a data-driven model of nucleosynthesis 
 where a set of latent factors (e.g., nucleosynthetic yields) contribute
-to all stars with different scores, and the clustering (e.g., chemical tagging) 
-is represented by a mixture model in a lower-dimensional latent space.
+to all stars with different scores, and clustering (e.g., chemical tagging) 
+is represented by a mixture model of multivariate normal distributions
+in a lower-dimensional latent space.
 We use an exact method to simultaneously estimate the factor scores for
@@
 latent factors and clusters.
-We perform experiments on literature chemical abundances
+We perform experiments on generated and real data, including literature chemical abundances
 of metal-poor stars, and those from the second \Galah\ data release. 
@@
 capture process, as well as nucleosynthesis in Type II supernovae. 
-We find \todo{two} clusters are preferred in the \Galah\ data of \todo{X} abundances
-and \todo{Y} stars, which are representative of the [$\alpha$/Fe]-poor and 
-[$\alpha$/Fe]-rich populations in the Milky Way disk. We expect more components
+We find \todo{two} clusters are preferred in the \Galah\ data of 16 abundances
+and 1,136 stars, which are representative of the [$\alpha$/Fe]-poor and 
+[$\alpha$/Fe]-rich populations in the Milky Way disk. 
+We expect more components
 would be preferred given complete abundance inventory of a larger sample of stars.
-We release software that scales \todo{as $\mathcal{O}(XX)$}, where $X$ is \todo{what},
-enabling model parameter estimation in minutes given $\sim$$10^7$
-chemical abundance measurements.
+We release accompanying software that scales well with the available data,
+allowing model parameters to be optimised in a few minutes even given $\sim$$10^{7.5}$ 
+abundance measurements.
+%We release software that scales \todo{as $\mathcal{O}(XX)$}, where $X$ is \todo{what},
+%enabling model parameter estimation in minutes given $\sim$$10^7$
+%chemical abundance measurements.
 \end{abstract}
@@
 of these experiments in Section~\ref{sec:discussion}, including the caveats with
-the model as described. We conclude in Section~\ref{sec:conclusion}.
+the model as described. We conclude in Section~\ref{sec:conclusions}.
 
@@
 of the log likelihood (Eq.~\ref{eq:log-likelihood}) to zero and solving for
-the parameter values. In doing so this guarantees that every updated
+the parameter values\footnote{Strictly this introduces a statistical inconsistency in that we should update our parameter estimates by setting the second derivative of our chosen objective function (Eq.~\ref{eq:message-length}) to zero instead of the log likelihood, but this difference is inconsistency becomes serious at small $N$.}
+In doing so this guarantees that every updated
 estimate of the model parameters is guaranteed to increase the log likelihood.
@@
 
-\begin{figure*}
-	\includegraphics[width=1.0\textwidth]{experiments/exp1-data-colour.pdf}
-    \caption{The generated data for Experiment~1, where points are coloured by their
-		     true associations to each component.}
-    \label{fig:exp1-data}
-\end{figure*}
-
 In practice calculating the message length can be a non-trivial task, 
@@
 %\todo{Dowe said something about MML reducing to BIC under some conditions, but I have not yet found the right paper to cite.}
+
+\begin{figure*}
+	\includegraphics[width=1.0\textwidth]{experiments/exp1-data-colour.pdf}
+    \caption{The generated data for Experiment~1, where points are coloured by their
+		     true associations to each component.}
+    \label{fig:exp1-data}
+\end{figure*}
 
@@
 	I(\factorloads) &=& -\log\left(\frac{p(\factorloads)}{\sqrt{|\mathcal{F}(\factorloads)|}}\right) \nonumber \\
-	&=& -\log{p(\factorloads)} + \todo{\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \nonumber \\
 I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} \nonumber \\
@@
 
-Combining equations \ref{eq:prior_J}, \ref{eq:prior_K}, \ref{eq:prior_pi}, \ref{eq:prior_xi_omega}, and \ref{eq:prior_L} with equation \ref{eq:mml} leads to the full message length
+Combining equations \ref{eq:prior_J}, \ref{eq:prior_K}, \ref{eq:prior_pi}, \ref{eq:prior_xi_omega}, and \ref{eq:prior_L} with equation \ref{eq:mml} leads to the full message length:
+\begin{widetext}
 \begin{eqnarray}
-	I(\vec\Psi, \vec\data) &=& -\log\likelihood(\vec\data|\vec\Psi) \nonumber \\
-	&& +\frac{1}{4}\left(\NumLatentFactors + 4\right)\left(\NumLatentFactors - 1\right)\sum_{\numcomponents=1}^\NumComponents\log\weight_\numcomponents + \left(\NumComponents - \frac{1}{2}\right)\log{\NumData} \nonumber \\
-	&& +\frac{1}{2}\NumDimensions\log|\textrm{Cov}\left(\factorloads\transpose\right)| - \frac{1}{2}\left(D-J-1\right)\log|\factorloads\transpose\factorloads| \nonumber \\
-	&& +\textrm{Tr}\left(\textrm{Cov}\left(\factorloads\transpose\right)^{-1}\factorloads\transpose\factorloads\right) - \left(\NumLatentFactors + \frac{3}{2}\right)\sum_{\numcomponents=1}^\NumComponents\log|\scorecovs_\numcomponents| \nonumber \\
-	&& -\log\Gamma\left(\NumComponents\right) - \Gamma\left(\frac{\NumDimensions}{2}\right) + \frac{Q}{2}\log\kappa_q \nonumber \\
-	&& +\frac{1}{2}\left[\NumLatentFactors(\NumDimensions + 2) + \NumComponents(2-\NumData)\right]\log{2} \nonumber \\
-	&& \todo{+\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} .
+	I(\vec\Psi, \vec\data) &=& -\log\likelihood(\vec\data|\vec\Psi)
+ +\frac{1}{4}\left(\NumLatentFactors + 4\right)\left(\NumLatentFactors - 1\right)\sum_{\numcomponents=1}^\NumComponents\log\weight_\numcomponents + \left(\NumComponents - \frac{1}{2}\right)\log{\NumData}
+ +\frac{1}{2}\NumDimensions\log|\textrm{Cov}\left(\factorloads\transpose\right)| \nonumber \\
+  && \cdots -\frac{1}{2}\left(D-J-1\right)\log|\factorloads\transpose\factorloads| + \textrm{Tr}\left(\textrm{Cov}\left(\factorloads\transpose\right)^{-1}\factorloads\transpose\factorloads\right) - \left(\NumLatentFactors + \frac{3}{2}\right)\sum_{\numcomponents=1}^\NumComponents\log|\scorecovs_\numcomponents|  -\log\Gamma\left(\NumComponents\right) - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\
+&& \cdots +\frac{Q}{2}\log\kappa_q +\frac{1}{2}\left[\NumLatentFactors(\NumDimensions + 2) + \NumComponents(2-\NumData)\right]\log{2} + \todo{\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \quad .
+\label{eq:message-length}
  \end{eqnarray}
+\end{widetext}
 
@@
 In this experiment we use detailed chemical abundances of metal-poor stars 
-\citep{Barklem:2005} to demonstrate the utility and limitations
+from \citet{Barklem:2005} to illustrate the utility and limitations
 in interpreting the latent factors as nucleosynthetic processes. The data 
@@
 abundances measured include Al (light odd-Z), Mg and Ca ($\alpha$-process), Sc, Ti, Cr, Mn, Fe, Co, and Ni (Fe-peak elements), as well as Sr, Y, Ba, and Eu ($s$/$r$-process). 
-The data vector $\data$ includes [Fe/H] abundance ratios and all other elements
-relative to Fe (e.g., [Ti/Fe]) after subtracting the mean value to allow the
-latent factors to describe the abundance patterns away from the mean.
+The data vector includes [Fe/H] abundance ratios and all other elements
+relative to Fe (e.g., [Ti/Fe]) after subtracting the mean value in each dimension.
 
@@
 The lowest BIC value is found for $J = 5$ and $K = 1$, indicating
-six latent factors are preferred, with no clustering. The shortest message length
+five latent factors are preferred with no clustering. The shortest message length
 is also found with no clustering,  but with four latent factors ($J = 4$, $K = 1$).
@@
 Despite the caveats in interpreting latent factors as nucleosynthetic processes
-(e.g. see Sections \ref{sec:exp-1} and \ref{sec:discussion}), here we will
+(e.g., see Sections~\ref{sec:exp-1} and \ref{sec:discussion}), here we will
 qualitatively describe the latent factors found in our selected model.
@@
 Fe-peak elements, have the largest contribution from this factor. We interpret this latent
-factor as contributions from core-collapse supernovae.
+factor as being most similar to expected contributions from core-collapse supernovae.
 The second latent factor $|\mathbf{L_1}|$ has negligible contribution at most elements,
@@
 interpret this latent factor as being most closely representative of the slow neutron
-capture process. The third latent factor has almost no contribution to Ba, but has
+capture process, which primarily occurs in asymptotic giant branch stars. The third latent factor has almost no contribution to Ba, but has
 non-zero contributions to other neutron-capture elements Sr, Y, and Eu. Since Eu has
@@
 representative of the rapid neutron capture process. The last latent factor $|\mathbf{L_3}|$
-has a noticeable peak at Fe and the surrounding Fe-peak elements, with small non-zero contributions
-at most elements.
+has a noticeable peak at Fe and the surrounding Fe-peak elements, with small non-zero contributions at most elements. \todo{What to say of $|\mathbf{L_3}|$?}
 
@@
 factors that contribute to those elements, but those latent factors were not included because
-the message length preferred a fewer number of factors. However, the scatter in 
-Fe and Ni is somewhat surprising. Those elements usually have strong and numerous atomic
-lines that can be easily measured in metal-poor stars.
+the message length preferred a fewer number of factors, or it could be limitations in
+the assumptions of our model (e.g., we assume mutually orthogonal latent factors, an assumption that is likely provably false; see Section~\ref{sec:discussion}).
+%However, the scatter in 
+%Fe and Ni is somewhat surprising. Those elements usually have strong and numerous atomic
+%lines that can be easily measured in metal-poor stars.
 
@@
 	\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-specific-scatter.pdf}
-	\caption{Specific scatter remaining in the \citet{Barklem:2005} data used in
+	\caption{Specific scatter $\sqrt{\specificvariance}$ remaining in the \citet{Barklem:2005} data used in
 			 Experiment~2 after accounting for the contributions by all latent
@@
 up to 23 chemical abundances reported for 342,682 stars.
-Some abundances are not reported, for a variety of 
+Some abundances are not reported for a variety of 
 reasons related to astrophysics or the adopted analyses
 (e.g., a star is too hot to estimate the abundance for a given
-transition, or the signal-to-noise of the spectrum is too low).
-Our method as described cannot handle missing data entries for
-a given star, and as such we are forced to exclude stars that
+electronic transition, or the signal-to-noise of the spectrum is too low).
+Our method as described cannot efficiently handle missing data 
+entries for a given star, and as such we are forced to exclude stars that
 do not have a full complement of abundances. However, we also
@@
 (e.g., $\teff$, $\logg$) are unreliable, and as a result the 
-detailed chemical abundances are untrustworthy. We then took all
+detailed chemical abundances would be untrustworthy. We then took all
 stars with no erroneous flags in the following elemental abundance
@@
 	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-specific-scatter.pdf}
-	\caption{Specific scatter remaining in the \Galah\ data \citep{Buder:2018}
+	\caption{Specific scatter $\sqrt{\specificvariance}$ remaining in the \Galah\ data \citep{Buder:2018}
 			 in Experiment~3 after accounting for the contributions by all
@@
 
-The relative weighting of the two components identified by this model using
-\Galah\ data is nearly equal. In Figure~\ref{fig:exp9-latent-space} we show 
-abundance ratios colored by their inferred component association.
+The two components identified by this model have nearly equal mixing proportions.
+In Figure~\ref{fig:exp9-latent-space} we show the latent space coloured by the
+inferred component association, and in Figure~\ref{fig:exp9-cluster} we project
+this to real space showing [Fe/H] abundance ratios relative to [Mg/Fe].
 It is clear that the component association in latent space does not directly
@@
 there is overlap in points between the two associations. 
-\todo{When we look at latent space these components are more clearly separated.}
-However, in general
-the two components could be identified as high and low-$\alpha$ element ratio
-components.
-
-\subsection{Experiment~4:\\Including missing data from the \Galah\ survey} \label{sec:experiment-4}
-
-\todo{Do we bother?}
+However, when we consider the clustering in latent space these components 
+are more clearly separated. In general the two components identified by the
+model could be described as high- and low-$\alpha$ element ratio stars in
+the disk of the Milky Way.
+
+With a larger sample of stars that have a complete abundance inventory we would expect to 
+identify more components, and perhaps unique star formation events. However
+because our model requires no missing abundances in any element, we would
+need to reduce the number of elemental abundances we include in order to increase
+the sample of stars. And by doing so we restrict the number of latent factors
+that can be inferred from the data.
 
@@
 
-\todo{data-driven model of nucleosynthesis}
-
-%We have presented a novel approach to chemical tagging that incorporates
-%latent factors common to all stars, and allows for clustering in the relative
-%contributions of those factors. This is analogous to simultaneously
-%inferring nucleosynthetic yields as part of chemical tagging.
-
-Earlier studies investigating the limitations of chemical tagging have shown
+Literature studies testing the limitations chemical tagging have shown
 that the effective dimensionality of chemical abundance space is lower than
@@
 This is expected. Theoretical models of stellar evolution and nucleosynthesis
-suggest that there are of order \todo{5} nucleosynthetic mechanisms that are
-predominately responsible for the production of the $\sim$20 or so observed 
-reported in those studies. As a consequence we can expect that the observed
-chemical abundances are correlated in some way (e.g., by a linear combination
-of a fewer number of nucleosynthetic pathways).  However, most approaches to
-chemical tagging to date have assumed that the observed chemical abundances
-are independent of each other \todo{\citep[e.g.,][]{who}}. Those inferences
-will be biased by that assumption in a non-trivial way. This problem cannot be
+would suggest that there are of order at most ten important nucleosynthetic 
+contributors that are responsible for the production of the $\sim$20--30
+chemical abundances typically reported. As a consequence we can expect that 
+the observed chemical abundances are correlated in some way (e.g., by some
+linear combination of a fewer number of nucleosynthetic pathways).  However, 
+most approaches to chemical tagging to date have assumed that the observed 
+chemical abundances are independent of each other. This problem cannot be
 resolved by simply lowering the number of abundances used for chemical tagging (e.g.,
@@
 effective dimensionality of chemical abundance space, and perform clustering
-in that lower dimensional space. Although our model allows us to estimate
-those latent factors that contribute to all stars, and cluster those stars by
-their relative contributions from those factors, there are issues with the
-model we adopt. The model we describe is very likely \emph{not} the correct
-model to use to describe chemical abundances of stars. We require the latent
-factors in our model to be mutually orthogonal to each other. This suggests
-an astrophysical context where the mean nucleosynthetic yields (over all star
-formation histories and stellar masses) of various
-processes (e.g., $r$-process, $s$-process) are mutually orthogonal to each
-other. Clearly this assumption is likely to be incorrect: the nuclear physics
-of one environment where elements are produced  can be very different from
-other environments, and there is no astrophysical constraint that those
-latent factors should be mutually orthogonal.
-A more flexible data-driven model of nucleosynthetic yields (as a function of mass,
-stellar metallicity, and other factors) that does not require the latent
-factors to be mutually orthogonal would be a worthy extension to this work.
+in that lower dimensional space. This provides a data-driven model of
+nucleosynthesis yields and chemical tagging that allows us to simultaneously
+estimate the latent factors that contribute to all stars, and cluster those 
+stars by their relative contributions from each factors. The results are
+encouraging in that we find latent factors that are representative of the
+expected yields from dominant nucleosynthetic channels. However, the model that
+we describe is very likely \emph{not} the correct model to use to represent 
+chemical abundances of stars. Here we discuss the limitations of our model 
+in detail.
+
+We require latent factors to be mutually orthogonal to in order to resolve
+the indeterminacy in our model. This suggests an astrophysical context where 
+the mean nucleosynthetic yields (integrated over all stellar masses and star
+formation histories) of various nucleosynthetic processes (e.g., $r$-process, 
+$s$-process) are mutually orthogonal to each other. Clearly this assumption 
+is likely to be incorrect: the nuclear physics
+of one environment where elements are produced will be very different from
+others, and there is no astrophysical constraint that those
+yields (or latent factors) should be mutually orthogonal.
+One could represent them by a hierarchical data-driven model where the yields contribute as a function of stellar mass, 
+metallicity, and other factors), but in principle to resolve the indeterminacy
+the mean of these yields would require mutual orthogonality.
+Introducing a constraint on the factor scores that resolves this indeterminacy and allows for more flexible latent factors would be a worthy extension to this work.
+
+The constraint of mutual orthogonality limits the inferences we want to make
+about stellar nucleosynthetic yields. For example, after accounting for all
+known sources of potassium production in the Milky Way, galactic chemical evolution
+models under-predict the level of potassium in the Milky Way by more than an
+order of magnitude \citep{Kobayashi:2006}. From our inferences in Experiment~3
+using \Galah\ data, we find multiple latent factors have non-zero contributions
+to potassium: $\factorloads_1$, $\factorloads_2$, and $\factorloads_3$. Let us consider
+$\factorloads_3$, which is the dominant contributor of the principally $r$-process element europium. If our model confidently and
+reliably associated the production of potassium with the rapid neutron capture
+process, it could help explain the peculiar abundances of stars enhanced in
+potassium and depleted in magnesium \citep{Mucciarelli:2012,Cohen:2012} -- a chemical
+abundance pattern that currently lacks explanation \citep{Kemp:2018}.
+However, is the potassium contribution that we infer physically realistic,
+or is it a consequence of requiring that the latent factors are mutually
+orthogonal? Distinguishing these possibilities is non-trivial, which is in
+part why caution is warranted when trying to interpret latent factor models.
+In this situation it is worth commenting that potassium has the largest specific
+scatter (Figure~\ref{fig:exp9-specific-scatter}), suggesting that the contributions
+of potassium are not as well described by the latent factor model as other
+elements.
 
@@
 from astrophysical processes (whilst remaining orthogonal), but here we are limited
-by what astrophysical processes we are \emph{expected} to find.
-
-This constrains our ability to identify new nucleosynthetic processes. For example,
+by what astrophysical processes we are \emph{expected} to find within the constraint
+of being mutually orthogonal.
+
+This in part constrains our ability to identify new nucleosynthetic processes. For example,
 let us consider a hypothetical situation where we would only expect there to be four 
-nucleosynthetic processes that predominately contribute to the observed \GALAH\ abundances,
+nucleosynthetic processes that predominately contribute to the observed \Galah\ abundances,
 but in practice we found that the data are best explained (with high significance) with
@@
 model parameters. When applied to chemical abundances measured from metal-poor stars 
-(Experiment~2) we find that \todo{seven latent factors and two clusters are preferred by
-the data.}
-
-\todo{discussion of experimental results}
-
-\section{Conclusions} \label{sec:conclusion}
-
-\todo{some}
+(Experiment~2) we find that four latent factors are preferred by the data, and
+those latent factors qualitatively re-appear in Experiment~3 using \Galah\ data. 
+That did not have to be the case: the latent factors between different experiments (data sets)
+might be entirely different, and those factors did not have to qualitatively match our
+expectations of nucleosynthetic yields. Indeed, the inferred factors -- even after a
+valid rotation -- could have made no astrophysical sense whatsoever. For this reason it
+is encouraging that there is some interpretability in the latent factors, even though 
+caution is warranted.
+
+The relatively small data set with complete abundances measured in \Galah\ both limits
+the number of components we can infer, and the efficacy of chemical tagging. Only two
+components were preferred from the data set of 1,136 stars, which we identified as
+being stars with low- and high-[$\alpha$/Fe] abundance ratios. A worthwhile extension
+of this work would be to incorporate missing and noisy data in order to allow for
+blind chemical tagging on an industrial scale, allowing for the latent factors to
+be efficiently estimated while chemical tagging occurs in the lower-dimensional
+latent space.
+
+\section{Conclusions} \label{sec:conclusions}
+
+We have introduced a data-driven model of nucleosynthesis by incorporating 
+latent factors that are common to all stars, and allowing for clustering in the
+lower-dimensional latent space. This approach simultaneously allows us to efficiently
+tag stars based on their chemical abundances, and to infer the contributions that are
+common to all stars (e.g., nucleosynthetic yields). Experiments with generated data
+demonstrate that MML is a useful principle for identifying the preferred number of
+latent factors and components. Experiments with real data reveal latent factors
+that are qualitatively similar to expected nucleosynthetic yields (e.g., products
+from the $s$-process, $r$-process, et cetera). Similar latent factors emerge 
+from chemical abundances of metal-poor stars and using data from the second \Galah\
+data release. We express caution in directly interpreting those latent
+factors, as our model explicitly requires mutually orthogonal latent factors.
+We highlight potential extensions of this work that would efficiently deal with
+missing and noisy data points.
+
+% This paper was built from reference XXX on github repo YYY
 
 \acknowledgements
-A.~R.~C. is supported in part by Australian Research Council
-Discovery Project DP160100637.
-The \GALAH\ survey is based on observations made at the Australian Astronomical Observatory, under programmes A/2013B/13,
+We acknowledge support from the Australian Research Council
+through Discovery Project DP160100637.
+The \Galah\ survey is based on observations made at the Australian Astronomical Observatory, under programmes A/2013B/13,
 A/2014A/25, A/2015A/19, A/2017A/18. We acknowledge the traditional owners of the land on which the AAT stands, the Gamilaraay
 people, and pay our respects to elders past and present.
-This work presents results from the European Space Agency (ESA) space mission Gaia.
+%This work presents results from the European Space Agency (ESA) space mission Gaia.
 Gaia data is being processed by the Gaia Data Processing and Analysis Consortium (DPAC).
@@
     \package{Jupyter Notebooks}\,\,\citep{jupyter-notebooks}\,\,
-}
+}    % TOPCAT
 

### 2019-05-08 article/mcfa.tex
standardising figs

@@
 on the precision with which we can measure those chemical abundances in stars. 
-Data-driven techniques are
-improving upon this precision, \citep{Ness:2015,Ness:2018a,Ness:2018b,
+Data-driven approaches to modelling stellar spectra are
+improving upon this precision \citep{Ness:2015,Ness:2018a,Ness:2018b,
 Casey:2016,Casey:2017,Ho:2017b,Ho:2017a,Leung;Bovy:2018}, but more work is
@@
 
+Test \the\textwidth
+
 \begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-specific-scatter.pdf}
+	%\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-specific-scatter.pdf}
+	\includegraphics[width=0.5\textwidth]{../experiments/exp2-specific-scatter.pdf}
 	\caption{Specific scatter $\sqrt{\specificvariance}$ remaining in the \citet{Barklem:2005} data used in

### 2019-05-08 experiments/exp2.py

@@ -1,0 +1,286 @@
+
+import os
+import sys
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import pickle
+import yaml
+from scipy import linalg
+
+from matplotlib.ticker import MaxNLocator
+
+sys.path.insert(0, "../")
+
+from mcfa import (mcfa, grid_search, mpl_utils, utils)
+
+matplotlib.style.use(mpl_utils.mpl_style)
+
+here = os.path.dirname(os.path.realpath(__file__))  
+
+
+with open("config.yml") as fp:
+    config = yaml.load(fp)
+
+
+np.random.seed(config.get("random_seed", 0))
+
+
+def savefig(fig, suffix):
+    prefix = os.path.basename(__file__)[:-3]
+    filename = os.path.join(here, f"{prefix}-{suffix}")
+    fig.savefig(f"{filename}.pdf", dpi=300)
+    fig.savefig(f"{filename}.png", dpi=150)
+    print(f"Created figures {filename}.png and {filename}.pdf")
+    
+
+from astropy.table import Table
+
+mcfa_kwds = dict(init_factors="random", init_components="random", tol=1e-5,
+                 max_iter=10000)
+
+
+
+periodic_table = """H                                                  He
+                    Li Be                               B  C  N  O  F  Ne
+                    Na Mg                               Al Si P  S  Cl Ar
+                    K  Ca Sc Ti V  Cr Mn Fe Co Ni Cu Zn Ga Ge As Se Br Kr
+                    Rb Sr Y  Zr Nb Mo Tc Ru Rh Pd Ag Cd In Sn Sb Te I  Xe
+                    Cs Ba Lu Hf Ta W  Re Os Ir Pt Au Hg Tl Pb Bi Po At Rn
+                    Fr Ra Lr Rf"""
+
+lanthanoids    =   "La Ce Pr Nd Pm Sm Eu Gd Tb Dy Ho Er Tm Yb"
+actinoids      =   "Ac Th Pa U  Np Pu Am Cm Bk Cf Es Fm Md No"
+
+periodic_table = periodic_table.replace(" Ba ", " Ba " + lanthanoids + " ") \
+                               .replace(" Ra ", " Ra " + actinoids + " ") \
+                               .split()
+
+t = Table.read(os.path.join(here, "../catalogs/barklem_t3.fits"))
+
+elements = np.unique(t["El"])
+stars = np.unique(t["Name"])
+
[…]

### 2019-05-08 experiments/exp3.py

@@ -1,0 +1,284 @@
+
+"""
+Experiment using all GALAH data.
+"""
+
+from __future__ import division # Just in case. Use Python 3.
+
+import os
+import sys
+import pickle
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import yaml
+from matplotlib.ticker import MaxNLocator
+from collections import Counter
+from scipy import linalg
+
+sys.path.insert(0, "../")
+
+from mcfa import (mcfa, grid_search, mpl_utils, utils)
+
+import galah_dr2 as galah
+
+
+matplotlib.style.use(mpl_utils.mpl_style)
+
+here = os.path.dirname(os.path.realpath(__file__))
+
+with open("config.yml") as fp:
+    config = yaml.load(fp)
+
+np.random.seed(config.get("random_seed", 0))
+
+prefix = os.path.basename(__file__)[:-3]
+
+def savefig(fig, suffix):
+    here = os.path.dirname(os.path.realpath(__file__))
+    filename = os.path.join(here, f"{prefix}-{suffix}")
+    fig.savefig(f"{filename}.png", dpi=150)
+    fig.savefig(f"{filename}.pdf", dpi=300)
+
+
+N_elements = 20
+use_galah_flags = config["use_galah_flags"]
+
+
+mcfa_kwds = dict(init_factors="svd", init_components="random", tol=1e-5,
+                 max_iter=10000)
+
+
+elements = ["Mg", "Al", "Ca", "Sc", "Ti", "Cr", "Mn", "Fe", "Co", "Ni", "Y", "Ba", "Eu"] + ["K", "Na", "Si"]
+
+
+print(elements)
+
+
+mask = galah.get_abundance_mask(elements, use_galah_flags=use_galah_flags)
+
+print(f"Number of stars: {sum(mask)}")
+
+
+
+X_H, label_names = galah.get_abundances_wrt_h(elements, mask=mask)
+
+
+
+
+
+print(f"Data shape: {X_H.shape}")
+
+
+def convert_xh_to_xy(X_H, label_names, y_label):
+
+    index = label_names.index(y_label)
+    y_h = X_H[:, index]
+
+    offsets = np.zeros_like(X_H)
+    for i, label_name in enumerate(label_names):
+        if label_name == y_label: continue
+        offsets[:, i] = y_h
+
+    return X_H - offsets
+ 
+
+if config["wrt_x_fe"]:
+    X = convert_xh_to_xy(X_H, label_names, "fe_h")
[…]

### 2019-05-08 experiments/galah_dr2.py

@@ -1,0 +1,141 @@
+
+
+import numpy as np
+from astropy.table import Table
+from collections import OrderedDict
+from operator import itemgetter
+import os
+#data = Table.read("GALAH_iDR3_OpenClusters.fits")
+
+here = os.path.dirname(os.path.realpath(__file__))
+
+data = Table.read(os.path.join(here, "../catalogs/GALAH_DR2.1_catalog.fits"))
+
+periodic_table = """H                                                  He
+                    Li Be                               B  C  N  O  F  Ne
+                    Na Mg                               Al Si P  S  Cl Ar
+                    K  Ca Sc Ti V  Cr Mn Fe Co Ni Cu Zn Ga Ge As Se Br Kr
+                    Rb Sr Y  Zr Nb Mo Tc Ru Rh Pd Ag Cd In Sn Sb Te I  Xe
+                    Cs Ba Lu Hf Ta W  Re Os Ir Pt Au Hg Tl Pb Bi Po At Rn
+                    Fr Ra Lr Rf"""
+
+lanthanoids    =   "La Ce Pr Nd Pm Sm Eu Gd Tb Dy Ho Er Tm Yb"
+actinoids      =   "Ac Th Pa U  Np Pu Am Cm Bk Cf Es Fm Md No"
+
+periodic_table = periodic_table.replace(" Ba ", " Ba " + lanthanoids + " ") \
+                               .replace(" Ra ", " Ra " + actinoids + " ") \
+                               .split()
+
+available_elements = list(set([ea.split("_")[0].title() \
+                        for ea in data.dtype.names if (ea.endswith("_fe") or ea == "fe_h") and ea.split("_")[0].lower() not in ("e", "flag", "alpha")]))
+
+indices = np.argsort([periodic_table.index(el) for el in available_elements])
+available_elements = [available_elements[i] for i in indices]
+
+def _get_elements(data):
+    return [ln.split("_")[1] for ln in data.dtype.names \
+            if ln.startswith("A_") and ln.split("_")[1] in periodic_table]
+
+def _abundance_label(element):
+    if element.lower() == "fe":
+        return "fe_h"
+    else:
+        return f"{element.lower()}_fe"
+
+def _abundance_flag_label(element):
+    if element.lower() == "fe":
[…]

### 2019-05-08 mcfa/mpl_utils.py

@@ -9,6 +9,7 @@ from scipy.stats import binned_statistic_2d
 from matplotlib.colors import (BoundaryNorm, LinearSegmentedColormap, LogNorm)
 from matplotlib.ticker import MaxNLocator, FormatStrFormatter, FuncFormatter
 from matplotlib.patches import Ellipse
+from matplotlib import ticker
 
 from .utils import find_rotation_matrix, givens_rotation_matrix
 
@@ -160,7 +161,7 @@ def plot_specific_scatter(model, y=None, scales=1,
                           xlabel=None, xticklabels=None, ylabel=None,
                           steps=False, fill=True, line_kwds=None, fill_kwds=None):
 
-    fig, ax = plt.subplots()
+    fig, ax = plt.subplots(figsize=(3.57, 3.57))
 
     if y is None:
         y = scales * np.sqrt(model.theta_[model.parameter_names.index("psi")])
@@ -193,7 +194,23 @@ def plot_specific_scatter(model, y=None, scales=1,
     ax.set_xticks(x)
 
     if xticklabels is not None:
+        ax.xaxis.set_tick_params(width=0)
+
+        ax.xaxis.set_minor_locator(ticker.FixedLocator(np.arange(x.size) + 0.5))
+
+        """
+        ax.xaxis.set_major_formatter(ticker.NullFormatter())
+        ax.xaxis.set_tick_params(width=00)
+        ax.xaxis.set_minor_formatter(ticker.FixedFormatter(xticklabels))
+        """
+
         ax.set_xticklabels(xticklabels)
+
+        for i, tick in enumerate(ax.get_xaxis().get_major_ticks()):
+            if i % 2 == 0: continue
+
+            tick.set_pad(15.)
+            tick.label1 = tick._get_text1()
 
     if xlabel is not None:
         ax.set_xlabel(xlabel)

### 2019-05-20 article/mcfa.tex
moving furniture and cleaning house

@@
 \usepackage{comment}
-%\usepackage{layouts}
+\usepackage{layouts}
 
@@
 \newcommand{\likelihood}{\mathcal{L}}
+\newcommand{\ExperimentHash}{96ff8}
 
@@
 present-day stellar abundances.
-Previous techniques have treated chemical abundances as being 
+Previous techniques have treated the abundances of each element as being
 statistically independent, despite the nucleosynthesis literature showing that
-most elements can be produced by more than one nucleosynthetic pathway.
+most elements can be produced by more than one pathway.
 In this work we introduce a data-driven model of nucleosynthesis 
@@
 to all stars with different scores, and clustering (e.g., chemical tagging) 
-is represented by a mixture model of multivariate normal distributions
+is represented by a mixture model of multivariate gaussians
 in a lower-dimensional latent space.
 We use an exact method to simultaneously estimate the factor scores for
-each star, the relative weighting (e.g., partial assignment) of each star
-to each cluster, and to infer the latent factors that are common to all stars.
+each star, the partial assignment of each star
+to each cluster, and the latent factors common to all stars.
 We use an information-theoretic Bayesian principle to estimate the number of
@@
 We release accompanying software that scales well with the available data,
-allowing model parameters to be optimised in a few minutes even given $\sim$$10^{7.5}$ 
-abundance measurements.
+allowing for model parameters to be optimised in seconds given a fixed number
+of latent factors, components, and $\sim$$10^{7.5}$ abundance measurements.
 %We release software that scales \todo{as $\mathcal{O}(XX)$}, where $X$ is \todo{what},
@@
 
-The Fisher information for $\factorloads$ \todo{is hard to compute. I have 
-the first derivative from the log likelihood function, but the second
-derivative is (surprisingly) causing some trouble.}
+\printinunitsof{in}\prntlen{\textheight}
+and 
+\printinunitsof{in}\prntlen{\textwidth}
+
+%\todo{Fisher:}
+%\begin{widetext}
+%\begin{eqnarray}
+%	\log\likelihood(\vecdata|\vec\Psi_k) = -\frac{1}{2}(\vecdata - \factorloads\vec\xi_k)\transpose(\factorloads\omega_k\factorloads\transpose + \diag{D})^{-1}) - \frac{1}{2}\log|(\factorloads\omega_k\factorloads\transpose + \diag{D})| -\frac{1}{2}\log{2\pi}
+%	\log\likelihood(\vecdata|\vec\Psi) &=& -\frac{1}{2}(\vecdata - \vec{M})\transpose\vec\Sigma^{-1}(\vecdata - \vec{M}) - \frac{1}{2}\log|\vec\Sigma| - \frac{1}{2}\log{2\pi} \quad \mathrm{where} \\
+%	\vec{M} &=& \factorloads\scoremeans \\
+%	\vec\Sigma &=& \factorloads\scorecovs\factorloads\transpose + \diag{\specificvariance}
+%\end{eqnarray}
+%\end{widetext}
+
+The Fisher information for $\factorloads$ \todo{needs to be checked.}
 Thus the message length to encode $\factorloads$ is given by
@@
 \begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-gridsearch-ll.pdf}
-	\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-gridsearch-bic.pdf}
-	\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-gridsearch-mml.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/exp2-\ExperimentHash-gridsearch-ll.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/exp2-\ExperimentHash-gridsearch-bic.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/exp2-\ExperimentHash-gridsearch-mml.pdf}
     \caption{The top panel shows the negative log likelihood 
@@
 \begin{figure*}
-	\includegraphics[width=\textwidth]{experiments/exp2-log-xfe-offset-latent-factors-visualize-abs.pdf}
+	\includegraphics[width=\textwidth]{experiments/exp2-\ExperimentHash-latent-factors-and-contributions.pdf}
 	\caption{Latent factors found in Experiment~2 using the \citet{Barklem:2005}
-			 data. The panels on the left show the absolute entries for each factor load. 
+			 data. \todo{The panels on the left show the absolute entries for each factor load.} 
 			 On the right we show the absolute fractional contributions to
@@
 \begin{figure}
-	%\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-specific-scatter.pdf}
-	\includegraphics[width=0.5\textwidth]{../experiments/exp2-specific-scatter.pdf}
+	\includegraphics[width=0.5\textwidth]{../experiments/exp2-\ExperimentHash-specific-scatter.pdf}
 	\caption{Specific scatter $\sqrt{\specificvariance}$ remaining in the \citet{Barklem:2005} data used in
@@
 \begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-gridsearch-ll.pdf}
-	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-gridsearch-bic.pdf}
-	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-gridsearch-mml.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/exp3-\ExperimentHash-gridsearch-ll.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/exp3-\ExperimentHash-gridsearch-bic.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/exp3-\ExperimentHash-gridsearch-mml.pdf}
     \caption{The top panel shows the negative log likelihood 
@@
 			 preferred number of latent factors and components.}
-    \label{fig:experiment-9-gridsearch}
+    \label{fig:exp3-gridsearch}
 \end{figure}
@@
 \begin{figure*}
-	\includegraphics[width=\textwidth]{experiments/exp9-log-xfe-offset-latent-factors-visualize-abs.pdf}
+	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-latent-factors-and-contributions.pdf}
 	\caption{Latent factors inferred in Experiment~3 using the \Galah\
-			 data. The panels on the left show the absolute entries for each
-			 factor load. On the right we show the absolute fractional contributions
+			 data. \todo{The panels on the left show the absolute entries for each
+			 factor load.} On the right we show the absolute fractional contributions
 			 to each element, illustrating the loads that contribute most to each
 			 element.}
-    \label{fig:exp9-factor-loads}
+    \label{fig:exp3-factor-loads}
 \end{figure*}
@@
 \begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp4-abundance-counts.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/exp3-abundance-counts.pdf}
 	\caption{The number of abundance measurements in the second
@@
 		     elemental abundances, which we use for Experiment~4.}
-    \label{fig:exp4-abundance-counts} 
+    \label{fig:exp3-abundance-counts} 
 \end{figure}
@@
 \begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-specific-scatter.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/exp3-\ExperimentHash-specific-scatter.pdf}
 	\caption{Specific scatter $\sqrt{\specificvariance}$ remaining in the \Galah\ data \citep{Buder:2018}
@@
 			 latent factors.}
-    \label{fig:exp9-specific-scatter}
+    \label{fig:exp3-specific-scatter}
 \end{figure}
 
-\begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-cluster.pdf}
-	\caption{Abundance ratios of [Fe/H] and [Mg/Fe] for all stars in Experiment~3,
-			 colored by their probability of being associated with the first (of two)
-			 components. This illustrates how the clustering in latent space is projected
-			 to real space, and indicates that the two components identified are largely
-			 consistent with being stars with low- and high-$\alpha$ element abundance
-			 ratios.}
-    \label{fig:exp9-cluster}
-\end{figure}
+\begin{figure*}
+	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-latent-space.pdf}
+	\caption{\todo{Caption}}
+    \label{fig:exp3-latent-space}
+\end{figure*}
+
+\begin{figure*}
+	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-data-space.pdf}
+	\caption{\todo{Caption}}
+    \label{fig:exp3-data-space}
+\end{figure*}
 

### 2019-05-20 article/experiments/exp1.py

@@ -7,7 +7,7 @@ import matplotlib.pyplot as plt
 from matplotlib.ticker import MaxNLocator
 from time import time
 
-sys.path.insert(0, "../../")
+sys.path.insert(0, "../")
 
 from mcfa import (mcfa, grid_search, mpl_utils, utils)
 
@@ -15,10 +15,10 @@ matplotlib.style.use(mpl_utils.mpl_style)
 colors = plt.rcParams['axes.prop_cycle'].by_key()['color']
 
 
-n_features = 15
-n_components = 10
-n_latent_factors = 5
-n_samples = 100_000
+n_features = D = 15
+n_components = K = 10
+n_latent_factors = J = 5
+n_samples = N = 10_000
 
 omega_scale = 1
 noise_scale = 1
@@ -66,9 +66,6 @@ tock = time()
 model.message_length(Y)
 
 print(f"Model took {tock - tick:.1f} seconds")
-
-
-
 
 
 # Plot the true latent factors w.r.t. the estimated ones, after rotation.

### 2019-05-20 article/experiments/exp2-linear.py

@@ -1,457 +1,0 @@
-
-import os
-import sys
-import numpy as np
-import matplotlib
-import matplotlib.pyplot as plt
-import pickle
-from scipy import linalg
-
-from matplotlib.ticker import MaxNLocator
-
-sys.path.insert(0, "../../")
-from mcfa import (mcfa, grid_search, mpl_utils, utils)
-
-matplotlib.style.use(mpl_utils.mpl_style)
-
-here = os.path.dirname(os.path.realpath(__file__))  
-
-random_seed = 10
-adopted_metric = "mml"
-
-np.random.seed(random_seed)
-
-def savefig(fig, suffix):
-    prefix = os.path.basename(__file__)[:-3]
-    filename = os.path.join(here, f"{prefix}-{suffix}")
-    fig.savefig(f"{filename}.pdf", dpi=300)
-    fig.savefig(f"{filename}.png", dpi=150)
-    print(f"Created figures {filename}.png and {filename}.pdf")
-    
-
-from astropy.table import Table
-
-
-periodic_table = """H                                                  He
-                    Li Be                               B  C  N  O  F  Ne
-                    Na Mg                               Al Si P  S  Cl Ar
-                    K  Ca Sc Ti V  Cr Mn Fe Co Ni Cu Zn Ga Ge As Se Br Kr
-                    Rb Sr Y  Zr Nb Mo Tc Ru Rh Pd Ag Cd In Sn Sb Te I  Xe
-                    Cs Ba Lu Hf Ta W  Re Os Ir Pt Au Hg Tl Pb Bi Po At Rn
-                    Fr Ra Lr Rf"""
-
-lanthanoids    =   "La Ce Pr Nd Pm Sm Eu Gd Tb Dy Ho Er Tm Yb"
-actinoids      =   "Ac Th Pa U  Np Pu Am Cm Bk Cf Es Fm Md No"
-
-periodic_table = periodic_table.replace(" Ba ", " Ba " + lanthanoids + " ") \
-                               .replace(" Ra ", " Ra " + actinoids + " ") \
-                               .split()
-
-t = Table.read(os.path.join(here, "../../catalogs/barklem_t3.fits"))
-
-elements = np.unique(t["El"])
-stars = np.unique(t["Name"])
-
-S, E = (len(stars), len(elements))
[…]

### 2019-05-20 article/experiments/exp2.py

@@ -5,32 +5,59 @@ import numpy as np
 import matplotlib
 import matplotlib.pyplot as plt
 import pickle
+import yaml
 from scipy import linalg
+from hashlib import md5
+
 
 from matplotlib.ticker import MaxNLocator
 
-sys.path.insert(0, "../../")
+sys.path.insert(0, "../")
+
 from mcfa import (mcfa, grid_search, mpl_utils, utils)
 
 matplotlib.style.use(mpl_utils.mpl_style)
 
 here = os.path.dirname(os.path.realpath(__file__))  
 
-random_seed = 10
-adopted_metric = "mml"
 
-np.random.seed(random_seed)
+with open("config.yml") as fp:
+    config = yaml.load(fp)
+
+
+np.random.seed(config.get("random_seed", 0))
+
+unique_hash = md5((f"{config}").encode("utf-8")).hexdigest()[:5]
+
+unique_config_path = f"{unique_hash}.yml"
+if os.path.exists(unique_config_path):
+    print(f"Warning: this configuration file already exists {unique_config_path}")
+
+with open(unique_config_path, "w") as fp:
+    yaml.dump(config, fp)
+
+with open(__file__, "r") as fp:
+    code = fp.read()
+
+with open(f"{unique_hash}-{__file__}", "w") as fp:
+    fp.write(code)
+
+prefix = os.path.basename(__file__)[:-3]
 
 def savefig(fig, suffix):
-    prefix = os.path.basename(__file__)[:-3]
-    filename = os.path.join(here, f"{prefix}-{suffix}")
-    fig.savefig(f"{filename}.pdf", dpi=300)
+    here = os.path.dirname(os.path.realpath(__file__))
+    filename = os.path.join(here, f"{prefix}-{unique_hash}-{suffix}")
     fig.savefig(f"{filename}.png", dpi=150)
+    fig.savefig(f"{filename}.pdf", dpi=300)
     print(f"Created figures {filename}.png and {filename}.pdf")
+
+
     
 
 from astropy.table import Table
 
+mcfa_kwds = dict()
+mcfa_kwds.update(config.get("mcfa_kwds", dict()))
 
 periodic_table = """H                                                  He
                     Li Be                               B  C  N  O  F  Ne
[…]

### 2019-05-20 article/experiments/exp3.py

@@ -1,126 +1,175 @@
 
 """
-Experiment using a few gravitationally-bound clusters observed by GALAH.
+Experiment using all GALAH data.
 """
 
 from __future__ import division # Just in case. Use Python 3.
 
 import os
 import sys
+import pickle
 import numpy as np
 import matplotlib
 import matplotlib.pyplot as plt
+import yaml
+from matplotlib.ticker import MaxNLocator
 from collections import Counter
 from scipy import linalg
+from hashlib import md5
 
-sys.path.insert(0, "../../")
+sys.path.insert(0, "../")
 
 from mcfa import (mcfa, grid_search, mpl_utils, utils)
 
-import galah
-
-np.random.seed(100)
+import galah_dr2 as galah
 
 
 matplotlib.style.use(mpl_utils.mpl_style)
 
+here = os.path.dirname(os.path.realpath(__file__))
+
+with open("config.yml") as fp:
+    config = yaml.load(fp)
+
+print(f"Config: {config}")
+
+np.random.seed(config["random_seed"])
+
+prefix = os.path.basename(__file__)[:-3]
+
+unique_hash = md5((f"{config}").encode("utf-8")).hexdigest()[:5]
+
+unique_config_path = f"{unique_hash}.yml"
+if os.path.exists(unique_config_path):
+    print(f"Warning: this configuration already exists: {unique_config_path}")
+
+with open(unique_config_path, "w") as fp:
+    yaml.dump(config, fp)
+
+with open(__file__, "r") as fp:
+    code = fp.read()
+
+with open(f"{unique_hash}-{__file__}", "w") as fp:
+    fp.write(code)
 
 def savefig(fig, suffix):
     here = os.path.dirname(os.path.realpath(__file__))
-    prefix = os.path.basename(__file__)[:-3]
-    filename = os.path.join(here, f"{prefix}-{suffix}")
+    filename = os.path.join(here, f"{prefix}-{unique_hash}-{suffix}")
     fig.savefig(f"{filename}.png", dpi=150)
     fig.savefig(f"{filename}.pdf", dpi=300)
 
 
-mcfa_kwds = dict(init_factors="svd", init_components="kmeans++", tol=1e-8,
-                 max_iter=10000)
+import os
+os.system("rm -f *.pkl")
+
+N_elements = 20
+use_galah_flags = config["use_galah_flags"]
+
+mcfa_kwds = dict()
[…]

### 2019-05-20 article/experiments/exp4-linear.py

@@ -1,496 +1,0 @@
-
-"""
-Experiment using all GALAH data.
-"""
-
-from __future__ import division # Just in case. Use Python 3.
-
-import os
-import sys
-import numpy as np
-import matplotlib
-import matplotlib.pyplot as plt
-from matplotlib.ticker import MaxNLocator
-from collections import Counter
-from scipy import linalg
-
-sys.path.insert(0, "../../")
-
-from mcfa import (mcfa, grid_search, mpl_utils, utils)
-
-import galah_dr3 as galah
-
-import pickle
-
-matplotlib.style.use(mpl_utils.mpl_style)
-
-here = os.path.dirname(os.path.realpath(__file__))
-
-prefix = os.path.basename(__file__)[:-3]
-
-def savefig(fig, suffix):
-    here = os.path.dirname(os.path.realpath(__file__))
-    filename = os.path.join(here, f"{prefix}-{suffix}")
-    fig.savefig(f"{filename}-test-mask.png", dpi=150)
-    fig.savefig(f"{filename}-test-mask.pdf", dpi=300)
-
-
-N_elements = 18
-use_galah_flags = True
-
-
-mcfa_kwds = dict(init_factors="svd", init_components="kmeans++", tol=1e-8,
-                 max_iter=10000)
-
-
-abundance_counts = galah.get_abundances_breakdown(galah.available_elements,
-                                                  use_galah_flags=use_galah_flags)
-
-xlabels = abundance_counts.keys()
-latex_labels = [r"{1}$\textrm{{{0}}}$".format(ea, "\n" if i % 2 else "") for i, ea in enumerate(xlabels)]
-x = np.arange(len(xlabels))
-y_independent = np.array(list(abundance_counts.values()))
-y_cumulative = np.zeros_like(y_independent)
-
-mask = np.ones(len(galah.data), dtype=bool)
-
-for i, xlabel in enumerate(xlabels):
-    mask *= galah._abundance_mask(xlabel, use_galah_flags)
-    y_cumulative[i] = np.sum(mask)
-
-
-def xystack(x, y, h_align="mid"):
-    xx = x.repeat(2)[1:]
-    # Now: the average x binwidth
-    xstep = np.repeat((x[1:] - x[:-1]), 2)
-    xstep = np.concatenate(([xstep[0]], xstep, [xstep[-1]]))
-    # Now: add one step at end of row.
-    xx = np.append(xx, xx.max() + xstep[-1])
-
-    # Make it possible to chenge step alignment.
-    if h_align == 'mid':
[…]

### 2019-05-20 article/experiments/exp7.py

@@ -1,463 +1,0 @@
-
-"""
-Experiment using all APOGEE DR12 data.
-"""
-
-from __future__ import division # Just in case. Use Python 3.
-
-import os
-import sys
-import numpy as np
-import pickle
-import matplotlib
-import matplotlib.pyplot as plt
-from collections import Counter
-from scipy import linalg
-from astropy.table import Table
-
-sys.path.insert(0, "../../")
-
-from mcfa import (mcfa, grid_search, mpl_utils, utils)
-
-#import galah_dr3 as galah
-
-matplotlib.style.use(mpl_utils.mpl_style)
-
-prefix = os.path.basename(__file__)[:-3]
-
-def savefig(fig, suffix):
-    filename = f"{prefix}-{suffix}.png"
-    fig.savefig(filename, dpi=150)
-
-
-mcfa_kwds = dict(init_factors="svd", init_components="kmeans++", tol=1e-2,
-                 max_iter=10000, random_seed=42)
-
-
-use_galah_flags = False
-require_flag_cannon_ok = True
-
-
-
-elements = ["Na", "Mg", "Fe", "Sc", "Ti", "Zn", "Mn", "Y", "Ca", "Ni", "Cr", "O",
-            "Si", "K", "Ba", "V", "Cu", "Al", "La", "Eu"]
-#            # "Ni", "Mn", "Cr", "Ti", "Al", "Mg", "Na", "K", "Sc", "Ca", "O", "Zn", "Ba", "Y", "Ba", "Eu"]
-
-
-barklem_elements = ['mg_h', 'al_h', 'ca_h', 'sc_h', 'ti_h', 'cr_h', 'mn_h', 'fe_h', 'co_h', 'ni_h', 'sr_h', 'y_h', 'ba_h', 'eu_h']
-missing_elements = ["Sr"]
-
-barklem_elements = [ea.split("_")[0].title() for ea in barklem_elements]
-barklem_elements = list(set(barklem_elements).difference(missing_elements))
-
-"""
-In [2]: set(elements).difference(barklem_elements)
-Out[2]: {'Cu', 'K', 'La', 'Na', 'O', 'Si', 'V', 'Zn'}
-
-In [3]: set(barklem_elements).difference(elements)
-Out[3]: {'Co', 'Sr'}
-"""
-
-elements = ["AL", "CA", "C", "FE", "K", "MG", "MN", "NA", "NI", "N", "O", "SI", "S", "TI", "V"]
-
[…]

### 2019-05-20 article/experiments/exp9-log-xfe-offset-subset.py

@@ -1,581 +1,0 @@
-
-"""
-Experiment using all GALAH data.
-"""
-
-from __future__ import division # Just in case. Use Python 3.
-
-import os
-import sys
-import pickle
-import numpy as np
-import matplotlib
-import matplotlib.pyplot as plt
-from matplotlib.ticker import MaxNLocator
-from collections import Counter
-from scipy import linalg
-
-sys.path.insert(0, "../../")
-
-from mcfa import (mcfa, grid_search, mpl_utils, utils)
-
-import galah_dr3 as galah
-
-
-matplotlib.style.use(mpl_utils.mpl_style)
-
-here = os.path.dirname(os.path.realpath(__file__))
-
-prefix = os.path.basename(__file__)[:-3]
-
-def savefig(fig, suffix):
-    here = os.path.dirname(os.path.realpath(__file__))
-    filename = os.path.join(here, f"{prefix}-{suffix}")
-    fig.savefig(f"{filename}.png", dpi=150)
-    fig.savefig(f"{filename}.pdf", dpi=300)
-
-
-N_elements = 20
-use_galah_flags = True
-
-
-mcfa_kwds = dict(init_factors="svd", init_components="kmeans++", tol=1e-5,
-                 max_iter=10000)
-
-
-abundance_counts = galah.get_abundances_breakdown(galah.available_elements,
-                                                  use_galah_flags=use_galah_flags)
-
-xlabels = abundance_counts.keys()
-latex_labels = [r"{1}$\textrm{{{0}}}$".format(ea, "\n" if i % 2 else "") for i, ea in enumerate(xlabels)]
-x = np.arange(len(xlabels))
-y_independent = np.array(list(abundance_counts.values()))
-y_cumulative = np.zeros_like(y_independent)
-
-mask = np.ones(len(galah.data), dtype=bool)
-
-for i, xlabel in enumerate(xlabels):
-    mask *= galah._abundance_mask(xlabel, use_galah_flags)
-    y_cumulative[i] = np.sum(mask)
-
-
-def xystack(x, y, h_align="mid"):
-    xx = x.repeat(2)[1:]
-    # Now: the average x binwidth
-    xstep = np.repeat((x[1:] - x[:-1]), 2)
-    xstep = np.concatenate(([xstep[0]], xstep, [xstep[-1]]))
-    # Now: add one step at end of row.
-    xx = np.append(xx, xx.max() + xstep[-1])
-
[…]

### 2019-05-20 article/experiments/exp9-log-xfe-offset.py

@@ -1,633 +1,0 @@
-
-"""
-Experiment using all GALAH data.
-"""
-
-from __future__ import division # Just in case. Use Python 3.
-
-import os
-import sys
-import pickle
-import numpy as np
-import matplotlib
-import matplotlib.pyplot as plt
-from matplotlib.ticker import MaxNLocator
-from collections import Counter
-from scipy import linalg
-
-sys.path.insert(0, "../../")
-
-from mcfa import (mcfa, grid_search, mpl_utils, utils)
-
-import galah_dr3 as galah
-
-
-matplotlib.style.use(mpl_utils.mpl_style)
-
-here = os.path.dirname(os.path.realpath(__file__))
-
-prefix = os.path.basename(__file__)[:-3]
-
-def savefig(fig, suffix):
-    here = os.path.dirname(os.path.realpath(__file__))
-    filename = os.path.join(here, f"{prefix}-{suffix}")
-    fig.savefig(f"{filename}.png", dpi=150)
-    fig.savefig(f"{filename}.pdf", dpi=300)
-
-
-N_elements = 20
-use_galah_flags = True
-
-
-mcfa_kwds = dict(init_factors="svd", init_components="kmeans++", tol=1e-5,
-                 max_iter=10000)
-
-
-abundance_counts = galah.get_abundances_breakdown(galah.available_elements,
-                                                  use_galah_flags=use_galah_flags)
-
-xlabels = abundance_counts.keys()
-latex_labels = [r"{1}$\textrm{{{0}}}$".format(ea, "\n" if i % 2 else "") for i, ea in enumerate(xlabels)]
-x = np.arange(len(xlabels))
-y_independent = np.array(list(abundance_counts.values()))
-y_cumulative = np.zeros_like(y_independent)
-
-mask = np.ones(len(galah.data), dtype=bool)
-
-for i, xlabel in enumerate(xlabels):
-    mask *= galah._abundance_mask(xlabel, use_galah_flags)
-    y_cumulative[i] = np.sum(mask)
-
-
-def xystack(x, y, h_align="mid"):
-    xx = x.repeat(2)[1:]
-    # Now: the average x binwidth
-    xstep = np.repeat((x[1:] - x[:-1]), 2)
-    xstep = np.concatenate(([xstep[0]], xstep, [xstep[-1]]))
-    # Now: add one step at end of row.
[…]

### 2019-05-20 article/experiments/galah-experiment-1.py

@@ -1,114 +1,0 @@
-
-
-
-import sys
-import numpy as np
-import matplotlib
-
-import pickle
-import operator
-
-sys.path.insert(0, "../../")
-from astropy.table import Table
-
-
-REQUIRE_ELEMENTS = ["fe", "eu", "ba"]
-EXCLUDE_LABEL_NAMES = ["o_fe"]
-
-galah = Table.read("../../catalogs/GALAH_DR2.1_catalog.fits")
-
-
-all_element_label_names = [ln for ln in galah.dtype.names \
-    if ln.endswith(("_fe", "fe_h")) and not ln.startswith(("flag_", "e_", "alpha_"))]
-
-
-# Select a subset of stars and abundances
-common_mask = (galah["flag_cannon"] == 0)
-for element in REQUIRE_ELEMENTS:
-    if element != "fe":
-        common_mask *= (galah[f"flag_{element}_fe"] == 0)
-
-
-is_ok = dict()
-for label_name in set(all_element_label_names).difference(EXCLUDE_LABEL_NAMES):
-    flag_name = "flag_cannon" if label_name == "fe_h" else f"flag_{label_name}"
-    is_ok[label_name] = sum((galah[flag_name] == 0) * common_mask)
-
-sorted_items = sorted(is_ok.items(), key=operator.itemgetter(1))[::-1]
-
-
-# Cumulative
-
-x = np.arange(len(all_element_label_names))
-y = np.zeros_like(x)
-y_cumulative = np.zeros_like(x)
-
-sorted_label_names = []
-
-for i, (k, v) in enumerate(sorted_items):
-    y[i] = v
-    sorted_label_names.append(k)
-
-    # Do cumulative count.
-    mask = (galah["flag_cannon"] == 0)
-    for k2, _ in sorted_items[:i + 1]:
-        if k2 != "fe_h":
-            mask *= (galah[f"flag_{k2}"] == 0)
-
-    y_cumulative[i] = np.sum(mask)
-
-
-D = 14
-
-# Generate data array with all things relative to [X/H]
-mask = (galah["flag_cannon"] == 0)
-for ln, v in sorted_items[:D]:
-
-    flag_label = f"flag_{ln}"
-    if flag_label in galah.dtype.names:
-        mask *= (galah[flag_label] == 0)
-
-N = sum(mask)
-print(f"There are {N} data points (D = {D})")
-
-# order by atomic number.
-
[…]

### 2019-05-20 article/experiments/galah_dr2.py

@@ -138,4 +138,23 @@ def get_abundances_wrt_h(elements, mask=None, cluster_names=None, use_galah_flag
 
     return (X_H, label_names_wrt_h)
 
+def get_abundances(elements, mask=None, cluster_names=None, use_galah_flags=False):
+
+
+    # Sort elements by atomic number.
+    indices = np.argsort([periodic_table.index(el) for el in elements])
+    sorted_elements = [elements[idx] for idx in indices]
+
+    if mask is None:
+        mask = get_abundance_mask(sorted_elements, cluster_names, 
+                                 use_galah_flags=use_galah_flags)
+
+    X_H = np.array([data[_abundance_label(el)][mask]  \
+                    for el in sorted_elements]).T
+
+    label_names_wrt_h = ["{0}_h".format(el).lower() for el in sorted_elements]
+
+    return (X_H, label_names_wrt_h)
+
+
 

### 2019-05-20 experiments/exp2.py

@@ -1,334 +1,0 @@
-
-import os
-import sys
-import numpy as np
-import matplotlib
-import matplotlib.pyplot as plt
-import pickle
-import yaml
-from scipy import linalg
-from hashlib import md5
-
-
-from matplotlib.ticker import MaxNLocator
-
-sys.path.insert(0, "../")
-
-from mcfa import (mcfa, grid_search, mpl_utils, utils)
-
-matplotlib.style.use(mpl_utils.mpl_style)
-
-here = os.path.dirname(os.path.realpath(__file__))  
-
-
-with open("config.yml") as fp:
-    config = yaml.load(fp)
-
-
-np.random.seed(config.get("random_seed", 0))
-
-unique_hash = md5((f"{config}").encode("utf-8")).hexdigest()[:5]
-
-unique_config_path = f"{unique_hash}.yml"
-if os.path.exists(unique_config_path):
-    print(f"Warning: this configuration file already exists {unique_config_path}")
-
-with open(unique_config_path, "w") as fp:
-    yaml.dump(config, fp)
-
-with open(__file__, "r") as fp:
-    code = fp.read()
-
-with open(f"{unique_hash}-{__file__}", "w") as fp:
-    fp.write(code)
-
-prefix = os.path.basename(__file__)[:-3]
-
-def savefig(fig, suffix):
-    here = os.path.dirname(os.path.realpath(__file__))
-    filename = os.path.join(here, f"{prefix}-{unique_hash}-{suffix}")
-    fig.savefig(f"{filename}.png", dpi=150)
-    fig.savefig(f"{filename}.pdf", dpi=300)
-    print(f"Created figures {filename}.png and {filename}.pdf")
-
-
-    
-
-from astropy.table import Table
-
-mcfa_kwds = dict()
-mcfa_kwds.update(config.get("mcfa_kwds", dict()))
-
-periodic_table = """H                                                  He
-                    Li Be                               B  C  N  O  F  Ne
-                    Na Mg                               Al Si P  S  Cl Ar
-                    K  Ca Sc Ti V  Cr Mn Fe Co Ni Cu Zn Ga Ge As Se Br Kr
-                    Rb Sr Y  Zr Nb Mo Tc Ru Rh Pd Ag Cd In Sn Sb Te I  Xe
-                    Cs Ba Lu Hf Ta W  Re Os Ir Pt Au Hg Tl Pb Bi Po At Rn
-                    Fr Ra Lr Rf"""
-
[…]

### 2019-05-20 experiments/exp3.py

@@ -1,372 +1,0 @@
-
-"""
-Experiment using all GALAH data.
-"""
-
-from __future__ import division # Just in case. Use Python 3.
-
-import os
-import sys
-import pickle
-import numpy as np
-import matplotlib
-import matplotlib.pyplot as plt
-import yaml
-from matplotlib.ticker import MaxNLocator
-from collections import Counter
-from scipy import linalg
-from hashlib import md5
-
-sys.path.insert(0, "../")
-
-from mcfa import (mcfa, grid_search, mpl_utils, utils)
-
-import galah_dr2 as galah
-
-
-matplotlib.style.use(mpl_utils.mpl_style)
-
-here = os.path.dirname(os.path.realpath(__file__))
-
-with open("config.yml") as fp:
-    config = yaml.load(fp)
-
-print(f"Config: {config}")
-
-np.random.seed(config["random_seed"])
-
-prefix = os.path.basename(__file__)[:-3]
-
-unique_hash = md5((f"{config}").encode("utf-8")).hexdigest()[:5]
-
-unique_config_path = f"{unique_hash}.yml"
-if os.path.exists(unique_config_path):
-    print(f"Warning: this configuration already exists: {unique_config_path}")
-
-with open(unique_config_path, "w") as fp:
-    yaml.dump(config, fp)
-
-with open(__file__, "r") as fp:
-    code = fp.read()
-
-with open(f"{unique_hash}-{__file__}", "w") as fp:
-    fp.write(code)
-
-def savefig(fig, suffix):
-    here = os.path.dirname(os.path.realpath(__file__))
-    filename = os.path.join(here, f"TEST-{prefix}-{unique_hash}-{suffix}")
-    fig.savefig(f"{filename}.png", dpi=150)
-    fig.savefig(f"{filename}.pdf", dpi=300)
-
-
-import os
-os.system("rm -f *.pkl")
-
-N_elements = 20
-use_galah_flags = config["use_galah_flags"]
-
-mcfa_kwds = dict()
-mcfa_kwds.update(config["mcfa_kwds"])
-
-elements = config[prefix]["elements"]
-elements = [el for el in elements if el not in config[prefix]["ignore_elements"]]
-
-print(elements)
-
-
-mask = galah.get_abundance_mask(elements, use_galah_flags=use_galah_flags)
-
-
-galah_cuts = config["exp3"]["galah_cuts"]
-if galah_cuts is not None:
[…]

### 2019-05-20 experiments/galah_dr2.py

@@ -1,141 +1,0 @@
-
-
-import numpy as np
-from astropy.table import Table
-from collections import OrderedDict
-from operator import itemgetter
-import os
-#data = Table.read("GALAH_iDR3_OpenClusters.fits")
-
-here = os.path.dirname(os.path.realpath(__file__))
-
-data = Table.read(os.path.join(here, "../catalogs/GALAH_DR2.1_catalog.fits"))
-
-periodic_table = """H                                                  He
-                    Li Be                               B  C  N  O  F  Ne
-                    Na Mg                               Al Si P  S  Cl Ar
-                    K  Ca Sc Ti V  Cr Mn Fe Co Ni Cu Zn Ga Ge As Se Br Kr
-                    Rb Sr Y  Zr Nb Mo Tc Ru Rh Pd Ag Cd In Sn Sb Te I  Xe
-                    Cs Ba Lu Hf Ta W  Re Os Ir Pt Au Hg Tl Pb Bi Po At Rn
-                    Fr Ra Lr Rf"""
-
-lanthanoids    =   "La Ce Pr Nd Pm Sm Eu Gd Tb Dy Ho Er Tm Yb"
-actinoids      =   "Ac Th Pa U  Np Pu Am Cm Bk Cf Es Fm Md No"
-
-periodic_table = periodic_table.replace(" Ba ", " Ba " + lanthanoids + " ") \
-                               .replace(" Ra ", " Ra " + actinoids + " ") \
-                               .split()
-
-available_elements = list(set([ea.split("_")[0].title() \
-                        for ea in data.dtype.names if (ea.endswith("_fe") or ea == "fe_h") and ea.split("_")[0].lower() not in ("e", "flag", "alpha")]))
-
-indices = np.argsort([periodic_table.index(el) for el in available_elements])
-available_elements = [available_elements[i] for i in indices]
-
-def _get_elements(data):
-    return [ln.split("_")[1] for ln in data.dtype.names \
-            if ln.startswith("A_") and ln.split("_")[1] in periodic_table]
-
-def _abundance_label(element):
-    if element.lower() == "fe":
-        return "fe_h"
-    else:
-        return f"{element.lower()}_fe"
-
-def _abundance_flag_label(element):
-    if element.lower() == "fe":
[…]

### 2019-05-20 mcfa/mpl_utils.py

@@ -159,7 +159,7 @@ def fill_between_steps(ax, x, y1, y2=0, h_align='mid', **kwargs):
 
 def plot_specific_scatter(model, y=None, scales=1, 
                           xlabel=None, xticklabels=None, ylabel=None,
-                          steps=False, fill=True, line_kwds=None, fill_kwds=None):
+                          steps=False, fill=True, line_kwds=None, fill_kwds=None, **kwargs):
 
     fig, ax = plt.subplots(figsize=(3.57, 3.57))
 
@@ -209,7 +209,7 @@ def plot_specific_scatter(model, y=None, scales=1,
         for i, tick in enumerate(ax.get_xaxis().get_major_ticks()):
             if i % 2 == 0: continue
 
-            tick.set_pad(15.)
+            tick.set_pad(kwargs.get("ticker_pad", 15.))
             tick.label1 = tick._get_text1()
 
     if xlabel is not None:
@@ -248,7 +248,11 @@ def plot_latent_space(model, X, ellipse_kwds=None, **kwargs):
 
     hard_associations = np.argmax(tau, axis=1)
 
-    fig = corner_scatter(v_mean, c=hard_associations, **kwargs)
+    kwds = dict()
+    kwds["c"] = np.argmax(tau, axis=1)
+    kwds.update(kwargs)
+
+    fig = corner_scatter(v_mean, **kwargs)
 
     xi = model.theta_[model.parameter_names.index("xi")] # latent factors, number of components.
     omega = model.theta_[model.parameter_names.index("omega")]

### 2019-05-20 article/mcfa.tex
words

@@
 present-day stellar abundances.
-Previous techniques have treated the abundances of each element as being
-statistically independent, despite the nucleosynthesis literature showing that
-most elements can be produced by more than one pathway.
+Previous techniques have treated each abundance dimension as being
+statistically independent, despite theoretical expectations that
+many elements can be produced by more than one nucleosynthetic process.
 In this work we introduce a data-driven model of nucleosynthesis 
@@
 of metal-poor stars, and those from the second \Galah\ data release. 
-We find that \todo{five} latent factors are preferred to explain the detailed 
-abundances of metal-poor stars, and \todo{eight} are preferred given \todo{X} abundances
-in \Galah\ data. In both experiments we recover latent factors that 
+We find that four latent factors are preferred to explain the detailed 
+abundances of metal-poor stars, and five latent factors are preferred given 
+16 chemical abundances in \Galah\ data. In both experiments we recover latent factors that 
 are representative of mean yields expected from the rapid- and slow-neutron
 capture process, as well as nucleosynthesis in Type II supernovae. 
-We find \todo{two} clusters are preferred in the \Galah\ data of 16 abundances
-and 1,136 stars, which are representative of the [$\alpha$/Fe]-poor and 
+However, in both experiments we also identify latent factors that link
+the production of light (e.g., K, Al) and heavy (e.g., Ba, Eu) elements.
+We find three clusters are preferred given 1,136 stars in \Galah,
+two of which are representative of the [$\alpha$/Fe]-poor and 
 [$\alpha$/Fe]-rich populations in the Milky Way disk. 
-We expect more components
-would be preferred given complete abundance inventory of a larger sample of stars.
+%The third is consistent
+%\todo{with what.}. 
 We release accompanying software that scales well with the available data,
@@
 
-%Chemical tagging seeks to identify star formation events using the present
-%day photospheric abundances of stars \citep{Freeman;Bland-Hawthorn:2002}.
-The detailed chemical abundances observable in a star's photosphere provide a
+The detailed chemical abundances that are observable in a star's photosphere provide a
 fossil record that carries with it information about where and when that star
@@
 would need to be before they could be tagged as belonging to the same
-star formation site \citep{Mitschang:2014}. However, there are also analysis
+star formation site \citep{Mitschang:2014}. However, there are analysis
 issues in understanding how precisely those chemical abundances can be measured
@@
 
-There are further chemical tagging obstacles to overcome once all the chemical 
-abundance information in a stellar spectrum can be extracted. Most notably is a 
-catalogue of precise detailed chemical abundances for a large number of stars,
+%There are further chemical tagging obstacles to overcome once all the chemical 
+%abundance information in a stellar spectrum can be extracted. Most notably is a 
+Chemical tagging experiments require a catalogue of precise chemical abundance
+measurements for a large number of stars,
 where those chemical abundances trace different nucleosynthetic pathways.
-This is the primary scientific goal of the \Galah\ survey \citep{DeSilva:2015,Martell:2017,Buder:2018},
+This is the primary goal of the \Galah\ survey \citep{DeSilva:2015,Martell:2017,Buder:2018},
 a stellar spectroscopic survey that uses the High Efficiency and Resolution 
@@
 through the slow (e.g., Ba) and rapid neutron-capture process
-(e.g., Eu). These data provide an unparalleled view on the production
-of chemical elements in the Milky Way.
+(e.g., Eu). No other spectroscopic survey provides an equivalent set of
+chemical abundances for a comparable number of stars.
+
+%These data provide an unparalleled view on the production
+%of chemical elements in the Milky Way.
 
@@
 quantities, and the effective dimensionality of stellar abundance datasets has been shown
-to be lower than the actual number of dimensions \citep{Price-Jones:2018}.
+to be lower than the actual number of dimensions \citep{Ting:2012,Price-Jones:2018}.
 Any clustering approach that treats each new elemental abundance as an 
@@
 types of stars, but quantifying the precise contribution of different processes
-to each star is an unsolved problem. For example, the [$\alpha$/Fe] `knee' in
+to each star is an unsolved problem. For example, the so-called [$\alpha$/Fe] `knee' in
 abundance ratios in the Milky Way can qualitatively be explained by 
 core-collapse supernovae being the predominant nucleosynthetic process in the
-early Milky Way before Type Ia contributions had a significant impact, but 
+early Milky Way before Type Ia supernovae had a significant contribution, but 
 efforts to date have not sought to try to explain the detailed abundances of a 
@@
 and is complicated by the differences in yield predictions that account from 
-prescriptions used in different codes.
+prescriptions used in different theoretical models.
 
@@
 two stars could have formed from the same star cluster, given some model of the
-Milky Way. In this work we focus on the former.
-
+Milky Way.
+
+In this work we focus on the former.
 Here we present a new approach to chemical tagging that allows us to identify 
@@
 dimensionality reduction tool to identify latent factors that contribute to
-the chemical abundances of stars \citep[e.g.,][]{Price-Jones:2018}, factor
+the chemical abundances of stars \citep[e.g.,][]{Ting:2012,Price-Jones:2018}, factor
 analysis cannot describe clustering in the data (or latent) space.
@@
 \noindent{}where $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$ 
-matrix of factor loads that is common to all data points, and the factor scores 
+matrix of factor loads that is common to all data points, $\NumLatentFactors$ is
+the number of latent factors, and the factor scores 
 for the $\numdata$th data point
@@
 Within the context of stellar abundances, the factor loads
-$\factorloads$ can be thought of as the \emph{typical} yields
+$\factorloads$ can be thought of as the \emph{mean} yields
 of nucleosynthetic
@@
 that we have a generative data-driven model that can 
-quantitatively account for nucleosynthetic yields, variations in
-turbulence and gas mixing, or star formation efficiency,
-and the parameters of this model can be estimated consistently
-given some data.
+quantitatively describe nucleosynthetic yields, and the
+factor scores describe the variance in turbulence and gas mixing,
+or star formation efficiency, and the parameters of this model
+can be simultaneously estimated in a self-consistent way.
 
@@
 We then initially assign each data point as belonging to one of the
-$\NumComponents$ components by drawing from a multinomial distribution
-and generating the $\NumData \times \NumComponents$ responsibility matrix $\vec\tau$,
-where $\sum_{k=1}^{K} \tau_{nk} = 1$ for any $n$th row. Given the initial
+$\NumComponents$ components using the \texttt{k-means++} algorithm \citep{Arthur:2007}.
+During the Experiments described in Section~\ref{sec:experiments} we repeat
+this initialisation procedure 25 times for every permutation of $\NumLatentFactors$ and $\NumComponents$. We run expectation-maximization (Section~\ref{sec:expectation-maximization})
+from each initialisation until convergence and adopt the model with the
+highest log likelihood as the preferred model for that permutation of $\NumLatentFactors$ and $\NumComponents$.
+
+Given the initial
 factor loads and assignments, we then estimate the relative weights
@@
 methods are available in the code associated with this \emph{Article}. 
-Throughout this work we found that the optimised model parameters appear
-to be insensitive to the choice of initialisation up until orthogonal
-rotation of the latent space.
+%Throughout this work we found that the optimised model parameters appear
+%to be insensitive to the choice of initialisation up until orthogonal
+%rotation of the latent space.
 
@@
 \citep{Dempster:1977}. With each expectation step we evaluate the log likelihood 
-given the model parameters $\vec\Psi$, and we re-calculate the $\NumData \times \NumComponents$ responsibility 
+given the model parameters $\vec\Psi$ and we calculate the $\NumData \times \NumComponents$ responsibility 
 matrix $\vec\tau$ whose entries are the posterior probability that the 
@@
 
-At the maximization step we update our estimates of the parameters,
+At the maximization step we update our estimates of the parameters $\vec\Psi$,
 conditioned on the data $\vecdata$ and the responsibility matrix $\vec\tau$.
@@
 of the log likelihood (Eq.~\ref{eq:log-likelihood}) to zero and solving for
-the parameter values\footnote{Strictly this introduces a statistical inconsistency in that we should update our parameter estimates by setting the second derivative of our chosen objective function (Eq.~\ref{eq:message-length}) to zero instead of the log likelihood, but this difference is inconsistency becomes serious at small $N$.}
+the parameter values\footnote{Strictly this introduces a statistical inconsistency in that we should update our parameter estimates by setting the second derivative of our chosen objective function (Eq.~\ref{eq:message-length}) to zero instead of the log likelihood, but this inconsistency only becomes serious with small $N$.}
 In doing so this guarantees that every updated
@@
 practice it is sufficient to run expectation-maximization from multiple
-initialisations in order to ensure that the global minima is reached.
+initialisations (as we do) in order to ensure that the global minima is reached.
 At the maximization step we first update our estimate of the relative weights 
@@
 \noindent{}where the $\vec{\Psi}\thisstep$ superscript refers to the current parameter estimates and $\vec{\Psi}\nextstep$ refers to the updated estimate for the next iteration.
-
 The updated estimates of the mean factor scores 
@@
 \end{equation} 
-\noindent{}where $Q$ is the number of model parameters, which for a mixture of
-factor analyzers is:
+\noindent{}where $Q$ is the number of model parameters:
 \begin{equation}
@@
 almost certain about. In other words, an outcome that has a probability close
-to unity has zero information content because nothing new is learned from it,
+to unity has nearly zero information content because almost nothing new is learned from it,
 whereas rarer events convey a much higher information content.
@@
 \end{equation}
-
-\printinunitsof{in}\prntlen{\textheight}
-and 
-\printinunitsof{in}\prntlen{\textwidth}
 
@@
 and   $\NumComponents_\textrm{max} = 40$ (e.g., twice $\NumLatentFactors_\textrm{true}$ and $\NumComponents_\textrm{true}$).
-For each $\NumLatentFactors$, $\NumComponents$ permutation we initialised the factor
-loads and component assignments randomly. We performed expectation-maximization 
-cycles until the relative log likelihood improved by less than $10^{-5}$.
-
+%For each $\NumLatentFactors$, $\NumComponents$ permutation we initialised the factor
+%loads and component assignments randomly. 
+We initialised the model as per Section~\ref{sec:initialisation} and performed expectation-maximization cycles until the relative log likelihood improved by less than $10^{-5}$.
+
+We recorded the \emph{negative} log likelihood, the BIC, and the message length for each permutation of $\NumLatentFactors$ and $\NumComponents$.
 These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
-Unsurprisingly the log likelihood increases with increasing numbers of latent
+Unsurprisingly the negative log likelihood increases with increasing numbers of latent
 factors $\NumLatentFactors$ and increasing numbers of components $\NumComponents$.
@@
 but the addition of just \emph{one latent factor} improvers the log likelihood more than adding
-\emph{twenty components}. Not much can be said for this example because the true data generating process 
-is drawn from a higher number of latent factors, but this does illustrate how 
+\emph{twenty components}. Not much more can be said for this example because the true data generating process 
+is drawn from a higher number of latent factors, but this toy model does illustrate how 
 clustering in high dimensional data can be better described by latent factors with 
@@
 
+\vspace{2em}
+
 \subsection{Experiment~2:\\Inferring astrophysically realistic latent factors}
@@
 abundances measured include Al (light odd-Z), Mg and Ca ($\alpha$-process), Sc, Ti, Cr, Mn, Fe, Co, and Ni (Fe-peak elements), as well as Sr, Y, Ba, and Eu ($s$/$r$-process). 
-The data vector includes [Fe/H] abundance ratios and all other elements
+The data vector $\vecdata$ includes [Fe/H] abundance ratios and all other elements
 relative to Fe (e.g., [Ti/Fe]) after subtracting the mean value in each dimension.
@@
 %and $\NumComponents$, and chose not to add any regularization term.
-We initialised
-the model parameters randomly for each permutation of $J$ and $K$, and continued
-the expectation-maximization algorithm
-until a convergence threshold of $10^{-5}$ was reached in the log likelihood. 
+%We initialised
+%the model parameters randomly for each permutation of $J$ and $K$, and continued
+%the expectation-maximization algorithm
+%until a convergence threshold of $10^{-5}$ was reached in the log likelihood. 
 In Figure~\ref{fig:exp2-gridsearch-contours} we show the negative log likelihood,
@@
 for each permutation of $J$ and $K$. The negative log likelihood continues to
-increase with increasing numbers of clusters, or increasing numbers of latent
+increase with increasing numbers of clusters, and increasing numbers of latent
 factors.
-The lowest BIC value is found for $J = 5$ and $K = 1$, indicating
-five latent factors are preferred with no clustering. The shortest message length
-is also found with no clustering,  but with four latent factors ($J = 4$, $K = 1$).
+The lowest BIC value is found for $J = 4$ and $K = 2$, indicating
+four latent factors are preferred with two clusters. The shortest message length
+is also found with four latent factors, but with no clustering ($J = 4$, $K = 1$).
 We adopt the message length as our objective function and chose $J = 4$ and $K = 1$
@@
 %loads. 
+We applied an orthogonal rotation to the estimated latent factors such that we
+could make qualitative comparisons between experiments (e.g., so we can discuss
+the closest estimated latent factors between different experiments).
 Despite the caveats in interpreting latent factors as nucleosynthetic processes
 (e.g., see Sections~\ref{sec:exp-1} and \ref{sec:discussion}), here we will
-qualitatively describe the latent factors found in our selected model.
+qualitatively describe the latent factors found in our selected model (Figure~\ref{fig:exp2-factor-loads}).
+The absolute\footnote{We discuss non-zero entries because a negative factor loads could have only negative factor scores, and their multiplication would produce a positive abundance.}
+entries of first factor load $\mathbf{L}_0$ has strong non-zero entries at Eu, Ba, Sr, and Al.
+The strong link with Eu is suggestive that this latent factor is consistent with being
+the rapid neutron-capture process.
+However, some of these factor load entries may be non-zero because we require the latent factors to
+be mutually orthogonal, and not because they truly contribute to the data. To try and
+disentangle these possibilities, we calculate the fractional contribution that factor
+load makes to the observed abundances relative to other factor loads. We define the
+fractional contribution of the $j$th factor load to the $d$th
+data dimension as:
+\begin{equation}
+% $|\mathbf{C}_{d}|$
+	\mathbf{C}_\textrm{d,j} = \frac{\sum^{N}|\factorloads_{j,d}\factorscores_{n,j}|}{\sum^J\sum^{N}|\factorloads_{j,d}\factorscores_{n,j}|} \quad .
+\end{equation}
+The fractional contributions to each element are shown in the right hand side of
+Figure~\ref{fig:exp2-factor-loads}.
+Here it is clear that despite that the first factor load $|\mathbf{L}_0|$ has support at Ba and Sr, it is not the dominant factor that contributes to those elements.
+However, we also find that Al abundances are primarily explained by this factor.
+
+The second factor load $\mathbf{L}_1$ closely represents the slow neutron capture process.
+Here the contributions to Ba, Y, and Sr are all highest for this factor load. We
+also find the Fe-peak element Ni to have the highest fractional contribution from this factor,
+but Ni has a near uniform mix of contributions from different factors.
+The third and fourth factors $\mathbf{L}_2$ show non-zero support for a mix of elements
+that are produced by $\alpha$-particle capture reactions. Mg, Ca, and Ti all have their
+highest contributions from $\mathbf{L}_3$, as do Sc and Co, whereas the iron-peak elements
+Cr, Mn, and Fe all have their largest contributions from $\mathbf{L}_2$.
+
+\begin{comment}
 The first latent factor $|\mathbf{L_0}|$ shows a non-zero contribution to most
@@
 
-\begin{comment}
 The first factor shows a positive contributions to most elements except for
@@
 	\caption{Latent factors found in Experiment~2 using the \citet{Barklem:2005}
-			 data. \todo{The panels on the left show the absolute entries for each factor load.} 
+			 data. The panels on the left show the absolute entries for each factor load. 
 			 On the right we show the absolute fractional contributions to
@@
 Figure~\ref{fig:exp2-specific-scatter}. The remaining scatter varies from as low 
-as 0.05~dex (Mg, Ca, Sc, Ti) up to 0.3~dex (C). It is perhaps not surprising that
-the largest specific scatter is found in C, since the present day carbon abundance
-is mostly affected by the evolutionary state of the star, rather than the natal
-abundance. After C, the elements with large specific scatter include Al, Fe, Ni, 
-and Ba. A large relative scatter in Al and Ba might suggest that there are other latent
+as 0.05~dex (Mg, Ca, Ti, Y) up to 0.25~dex (Ba). 
+ A large relative scatter might suggest that there are other latent
 factors that contribute to those elements, but those latent factors were not included because
 the message length preferred a fewer number of factors, or it could be limitations in
-the assumptions of our model (e.g., we assume mutually orthogonal latent factors, an assumption that is likely provably false; see Section~\ref{sec:discussion}).
+the assumptions of our model (e.g., we assume mutually orthogonal latent factors).
 %However, the scatter in 
@@
 %interpretability.
-
-Test \the\textwidth
 
@@
 	\caption{Latent factors inferred in Experiment~3 using the \Galah\
-			 data. \todo{The panels on the left show the absolute entries for each
-			 factor load.} On the right we show the absolute fractional contributions
+			 data. The panels on the left show the absolute entries for each
+			 factor load. On the right we show the absolute fractional contributions
 			 to each element, illustrating the loads that contribute most to each
@@
 stars with no erroneous flags in the following elemental abundance
-measurements: Na, Mg, Al, Si, K, Ca, Sc, Ti, Cr, Mn, Fe, Co, Ni, Y, Ba, and Eu.
+measurements: Na, Al, Si, K, Ca, Sc, Ti, V, Mn, Fe, Ni, Cu, Zn, Ba, La, and Eu.
 These elements are produced from multiple nucleosynthetic
@@
 our initial chemical experiment with \Galah\ data.
-There were 1,153 stars that met these criteria.
+There were 1,136 stars that met these criteria.
 
 %["Mg", "Al", "Ca", "Sc", "Ti", "Cr", "Mn", "Fe", "Co", "Ni", "Y", "Ba", "Eu"]
-
 \begin{comment}
@@
 Starting with $\NumLatentFactors = 1$ and $\NumComponents = 1$, we trialled each
-permutation of $\NumLatentFactors$ and $\NumComponents$ up until $\NumLatentFactors = 10$
-and $\NumComponents = 10$. For each permutation we initialised the model randomly,
-and continued expectation-maximization until a tolerance of $10^{-5}$ was achieved
-in the log likelihood. The results of this grid search are shown in Figure~\ref{fig:experiment-9-gridsearch},
+permutation of $\NumLatentFactors$ and $\NumComponents$ up until $\NumLatentFactors = 7$
+and $\NumComponents = 5$. 
+The results of this grid search are shown in Figure~\ref{fig:exp3-gridsearch},
 where we show the negative log likelihood, the BIC, and message length found for each permutation.
@@
 	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-latent-space.pdf}
-	\caption{\todo{Caption}}
+	\caption{The factor scores $\factorscores$ estimated in Experiment~3 using the \Galah\ data \citep{Buder:2018}, where each star is coloured by its inferred component.}
     \label{fig:exp3-latent-space}
@@
 	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-data-space.pdf}
-	\caption{\todo{Caption}}
+	\caption{Detailed chemical abundances from \Galah\ \citep{Buder:2018} that were used in Experiment~3. Each star is coloured by its inferred component in latent space.}
     \label{fig:exp3-data-space}
@@
 
-The latent factors inferred from this model are shown in Figure~\ref{fig:exp9-factor-loads},
+The latent factors inferred from this model are shown in Figure~\ref{fig:exp3-factor-loads},
 where we observe some similarities to the factors identified using the \citet{Barklem:2005}
-data of metal-poor stars. The first factor, $|\mathbf{L}_0|$, has non-zero contributions
-to most elements, with a peak near the Fe-peak elements of Fe, Mn, Cr, Co, and Sc. 
-Ca, an $\alpha$-capture element, and Ti, which is observed to have similar trends to
-$\alpha$-capture elements, both have non-zero contributions from $|\mathbf{L}_0|$,
-although both of them have their primary contributions by $|\mathbf{L}_1|$. $|\mathbf{L}_1|$
-contributes most to Mg, another $\alpha$-capture element, and contributes to the light
-odd-Z elements Na, Al, K. 
-The element contributions by the third latent factor, $|\mathbf{L}_2|$, are more ambiguous.
-It contributes in reasonable amounts to many elements, including light elements (Al, Si, K),
-only some Fe-peak elements (notably Sc, Ti, Mn, an Co), and to a couple of neutron-capture
-elements (Ba, Eu). However, $|\mathbf{L}_2|$ is not the dominant factor for any single element:
-it contributes only partially to many elements.
-
-The fourth factor $|\mathbf{L}_3|$ has strong contributions to Ba and Y, elements that are
-typically produced through the slow neutron capture process. However, this factor is also
-the dominant contributor to Ni, and has non-zero contributions to Mn, and a few other light
-elements. Similarly, $|\mathbf{L}_4|$ is the dominant contributor to Eu, an
-element almost entirely produced by the rapid neutron capture process, but it too has
-non-zero contributions to a handful of other elements, including Na and K.
+data of metal-poor stars. The first factor $\mathbf{L}_0$ is the dominant contributor to
+Eu and La, both heavy elements, but is also the dominant contributor for the Fe-peak
+element Sc. Unlike experiment 2 where $\mathbf{L}_0$ is the dominant contributor to Al, here $\mathbf{L}_0$ has a negligible contribution to Al.
+
+The second latent factor $\mathbf{L}_1$ here is most representative of the slow neutron capture process, with dominant contributions to Ba and Zn. This latent factor has some support at V, but with
+relatively little contribution.
+However, interestingly we find that the production of K (a light odd-Z element) is almost entirely dominated by this latent factor.
+Indeed, the fractional contribution this
+factor makes to K is higher than the fractional contributions it makes to Ba or Zn.
+
+The Fe-peak elements V, Mn, Fe, Ni, and Cu are all dominated by the third latent factor $\mathbf{L}_2$.
+Although the entries $\mathbf{L}_2$ are lower for light odd-Z elements Na and Al, we surprisingly find that this
+factor also makes the dominant contributions to Na and Al.
+For the typically used tracers of $\alpha$-element production -- Si, Ca, and Ti -- we find that all three elements have strong dominant contributions by $\mathbf{L}_3$.
+The last factor, $\mathbf{L}_4$ has non-zero support that is highest at the slow neutron
+capture element tracers Zn, Na, and La, but this latent factor is not a dominant contributor to any single element.
 
@@
 
-The two components identified by this model have nearly equal mixing proportions.
-In Figure~\ref{fig:exp9-latent-space} we show the latent space coloured by the
-inferred component association, and in Figure~\ref{fig:exp9-cluster} we project
-this to real space showing [Fe/H] abundance ratios relative to [Mg/Fe].
-It is clear that the component association in latent space does not directly
-translate to a clear separation in data space: in this [Fe/H]-[Mg/Fe] plane
-there is overlap in points between the two associations. 
-However, when we consider the clustering in latent space these components 
-are more clearly separated. In general the two components identified by the
-model could be described as high- and low-$\alpha$ element ratio stars in
-the disk of the Milky Way.
-
-With a larger sample of stars that have a complete abundance inventory we would expect to 
-identify more components, and perhaps unique star formation events. However
+In Figure~\ref{fig:exp3-latent-space} we show the inferred clustering in latent space,
+where the separation between components is arguably best seen in the splitting between $\factorscores_1$ with $\factorscores_2$ and $\factorscores_3$.
+When projected to data space (Figure~\ref{fig:exp3-data-space}) the third component (purple) is seen to have relatively higher abundance ratios of [K,Ba,Zn/Fe] at a given [Fe/H],
+and lower abundance ratios of [V/Fe].
+This is consistent with the clustering in latent space, where $\factorloads_1$ is the dominant contributor to K, Ba, and Zn, and $\factorloads_2$ is the dominant contributor to the iron-peak elements (including V).
+
+With a larger sample of stars that have a complete abundance inventory we might expect to 
+identify more components. However
 because our model requires no missing abundances in any element, we would
@@
 that can be inferred from the data.
+The specific scatter after accounting for all latent factors is smallest for Fe (0.01~dex), with
+K and Ba showing the largest scatter (0.13~dex; Figure~\ref{fig:exp3-specific-scatter}).
 
@@
 that the effective dimensionality of chemical abundance space is lower than
-the number of reported abundances \citep[e.g.,][]{Price-Jones:2018}. 
+the number of reported abundances \citep[e.g.,][]{Ting:2012, Price-Jones:2018}. 
 This is expected. Theoretical models of stellar evolution and nucleosynthesis
@@
 order of magnitude \citep{Kobayashi:2006}. From our inferences in Experiment~3
-using \Galah\ data, we find multiple latent factors have non-zero contributions
-to potassium: $\factorloads_1$, $\factorloads_2$, and $\factorloads_3$. Let us consider
-$\factorloads_3$, which is the dominant contributor of the principally $r$-process element europium. If our model confidently and
-reliably associated the production of potassium with the rapid neutron capture
-process, it could help explain the peculiar abundances of stars enhanced in
+using \Galah\ data, we find that $\factorloads_1$ is the dominant contributor to
+potassium, but this factor load is also the dominant contributor to Ba and Zn,
+elements that primarily trace the slow neutron capture process. Does this
+suggest that the production of potassium is linked to the s-process?
+If our model could confidently and
+reliably associate the production of potassium with the slow neutron capture
+process then it could help explain the peculiar abundances of stars enhanced in
 potassium and depleted in magnesium \citep{Mucciarelli:2012,Cohen:2012} -- a chemical
@@
 In this situation it is worth commenting that potassium has the largest specific
-scatter (Figure~\ref{fig:exp9-specific-scatter}), suggesting that the contributions
+scatter (Figure~\ref{fig:exp3-specific-scatter}), suggesting that the contributions
 of potassium are not as well described by the latent factor model as other
 elements.
+
+The same arguments hold for the contributions to aluminium by $\factorloads_0$ in 
+Experiment~2. This factor is the dominant contributor to Eu, an element that
+is almost entirely produced by the r-process, yet is also the dominant contributor
+to Al, a light odd-Z element. The specific scatter for Al in Experiment 2 is 
+among the highest observed (nearly 0.2~dex), which would suggest that the link
+between Al and Eu is not physically realistic and that Al is not well-described
+by the combination of latent factors in Experiment 2. Indeed, when we make use
+of a larger set of precise Al measurements in Experiment 3, we find that Al has
+almost no contribution by the 'r-process factor load` $\mathbf{L}_0$, and its
+dominant contributor is $\mathbf{L}_2$, the same factor load that dominates
+the contributions to the other odd-Z element Na.
 
@@
 nucleosynthetic processes that predominately contribute to the observed \Galah\ abundances,
-but in practice we found that the data are best explained (with high significance) with
+but in practice we found that the data are best explained with
 five latent factors. We construct a rotation matrix where the first four latent factors
@@
 
-Notwithstanding these issues, we have presented numerous experiments that demonstrate
-that a latent factor model that allows for clustering in latent space can adequately
+Notwithstanding these issues, we have presented experiments that demonstrate
+that a latent factor model which allows for clustering in latent space can adequately
 describe chemical abundance data. Our experiments with toy data (Experiment~1) reliably
@@
 The relatively small data set with complete abundances measured in \Galah\ both limits
-the number of components we can infer, and the efficacy of chemical tagging. Only two
+the number of components we can infer, and the efficacy of chemical tagging. Only three
 components were preferred from the data set of 1,136 stars, which we identified as
-being stars with low- and high-[$\alpha$/Fe] abundance ratios. A worthwhile extension
+being stars with low- and high-[$\alpha$/Fe] abundance ratios, and a third component that
+primarily differs in its abundances of K, Ba, Zn, and V. A worthwhile extension
 of this work would be to incorporate missing and noisy data in order to allow for
@@
 from chemical abundances of metal-poor stars and using data from the second \Galah\
-data release. We express caution in directly interpreting those latent
-factors, as our model explicitly requires mutually orthogonal latent factors.
-We highlight potential extensions of this work that would efficiently deal with
-missing and noisy data points.
+data release. While we express caution in physically interpreting those latent factors, our model does provide the first data-driven approach to nucleosynthesis and chemical tagging.
 

### 2019-05-20 article/mcfa.tex
stupid bugz

@@
 \newcommand{\likelihood}{\mathcal{L}}
-\newcommand{\ExperimentHash}{96ff8}
+\newcommand{\ExperimentHash}{8761b}
 
@@
 
-\author[0000-0003-0174-0564]{Andrew R. Casey}
+%[0000-0003-0174-0564]
+\author{Some combination (TBD) of: Andrew R. Casey}
 \affiliation{School of Physics \& Astronomy, 
@@
 to all stars with different scores, and clustering (e.g., chemical tagging) 
-is represented by a mixture model of multivariate gaussians
-in a lower-dimensional latent space.
+is modelled by a mixture of multivariate gaussians
+in lower-dimensional latent space.
 We use an exact method to simultaneously estimate the factor scores for
@@
 latent factors and clusters.
-We perform experiments on generated and real data, including literature chemical abundances
-of metal-poor stars, and those from the second \Galah\ data release. 
-We find that four latent factors are preferred to explain the detailed 
-abundances of metal-poor stars, and five latent factors are preferred given 
-16 chemical abundances in \Galah\ data. In both experiments we recover latent factors that 
-are representative of mean yields expected from the rapid- and slow-neutron
-capture process, as well as nucleosynthesis in Type II supernovae. 
-However, in both experiments we also identify latent factors that link
-the production of light (e.g., K, Al) and heavy (e.g., Ba, Eu) elements.
-We find three clusters are preferred given 1,136 stars in \Galah,
+We find that five latent factors are preferred to explain 18 chemical elements
+in the second \Galah\ data release.
+From these latent factors we identify the rapid- and slow-neutron capture 
+processes, as well as latent factors consistent with Fe-peak and 
+$\alpha$-element production, and a fifth latent factor where only K and Zn
+dominate. We discuss examples where the inferred latent factors deviate from 
+expected yields, which includes an  indication that Sc is associated with the r-process.
+We find three clusters are preferred given 1,072 stars in \Galah,
 two of which are representative of the [$\alpha$/Fe]-poor and 
 [$\alpha$/Fe]-rich populations in the Milky Way disk. 
-%The third is consistent
-%\todo{with what.}. 
 We release accompanying software that scales well with the available data,
 allowing for model parameters to be optimised in seconds given a fixed number
-of latent factors, components, and $\sim$$10^{7.5}$ abundance measurements.
-%We release software that scales \todo{as $\mathcal{O}(XX)$}, where $X$ is \todo{what},
-%enabling model parameter estimation in minutes given $\sim$$10^7$
-%chemical abundance measurements.
+of latent factors, components, and  $\sim3 \times 10^7$ abundance measurements.
 \end{abstract}
@@
 
-Here we describe the default initialisation of the model parameters, which
-we have found to be robust in a large set of experiments. Given a number of
-$\NumLatentFactors$ latent factors we initialise the factor load
-$\factorloads$ entries to be drawn from a uniform distribution
-$\factorloads \sim \mathcal{U}\left(-1, +1\right)$ and then ensure that
-the factor loads are orthogonal such that
-%TODO: Should we re-write this so that we just draw from the scipy special ortho group?
-\begin{equation}
-	\factorloads \factorloads\transpose = \eye \quad .
-\end{equation}
+We randomly draw a $\NumDimensions \times \NumDimensions$ matrix from a Haar distribution \citep{Haar:1933},
+which is uniform on the special orthogonal group $\textrm{SO}(n)$ and therefore guaranteed to return an orthogonal
+matrix with a determinant of unity \citep{Stewart:1980}.
+We denote the $\NumLatentFactors \times \NumDimensions$ left-most region of this
+matrix to be $\mathbf{H}$, and by taking $\factorloads_\ast$ to be the Cholesky decomposition of $\mathbf{H}\transpose \mathbf{H}$, we initialise the factor loads as
+\begin{eqnarray}
+ 	\factorloads = \mathbf{H}\left(\left(\factorloads_\ast\right)^{-1}\eye\right)
+\end{eqnarray}
+\noindent{}which ensures that $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$
+matrix of mutually orthogonal vectors and that $\factorloads\transpose \factorloads = \eye$.
 We then initially assign each data point as belonging to one of the
-$\NumComponents$ components using the \texttt{k-means++} algorithm \citep{Arthur:2007}.
-During the Experiments described in Section~\ref{sec:experiments} we repeat
-this initialisation procedure 25 times for every permutation of $\NumLatentFactors$ and $\NumComponents$. We run expectation-maximization (Section~\ref{sec:expectation-maximization})
-from each initialisation until convergence and adopt the model with the
-highest log likelihood as the preferred model for that permutation of $\NumLatentFactors$ and $\NumComponents$.
-
-Given the initial
+$\NumComponents$ components using the \texttt{k-means++} algorithm \citep{Arthur:2007}
+in the pseudo-latent space $\vecdata\factorloads$. Given the initial
 factor loads and assignments, we then estimate the relative weights
@@
 methods are available in the code associated with this \emph{Article}.
+
+Throughout this work we repeat this initialisation procedure 25 times for
+every trial of $\NumLatentFactors$ and $\NumComponents$ for a given data set. 
+We then run expectation-maximization (Section~\ref{sec:expectation-maximization})
+from each initialisation until the log likelihood improves by less than $10^{-5}$
+per step, and we adopt the model with the highest log likelihood as the preferred 
+model given that trial of $\NumLatentFactors$, $\NumComponents$, and the data. Although
+this optimisation procedure is not convex, in practice it is normally sufficient to
+initialise from many points to avoid local minima.
+
 %Throughout this work we found that the optimised model parameters appear
@@
 
-\noindent{}where $\odot$ denotes is the entry-wise (Hadamard) product. We repeat
-the expectation-maximization cycle for up to 10,000 steps or until the log likelihood
-improved by less than $10^{-5}$ between successive iterations. 
-Our software implementation allows for a positive regularisation term to be added
-along the diagonal of covariance matrices in latent space $\scorecovs$, for all experiments
-here we set this term to zero. Similarly, we assume that the data are noiseless and
+\noindent{}where $\odot$ denotes is the entry-wise (Hadamard) product. 
+%We repeat
+%the expectation-maximization cycle for up to 10,000 steps or until the log likelihood
+%improved by less than $10^{-5}$ between successive iterations. 
+%Our software implementation allows for a positive regularisation term to be added
+%along the diagonal of covariance matrices in latent space $\scorecovs$, for all experiments
+%In all experiments 
+Throughout this work we assume that the data are noiseless and
 we do not add any observed errors to the constructed covariance matrices.
@@
 number of latent factors $\NumLatentFactors$ and $\NumComponents$. In the next 
-Section we describe experiments using generated and real data, and in all cases we will
-assume that the true number of latent factors and components are not known. 
-Hence we require some criterion to decide how many latent factors, and how many
-components, are preferred given some data. An increasing number of factors
+section we describe a toy model using generated and real data, where we will assume
+that the true number of latent factors and components are not known. 
+Hence we require some heuristic to decide how many latent factors and components are 
+preferred given some data. An increasing number of factors
 and components will undoubtedly increase the log likelihood of the model given
@@
 \end{equation}
-\noindent{}We evaluate the BIC for all experiments performed in this work.
-However, while the BIC does include a penalisation term for the number of
+\noindent{}While the BIC does include a penalisation term for the number of
 parameters (which scales with $\log{N}$), it does not describe for the
@@
 
-For these reasons we also calculate the Minimum Message Length \citep[MML;][]{Wallace:2005}
-for every model as a criterion for model selection and evaluation. 
-The classically-described princple of MML is that the best explanation of the
+For these reasons we also apply the Minimum Message Length \citep[MML;][]{Wallace:2005}
+principle as a criterion for model selection and evaluation. 
+The classically-described principle of MML is that the best explanation of the
 data given a model is the one that leads to the shortest so-called two-part message~\citep{Wallace:2005}, 
@@
 
-\begin{figure*}
-	\includegraphics[width=1.0\textwidth]{experiments/exp1-data-colour.pdf}
-    \caption{The generated data for Experiment~1, where points are coloured by their
-		     true associations to each component.}
-    \label{fig:exp1-data}
-\end{figure*}
+%\begin{figure*}
+%	\includegraphics[width=1.0\textwidth]{experiments/exp1-data-colour.pdf}
+%    \caption{The generated data for Experiment~1, where points are coloured by their
+%		     true associations to each component.}
+%    \label{fig:exp1-data}
+%\end{figure*}
 
@@
 Only $\NumComponents - 1$ of the relative weights $\vec\weight$ need encoding because 
-$\sum_{\numcomponents=1}^{\NumComponents} = 1$. We assume a uniform prior on 
+$\sum_{\numcomponents=1}^{\NumComponents}\weight_\numcomponents = 1$. We assume a uniform prior on 
 individual weights,
@@
 addressed the indeterminacy between the factor loads and factor scores by
-placing a joint prior on the product of factor loads and scores, whilst
-ensuring that the factor loads themselves remain mutually orthogonal \citep{WallaceMLF}.
+placing a joint prior on the \emph{product} of factor loads and scores \citep{WallaceMLF}.
 Adopting the same prior density in our model is not practical because 
@@
 
-The Fisher information for $\factorloads$ \todo{needs to be checked.}
+The Fisher information for $\factorloads$ \todo{needs to be checked, but the effect is small.}
 Thus the message length to encode $\factorloads$ is given by
@@
 
-\subsection{Experiment 1: Toy model with generated data} \label{sec:exp-1}
+\subsection{Toy model with generated data} \label{sec:exp-1}
 
@@
 $\NumLatentFactors = 5$ ($\NumLatentFactors \times \NumDimensions$) factor loads, 
-with $\NumComponents = 20$ clusters in the latent space. The relative weights $\vec\weight$
+with $\NumComponents = 20$ clusters in the latent space. 
+We generated the factor loads in the same way that we initialise the optimisation (Section~\ref{sec:initialisation}). The relative weights $\vec\weight$
 are drawn from a multinomial distribution and the means of the clusters
@@
 each dimension $\specificvariance$ are also drawn $\specificvariance \sim \vec\Gamma\left(1\right)$.
-We randomly draw a  $\NumDimensions \times \NumDimensions$ matrix from a Haar distribution \citep{Haar:1933},
-which is uniform on the special orthogonal group $\textrm{SO}(n)$ and therefore guaranteed to return an orthogonal
-matrix with a determinant of unity \citep{Stewart:1980}.
-We denote the $\NumLatentFactors \times \NumDimensions$ left-most region of this
-matrix to be $\mathbf{H}$, and by taking $\factorloads_\ast$ to be the Cholesky decomposition of $\mathbf{H}\transpose \mathbf{H}$, we find the factor loads
-\begin{eqnarray}
- 	\factorloads = \mathbf{H}\left(\left(\factorloads_\ast\right)^{-1}\eye\right)
-\end{eqnarray}
-\noindent{}which ensures that the true factor loads is a $\NumLatentFactors \times \NumDimensions$
-matrix of mutually orthogonal vectors and that $\factorloads\transpose \factorloads = \eye$.
 The $\numdata$th data point (which belongs to the $\numcomponents$th cluster) is then
@@
 			 $\NumLatentFactors$ and number of clusters 
-			 $\NumComponents$ using the generated data in
-			 Experiment~1.  The middle panel shows 
+			 $\NumComponents$ using the generated data in our toy model.
+			 The middle panel shows 
 			 the BIC (Eq.~\ref{eq:bic}) for those 
@@
 	\includegraphics[width=\textwidth]{experiments/exp1-latent.pdf}
-    \caption{The inferred latent scores $\factorscores$ for Experiment~1
-    		 from a model with $\NumLatentFactors = 5$ latent factors and
-		 	 $\NumComponents = 20$ multivariate normal components in
-			 latent space.
+    \caption{The inferred latent scores $\factorscores$ for our toy model
+     with $\NumLatentFactors = 5$ latent factors and
+		 	 $\NumComponents = 20$ components.
              Points are coloured by the inferred component.} 
@@
 %loads and component assignments randomly. 
-We initialised the model as per Section~\ref{sec:initialisation} and performed expectation-maximization cycles until the relative log likelihood improved by less than $10^{-5}$.
-
-We recorded the \emph{negative} log likelihood, the BIC, and the message length for each permutation of $\NumLatentFactors$ and $\NumComponents$.
+%We initialised the model as per Section~\ref{sec:initialisation} and performed expectation-maximization cycles until the relative log likelihood improved by less than $10^{-5}$.
+
+We recorded the \emph{negative} log likelihood, the BIC, and the message length (omitting constant terms) for each permutation of $\NumLatentFactors$ and $\NumComponents$.
 These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
@@
 
-We show a corner plot of the generated data in Figure~\ref{fig:exp1-data},
-demonstrating the clustering (colour) and structure varying structure in data space.
-Despite this clustering in data space, it is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
-and allowing clustering in the latent space provides a better description of the (generated) data.
-Adding components to the model improves the log likelihood, even with a single latent factor,
-but the addition of just \emph{one latent factor} improvers the log likelihood more than adding
+It is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
+and clustering in the latent space provides a better description of the (generated) data.
+Adding components to the model does improves the log likelihood, even with a single latent factor,
+but the addition of just \emph{one latent factor} improves the log likelihood more so than adding
 \emph{twenty components}. Not much more can be said for this example because the true data generating process 
@@
 
-Some technical discussion is warranted before we compare our estimated model
+Some technical background is warranted before we compare our estimated model
 parameters to the true values. The latent factors in this model are only
@@
 
+\begin{comment}
 \subsection{Experiment~2:\\Inferring astrophysically realistic latent factors}
@@
 
-\begin{comment}
-With this model we sought to find a rotation matrix that would allow us to 
-identify plausible astrophysical processes in the factor loads.
-Although having to perform this rotation is comfortable, there is utility in
-trying to identify the latent factors because the factors can only be rotated by
-some valid rotation matrix. That is to say that we will not only exactly what
-we have put in: the rotated factors will simply be the closest set of factor
-loads that are consistent with the data, allowing us to identify the factor
-loads that are astrophysically plausible. We constructed
-a $J \times D$ astrophysical factor load matrix $\factorloads_\textrm{A}$ where
-all entries were zero except for the following entries:
-%\begin{itemize}
-%	\item the entry corresponding to Al in the first latent factor to represent
-%		  odd-Z element production,
-%	\item the entries corresponding to Ca and Mg in the second latent factor to
-%		  represent $\alpha$ particle production,
-%	\item the entries corresponding to Ni, Co, Fe, Mn, Cr, Ti, and Sc in the third
-%		  latent factor to represent the Fe-peak group,
-%	\item the entries corresponding to Sr, Y, and Ba in the fourth latent factor
-%		  to represent products of the slow neutron capture process, and
-%	\item the entries corresponding to Eu in the fifth latent factor to represent
-%		  products of the rapid nucleosynthesis process.
-%\end{itemize}
-\todo{\begin{itemize}
-	\item the entries corresponding to C, Al, Ca, Mg, Ni, Co, Fe, Mn, Cr, Ti, and Sc in the first
-		  latent factor to represent the products of core-collapse supernovae,
-	\item the entries corresponding to Sr, Y, Ba, and Eu in the second latent factor
-		  to represent products of the neutron capture process.
-\end{itemize}
-}
-These entries were set to $1/\sqrt{E}$ where $E$ is the number of non-zero entries
-in that latent factor, in order to ensure that $\factorloads_\textrm{A}$ is
-mutually orthogonal. We then found the valid rotation matrix $\vec{R}$ that most
-closely approximated the astrophysical loads $\factorloads_\textrm{A}$.
-In Figure~\ref{fig:exp2-factor-loads} we show the rotated factor loads found by our
-preferred model, and illustrate the relative contributions (positive or negative)
-to each element.
-\end{comment}
-
 \begin{figure}
@@
 \end{figure}
-
-\begin{comment}
-The rotated loads are not precisely the same as the target astrophysical loads
-in part because the target loads only seek to \emph{identify} and \emph{order}
-factors that can be interpreted within an astrophysical context. 
-We discuss this in more detail in Section~\ref{sec:discussion}.
-\end{comment}
 
@@
 
-\begin{comment}
-The first latent factor $|\mathbf{L_0}|$ shows a non-zero contribution to most
-elements (Figure~\ref{fig:exp2-factor-loads}). The elements Ca, Sc, Ti, Cr, and Co, mostly
-Fe-peak elements, have the largest contribution from this factor. We interpret this latent
-factor as being most similar to expected contributions from core-collapse supernovae.
-The second latent factor $|\mathbf{L_1}|$ has negligible contribution at most elements,
-with strong contributions for C, and the neutron-capture elements Sr, Y, Ba, and Eu. We
-interpret this latent factor as being most closely representative of the slow neutron
-capture process, which primarily occurs in asymptotic giant branch stars. The third latent factor has almost no contribution to Ba, but has
-non-zero contributions to other neutron-capture elements Sr, Y, and Eu. Since Eu has
-the strongest contribution by $|\mathbf{L_2}|$, we interpret this factor as being most
-representative of the rapid neutron capture process. The last latent factor $|\mathbf{L_3}|$
-has a noticeable peak at Fe and the surrounding Fe-peak elements, with small non-zero contributions at most elements. \todo{What to say of $|\mathbf{L_3}|$?}
-
-The first factor shows a positive contributions to most elements except for
-neutron-capture products, where the contribution decreases with increasing
-atomic number. The second latent factor shows strong contributions to neutron
-capture process elements -- with this latent factor being responsible for almost
-all Eu -- and shows a predominant contribution to Sc. Scandium is one of the most
-discrepant elements between observations and models of galactic chemical evolution,
-as models under-predict [Sc/Fe] ratios by a factor of ten \citep[e.g.,][]{Kobayashi:2006,Casey:2015}.
-\end{comment}
-
 % Sr, Y, Ba, and to a
@@
 \begin{figure*}
-	\includegraphics[width=\textwidth]{experiments/exp2-\ExperimentHash-latent-factors-and-contributions.pdf}
+	\includegraphics[width=\textwidth]{experiments/exp2-\ExperimentHash-latent-factors-and-contributions-with-targets.pdf}
 	\caption{Latent factors found in Experiment~2 using the \citet{Barklem:2005}
@@
 	\includegraphics[width=0.5\textwidth]{../experiments/exp2-\ExperimentHash-specific-scatter.pdf}
-	\caption{Specific scatter $\sqrt{\specificvariance}$ remaining in the \citet{Barklem:2005} data used in
+	\caption{Specific scatter (e.g., $\sqrt{\specificvariance}$) remaining in the \citet{Barklem:2005} data used in
 			 Experiment~2 after accounting for the contributions by all latent
@@
 
-\subsection{Experiment 3:\\Chemical abundances in the \Galah\ survey}
+\end{comment}
+
+\subsection{The \Galah\ survey}
 \label{sec:exp4}
@@
 \begin{figure*}
-	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-latent-factors-and-contributions.pdf}
-	\caption{Latent factors inferred in Experiment~3 using the \Galah\
-			 data. The panels on the left show the absolute entries for each
-			 factor load. On the right we show the absolute fractional contributions
-			 to each element, illustrating the loads that contribute most to each
+	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-latent-factors-and-contributions-with-targets.pdf}
+	\caption{Latent factors inferred from the \Galah\
+			 data \citep[][; thick lines]{Buder:2018}. The panels on the left show the absolute entries for each
+			 factor load, where the thin lines indicate the target latent factors. On the right we show the absolute fractional contributions
+			 to each element, ordered by the loads that contribute most to each
 			 element.}
@@
 detailed chemical abundances would be untrustworthy. We then took all
-stars with no erroneous flags in the following elemental abundance
-measurements: Na, Al, Si, K, Ca, Sc, Ti, V, Mn, Fe, Ni, Cu, Zn, Ba, La, and Eu.
-These elements are produced from multiple nucleosynthetic
-pathways, and it is this subsample that we will use for 
-our initial chemical experiment with \Galah\ data.
-There were 1,136 stars that met these criteria.
+stars with a signal-to-noise ratio exceeding 40 in the blue arm (\texttt{snr\_c1} $>$ 40), and
+stars with no erroneous flags in all of the following abundances: Na, Al, Si, K, Ca, Sc, Ti, V, Mn, Fe, Ni, Cu, Zn, Y, Ba, La, and Eu.
+%These elements are produced from multiple nucleosynthetic
+%pathways, and 
+%It is this subsample that we will use for 
+%our initial chemical experiment with \Galah\ data.
+These elements were chosen because they trace multiple nucleosynthetic pathways, and they are more commonly reported in the \Galah\ data release, allowing for a larger number of stars with a complete abundance inventory.
+There are 1,072 stars that met these criteria.
 
@@
 where we show the negative log likelihood, the BIC, and message length found for each permutation.
-The model with $J = 5$ and $K = 2$ is found to have the shortest message length.
+The behaviour of the BIC and the message length are very different here, unlike what was observed
+in our toy model. Here the BIC behaviour appears similar to the negative log likelihood in that the BIC
+prefers higher components and latent factors than the extent of the grid (e.g., $J > 7$ and $K > 5$).
+The model with $J = 5$ and $K = 3$ is found to have the shortest message length, which we take as our preferred model for these data.
 
@@
 	\includegraphics[width=0.45\textwidth]{experiments/exp3-\ExperimentHash-specific-scatter.pdf}
-	\caption{Specific scatter $\sqrt{\specificvariance}$ remaining in the \Galah\ data \citep{Buder:2018}
-			 in Experiment~3 after accounting for the contributions by all
+	\caption{Specific scatter (e.g., $\sqrt{\specificvariance}$) remaining in the \Galah\ data \citep{Buder:2018}
+			 after accounting for the contributions by all
 			 latent factors.}
@@
 	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-data-space.pdf}
-	\caption{Detailed chemical abundances from \Galah\ \citep{Buder:2018} that were used in Experiment~3. Each star is coloured by its inferred component in latent space.}
+	\caption{Detailed chemical abundances from \Galah\ \citep{Buder:2018} used in this work. Each star is coloured by its inferred component in latent space.}
     \label{fig:exp3-data-space}
@@
 
-The latent factors inferred from this model are shown in Figure~\ref{fig:exp3-factor-loads},
-where we observe some similarities to the factors identified using the \citet{Barklem:2005}
-data of metal-poor stars. The first factor $\mathbf{L}_0$ is the dominant contributor to
-Eu and La, both heavy elements, but is also the dominant contributor for the Fe-peak
-element Sc. Unlike experiment 2 where $\mathbf{L}_0$ is the dominant contributor to Al, here $\mathbf{L}_0$ has a negligible contribution to Al.
-
-The second latent factor $\mathbf{L}_1$ here is most representative of the slow neutron capture process, with dominant contributions to Ba and Zn. This latent factor has some support at V, but with
-relatively little contribution.
-However, interestingly we find that the production of K (a light odd-Z element) is almost entirely dominated by this latent factor.
-Indeed, the fractional contribution this
-factor makes to K is higher than the fractional contributions it makes to Ba or Zn.
-
-The Fe-peak elements V, Mn, Fe, Ni, and Cu are all dominated by the third latent factor $\mathbf{L}_2$.
-Although the entries $\mathbf{L}_2$ are lower for light odd-Z elements Na and Al, we surprisingly find that this
-factor also makes the dominant contributions to Na and Al.
-For the typically used tracers of $\alpha$-element production -- Si, Ca, and Ti -- we find that all three elements have strong dominant contributions by $\mathbf{L}_3$.
-The last factor, $\mathbf{L}_4$ has non-zero support that is highest at the slow neutron
-capture element tracers Zn, Na, and La, but this latent factor is not a dominant contributor to any single element.
-
-%The specific scatter after accounting for these latent factors is smallest
-%for Fe and largest for K and Y (Figure~\ref{fig:exp9-specific-scatter}).
+Earlier we described how the latent factors we estimate can only be identified up until an orthogonal rotation.
+If we want to interpret the latent factors estimated from \Galah\ data, then we must specify
+some target factor loads such that we can identify which factors are most similar to the yields we expect.
+We specified the following target latent factors where:
+\begin{itemize}
+	\item The first factor load should have non-zero entries in Eu and La (e.g., the $r$-process).
+	\item The second factor load should have non-zero entries in Ba, Y, and La (e.g., the $s$-process).
+	\item The third factor load should have non-zero entries in Fe-peak elements.
+	\item The fourth factor load should have non-zero entries in the $\alpha$-element tracers Si, Ca, and Ti.
+	\item The fifth factor load should have non-zero entries in the light odd-Z elements Na, Al, and K.
+\end{itemize}
+
+We initially set each non-zero entry in these target factor loads $\factorloads_\textrm{target}$ to $E^{-\frac{1}{2}}$, where $E$ is the number of non-zero entries in that factor load, to ensure that $\factorloads_\textrm{target}$ is mutually orthogonal.
+We solved for the $\NumLatentFactors$ unknown angles to produce a valid rotation matrix $\vec{R}$ that would make our estimated loads $\factorloads$ as close as possible to the target loads $\factorloads_\textrm{target}$, and then applied that rotation to the model.
+The target loads and (rotated) estimated loads are shown in Figure~\ref{fig:exp3-factor-loads}.
+Note that the purpose of this procedure is not to `find' the target loads that we expect, but to provide as little information needed in order to identify and describe all factor loads within an astrophysical context.
+This procedure still requires that the factors be mutually orthogonal and that they describe the data.
+For these reasons, we will not always recover the exact target loads we seek: we will only be able to identify factor loads that are closest to the target loads.
+
+This is demonstrated in Figure~\ref{fig:exp3-factor-loads}, where some estimated factor loads match
+closely to the target load (e.g., $\mathbf{L}_1$ which we identify as the s-process), and some barely
+match at all (e.g., $\mathbf{L}_4$). Here we show the absolute entry of the factor loads because even
+if an entry is negative, the corresponding factor scores could also be negative, and their product will contribute to the observed abundances.
+For this reason the factor load sign does not matter here.
+
+Some of these factor load entries may be non-zero because we require the latent factors to
+be mutually orthogonal, and not because they truly contribute to the data. To try and
+disentangle these possibilities, we calculate the fractional contribution that factor
+load makes to the observed abundances relative to other factor loads. We define the
+fractional contribution of the $j$th factor load to the $d$th
+data dimension as:
+\begin{equation}
+% $|\mathbf{C}_{d}|$
+	\mathbf{C}_\textrm{d,j} = \frac{\sum^{N}|\factorloads_{j,d}\factorscores_{n,j}|}{\sum^J\sum^{N}|\factorloads_{j,d}\factorscores_{n,j}|} \quad .
+\end{equation}
+The fractional contributions to each element are shown in the right hand side of
+Figure~\ref{fig:exp3-factor-loads}. As targeted, we identify the first factor $\mathbf{L}_0$ 
+as being most similar to the r-process, and here it is the dominant contributor to Eu and La, both heavy elements that are primarily r-process tracers.
+Surprisingly we also find that this factor load is the dominant contributor to the Fe-peak element Sc.
+The specific scatter in Sc is 0.03~dex (Figure~\ref{fig:exp3-specific-scatter}), suggesting that the Sc abundances are well-described by this latent factor model.
+
+The second latent factor $\mathbf{L}_1$ here is most representative of the slow neutron capture process, with dominant contributions to Ba and Y.
+This factor has some support at other elements, including Cu and Sc, but with very little effective contributions.
+Most Cu is contributed by $\mathbf{L}_2$, our Fe-peak latent factor.
+In fact $\mathbf{L}_2$ is the primary contributor to nearly all Fe-peak elements, with close to negligible contributions from other factors.
+The exception here is Zn, where slightly more contribution comes from $\mathbf{L}_4$ than it does from $\mathbf{L}_2$.
+Recall that while $\mathbf{L}_4$ was described as our `light odd-Z element' factor load, it is the dominant producer of both K and Zn, and secondary contributor to Ti, V, and to a lesser extent, La.
+The fourth latent factor $\mathbf{L}_3$ is better behaved.
+It is the dominant contributor to the $\alpha$-element tracers Si, Ca, and Ti, and nothing else.
+It contributes a reasonable fraction to Al and Zn, but with negligible contributions elsewhere.
+The specific scatter after accounting for these latent factors is smallest
+for Fe and largest for K (Figure~\ref{fig:exp3-specific-scatter}). The typical scatter is about 0.05~dex.
 
 In Figure~\ref{fig:exp3-latent-space} we show the inferred clustering in latent space,
-where the separation between components is arguably best seen in the splitting between $\factorscores_1$ with $\factorscores_2$ and $\factorscores_3$.
+where the separation between components is arguably best seen in the splitting between $\factorscores_4$ with respect to $\factorscores_1$ or $\factorscores_2$.
 When projected to data space (Figure~\ref{fig:exp3-data-space}) the third component (purple) is seen to have relatively higher abundance ratios of [K,Ba,Zn/Fe] at a given [Fe/H],
 and lower abundance ratios of [V/Fe].
-This is consistent with the clustering in latent space, where $\factorloads_1$ is the dominant contributor to K, Ba, and Zn, and $\factorloads_2$ is the dominant contributor to the iron-peak elements (including V).
-
-With a larger sample of stars that have a complete abundance inventory we might expect to 
-identify more components. However
-because our model requires no missing abundances in any element, we would
-need to reduce the number of elemental abundances we include in order to increase
-the sample of stars. And by doing so we restrict the number of latent factors
-that can be inferred from the data.
-The specific scatter after accounting for all latent factors is smallest for Fe (0.01~dex), with
-K and Ba showing the largest scatter (0.13~dex; Figure~\ref{fig:exp3-specific-scatter}).
+This is consistent with the clustering in latent space, where $\factorloads_4$ is the dominant contributor to K and Zn, and $\factorloads_{1,2}$ are the dominant contributor to Ba and V, respectively.
 
@@
 
-Literature studies testing the limitations chemical tagging have shown
-that the effective dimensionality of chemical abundance space is lower than
-the number of reported abundances \citep[e.g.,][]{Ting:2012, Price-Jones:2018}. 
-This is expected. Theoretical models of stellar evolution and nucleosynthesis
-would suggest that there are of order at most ten important nucleosynthetic 
-contributors that are responsible for the production of the $\sim$20--30
-chemical abundances typically reported. As a consequence we can expect that 
-the observed chemical abundances are correlated in some way (e.g., by some
-linear combination of a fewer number of nucleosynthetic pathways).  However, 
-most approaches to chemical tagging to date have assumed that the observed 
-chemical abundances are independent of each other. This problem cannot be
-resolved by simply lowering the number of abundances used for chemical tagging (e.g.,
-down to something near the effective dimensionality), either, because those
-chemical abundances are still produced by multiple pathways.
-
-Here we have introduced a model to simultaneously account for the lower
+%Literature studies testing the limitations chemical tagging have shown
+%that the effective dimensionality of chemical abundance space is lower than
+%the number of reported abundances \citep[e.g.,][]{Ting:2012, Price-Jones:2018}. 
+%This is expected. Theoretical models of stellar evolution and nucleosynthesis
+%would suggest that there are of order at most ten important nucleosynthetic 
+%contributors that are responsible for the production of the $\sim$20--30
+%chemical abundances typically reported. As a consequence we can expect that 
+%the observed chemical abundances are correlated in some way (e.g., by some
+%linear combination of a fewer number of nucleosynthetic pathways).  However, 
+%most approaches to chemical tagging to date have assumed that the observed 
+%chemical abundances are independent of each other. This problem cannot be
+%resolved by simply lowering the number of abundances used for chemical tagging (e.g.,
+%down to something near the effective dimensionality), either, because those
+%chemical abundances are still produced by multiple pathways.
+
+We have introduced a model to simultaneously account for the lower
 effective dimensionality of chemical abundance space, and perform clustering
@@
 estimate the latent factors that contribute to all stars, and cluster those 
-stars by their relative contributions from each factors. The results are
+stars by their relative contributions from each factor. The results are
 encouraging in that we find latent factors that are representative of the
@@
 We require latent factors to be mutually orthogonal to in order to resolve
-the indeterminacy in our model. This suggests an astrophysical context where 
+an indeterminacy. This suggests an astrophysical context where 
 the mean nucleosynthetic yields (integrated over all stellar masses and star
@@
 yields (or latent factors) should be mutually orthogonal.
-One could represent them by a hierarchical data-driven model where the yields contribute as a function of stellar mass, 
-metallicity, and other factors), but in principle to resolve the indeterminacy
-the mean of these yields would require mutual orthogonality.
+In principle one could represent the latent factors using a hierarchical data-driven model where the yields contribute as a function of stellar mass, 
+metallicity, and other factors, but in principle to resolve the indeterminacy
+the mean of these yields would still require mutual orthogonality in this model.
 Introducing a constraint on the factor scores that resolves this indeterminacy and allows for more flexible latent factors would be a worthy extension to this work.
@@
 known sources of potassium production in the Milky Way, galactic chemical evolution
-models under-predict the level of potassium in the Milky Way by more than an
-order of magnitude \citep{Kobayashi:2006}. From our inferences in Experiment~3
-using \Galah\ data, we find that $\factorloads_1$ is the dominant contributor to
-potassium, but this factor load is also the dominant contributor to Ba and Zn,
-elements that primarily trace the slow neutron capture process. Does this
-suggest that the production of potassium is linked to the s-process?
+models under-predict the level of K in the Milky Way by more than an
+order of magnitude \citep{Kobayashi:2006}. From our inferences 
+using \Galah\ data, we find that $\factorloads_4$ is the dominant contributor to
+potassium, but this factor load is also the dominant contributor to Zn, 
+an element that is typically associated with the iron-peak group. That factor load
+is also a reasonable contributor to Ti, V, and to a lesser extent, La. Does this
+suggest production of K is linked to the production of much heavier nuclei?
 If our model could confidently and
-reliably associate the production of potassium with the slow neutron capture
-process then it could help explain the peculiar abundances of stars enhanced in
-potassium and depleted in magnesium \citep{Mucciarelli:2012,Cohen:2012} -- a chemical
+reliably associate the production of K with other elements or sites then it could help explain the peculiar abundances of stars enhanced in
+K and depleted in Mg \citep{Mucciarelli:2012,Cohen:2012} -- a chemical
 abundance pattern that currently lacks explanation \citep{Kemp:2018}.
-However, is the potassium contribution that we infer physically realistic,
+However, is the K contribution that we infer physically realistic,
 or is it a consequence of requiring that the latent factors are mutually
@@
 part why caution is warranted when trying to interpret latent factor models.
-In this situation it is worth commenting that potassium has the largest specific
+In this situation it is worth commenting that K has the largest specific
 scatter (Figure~\ref{fig:exp3-specific-scatter}), suggesting that the contributions
-of potassium are not as well described by the latent factor model as other
+of K are not as well described by the latent factor model as other
 elements.
 
-The same arguments hold for the contributions to aluminium by $\factorloads_0$ in 
-Experiment~2. This factor is the dominant contributor to Eu, an element that
-is almost entirely produced by the r-process, yet is also the dominant contributor
-to Al, a light odd-Z element. The specific scatter for Al in Experiment 2 is 
-among the highest observed (nearly 0.2~dex), which would suggest that the link
-between Al and Eu is not physically realistic and that Al is not well-described
-by the combination of latent factors in Experiment 2. Indeed, when we make use
-of a larger set of precise Al measurements in Experiment 3, we find that Al has
-almost no contribution by the 'r-process factor load` $\mathbf{L}_0$, and its
-dominant contributor is $\mathbf{L}_2$, the same factor load that dominates
-the contributions to the other odd-Z element Na.
+A similar argument could be made for Sc, where $\mathbf{L}_0$ -- a factor load we
+identify as the r-process -- is the dominant contributor. Indeed, the second largest
+contributor to Sc is the factor load we identify as the s-process. Sc is under-produced in 
+galactic chemical evolution models relative to observations \citep{Kobayashi:2006,Casey:2015}.
+Based on this work, is the production of Sc linked to the production of heavy nuclei? 
+Unlike K, the specific scatter in Sc is remarkably low: just 0.03~dex, among the
+best-described elements after Ti and Fe (0.01~dex). This would suggest that the
+latent factor model is a very good description for the production of Sc, but it
+does not prove that it is \emph{the} description for the production of Sc.
 
@@
 five latent factors. We construct a rotation matrix where the first four latent factors
-describe the mean nucleosynthetic processes we expect to find. What of the fifth latent
+describe the nucleosynthetic processes we expect to find. What of the fifth latent
 factor? We can constrain the possible values of the fifth latent factor conditioned on
@@
 
-Notwithstanding these issues, we have presented experiments that demonstrate
-that a latent factor model which allows for clustering in latent space can adequately
-describe chemical abundance data. Our experiments with toy data (Experiment~1) reliably
-finds the true number of latent factors and cluster components when we adopt an information-theoretic approach to the objective function (e.g., BIC or MML), and correctly estimates the latent
-model parameters. When applied to chemical abundances measured from metal-poor stars 
-(Experiment~2) we find that four latent factors are preferred by the data, and
-those latent factors qualitatively re-appear in Experiment~3 using \Galah\ data. 
-That did not have to be the case: the latent factors between different experiments (data sets)
-might be entirely different, and those factors did not have to qualitatively match our
-expectations of nucleosynthetic yields. Indeed, the inferred factors -- even after a
+Notwithstanding these issues, we have shown that a latent factor model which allows for clustering in latent space can adequately
+describe chemical abundance data. 
+We find that five latent factors from \Galah\ data, and
+those latent factors can qualitatively be described within the context of astrophysical yields. 
+That did not have to be the case: the mutually orthogonal latent factors could be entirely 
+different from our expectations such that they did not have to match our expectations of
+nucleosynthetic yields. Indeed, the inferred factors -- even after a
 valid rotation -- could have made no astrophysical sense whatsoever. For this reason it
-is encouraging that there is some interpretability in the latent factors, even though 
-caution is warranted.
+is encouraging that there is some interpretability in the latent factors.
+Indeed, in the elements where we find surprising associations (e.g., Sc with what we identify as the r-process, K and Zn linked by a common process), these are elements where galactic chemical evolution models are most discrepant from observations, even after accounting for systematic errors in abundance measurements.
 
@@
 the number of components we can infer, and the efficacy of chemical tagging. Only three
-components were preferred from the data set of 1,136 stars, which we identified as
+components were preferred from the data set of 1,072 stars, which we identified as
 being stars with low- and high-[$\alpha$/Fe] abundance ratios, and a third component that
@@
 common to all stars (e.g., nucleosynthetic yields). Experiments with generated data
-demonstrate that MML is a useful principle for identifying the preferred number of
-latent factors and components. Experiments with real data reveal latent factors
-that are qualitatively similar to expected nucleosynthetic yields (e.g., products
-from the $s$-process, $r$-process, et cetera). Similar latent factors emerge 
-from chemical abundances of metal-poor stars and using data from the second \Galah\
-data release. While we express caution in physically interpreting those latent factors, our model does provide the first data-driven approach to nucleosynthesis and chemical tagging.
+demonstrate that MML is a useful principle for selecting the appropriate number of
+latent factors and components. Experiments with \Galah\ data reveal latent factors
+that are qualitatively and quantitatively similar to expected nucleosynthetic yields (e.g., products
+from the $s$-process, $r$-process, et cetera). Interestingly we find that deviations from
+expected yields occur in elements where observations and galactic chemical evolution models
+are most discrepant (e.g., K, Sc). While we advise caution in directly interpreting
+those latent factors as being nucleosynthetic yields, our
+our model does provide the first data-driven approach to nucleosynthesis and chemical tagging.
+
+With a larger sample of stars that have a complete abundance inventory we would expect to 
+identify more latent factors and/or components.
+Alternatively, a useful extension of this work would be to allow for missing and/or noisy data, such that realistic data sets of chemical abundances can be used to perform chemical tagging experiments on a larger scale.
 
@@
     \package{scipy}\,\,\citep{scipy},\,\,
-    \package{Stan}\,\,\citep{stan},\,\,
+%    \package{Stan}\,\,\citep{stan},\,\,
     \package{Jupyter Notebooks}\,\,\citep{jupyter-notebooks}\,\,

### 2019-05-20 article/experiments/exp2.py

@@ -12,7 +12,7 @@ from hashlib import md5
 
 from matplotlib.ticker import MaxNLocator
 
-sys.path.insert(0, "../")
+sys.path.insert(0, "../../")
 
 from mcfa import (mcfa, grid_search, mpl_utils, utils)
 
@@ -74,7 +74,7 @@ periodic_table = periodic_table.replace(" Ba ", " Ba " + lanthanoids + " ") \
                                .replace(" Ra ", " Ra " + actinoids + " ") \
                                .split()
 
-t = Table.read(os.path.join(here, "../catalogs/barklem_t3.fits"))
+t = Table.read(os.path.join(here, "../../catalogs/barklem_t3.fits"))
 
 elements = np.unique(t["El"])
 stars = np.unique(t["Name"])

### 2019-05-20 article/experiments/galah_dr2.py

@@ -9,7 +9,7 @@ import os
 
 here = os.path.dirname(os.path.realpath(__file__))
 
-data = Table.read(os.path.join(here, "../catalogs/GALAH_DR2.1_catalog.fits"))
+data = Table.read(os.path.join(here, "../../catalogs/GALAH_DR2.1_catalog.fits"))
 
 periodic_table = """H                                                  He
                     Li Be                               B  C  N  O  F  Ne
@@ -103,6 +103,7 @@ def suggest_abundances_to_include(mask, elements, use_galah_flags=False,
 
     return OrderedDict(sorted(updated.items(), key=itemgetter(1), reverse=True))
 
+
 def get_abundances_wrt_h(elements, mask=None, cluster_names=None, use_galah_flags=False):
 
     # Prepare data array.
@@ -131,12 +132,21 @@ def get_abundances_wrt_h(elements, mask=None, cluster_names=None, use_galah_flag
         mask = get_abundance_mask(sorted_elements, cluster_names, 
                                  use_galah_flags=use_galah_flags)
 
-    X_H = np.array([data[_abundance_label(el)][mask] - asplund_2009[el] \
-                    for el in sorted_elements]).T
+    #X_H = np.array([data[_abundance_label(el)][mask] - asplund_2009[el] \
+    #                for el in sorted_elements]).T
+    X_H = np.zeros((sum(mask), len(sorted_elements)), dtype=float)
+    for i, element in enumerate(sorted_elements):
+
+        if element.lower() == "fe":
+            X_H[:, i] = data[_abundance_label(element)][mask]
+        else:
+            X_H[:, i] = data[_abundance_label(element)][mask] + data[_abundance_label("Fe")][mask]
 
     label_names_wrt_h = ["{0}_h".format(el).lower() for el in sorted_elements]
 
     return (X_H, label_names_wrt_h)
+
+
 
 def get_abundances(elements, mask=None, cluster_names=None, use_galah_flags=False):
 

### 2019-05-20 mcfa/mpl_utils.py

@@ -656,7 +656,7 @@ def plot_fractional_positive_factor_contributions(model, Y, label_names=None, co
 
 
 def plot_factor_loads_and_contributions(model, Y, label_names=None, colors=None, line_kwds=None,
-                                        absolute_only=True, target_loads=None):
+                                        absolute_only=True, target_loads=None, **kwargs):
 
     L = model.theta_[model.parameter_names.index("A")]
     S = model.factor_scores(Y)[1]
@@ -720,12 +720,21 @@ def plot_factor_loads_and_contributions(model, Y, label_names=None, colors=None,
                 ax.set_xticks(x.astype(int))
                 ax.set_xticklabels(latex_label_names)
 
+
+
         else:
             ax.set_xticks([])
 
         ax.set_xlim(x[0] - 0.5, x[-1] + 0.5)
 
     # On last figure, show visualisation.
+
+    for i, tick in enumerate(ax.get_xaxis().get_major_ticks()):
+        if i % 2 == 0: continue
+
+        tick.set_pad(kwargs.get("ticker_pad", 20.))
+        tick.label1 = tick._get_text1()
+
 
     F = np.abs(f)/np.atleast_2d(np.sum(np.abs(f), axis=1)).T
     indices = np.argsort(1.0/F, axis=1)
@@ -752,7 +761,7 @@ def plot_factor_loads_and_contributions(model, Y, label_names=None, colors=None,
 
     ax.set_frame_on(False)
 
-    ax.set_xlabel(r"${|\mathbf{C}_\textrm{d}|} / {\sum_{j}|\mathbf{C}_\textrm{j,d}|}$")
+    #ax.set_xlabel(r"${|\mathbf{C}_\textrm{d}|} / {\sum_{j}|\mathbf{C}_\textrm{j,d}|}$")
 
     fig.tight_layout()
 

### 2019-05-20 article/mcfa.tex
minor

@@
 allowing for model parameters to be optimised in seconds given a fixed number
-of latent factors, components, and  $\sim3 \times 10^7$ abundance measurements.
+of latent factors, components, and $\sim10^6$ abundance measurements.
 \end{abstract}
@@
 with \Galah\ data do not include lithium or carbon abundances because the
-photospheric varies throughout a star's lifetime (e.g., see Experiment 2). This is true
+photospheric values will vary throughout a star's lifetime \citep[e.g.,][]{Casey:2016b,Casey:2019}. This is true
 to a small degree for many elements \citep[e.g.,][]{Dotter:2017},
@@
 detailed chemical abundances would be untrustworthy. We then took all
-stars with a signal-to-noise ratio exceeding 40 in the blue arm (\texttt{snr\_c1} $>$ 40), and
+stars with a signal-to-noise ratio exceeding 40 in the blue arm (\texttt{snr\_c1 > 40}), and
 stars with no erroneous flags in all of the following abundances: Na, Al, Si, K, Ca, Sc, Ti, V, Mn, Fe, Ni, Cu, Zn, Y, Ba, La, and Eu.
@@
 
-% This paper was built from reference XXX on github repo YYY
+% TODO:
+%\vspace{2em}
+%This \emph{Article} was compiled from the \texttt{git} commit \todo{\texttt{HASH}}.
+%The analysis, results, and figures presented in this \emph{Article} can be reproduced in full 
+%using the following commands in a modern terminal:\vspace{1em}\\
+%\texttt{git clone https://github.com/andycasey/mcfa}\\
+%\texttt{cd mcfa/}\\
+%\todo{\texttt{sh article/build.sh}}
 
@@
 
+\clearpage
+
 \software{
-	\package{Astropy}\,\,\citep{astropy:v1,astropy:v2},\,\,
-    \package{IPython}\,\,\citep{ipython},\,\,
-    \package{matplotlib}\,\,\citep{mpl},\,\,
-    \package{numpy}\,\,\citep{numpy},\,\,
-    \package{scipy}\,\,\citep{scipy},\,\,
+	\package{Astropy} \citep{astropy:v1,astropy:v2},\,\,
+    \package{IPython} \citep{ipython},\,\,
+    \package{matplotlib} \citep{mpl},\,\,
+    \package{numpy} \citep{numpy},\,\,
+    \package{scipy} \citep{scipy},\,\,
 %    \package{Stan}\,\,\citep{stan},\,\,
-    \package{Jupyter Notebooks}\,\,\citep{jupyter-notebooks}\,\,
+    \package{Jupyter Notebooks} \citep{jupyter-notebooks}
 }    % TOPCAT

### 2019-05-24 article/mcfa.tex
minor

@@
 issues in understanding how precisely those chemical abundances can be measured
-\citep{Bovy:2016}, and how chemically similar stars can be which did not form 
+\citep{Bovy:2016}, and how chemically similar stars can be that did not form 
 together \citep[doppleg\"angers;][]{Ness:2018}.
@@
 needed: astronomers have not yet developed unbiased estimators of chemical
-abundances that saturate the Cram\'er-Rao bound.
+abundances that saturate the Cram\'er-Rao bound \citep{Cramer:1946,Rao:1945}.
 
@@
 Given these data and the most favourable assumptions in chemical tagging 
--- that star clusters are truely chemically homogenous, and we can measure 
-those abundances with infinite precision, and those abundances are 
+-- that star clusters are truely chemically homogenous, that we can measure 
+those abundances with infinite precision, and that those abundances are 
 differentiable between star clusters -- then chemical
@@
 chemical tagging thus far have assumed that the data dimensions are independent. That is to say
-that adding a dimension of [Ni/H] provides entirely new, independent information
+that adding a dimension of say [Ni/H] provides entirely new, independent information
 that could not have been predicted from other elemental abundances.
@@
 quantities, and the effective dimensionality of stellar abundance datasets has been shown
-to be lower than the actual number of dimensions \citep{Ting:2012,Price-Jones:2018}.
+to be lower than the actual number of abundance dimensions \citep{Ting:2012,Price-Jones:2018}.
 Any clustering approach that treats each new elemental abundance as an 
@@
 %Alternative approaches are perhaps more challenging, as i
-It is not trivial to estimate
-the nucleosynthetic yields that have contributed to the chemical abundances of each star
-with great confidence. There are
+It is not trivial to confidently estimate
+the nucleosynthetic yields that have contributed to the chemical abundances of each star. There are
 qualitative statements that can be made for large numbers of stars, or particular
@@
 core-collapse supernovae being the predominant nucleosynthetic process in the
-early Milky Way before Type Ia supernovae had a significant contribution, but 
+early Milky Way before Type Ia supernovae made a significant contribution, but 
 efforts to date have not sought to try to explain the detailed abundances of a 
@@
 Similarly, clustering techniques applied to chemical abundances to date 
-\citep[e.g.,][]{Hogg:2016} do not account for the lower dimensionality of the
-data.
+\citep[e.g.,][]{Hogg:2016} do not account for the lower effective dimensionality in
+elemental abundances.
 
@@
 \end{equation}
-\noindent{}are drawn from the $\numcomponents$th multivariate normal distribution.
+\noindent{}are drawn from\footnote{For clarifying nomenclature across disciplines, the terminology e$z \sim \mathcal{N}(0, 1)$ indicates that the variable $z$ is drawn from a standard normal distribution.} the $\numcomponents$th multivariate normal distribution.
 The mean vector $\vec\mu$ describes the mean datum in each dimension.
@@
 
-The Fisher information for $\factorloads$ \todo{needs to be checked, but the effect is small.}
+The Fisher information for $\factorloads$ \todo{needs to be double checked.}
 Thus the message length to encode $\factorloads$ is given by
@@
 $\NumDimensions = 15$ dimensions. We adopted a latent dimensional space of 
-$\NumLatentFactors = 5$ ($\NumLatentFactors \times \NumDimensions$) factor loads, 
+$\NumLatentFactors = 5$ factor loads such that the vector $\factorloads$ has
+shape $\NumLatentFactors \times \NumDimensions$,
 with $\NumComponents = 20$ clusters in the latent space. 
@@
 until $\NumLatentFactors_\textrm{max} = 10$
-and   $\NumComponents_\textrm{max} = 40$ (e.g., twice $\NumLatentFactors_\textrm{true}$ and $\NumComponents_\textrm{true}$).
+and   $\NumComponents_\textrm{max} = 40$ (e.g., twice the true values of $\NumLatentFactors_\textrm{true}$ and $\NumComponents_\textrm{true}$).
 %For each $\NumLatentFactors$, $\NumComponents$ permutation we initialised the factor
@@
 			 data \citep[][; thick lines]{Buder:2018}. The panels on the left show the absolute entries for each
-			 factor load, where the thin lines indicate the target latent factors. On the right we show the absolute fractional contributions
+			 factor load, where the thin lines indicate the target latent factors (see Section~\ref{sec:exp4}). On the right we show the absolute fractional contributions
 			 to each element, ordered by the loads that contribute most to each
@@
 prefers higher components and latent factors than the extent of the grid (e.g., $J > 7$ and $K > 5$).
-The model with $J = 5$ and $K = 3$ is found to have the shortest message length, which we take as our preferred model for these data.
+The model with five latent factors and three components ($J = 5$, $K = 3$) is found to have the shortest message length, which we take as our preferred model for these data.
 
@@
 	\item The second factor load should have non-zero entries in Ba, Y, and La (e.g., the $s$-process).
-	\item The third factor load should have non-zero entries in Fe-peak elements.
+	\item The third factor load should have non-zero entries in Fe-peak elements Sc, V, Mn, Fe, Ni, Cu, and Zn.
 	\item The fourth factor load should have non-zero entries in the $\alpha$-element tracers Si, Ca, and Ti.
@@
 
-We require latent factors to be mutually orthogonal to in order to resolve
+We require latent factors to be mutually orthogonal in order to resolve
 an indeterminacy. This suggests an astrophysical context where 
@@
 K and depleted in Mg \citep{Mucciarelli:2012,Cohen:2012} -- a chemical
-abundance pattern that currently lacks explanation \citep{Kemp:2018}.
+abundance pattern that currently lacks explanation \citep{Iliadis:2016,Kemp:2018}.
+In the \citep{Cohen:2012} sample their high [K/Fe] stars also tend
+to be high in [Zn/Fe] (for which $\factorloads_4$ also dominates), but there
+are also numerous abundance correlations present.
 However, is the K contribution that we infer physically realistic,
@@
 latent factors up until this rotation. We have sought to address this by constructing
-rotation matrices where the entries each latent factor correspond to our expectations
+rotation matrices where the entries for each latent factor correspond to our expectations
 from astrophysical processes (whilst remaining orthogonal), but here we are limited
-by what astrophysical processes we are \emph{expected} to find within the constraint
+by what astrophysical processes we are \emph{expecting} to find within the constraint
 of being mutually orthogonal.
@@
 describe chemical abundance data. 
-We find that five latent factors from \Galah\ data, and
+We find five latent factors from \Galah\ data, and
 those latent factors can qualitatively be described within the context of astrophysical yields. 
@@
 being stars with low- and high-[$\alpha$/Fe] abundance ratios, and a third component that
-primarily differs in its abundances of K, Ba, Zn, and V. A worthwhile extension
+primarily differs in its abundances of K, Ba, Zn, and V (e.g., given some [Fe/H] and [$\alpha$/Fe]). A worthwhile extension
 of this work would be to incorporate missing and noisy data in order to allow for

### 2019-06-20 article/mcfa.tex
things

@@
 \newcommand{\todo}[1]{\textcolor{red}{#1}}
+\newcommand{\consider}[1]{\textcolor{blue}{#1}}
 
@@
 \newcommand{\likelihood}{\mathcal{L}}
-\newcommand{\ExperimentHash}{8761b}
+\newcommand{\ExperimentHash}{89dab}
+
+% NEED: Final model with # of loads and # of clusters so we can:
+% 		[ ] Update figure of loads (exp10-comparison)
+% 		[ ] Make a clustering plot in abundances
+% 		[ ] Check to see if any chemical clusters correspond to known clusters.
+% 		[ ] Update discussion w.r.t. final model loads, etc.
+% 		[ ] Update conclusion w.r.t. final model loads/clusters, etc, and missing data.
+% 		[ ] Update abstract w.r.t. final model loads/clusters.
+% 		[ ] Upload PDF to Galah survey website
+%		[ ] Send email to Galah team for comment
 
@@
 
-%[0000-0003-0174-0564]
-\author{Some combination (TBD) of: Andrew R. Casey}
+\author[0000-0003-0174-0564]{Andrew R. Casey}
 \affiliation{School of Physics \& Astronomy, 
@@
 
-\author{John Lattanzio}
+\author[0000-0003-2952-859X]{John C. Lattanzio}
 \affiliation{School of Physics \& Astronomy, 
@@
 
-\author{Aldeida Aleti}
+\author[0000-0002-1716-690X]{Aldeida Aleti}
 \affiliation{Faculty of Information Technology, 
@@
 
-\author{David Dowe}
+\author[0000-0002-0583-5918]{David Dowe}
 \affiliation{Faculty of Information Technology, 
@@
 latent factors and clusters.
-We find that five latent factors are preferred to explain 18 chemical elements
-in the second \Galah\ data release.
+\consider{We find that five latent factors are preferred to explain 18 chemical elements
+in the second \Galah\ data release.}
 From these latent factors we identify the rapid- and slow-neutron capture 
 processes, as well as latent factors consistent with Fe-peak and 
-$\alpha$-element production, and a fifth latent factor where only K and Zn
-dominate. We discuss examples where the inferred latent factors deviate from 
-expected yields, which includes an  indication that Sc is associated with the r-process.
-We find three clusters are preferred given 1,072 stars in \Galah,
+$\alpha$-element production, and a \consider{fifth latent factor where only K and Zn
+dominate}. We discuss examples where the inferred latent factors deviate from 
+expected yields, which includes an \consider{indication that Sc is associated with the r-process}.
+\consider{We find three clusters are preferred given 1,072 stars in \Galah,
 two of which are representative of the [$\alpha$/Fe]-poor and 
-[$\alpha$/Fe]-rich populations in the Milky Way disk. 
+[$\alpha$/Fe]-rich populations in the Milky Way disk.}
 We release accompanying software that scales well with the available data,
 allowing for model parameters to be optimised in seconds given a fixed number
-of latent factors, components, and $\sim10^6$ abundance measurements.
+of latent factors, components, and $\sim10^7$ abundance measurements.
 \end{abstract}
 
-\keywords{methods: statistical}
+\keywords{Bayesian statistics (1900), Chemical abundances (224), Galaxy chemical evolution (580)}
 
@@
 early Milky Way before Type Ia supernovae made a significant contribution, but 
-efforts to date have not sought to try to explain the detailed abundances of a 
-single star as a contribution of yields from different systems (e.g., 20\% of the
-Ba was produced by 3 M$_\odot$ asymptotic giant branch stars). This is in part 
+efforts to date have not sought to try to explain the detailed abundances of 
+stars as a contribution of yields from different systems \citep[however see][]{West:2013}.
+%e.g., 20\% of the
+%Ba was produced by 3 M$_\odot$ asymptotic giant branch stars).
+This is in part 
 because of the challenging and degenerate nature of the problem as described, 
@@
 \end{equation}
-\noindent{}are drawn from\footnote{For clarifying nomenclature across disciplines, the terminology e$z \sim \mathcal{N}(0, 1)$ indicates that the variable $z$ is drawn from a standard normal distribution.} the $\numcomponents$th multivariate normal distribution.
+\noindent{}are drawn from\footnote{For clarifying nomenclature across disciplines, the terminology $z \sim \mathcal{N}(0, 1)$ indicates that the $z$ variable \emph{is drawn from} a standard normal distribution.} the $\numcomponents$th multivariate normal distribution.
 The mean vector $\vec\mu$ describes the mean datum in each dimension.
@@
 
+\vspace{3em}
 %Throughout this work we found that the optimised model parameters appear
@@
 %not add any observed errors to the constructed covariance matrices.
+
+\subsection{Missing data}
+
+The Expectation-Maximization procedure as described requires that there be no 
+missing data entries in order to update our estimates of the model parameters.
+In practice, however, there will often be abundance measurements that are missing
+for some subset of stars. There are many potential reasons for this, including
+astrophysical explanations (e.g., an absorption line was not present above the noise),
+observational limitations (e.g., the signal-to-noise ratio was too low, or 
+contamination by a cosmic ray), or various other reasons.
+
+In this work we will assume that any missing data measurements are
+missing at random. The missing data points can then be treated as unknown
+parameters that must be solved for (and updated) at each iteration. Initially
+we impute zeros for missing data entries, and at each iteration we update
+this imputed value with our estimate of what the missing data values
+are, given the current model parameters. Similarly, with each update we inflate
+our estimates of the specific variances $\vec\psi$ based on the fraction of missing
+data points in each dimension
+\begin{equation}
+\specificvariance\nextstep = \specificvariance\nextstep\left(\frac{N}{N-\vec{M}}\right)
+\end{equation}
+\noindent{}where $\vec{M}$ is a vector that contains the number of missing data entries in
+all $D$ dimensions.
+In Section \ref{sec:toy-model-missing-data} we show with a toy model that the latent
+factor loads and scores can be reliably estimated even in the presence of high fractions
+of missing data, with varying fractions of missing data in each dimension.
 
@@
 	I(\scoremeans,\scorecovs) &=& -\sum_{\numcomponents=1}^{\NumComponents}\log{p(\scoremeans_k,\scorecovs_k)} + \frac{1}{2}\sum_{\numcomponents=1}^{\NumComponents}\log{|\mathcal{F}(\scoremeans_k,\scorecovs_k)|} \nonumber \\
-	&=& \frac{1}{2}\sum_{\numcomponents=1}^{\NumComponents}\log\left[(\NumData\weight_k)^{\frac{1}{2}\NumLatentFactors(\NumLatentFactors+3)}2^{-\NumData}|\scorecovs_k|^{-(\NumData + 2)}\right] \nonumber \\
-	&& \cdots \,\, -\sum_{\numcomponents=1}^{\NumComponents}\log{|\scorecovs_k|}^{\frac{1}{2}(\NumLatentFactors + 1)} \\
+%	&=& \frac{1}{2}\sum_{\numcomponents=1}^{\NumComponents}\log\left[(\NumData\weight_k)^{\frac{1}{2}\NumLatentFactors(\NumLatentFactors+3)}2^{-\NumData}|\scorecovs_k|^{-(\NumData + 2)}\right] \nonumber \\
+%	&& \cdots \,\, -\sum_{\numcomponents=1}^{\NumComponents}\log{|\scorecovs_k|}^{\frac{1}{2}(\NumLatentFactors + 1)} \\
 I(\scoremeans,\scorecovs) &=& \frac{1}{4}\NumLatentFactors(\NumLatentFactors+3)\sum_{\numcomponents=1}^\NumComponents\log{\NumData\weight_k} - \frac{KD}{2}\log{2} \nonumber \\ 
@@
 
-The Fisher information for $\factorloads$ \todo{needs to be double checked.}
+%The Fisher information for $\factorloads$ \todo{needs to be double checked.}
 Thus the message length to encode $\factorloads$ is given by
@@
 I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} \nonumber \\
-	&& \cdots \,\, + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\
-	&& \cdots \todo{+\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \quad . \label{eq:prior_L}
+	&& \cdots \,\, + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\ \label{eq:prior_L}%\nonumber \\
+%	&& \cdots \todo{+\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \quad . \label{eq:prior_L}
 \end{eqnarray}
+
+%\todo{As...}% TODO: 
+\vspace{10em}
 
@@
   && \cdots -\frac{1}{2}\left(D-J-1\right)\log|\factorloads\transpose\factorloads| + \textrm{Tr}\left(\textrm{Cov}\left(\factorloads\transpose\right)^{-1}\factorloads\transpose\factorloads\right) - \left(\NumLatentFactors + \frac{3}{2}\right)\sum_{\numcomponents=1}^\NumComponents\log|\scorecovs_\numcomponents|  -\log\Gamma\left(\NumComponents\right) - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\
-&& \cdots +\frac{Q}{2}\log\kappa_q +\frac{1}{2}\left[\NumLatentFactors(\NumDimensions + 2) + \NumComponents(2-\NumData)\right]\log{2} + \todo{\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \quad .
-\label{eq:message-length}
+&& \cdots +\frac{Q}{2}\log\kappa_q +\frac{1}{2}\left[\NumLatentFactors(\NumDimensions + 2) + \NumComponents(2-\NumData)\right]\log{2}  \quad . \nonumber \label{eq:message-length}
  \end{eqnarray}
 \end{widetext}
+%\vspace{10em}\clearpage
+% \todo{\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}}
 
@@
 % + Q/2\log\kappa(q)
-
-%\clearpage
-
-\section{Experiments} \label{sec:experiments}
-
-\subsection{Toy model with generated data} \label{sec:exp-1}
-
-We generated a data set with ${\NumData = 1}$00,000 data points, each with
-$\NumDimensions = 15$ dimensions. We adopted a latent dimensional space of 
-$\NumLatentFactors = 5$ factor loads such that the vector $\factorloads$ has
-shape $\NumLatentFactors \times \NumDimensions$,
-with $\NumComponents = 20$ clusters in the latent space. 
-We generated the factor loads in the same way that we initialise the optimisation (Section~\ref{sec:initialisation}). The relative weights $\vec\weight$
-are drawn from a multinomial distribution and the means of the clusters
-in factor scores $\scoremeans$ are drawn from a standard normal
-distribution.
-The off-diagonal entries in the covariance matrices in factor scores $\scorecovs$ are drawn from a gamma distribution $\scorecovs_{\numcomponents,i,i} \sim \vec\Gamma\left(1\right)$.
-The variance in
-each dimension $\specificvariance$ are also drawn $\specificvariance \sim \vec\Gamma\left(1\right)$.
-The $\numdata$th data point (which belongs to the $\numcomponents$th cluster) is then
-generated by drawing $\factorscores_{\numdata} \sim \mathcal{N}(\scoremeans_\numcomponents,\scorecovs_\numcomponents)$, projecting by the factor loads $\factorloads$, and adding variance $\specificvariance$.
 
@@
 
-\begin{figure*}
-	\includegraphics[width=\textwidth]{experiments/exp1-latent.pdf}
-    \caption{The inferred latent scores $\factorscores$ for our toy model
-     with $\NumLatentFactors = 5$ latent factors and
-		 	 $\NumComponents = 20$ components.
-             Points are coloured by the inferred component.} 
-    \label{fig:exp1-latent}
+%\clearpage
+
+\section{Experiments} \label{sec:experiments}
+
+\subsection{A toy model} \label{sec:exp-1}
+
+Here we introduce a toy model where we use generated data to verify that
+we recover the true data-generating parameters given some data, and to
+ensure that the method is yielding consistent results.
+We generated a data set with ${\NumData = 1}$00,000 data points, each with
+$\NumDimensions = 15$ dimensions. We adopted a latent dimensional space of 
+$\NumLatentFactors = 5$ factor loads such that the vector $\factorloads$ has
+shape $\NumLatentFactors \times \NumDimensions$,
+with $\NumComponents = 20$ clusters in the latent space. 
+We generated the factor loads in the same way that we initialise the optimisation (Section~\ref{sec:initialisation}). The relative weights $\vec\weight$
+are drawn from a multinomial distribution and the means of the clusters
+in factor scores $\scoremeans$ are drawn from a standard normal
+distribution.
+The off-diagonal entries in the covariance matrices in factor scores $\scorecovs$ are drawn from a gamma distribution $\scorecovs_{\numcomponents,i,i} \sim \vec\Gamma\left(1\right)$.
+The variance in
+each dimension $\specificvariance$ are also drawn $\specificvariance \sim \vec\Gamma\left(1\right)$.
+The $\numdata$th data point (which belongs to the $\numcomponents$th cluster) is then
+generated by drawing $\factorscores_{\numdata} \sim \mathcal{N}(\scoremeans_\numcomponents,\scorecovs_\numcomponents)$, projecting by the factor loads $\factorloads$, and adding variance $\specificvariance$.
+
+%\begin{figure*}
+%	\includegraphics[width=\textwidth]{experiments/exp1-latent.pdf}
+%    \caption{The inferred latent scores $\factorscores$ for our toy model
+%     with $\NumLatentFactors = 5$ latent factors and
+%		 	 $\NumComponents = 20$ components.
+%             Points are coloured by the inferred component.} 
+%    \label{fig:exp1-latent}
+%\end{figure*}
+
+\begin{figure*}[t!]
+	\includegraphics[width=\textwidth]{experiments/exp1-compare-all.pdf}
+    \caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
+    		 and specific variances $\specificvariance$ (right) compared to the 
+		 	 true data generating values
+		 	 for Experiment~1. The agreement is excellent.}
+    \label{fig:exp1-compare}
 \end{figure*}
@@
 and $\NumComponents = 20$, identical to the true values.
-
-\begin{figure*}
-	\includegraphics[width=\textwidth]{experiments/exp1-compare-all.pdf}
-    \caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
-    		 and specific variances $\specificvariance$ (right) compared to the 
-		 	 true data generating values
-		 	 for Experiment~1. The agreement is excellent.}
-    \label{fig:exp1-compare}
-\end{figure*}
-
 It is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
@@
 clustering in the lower dimensional latent space.
-The inferred clustering in latent space is shown in Figure~\ref{fig:exp1-latent},
-where each point is coloured by the inferred associated component.
+%The inferred clustering in latent space is shown in Figure~\ref{fig:exp1-latent},
+%where each point is coloured by the inferred associated component.
 
@@
 
-\begin{figure*}
+\begin{figure*}[t!]
 	\includegraphics[width=\textwidth]{experiments/exp2-\ExperimentHash-latent-factors-and-contributions-with-targets.pdf}
@@
 
+\subsection{A toy model with data missing at random}
+\label{sec:toy-model-missing-data}
+
+Here we repeat the toy model used in the previous Section, but we discard an increasing fraction
+of the data and evaluate the model's performance in the presence
+of incomplete data. We considered missing data fractions from 1\% to
+40\%. In each case we treated the model parameters as unknown, assumed
+the missing data points were missing at random, and initialised the
+model as per Section~\ref{sec:exp-1}.
+
+\begin{figure*}[t!]
+	\includegraphics[width=\textwidth]{experiments/exp-missing-data-not-at-random-inflated-40percent.png}
+    \caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
+    		 and specific variances $\specificvariance$ (right) compared to the 
+		 	 true data generating values
+		 	 for Experiment~2. Here 40\% of the data are missing at random.
+			 The agreement remains excellent,
+			 despite the large fraction of missing data entries.}
+    \label{fig:exp1-missing-data}
+\end{figure*}
+
+In Figure~\ref{fig:exp1-missing-data} we show the results of this
+experiment for our worst considered case, where 40\% of the data
+entries are randomly discarded. We find that despite the high fraction
+of missing entries, our estimates of the model parameters remain unbiased
+in this example using a toy model. The corrections to our estimates of the
+specific variances are sufficient, in that the specific variance in each
+dimension is not systematically under-estimated from the true values, 
+despite 40\% of the data entries to be missing.
+
 \subsection{The \Galah\ survey}
@@
 up to 23 chemical abundances reported for 342,682 stars.
-Some abundances are not reported for a variety of 
-reasons related to astrophysics or the adopted analyses
-(e.g., a star is too hot to estimate the abundance for a given
-electronic transition, or the signal-to-noise of the spectrum is too low).
-Our method as described cannot efficiently handle missing data 
-entries for a given star, and as such we are forced to exclude stars that
-do not have a full complement of abundances. However, we also
-chose to exclude some elemental abundances from our chemical
-tagging experiments. For example, our experiments
-with \Galah\ data do not include lithium or carbon abundances because the
+%Some abundances are not reported for a variety of 
+%reasons related to astrophysics or the adopted analyses
+%(e.g., a star is too hot to estimate the abundance for a given
+%electronic transition, or the signal-to-noise of the spectrum is too low).
+%Our method as described cannot efficiently handle missing data 
+%entries for a given star, and as such we are forced to exclude stars that
+%do not have a full complement of abundances. However, we also
+%chose to exclude some elemental abundances from our chemical
+In this example we chose to restrict ourselves to stars with a
+complete set of abundance measurements for a subset of those 23 elements
+(i.e., no missing data entries).
+%tagging experiments. 
+%For example, our experiments
+For example, here we will exclude
+%The element subset should not 
+%with \Galah\ data do not include 
+lithium and carbon abundances because the
 photospheric values will vary throughout a star's lifetime \citep[e.g.,][]{Casey:2016b,Casey:2019}. This is true
@@
 	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-latent-factors-and-contributions-with-targets.pdf}
-	\caption{Latent factors inferred from the \Galah\
-			 data \citep[][; thick lines]{Buder:2018}. The panels on the left show the absolute entries for each
+	\caption{Latent factors inferred from 1,072 stars in \Galah\
+			 \citep[][thick lines]{Buder:2018} with 17 abundance measurements. Left panels show the absolute entries for each
 			 factor load, where the thin lines indicate the target latent factors (see Section~\ref{sec:exp4}). On the right we show the absolute fractional contributions
-			 to each element, ordered by the loads that contribute most to each
-			 element.}
+			 to each element, ordered by the loads that contribute most.}
     \label{fig:exp3-factor-loads}
@@
 This is consistent with the clustering in latent space, where $\factorloads_4$ is the dominant contributor to K and Zn, and $\factorloads_{1,2}$ are the dominant contributor to Ba and V, respectively.
+
+\subsection{\Galah\ survey data with an increasing number of stars with missing data entries} \label{sec:exp10}
+Here we extend our experiment in Section~\ref{sec:exp4} to progressively
+include more stars, even though those stars have some abundance measurements
+missing. We assume those measurements are missing at random, and discuss the
+impact of that assumption (and others) in Section~\ref{sec:discussion}.
+Specifically we started with the same subset of 1,072 stars in Section~\ref{sec:exp4}
+and added a random set of stars that met our criteria of \texttt{flag\_cannon = 0} and
+\texttt{snr\_c1 > 40}.
+
+We initially added 100 stars to give a sample of $N =$ 1,172, then repeated
+the grid search for the number of latent factors and components, and recorded
+the model with the lowest message length. We then repeated this
+procedure using 1,000 stars ($N =$ 2,072), then again with 10,000 ($N =$ 11,072),
+and finally using all 99,174 stars that met the criteria of \texttt{flag\_cannon = 0} and
+\texttt{snr\_c1 > 40} to give a total sample size of $N =$ 100,246 stars.
+
+\consider{For sample sizes up to $N = 11,072$ we find that the number of preferred latent
+factors remains at $J = 5$. The factor loads inferred from each subset of stars
+is shown in Figure~\ref{fig:exp10-comparison-loads}, illustrating that the
+first\footnote{`First' has no concept here in terms of factor load ordering, but
+for the purposes of comparing inferred loads from different data sets we have ordered
+the loads to be as close to those inferred in Section~\ref{sec:exp4}.} inferred factor
+loads remain largely unchanged despite the increasing sample size, with an increasing
+fraction of missing data entries. It is not until the data set reaches $N = 11,072$ 
+stars where we find that $J = 8$ factor loads are preferred by the data. Again, the
+first five latent factors are extremely similar to those inferred from just 1,072
+stars with complete abundances, except for the notable difference in Zn loads in
+factor load $|\mathbf{L}_4|$, which disappears for $N = 11,072$ and re-appears
+in the sixth factor load $|\mathbf{L}_5|$ -- indicative that the additional data
+prefers the fourth latent factor to split into an additional latent factor where
+Zn contributions dominate.}
+
+\begin{figure*}[t!]
+	\includegraphics[width=0.95\textwidth]{experiments/exp10-comparison.pdf}
+	\caption{Latent factors found in Section~\ref{sec:exp10} using different subsets
+	         of \Galah\ data \citep{Buder:2018}. The thin line shows the result from
+	         Section~\ref{sec:exp4} with $N=1,072$ stars with 17 abundances and no
+	         missing data. Dotted line shows that data set complemented with 100 random
+	         stars that have some abundances missing. Dot-and-dashed line shows the
+	         original 1,072 stars with 1,000 random stars with some missing abundances,
+	         and the thick line shows the original 1,072 with 10,000 stars with
+	         missing abundances, where eight latent factors were preferred over five.}
+    \label{fig:exp10-comparison}
+\end{figure*}
+
+\todo{If we use ALL data, how many? Then just comment on clustering in the N = XXX,XXX data}
+
+% Something about the clustering in the large example.
 
@@
 primarily differs in its abundances of K, Ba, Zn, and V (e.g., given some [Fe/H] and [$\alpha$/Fe]). A worthwhile extension
-of this work would be to incorporate missing and noisy data in order to allow for
+of this work would be to incorporate missing and noisy data, either by imputing the
+missing data values or by additive variance in the constructed
+covariance matrices. In doing so would allow for
 blind chemical tagging on an industrial scale, allowing for the latent factors to
@@
 
-\clearpage
-
 \software{

### 2019-06-20 article/experiments/exp3.py

@@ -194,8 +194,9 @@ latex_label_names = [r"$\textrm{{{0}}}$".format(ea.split("_")[0].title()) for ea
 
 # Draw unrotated.
 J_max = config["max_n_latent_factors_for_colormap"]
-cmap = mpl_utils.discrete_cmap(J_max, base_cmap="Spectral_r")
-colors = [cmap(j) for j in range(J_max)][::-1]
+J_max = 8
+cmap = mpl_utils.discrete_cmap(J_max, base_cmap="Spectral")
+colors = [cmap(j) for j in range(J_max)]#[::-1]
 
 
 A_est = model.theta_[model.parameter_names.index("A")]
@@ -225,23 +226,35 @@ if config["correct_A_astrophysical"]:
     A_astrophysical = A_astrophysical @ linalg.solve(AL, np.eye(model.n_latent_factors))
 
 
-R, p_opt, cov, *_ = utils.find_rotation_matrix(A_astrophysical, A_est, 
-                                               full_output=True)
 
-R_opt = utils.exact_rotation_matrix(A_astrophysical, A_est, 
-                                    p0=np.random.uniform(-np.pi, np.pi, model.n_latent_factors**2))
+max_n_rotations = 3
 
-# WTF check R_opt.
-AL = linalg.cholesky(R_opt.T @ R_opt)
-R_opt2 = R_opt @ linalg.solve(AL, np.eye(model.n_latent_factors))
+for each in range(max_n_rotations):
 
-chi1 = np.sum(np.abs(A_est @ R - A_astrophysical))
-chi2 = np.sum(np.abs(A_est @ R_opt2 - A_astrophysical))
+    A_est = model.theta_[model.parameter_names.index("A")]
 
-R = R_opt2 if chi2 < chi1 else R
+    R, p_opt, cov, *_ = utils.find_rotation_matrix(A_astrophysical, A_est, 
+                                                   full_output=True)
 
-# Now make it a valid rotation matrix.
-model.rotate(R, X=X, ensure_valid_rotation=True)
+    R_opt = utils.exact_rotation_matrix(A_astrophysical, A_est, 
+                                        p0=np.random.uniform(-np.pi, np.pi, model.n_latent_factors**2))
+
+    # WTF check R_opt.
+    AL = linalg.cholesky(R_opt.T @ R_opt)
+    R_opt2 = R_opt @ linalg.solve(AL, np.eye(model.n_latent_factors))
+
[…]

### 2019-06-20 article/experiments/plot-exp10-comparison.py

@@ -5,8 +5,13 @@ Plot results from exp10
 
 import os
 import pickle
+import sys
 import matplotlib.pyplot as plt
 import yaml
+import numpy as np
+from scipy import linalg
+
+sys.path.insert(0, "../../")
 
 from mcfa import (mcfa, grid_search, mpl_utils, utils)
 
@@ -14,7 +19,7 @@ np.random.seed(0)
 
 
 # Get original model.
-with open("8761b-exp3-model.pkl", "rb") as fp:
+with open("89dab-exp3-model.pkl", "rb") as fp:
     original_model = pickle.load(fp)
 
 
@@ -25,6 +30,7 @@ comparison_model_paths = [
     "exp10-models-size-10000.pkl",
 ]
 
+# 89dab is for all intents and purposes identical to 8761b
 with open("8761b.yml") as fp:
     config = yaml.load(fp)
 
@@ -127,6 +133,7 @@ plot_factor_loads_kwds = dict(colors=colors, separate_axes=True, lw=1, xlabel=r"
                               xticklabels=[r"$\textrm{{{0}}}$".format(ea.split("_")[0].title()) for ea in label_names])
 
 fig = mpl_utils.plot_factor_loads(viz_load(_fill_factors(A_original, J_max)), 
+                                  figsize=(6.75, 9.50),
                                   **plot_factor_loads_kwds)
 plot_factor_loads_kwds.update(fig=fig)
 
@@ -215,11 +222,11 @@ for i, comparison_model_path in enumerate(comparison_model_paths):
 if ABS_ONLY:
     for i, ax in enumerate(fig.axes):
         ax.set_ylim(0, ax.get_ylim()[1])
-        ax.set_ylabel(r"$|\mathbf{{L}}_{{{0}}}|$".format(i))
+        ax.set_ylabel(r"$|\mathbf{{L}}_{{{0}}}|$".format(i + 1))
 
 
-fig.set_figwidth(6.75)
-fig.set_figheight(9.50)
+#fig.set_figwidth(6.75)
+#fig.set_figheight(9.50)
 
 fig.tight_layout()
 

### 2019-06-20 mcfa/mpl_utils.py

@@ -706,12 +706,12 @@ def plot_factor_loads_and_contributions(model, Y, label_names=None, colors=None,
         ax.set_xlim(0, D)
 
         if absolute_only:
-            ax.set_ylabel(r"$|\mathbf{{L}}_{{{0}}}|$".format(j))
+            ax.set_ylabel(r"$|\mathbf{{L}}_{{{0}}}|$".format(j + 1))
             ax.set_ylim(-0.1, 1.1)
             ax.set_yticks([0, 1])
 
         else:
-            ax.set_ylabel(r"$\mathbf{{L}}_{{{0}}}$".format(j))
+            ax.set_ylabel(r"$\mathbf{{L}}_{{{0}}}$".format(j + 1))
             ax.set_ylim(-1.1, 1.1)
             ax.set_yticks([-1, 0, 1])
 

### 2019-06-21 article/mcfa.tex
as posted to galah wiki

@@
 
-\title{A data-driven model of nucleosynthesis with chemical tagging in latent space}
+\title{A data-driven model of nucleosynthesis with chemical tagging in a lower-dimensional latent space}
 
@@
 is modelled by a mixture of multivariate gaussians
-in lower-dimensional latent space.
+in a lower-dimensional latent space.
 We use an exact method to simultaneously estimate the factor scores for
 each star, the partial assignment of each star
-to each cluster, and the latent factors common to all stars.
+to each cluster, and the latent factors common to all stars, even in the
+presence of missing data entries.
 We use an information-theoretic Bayesian principle to estimate the number of
 latent factors and clusters.
-\consider{We find that five latent factors are preferred to explain 18 chemical elements
-in the second \Galah\ data release.}
-From these latent factors we identify the rapid- and slow-neutron capture 
+Using the second \Galah\ data release we find that five latent factors are
+preferred to explain $N =$ 1,072 stars with 18 chemical abundances. 
+We identify the rapid- and slow-neutron capture 
 processes, as well as latent factors consistent with Fe-peak and 
-$\alpha$-element production, and a \consider{fifth latent factor where only K and Zn
-dominate}. We discuss examples where the inferred latent factors deviate from 
-expected yields, which includes an \consider{indication that Sc is associated with the r-process}.
-\consider{We find three clusters are preferred given 1,072 stars in \Galah,
-two of which are representative of the [$\alpha$/Fe]-poor and 
-[$\alpha$/Fe]-rich populations in the Milky Way disk.}
+$\alpha$-element production, and another where only K and Zn dominate.
+When we consider $N \sim$ 100,000 stars with missing abundances we find 
+another seven factors, as well as 16 components in latent space.
+Despite these components showing separation in chemistry that is explained
+through different yield contributions, none show significant structure in 
+their positions or motions. We argue that more data, and joint priors 
+on cluster membership that are constrained by dynamical models, are necessary
+to realise chemical tagging at a galactic-scale.
 We release accompanying software that scales well with the available data,
@@
 chemical tagging thus far have assumed that the data dimensions are independent. That is to say
-that adding a dimension of say [Ni/H] provides entirely new, independent information
+that adding a dimension of say [Ni/H] provides independent information
 that could not have been predicted from other elemental abundances.
@@
 but its rarity would be represented by associated factor
-scores being zero for most stars and thus have negligible contribution
+scores being zero for most stars and thus have no contribution
 to the observed abundances. In practice the factor loads can only be 
@@
 quantitatively describe nucleosynthetic yields, and the
-factor scores describe the variance in turbulence and gas mixing,
+factor scores can explain the variance in turbulence and gas mixing,
 or star formation efficiency, and the parameters of this model
-can be simultaneously estimated in a self-consistent way.
+can be simultaneously estimated in a self-consistent way with
+a single scalar-justified objective function.
 
@@
 and $\weight_\numcomponents$ describes the relative weighting of the $\numcomponents$th
-component in latent space and $\sum\vec\weight = 1$.
+component in latent space and $\sum\vec\weight^K = 1$.
 The log likelihood is then given by
@@
 
-We randomly draw a $\NumDimensions \times \NumDimensions$ matrix from a Haar distribution \citep{Haar:1933},
+Here we describe how the model parameters are initialised.\footnote{This describes the default initialisation approach. Other approaches are available in the accompanying software.}
+To initialise the factor loads $\factorloads$ we start by randomly drawing a $\NumDimensions \times \NumDimensions$ matrix from a Haar distribution \citep{Haar:1933},
 which is uniform on the special orthogonal group $\textrm{SO}(n)$ and therefore guaranteed to return an orthogonal
@@
 \noindent{}which ensures that $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$
-matrix of mutually orthogonal vectors and that $\factorloads\transpose \factorloads = \eye$.
+matrix of mutually orthogonal vectors.
+%and that it is a valid rotation matrix
+%$\factorloads\transpose \factorloads = \eye$.
 We then initially assign each data point as belonging to one of the
@@
 dimension as the variance in each data dimension. Other initialisation 
-methods are available in the code associated with this \emph{Article}.
+methods for the latent factors include singular value decomposition \citep{Golub:1970}
+or generating random noise with orthogonal constraints, and random assignment
+is an alternative method that is available for initialising assignments.
 
@@
 
-\vspace{3em}
 %Throughout this work we found that the optimised model parameters appear
@@
 of the log likelihood (Eq.~\ref{eq:log-likelihood}) to zero and solving for
-the parameter values\footnote{Strictly this introduces a statistical inconsistency in that we should update our parameter estimates by setting the second derivative of our chosen objective function (Eq.~\ref{eq:message-length}) to zero instead of the log likelihood, but this inconsistency only becomes serious with small $N$.}
+the parameter values.\footnote{Strictly this introduces a statistical inconsistency in that we should update our parameter estimates by setting the second derivative of our information-theoretic objective function (Eq.~\ref{eq:message-length}) to zero instead of the log likelihood, but this inconsistency only becomes serious with small $N$ -- precisely the opposite situation of chemical tagging!}
 In doing so this guarantees that every updated
@@
 
-Finally, the updated estimate of the specific variances $\specificvariance\nextstep$ is given
+Finally, the updated estimate of the specific variances $\specificvariance\nextstep$ are given
 by
@@
 
-The Expectation-Maximization procedure as described requires that there be no 
-missing data entries in order to update our estimates of the model parameters.
+The expectation-maximization procedure as described requires that there be no 
+missing data entries in order to update our estimates of the responsibility matrix $\vec\tau$
+and the model parameters $\vec\Psi$.
 In practice, however, there will often be abundance measurements that are missing
@@
 observational limitations (e.g., the signal-to-noise ratio was too low, or 
-contamination by a cosmic ray), or various other reasons.
+contamination by a cosmic ray), or various other reasons that cannot be inferred
+from the available information.
 
@@
 parameters that must be solved for (and updated) at each iteration. Initially
-we impute zeros for missing data entries, and at each iteration we update
-this imputed value with our estimate of what the missing data values
-are, given the current model parameters. Similarly, with each update we inflate
-our estimates of the specific variances $\vec\psi$ based on the fraction of missing
+we impute zeros for missing data entries in $\vec\data$, and at each iteration we update
+these imputed valuee with our estimate of what the missing data values
+are given the current model parameters. This ensures that the log-likelihood increases
+with each iteration. Similarly, with each update we inflate
+our estimates of the specific variances based on the fraction of missing
 data points in each dimension
 \begin{equation}
-\specificvariance\nextstep = \specificvariance\nextstep\left(\frac{N}{N-\vec{M}}\right)
-\end{equation}
-\noindent{}where $\vec{M}$ is a vector that contains the number of missing data entries in
-all $D$ dimensions.
+\specificvariance_d\nextstep = \specificvariance_d\nextstep\left(\frac{N}{N-M}\right)
+\end{equation}
+\noindent{}where $M$ is a the number of missing data entries in the $d$th dimension.
 In Section \ref{sec:toy-model-missing-data} we show with a toy model that the latent
 factor loads and scores can be reliably estimated even in the presence of high fractions
-of missing data, with varying fractions of missing data in each dimension.
+of missing data (e.g., 40\%), conditioned on our assumption that the data are missing
+at random.
 
@@
 number of latent factors $\NumLatentFactors$ and $\NumComponents$. In the next 
-section we describe a toy model using generated and real data, where we will assume
+Section we describe a toy model using generated data where we will assume
 that the true number of latent factors and components are not known. 
-Hence we require some heuristic to decide how many latent factors and components are 
+We require some heuristic to decide how many latent factors and components are 
 preferred given some data. An increasing number of factors
@@
 \end{equation} 
-\noindent{}where $Q$ is the number of model parameters:
+\noindent{}where $Q$ is the number of parameters in this model:
 \begin{equation}
@@
 
-For these reasons we also apply the Minimum Message Length \citep[MML;][]{Wallace:2005}
+For these reasons we use the Minimum Message Length \citep[MML;][]{Wallace:2005}
 principle as a criterion for model selection and evaluation. 
@@
 through the $\log\kappa_Q$ term. However, the model complexity is also 
-described through the priors and the Fisher information matrix, which
+described through the priors and the Fisher information, which
 describes the curvature of the log likelihood with respect to the model
@@
 $p(\NumComponents) \propto 2^{-\NumComponents}$ respectively,
-such that fewer numbers are preferred and the optimal lossless message to 
-encode the number of each is \citep[p279, sec. 6.8.2;][]{Knorr-Held:2000},
+such that fewer numbers are preferred. The optimal lossless message to 
+encode each is \citep[Sec. 6.8.2;][]{Knorr-Held:2000},
 \begin{eqnarray}
@@
 We assume uniform priors for the component means in latent space $\scoremeans$, 
-and a conjugate inverted Wishart prior for the component covariance matrices
+where the bounds are large enough outside the range of observable values such
+that those priors are proper (integrable) -- a necessary condition for the MML principle -- 
+and only add constant terms to the message length, which can be ignored. We assume 
+a conjugate inverted Wishart prior for the component covariance matrices
 $\scorecovs$ \citep[Section 5.2.3;][]{Knorr-Held:2000},
@@
 \end{equation}
+
 \noindent{}such that
+
 % todo put line in equation.
@@
 	I(\scoremeans,\scorecovs) &=& -\sum_{\numcomponents=1}^{\NumComponents}\log{p(\scoremeans_k,\scorecovs_k)} + \frac{1}{2}\sum_{\numcomponents=1}^{\NumComponents}\log{|\mathcal{F}(\scoremeans_k,\scorecovs_k)|} \nonumber \\
-%	&=& \frac{1}{2}\sum_{\numcomponents=1}^{\NumComponents}\log\left[(\NumData\weight_k)^{\frac{1}{2}\NumLatentFactors(\NumLatentFactors+3)}2^{-\NumData}|\scorecovs_k|^{-(\NumData + 2)}\right] \nonumber \\
-%	&& \cdots \,\, -\sum_{\numcomponents=1}^{\NumComponents}\log{|\scorecovs_k|}^{\frac{1}{2}(\NumLatentFactors + 1)} \\
+	&=& \frac{1}{2}\sum_{\numcomponents=1}^{\NumComponents}\log\left[(\NumData\weight_k)^{\frac{1}{2}\NumLatentFactors(\NumLatentFactors+3)}2^{-\NumData}|\scorecovs_k|^{-(\NumData + 2)}\right] \nonumber \\
+	&& \cdots \,\, -\sum_{\numcomponents=1}^{\NumComponents}\log{|\scorecovs_k|}^{\frac{1}{2}(\NumLatentFactors + 1)} \\
 I(\scoremeans,\scorecovs) &=& \frac{1}{4}\NumLatentFactors(\NumLatentFactors+3)\sum_{\numcomponents=1}^\NumComponents\log{\NumData\weight_k} - \frac{KD}{2}\log{2} \nonumber \\ 
@@
 %The Fisher information for $\factorloads$ \todo{needs to be double checked.}
-Thus the message length to encode $\factorloads$ is given by
+\noindent{}Thus the message length to encode $\factorloads$ is given by
 \begin{eqnarray}
@@
 
-%\todo{As...}% TODO: 
-\vspace{10em}
-
-Combining equations \ref{eq:prior_J}, \ref{eq:prior_K}, \ref{eq:prior_pi}, \ref{eq:prior_xi_omega}, and \ref{eq:prior_L} with equation \ref{eq:mml} leads to the full message length:
+Combining equations \ref{eq:prior_J}, \ref{eq:prior_K}, \ref{eq:prior_pi}, \ref{eq:prior_xi_omega}, and \ref{eq:prior_L} with equation \ref{eq:mml} leads to the full message length:\clearpage
 \begin{widetext}
@@
   && \cdots -\frac{1}{2}\left(D-J-1\right)\log|\factorloads\transpose\factorloads| + \textrm{Tr}\left(\textrm{Cov}\left(\factorloads\transpose\right)^{-1}\factorloads\transpose\factorloads\right) - \left(\NumLatentFactors + \frac{3}{2}\right)\sum_{\numcomponents=1}^\NumComponents\log|\scorecovs_\numcomponents|  -\log\Gamma\left(\NumComponents\right) - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\
-&& \cdots +\frac{Q}{2}\log\kappa_q +\frac{1}{2}\left[\NumLatentFactors(\NumDimensions + 2) + \NumComponents(2-\NumData)\right]\log{2}  \quad . \nonumber \label{eq:message-length}
+&& \cdots +\frac{Q}{2}\log\kappa_q +\frac{1}{2}\left[\NumLatentFactors(\NumDimensions + 2) + \NumComponents(2-\NumData)\right]\log{2}  \quad . \label{eq:message-length}
  \end{eqnarray}
@@
 Here we introduce a toy model where we use generated data to verify that
-we recover the true data-generating parameters given some data, and to
-ensure that the method is yielding consistent results.
+we recover the true model parameters given some data, and to
+ensure that the expectation-maximization method is yielding consistent results.
 We generated a data set with ${\NumData = 1}$00,000 data points, each with
@@
 with $\NumComponents = 20$ clusters in the latent space. 
-We generated the factor loads in the same way that we initialise the optimisation (Section~\ref{sec:initialisation}). The relative weights $\vec\weight$
+We generated the random factor loads in the same way that we initialise the optimisation (Section~\ref{sec:initialisation}). The relative weights $\vec\weight$
 are drawn from a multinomial distribution and the means of the clusters
@@
 		 	 true data generating values
-		 	 for Experiment~1. The agreement is excellent.}
+		 	 for Experiment~1 (Section~\ref{sec:exp-1}). The agreement is excellent.}
     \label{fig:exp1-compare}
@@
 
-Here we treat the generated data set as if the true number of latent factors
-and the true number of components are not known. Starting with $\NumLatentFactors = 1$
+We treat the generated data set as if the number of latent factors
+and components are not known. Starting with $\NumLatentFactors = 1$
 and $\NumComponents = 1$, we trialled each permutation of $\NumLatentFactors$ and $\NumComponents$
@@
 
-We recorded the \emph{negative} log likelihood, the BIC, and the message length (omitting constant terms) for each permutation of $\NumLatentFactors$ and $\NumComponents$.
+We recorded the \emph{negative} log likelihood, the BIC, and the message length\footnote{Omitting constant terms such that negative message lengths are allowed.}  for each permutation of $\NumLatentFactors$ and $\NumComponents$.
 These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
@@
 It is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
-and clustering in the latent space provides a better description of the (generated) data.
+and clustering in the latent space provides a better description of the (generated) data than a gaussian mixture model without latent factors.
 Adding components to the model does improves the log likelihood, even with a single latent factor,
@@
 \emph{twenty components}. Not much more can be said for this example because the true data generating process 
-is drawn from a higher number of latent factors, but this toy model does illustrate how 
+is known, but this toy model does illustrate how 
 clustering in high dimensional data can be better described by latent factors with 
@@
 Some technical background is warranted before we compare our estimated model
-parameters to the true values. The latent factors in this model are only
+parameters to the true values. We previously stated that the latent factors in this model are only
 identifiable up to an orthogonal rotation. That is to say that
@@
 could be different from the true factors, and the ordering of the dimensionality
-in latent space would be correspondingly different. Since no constraint is
+in latent space would then be accordingly different. Since no constraint is
 placed on the ordering of the factor loads during expectation-maximization,
@@
 general property that the factor loads and factor scores can be orthogonally 
-rotated by \emph{any valid rotation matrix}\footnote{A rotation matrix is valid if 
-$\vec{R}\vec{R}\transpose = \vec{I}$} $\vec{R}$. The estimated factor loads 
+rotated by \emph{any valid rotation matrix}\footnote{Recall that a rotation matrix is valid if 
+$\vec{R}\vec{R}\transpose = \vec{I}\,\,$.} $\vec{R}$. The estimated factor loads 
 $\factorloads_\textrm{est}$ could therefore appear very different from the true 
@@
 to construct a Givens rotation matrix \citep{Givens:1958}, and then we take the product of those Givens
-matrices. This process reduces to Euler angle rotation in three or fewer dimensions.
+matrices to produce a valid rotation matrix $\vec{R}$. This process reduces to Euler angle rotation in three or fewer dimensions.
 This process rotates the latent space
 ($\factorloads$, $\scoremeans$, $\scorecovs$), but has no effect on the model's 
-predictive power: the generated data, evaluated log likelihood, or the Kullback-Leibler divergence \citep{Kullback:1951} under the
+predictive power: the evaluated log likelihood or the Kullback-Leibler divergence \citep{Kullback:1951} under the
 rotated model is indistinguishable from the unrotated model.
@@
 
-Here we repeat the toy model used in the previous Section, but we discard an increasing fraction
-of the data and evaluate the model's performance in the presence
+Here we repeat the toy model used in the previous experiment, but we discard an increasing fraction
+of the data and evaluate the performance and accuracy of our method in the presence
 of incomplete data. We considered missing data fractions from 1\% to
@@
 		 	 true data generating values
-		 	 for Experiment~2. Here 40\% of the data are missing at random.
+		 	 for Experiment~2 (Section~\ref{sec:toy-model-missing-data}. Here 40\% of the data are missing at random.
 			 The agreement remains excellent,
@@
 dimension is not systematically under-estimated from the true values, 
-despite 40\% of the data entries to be missing.
+despite that 40\% of the data entries are missing.
 
@@
 	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-latent-space.pdf}
-	\caption{The factor scores $\factorscores$ estimated in Experiment~3 using the \Galah\ data \citep{Buder:2018}, where each star is coloured by its inferred component.}
+	\caption{The factor scores $\factorscores$ estimated in Experiment~3 (Section~\ref{sec:exp4} using $N =$ 1,072 stars in the \Galah\ data \citep{Buder:2018} that have 18 abundance measurements.
+Here each star is coloured by its inferred component.}
     \label{fig:exp3-latent-space}
@@
 	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-data-space.pdf}
-	\caption{Detailed chemical abundances from \Galah\ \citep{Buder:2018} used in this work. Each star is coloured by its inferred component in latent space.}
+	\caption{Detailed chemical abundances from the $N =$ 1,072 stars in \Galah\ \citep{Buder:2018} that have 18 chemical abundances (Section~\ref{sec:exp4}).
+Each star is coloured by its inferred component from the lower-dimensional latent space.}
     \label{fig:exp3-data-space}
@@
 This is demonstrated in Figure~\ref{fig:exp3-factor-loads}, where some estimated factor loads match
-closely to the target load (e.g., $\mathbf{L}_1$ which we identify as the s-process), and some barely
-match at all (e.g., $\mathbf{L}_4$). Here we show the absolute entry of the factor loads because even
-if an entry is negative, the corresponding factor scores could also be negative, and their product will contribute to the observed abundances.
-For this reason the factor load sign does not matter here.
+closely to the target load (e.g., $\mathbf{L}_2$ which we identify as the s-process), and some barely
+match at all (e.g., $\mathbf{L}_5$). Here we show the absolute entry of the factor loads because even
+if an entry is negative, the corresponding factor scores could also be negative, and their product will contribute to the observed abundances. For this reason the sign does not matter here.
 
@@
 The fractional contributions to each element are shown in the right hand side of
-Figure~\ref{fig:exp3-factor-loads}. As targeted, we identify the first factor $\mathbf{L}_0$ 
-as being most similar to the r-process, and here it is the dominant contributor to Eu and La, both heavy elements that are primarily r-process tracers.
-Surprisingly we also find that this factor load is the dominant contributor to the Fe-peak element Sc.
+Figure~\ref{fig:exp3-factor-loads}. We identify the first factor $\mathbf{L}_1$ 
+as being most similar to the r-process, and here it is the dominant contributor to Eu, a typical r-process tracer.
+Surprisingly we also find that this factor load is a reasonable  contributor to the Fe-peak element Sc.
 The specific scatter in Sc is 0.03~dex (Figure~\ref{fig:exp3-specific-scatter}), suggesting that the Sc abundances are well-described by this latent factor model.
 
-The second latent factor $\mathbf{L}_1$ here is most representative of the slow neutron capture process, with dominant contributions to Ba and Y.
-This factor has some support at other elements, including Cu and Sc, but with very little effective contributions.
-Most Cu is contributed by $\mathbf{L}_2$, our Fe-peak latent factor.
-In fact $\mathbf{L}_2$ is the primary contributor to nearly all Fe-peak elements, with close to negligible contributions from other factors.
-The exception here is Zn, where slightly more contribution comes from $\mathbf{L}_4$ than it does from $\mathbf{L}_2$.
-Recall that while $\mathbf{L}_4$ was described as our `light odd-Z element' factor load, it is the dominant producer of both K and Zn, and secondary contributor to Ti, V, and to a lesser extent, La.
-The fourth latent factor $\mathbf{L}_3$ is better behaved.
-It is the dominant contributor to the $\alpha$-element tracers Si, Ca, and Ti, and nothing else.
-It contributes a reasonable fraction to Al and Zn, but with negligible contributions elsewhere.
+The second latent factor $\mathbf{L}_2$ here is most representative of the slow neutron capture process (s-process), with dominant contributions to Ba, Y, and La.
+This factor has some support at other elements, notably K.
+Most Sc is contributed by $\mathbf{L}_3$, our Fe-peak latent factor.
+In fact $\mathbf{L}_3$ is the primary contributor to nearly all Fe-peak elements, with close to negligible contributions from other factors.
+The exception here is Zn, where a near-equal contribution comes from $\mathbf{L}_5$.
+% Recall that while $\mathbf{L}_5$ was described as our `light odd-Z element' factor load, 
+%it is also the dominant producer (just barely) of V, a Fe-peak element, and secondary contributor to K. 
+The third latent factor $\mathbf{L}_4$ is reasonably well behaved.
+It is the dominant contributor to the $\alpha$-element tracers Si, Ca, and Ti, and surprisingly, Al.
+It also seems to contribute a non-zero fraction to Eu, with negligible contributions elsewhere.
 The specific scatter after accounting for these latent factors is smallest
-for Fe and largest for K (Figure~\ref{fig:exp3-specific-scatter}). The typical scatter is about 0.05~dex.
+for Fe (0.01~dex) and largest for K (0.13~dex; Figure~\ref{fig:exp3-specific-scatter}). The typical scatter in most elements is about 0.05~dex.
 
 In Figure~\ref{fig:exp3-latent-space} we show the inferred clustering in latent space,
-where the separation between components is arguably best seen in the splitting between $\factorscores_4$ with respect to $\factorscores_1$ or $\factorscores_2$.
-When projected to data space (Figure~\ref{fig:exp3-data-space}) the third component (purple) is seen to have relatively higher abundance ratios of [K,Ba,Zn/Fe] at a given [Fe/H],
+where the separation between components is arguably best seen in the splitting between $\factorscores_5$ with respect to $\factorscores_2$ or $\factorscores_3$.
+When projected to data space (Figure~\ref{fig:exp3-data-space}) the third component (dark green) is seen to have relatively higher abundance ratios of [K,Ba,Zn/Fe] at a given [Fe/H],
 and lower abundance ratios of [V/Fe].
-This is consistent with the clustering in latent space, where $\factorloads_4$ is the dominant contributor to K and Zn, and $\factorloads_{1,2}$ are the dominant contributor to Ba and V, respectively.
+This is consistent with the clustering in latent space.%, where $\factorloads_5$ is the dominant contributor to K and Zn, and $\factorloads_{2,3}$ are the dominant contributor to Ba and V, respectively.
 
@@
 include more stars, even though those stars have some abundance measurements
-missing. We assume those measurements are missing at random, and discuss the
-impact of that assumption (and others) in Section~\ref{sec:discussion}.
+missing. 
+%We assume those measurements are missing at random, and discuss the
+%impact of that assumption (and others) in Section~\ref{sec:discussion}.
 Specifically we started with the same subset of 1,072 stars in Section~\ref{sec:exp4}
@@
 the model with the lowest message length. We then repeated this
-procedure using 1,000 stars ($N =$ 2,072), then again with 10,000 ($N =$ 11,072),
+procedure using 1,000 stars ($N =$ 2,072), again with 10,000 ($N =$ 11,072),
 and finally using all 99,174 stars that met the criteria of \texttt{flag\_cannon = 0} and
@@
 
-\consider{For sample sizes up to $N = 11,072$ we find that the number of preferred latent
-factors remains at $J = 5$. The factor loads inferred from each subset of stars
-is shown in Figure~\ref{fig:exp10-comparison-loads}, illustrating that the
+For sample sizes up to $N \sim $2,000 we found that five latent factors were
+preferred, and these factors shared common features (Figure~\ref{fig:exp10-comparison-loads}). This illustrates that the
 first\footnote{`First' has no concept here in terms of factor load ordering, but
 for the purposes of comparing inferred loads from different data sets we have ordered
-the loads to be as close to those inferred in Section~\ref{sec:exp4}.} inferred factor
-loads remain largely unchanged despite the increasing sample size, with an increasing
-fraction of missing data entries. It is not until the data set reaches $N = 11,072$ 
-stars where we find that $J = 8$ factor loads are preferred by the data. Again, the
-first five latent factors are extremely similar to those inferred from just 1,072
-stars with complete abundances, except for the notable difference in Zn loads in
-factor load $|\mathbf{L}_4|$, which disappears for $N = 11,072$ and re-appears
-in the sixth factor load $|\mathbf{L}_5|$ -- indicative that the additional data
-prefers the fourth latent factor to split into an additional latent factor where
-Zn contributions dominate.}
+the loads to be as close to those inferred in Section~\ref{sec:exp4}.} set of inferred factor
+loads inferred from a smaller, complete data set, remain largely unchanged despite the increasing sample size and the increasing number of missing data entries.
+When the sample size reaches $N = $11,072 we find another three latent factors are
+required to best explain the data.
+
+When $N \sim$ 100,000, the preferred number of latent factors rises to twelve. Of note among these factors is $\factorloads_9$, where
+no `target load' was prescribed\footnote{Although no `target load' was prescribed here, this statement should be interpreted with caution because the mutual orthogonality constraint exists and the remaining factors do have target loads.}, and the non-zero entries mimic what might be expected from a
+light s-process production. Similarly, $\factorloads_{10}$ has near zero contributions
+everywhere except among the light elements Al, K, Ca, and Sc. 
+With this data set we find that $K = 16$ components are preferred in latent space. None of these components appear coherently structured in their positions or motions.
 
@@
 	\caption{Latent factors found in Section~\ref{sec:exp10} using different subsets
-	         of \Galah\ data \citep{Buder:2018}. The thin line shows the result from
-	         Section~\ref{sec:exp4} with $N=1,072$ stars with 17 abundances and no
-	         missing data. Dotted line shows that data set complemented with 100 random
-	         stars that have some abundances missing. Dot-and-dashed line shows the
-	         original 1,072 stars with 1,000 random stars with some missing abundances,
-	         and the thick line shows the original 1,072 with 10,000 stars with
-	         missing abundances, where eight latent factors were preferred over five.}
+	         of \Galah\ data \citep{Buder:2018}. The thin solid line shows the result from
+	         Section~\ref{sec:exp4} with $N=1,072$ stars with 18 abundances and no
+	         missing data. Increasing thickness indicates larger samples, up to the solid thick line with $N \sim$ 100,000 stars, where twelve latent factors are preferred.}
     \label{fig:exp10-comparison}
 \end{figure*}
-
-\todo{If we use ALL data, how many? Then just comment on clustering in the N = XXX,XXX data}
 
@@
 metallicity, and other factors, but in principle to resolve the indeterminacy
-the mean of these yields would still require mutual orthogonality in this model.
+\emph{in this model} would still require mutual orthogonality on the mean yields.
 Introducing a constraint on the factor scores that resolves this indeterminacy and allows for more flexible latent factors would be a worthy extension to this work.
@@
 order of magnitude \citep{Kobayashi:2006}. From our inferences 
-using \Galah\ data, we find that $\factorloads_4$ is the dominant contributor to
-potassium, but this factor load is also the dominant contributor to Zn, 
-an element that is typically associated with the iron-peak group. That factor load
-is also a reasonable contributor to Ti, V, and to a lesser extent, La. Does this
+using \Galah\ data, we find that $\factorloads_1$ -- the factor we identify as the s-process --  is the dominant contributor to
+potassium. This latent factor persists even in the presence of missing data, and
+a sample size two orders of magnitude larger.
+Does this
 suggest production of K is linked to the production of much heavier nuclei?
@@
 In the \citep{Cohen:2012} sample their high [K/Fe] stars also tend
-to be high in [Zn/Fe] (for which $\factorloads_4$ also dominates), but there
+to be high in heavier elements, but there
 are also numerous abundance correlations present.
@@
 scatter (Figure~\ref{fig:exp3-specific-scatter}), suggesting that the contributions
-of K are not as well described by the latent factor model as other
-elements.
-
-A similar argument could be made for Sc, where $\mathbf{L}_0$ -- a factor load we
-identify as the r-process -- is the dominant contributor. Indeed, the second largest
-contributor to Sc is the factor load we identify as the s-process. Sc is under-produced in 
+of K are perhaps not as well described by the latent factor model as other
+elements. Still, the specific model scatter is far less than the observed scatter,
+indicative that the model does have some predictive power.
+
+A similar argument could be made for Sc, where $\mathbf{L}_1$ -- a factor load we
+identify as the r-process -- is the secondary contributor. Sc is under-produced in 
 galactic chemical evolution models relative to observations \citep{Kobayashi:2006,Casey:2015}.
@@
 orthogonal basis. As we have seen in our experiments, the ordering and sign 
-of the latent factors is not described \emph{a priori}. That means that we
+of the latent factors is not described \emph{a priori}. This is both a feature
+and a bug: unrestricted ordering and signs allow for the model parameters to be
+estimated more efficiently because they can freely rotate as the model
+parameters are updated, but it does mean that we
 must `assign' the latent factors we infer as being described by an astrophysical
-process (e.g., the zeroth-latent factor is r-process). A more general limitation
+process (e.g., the first latent factor is r-process). A more general limitation
 of this is that the latent factors can be multiplied by some arbitrary rotation
@@
 describe chemical abundance data. 
-We find five latent factors from \Galah\ data, and
+We find five latent factors from a small subset of \Galah\ data with complete abundances, and
 those latent factors can qualitatively be described within the context of astrophysical yields. 
+Those latent factors are recovered in larger samples where the data are incomplete.
 That did not have to be the case: the mutually orthogonal latent factors could be entirely 
@@
 is encouraging that there is some interpretability in the latent factors.
-Indeed, in the elements where we find surprising associations (e.g., Sc with what we identify as the r-process, K and Zn linked by a common process), these are elements where galactic chemical evolution models are most discrepant from observations, even after accounting for systematic errors in abundance measurements.
-
-The relatively small data set with complete abundances measured in \Galah\ both limits
-the number of components we can infer, and the efficacy of chemical tagging. Only three
-components were preferred from the data set of 1,072 stars, which we identified as
-being stars with low- and high-[$\alpha$/Fe] abundance ratios, and a third component that
-primarily differs in its abundances of K, Ba, Zn, and V (e.g., given some [Fe/H] and [$\alpha$/Fe]). A worthwhile extension
-of this work would be to incorporate missing and noisy data, either by imputing the
-missing data values or by additive variance in the constructed
-covariance matrices. In doing so would allow for
-blind chemical tagging on an industrial scale, allowing for the latent factors to
-be efficiently estimated while chemical tagging occurs in the lower-dimensional
-latent space.
+Indeed, in the elements where we find surprising associations (e.g., Sc and K), these are elements where galactic chemical evolution models are most discrepant from observations, even after accounting for systematic errors in abundance measurements (e.g., violations to the assumption of local thermodynamic equilibrium).
+
+In the subset of \Galah\ data with complete abundances we find that three components are
+preferred.
+These components can be described as those with (1) low- and (2) high-[$\alpha$/Fe] abundance ratios, and another (3) primarily differing in K, Ba, Zn, and V abundances at a given [Fe/H] and [$\alpha$/Fe] abundance ratio.
+When we include $\sim$100,000 stars with up to 18 abundances, and assume the incomplete abundances are missing at random, we find that 16 components in latent space are preferred to explain the data.
+By construction these  components
+are structured in their chemical abundances because of the projection from the latent
+space, and by extension of each component having similar chemistry, each component
+occupies realistic locations in a Hertzsprung-Russell diagram. When we
+project these component associations to the data space we find that none of the inferred
+components are structured or coherent in their positions or motions.
+However, in this sample of stars there are no gravitationally bound clusters where a reasonable (e.g.
+$\sim$30) number of stars have been observed.
+Clearly, more data would help to resolve a higher number of components.
+
+Perhaps it is not so discouraging that none of the inferred components are structured in their positions or motions because there are no gravitationally bound clusters in the data.
+But there is clearly more that can be done in chemical tagging.
+Some components we infer have stars with positions and galactic orbits that would imply that they cannot have formed in the same star cluster.
+In these situations there is likely significant value in including joint probabilities on whether two stars could be associated to the same star formation site based on their dynamic properties.
+Similarly, although stellar ages are historically difficult to estimate precisely, can this imprecise information help inform weak priors or probabilities of two stars having the same association?
+There is an incredible amount of dynamical information available from \project{Gaia}, particularly for stars in the \Galah\ survey, and weakly informative priors might be sufficient to help improve the granularity of chemical tagging without being overly constraining on the dynamical and star formation history we seek to infer.
+
+%The relatively small data set with complete abundances measured in \Galah\ both limits
+%the number of components we can infer, and the efficacy of chemical tagging. Only three
+%components were preferred from the data set of 1,072 stars, which we identified as
+%being stars with low- and high-[$\alpha$/Fe] abundance ratios, and a third component that
+%primarily differs in its abundances of K, Ba, Zn, and V (e.g., given some [Fe/H] and [$\alpha$/Fe]). A worthwhile extension
+%of this work would be to incorporate missing and noisy data, either by imputing the
+%missing data values or by additive variance in the constructed
+%covariance matrices. In doing so would allow for
+%blind chemical tagging on an industrial scale, allowing for the latent factors to
+%be efficiently estimated while chemical tagging occurs in the lower-dimensional
+%latent space.
 
@@
 our model does provide the first data-driven approach to nucleosynthesis and chemical tagging.
-
-With a larger sample of stars that have a complete abundance inventory we would expect to 
-identify more latent factors and/or components.
+We advocate that more data, and the inclusion of weakly informative priors -- joint probabilities using astrometry and a simplified model of the Milky Way -- would help in realising the full potential of chemical tagging.
+
+%With a larger sample of stars that have a complete abundance inventory we would expect to 
+%identify more latent factors and/or components.
 Alternatively, a useful extension of this work would be to allow for missing and/or noisy data, such that realistic data sets of chemical abundances can be used to perform chemical tagging experiments on a larger scale.

### 2019-06-21 article/experiments/exp3.py

@@ -194,7 +194,7 @@ latex_label_names = [r"$\textrm{{{0}}}$".format(ea.split("_")[0].title()) for ea
 
 # Draw unrotated.
 J_max = config["max_n_latent_factors_for_colormap"]
-J_max = 8
+J_max = 12
 cmap = mpl_utils.discrete_cmap(J_max, base_cmap="Spectral")
 colors = [cmap(j) for j in range(J_max)]#[::-1]
 
@@ -321,7 +321,7 @@ fig_fac = mpl_utils.plot_factor_loads_and_contributions(model, X,
                                                         label_names=latex_label_names, colors=colors)
 savefig(fig_fac, "latent-factors-and-contributions")
 
-
+raise a
 
 
 

### 2019-06-21 article/experiments/plot-exp10-comparison.py

@@ -28,6 +28,7 @@ comparison_model_paths = [
     "exp10-models-size-100.pkl",
     "exp10-models-size-1000.pkl",
     "exp10-models-size-10000.pkl",
+    "exp10-models-size-99174.pkl"
 ]
 
 # 89dab is for all intents and purposes identical to 8761b
@@ -79,6 +80,7 @@ for comparison_model_path in comparison_model_paths:
 
 # TODO: Set this from config, or something crazy to keep colours consistent?
 J_max = np.max(n_latent_factors)
+#J_max = 8
 
 cmap = mpl_utils.discrete_cmap(J_max, base_cmap="Spectral_r")
 colors = [cmap(j) for j in range(J_max)][::-1]
@@ -122,6 +124,7 @@ def _fill_factors(A, J_max):
     return np.hstack([A, np.ones((D, J_max - J)) * np.nan])
 
 
+max_rotations = 3
 
 ABS_ONLY = True
 if ABS_ONLY:
@@ -138,9 +141,10 @@ fig = mpl_utils.plot_factor_loads(viz_load(_fill_factors(A_original, J_max)),
 plot_factor_loads_kwds.update(fig=fig)
 
 plot_kwds = {
-    "exp10-models-size-100.pkl": dict(linestyle=":", lw=2, alpha=0.5),
-    "exp10-models-size-1000.pkl": dict(linestyle="-.", lw=3, alpha=0.75),
-    "exp10-models-size-10000.pkl": dict(linestyle="-", lw=4, alpha=1),
+    "exp10-models-size-100.pkl": dict(linestyle=":", lw=1, alpha=0.75),
+    "exp10-models-size-1000.pkl": dict(linestyle="-", lw=1, alpha=0.75),
+    "exp10-models-size-10000.pkl": dict(linestyle=":", lw=2, alpha=1),
+    "exp10-models-size-99174.pkl": dict(linestyle="-", lw=2, alpha=1)
 }
 
 
@@ -149,6 +153,7 @@ A_target = A_original
 #A_target[:-2, 0] = 0
 #A_target[-2:, 0] = 1
 
+stored_models = []
 
 for i, comparison_model_path in enumerate(comparison_model_paths):
 
@@ -165,49 +170,63 @@ for i, comparison_model_path in enumerate(comparison_model_paths):
     A_est = comparison_model.theta_[comparison_model.parameter_names.index("A")]
     if A_est.shape == A_target.shape:
 
[…]

### 2019-07-15 article/mcfa.tex
updates

@@
 
-% NEED: Final model with # of loads and # of clusters so we can:
-% 		[ ] Update figure of loads (exp10-comparison)
-% 		[ ] Make a clustering plot in abundances
-% 		[ ] Check to see if any chemical clusters correspond to known clusters.
-% 		[ ] Update discussion w.r.t. final model loads, etc.
-% 		[ ] Update conclusion w.r.t. final model loads/clusters, etc, and missing data.
-% 		[ ] Update abstract w.r.t. final model loads/clusters.
-% 		[ ] Upload PDF to Galah survey website
-%		[ ] Send email to Galah team for comment
-
 \received{2019}
@@
 
-\author[0000-0002-0583-5918]{David Dowe}
+\author[0000-0002-0583-5918]{David L. Dowe}
 \affiliation{Faculty of Information Technology, 
@@
 
-\author{the GALAH team}
+\author[0000-0001-7516-4016]{Joss Bland-Hawthorn}
+\affiliation{Sydney Institute for Astronomy, School of Physics,
+			 A28, The University of Sydney, NSW 2006, Australia}
+\affiliation{Center of Excellence for Astrophysics in Three Dimensions (ASTRO-3D),
+			 Australia}
+\affiliation{Miller Professor, Miller Institute, 
+			 UC Berkeley, 
+			 Berkeley, CA 94720, USA}
+
+\author[0000-0002-4031-8553]{Sven Buder}
+\affiliation{Research School of Astronomy and Astrophysics,
+			 Australian National University,
+			 Canberra, ACT 2611, Australia}
+
+%\author{Janez Kos}
+%\affiliation{Sydney Institute for Astronomy, School of Physics,
+%			 A28, The University of Sydney, NSW 2006, Australia}
+
+\author[0000-0003-3081-9319]{Geraint F. Lewis}
+\affiliation{Sydney Institute for Astronomy, School of Physics,
+			 A28, The University of Sydney, NSW 2006, Australia}
+
+\author[0000-0002-3430-4163]{Sarah L. Martell}
+\affiliation{School of Physics,
+			 University of New South Wales, 
+			 Sydney, NSW 2052, Australia}
+
+\author[0000-0001-5344-8069]{Thomas Nordlander}			 
+\affiliation{Research School of Astronomy and Astrophysics,
+			 Australian National University,
+			 Canberra, ACT 2611, Australia}
+\affiliation{Center of Excellence for Astrophysics in Three Dimensions (ASTRO-3D), 
+			 Australia}
+
+\author[0000-0002-8165-2507]{Jeffrey D. Simpson}
+\affiliation{School of Physics,
+			 University of New South Wales, 
+			 Sydney, NSW 2052, Australia}
+
+\author[0000-0003-1124-8477]{Daniel B. Zucker}
+\affiliation{Department of Physics and Astronomy,
+			 Macquarie University, 
+			 Sydney, NSW 2109, Australia}
 
@@
 to all stars with different scores, and clustering (e.g., chemical tagging) 
-is modelled by a mixture of multivariate gaussians
+is modelled by a mixture of multivariate Gaussians
 in a lower-dimensional latent space.
@@
 would need to be before they could be tagged as belonging to the same
-star formation site \citep{Mitschang:2014}. However, there are analysis
+star formation site \citep{Bland-Hawthorn:2010b,Bland-Hawthorn:2010a,Mitschang:2014}. However, there are analysis
 issues in understanding how precisely those chemical abundances can be measured
@@
 where those chemical abundances trace different nucleosynthetic pathways.
-This is the primary goal of the \Galah\ survey \citep{DeSilva:2015,Martell:2017,Buder:2018},
+This is the primary goal of the Galactic Archaeology with \project{HERMES} (\Galah) survey \citep{DeSilva:2015,Martell:2017,Buder:2018},
 a stellar spectroscopic survey that uses the High Efficiency and Resolution 
-Multi-Element Spectrograph \citep[\project{HERMES};][]{Sheinis:2015} on the Australian 
-Astronomical Telescope (AAT).  \Galah\ will observe up to $10^6$ stars in the 
-Milky Way, and measure up to 30 chemical abundances for each star. This includes
+Multi-Element Spectrograph \citep[\project{HERMES};][]{Sheinis:2015} on the 3.9~m Anglo-Australian 
+Telescope (AAT).  \Galah\ will observe up to $10^6$ stars in the 
+Milky Way, and measure up to 30 chemical abundances for each star \citep{Bland-Hawthorn:2016}. This includes
 light odd-Z elements (e.g., Na, K), elements produced through
@@
 through the slow (e.g., Ba) and rapid neutron-capture process
-(e.g., Eu). No other spectroscopic survey provides an equivalent set of
+(e.g., Eu). No other current or planned spectroscopic survey provides an equivalent set of
 chemical abundances for a comparable number of stars.
@@
 
-Factor analysis is a common statistical approach for describing correlated 
+Latent factor analysis is a common statistical approach for describing correlated 
 observations with a lower number of latent variables \citep[e.g.,][]{Thompson:2004}.
@@
 the chemical abundances of stars \citep[e.g.,][]{Ting:2012,Price-Jones:2018}, factor
-analysis cannot describe clustering in the data (or latent) space.
+analysis cannot describe clustering in the data (or latent) space. As a result,
+some works have performed clustering and then required different latent factors for each (totally assigned) component \citep[e.g.,][]{EdwardsDowe1998}.
 Similarly, clustering techniques applied to chemical abundances to date 
@@
 the number of latent factors, and the factor scores 
-for the $\numdata$th data point
+for the $\numdata^\mathrm{th}$ data point
 \begin{equation}
@@
 \end{equation}
-\noindent{}are drawn from\footnote{For clarifying nomenclature across disciplines, the terminology $z \sim \mathcal{N}(0, 1)$ indicates that the $z$ variable \emph{is drawn from} a standard normal distribution.} the $\numcomponents$th multivariate normal distribution.
+\noindent{}are drawn from\footnote{For clarifying nomenclature across disciplines, the terminology $z \sim \mathcal{N}(0, 1)$ indicates that the $z$ variable \emph{is drawn from} a standard normal distribution.} the $\numcomponents^\mathrm{th}$ multivariate
+normal distribution, where each $\numcomponents^\mathrm{th}$ component has a $\NumLatentFactors$-dimensional mean and a dense $\NumLatentFactors \times \NumLatentFactors$ covariance matrix.
 The mean vector $\vec\mu$ describes the mean datum in each dimension.
@@
 are \emph{clustered in the latent space} $\factorscores$ and projected
-into the data space by the factor loads $\factorloads$.
+into the data space by the factor loads $\factorloads$. In a sense we
+are using latent factor analysis as a form of dimensionality reduction and
+simultaneously performing clustering in latent space.
 
 We assume that the latent space is lower dimensionality than the
-data space (e.g., $\NumLatentFactors < \NumDimensions$).
+data space (i.e., $\NumLatentFactors < \NumDimensions$).
 Within the context of stellar abundances, the factor loads
@@
 Without loss of generality the density of the mean-subtracted 
-data $\vecdataunscaled - \vec\mu$ (which we hereafter will refer to simply as $\vecdata$) can be described as
+data $\vecdataunscaled - \vec\mu$ (which we hereafter will refer to simply as $\vecdata$) can be described as \citep[Sec. 5.2.9, p236;][]{Wallace:2005}
 \begin{equation}
@@
 $\vec\Psi$ includes $\{\factorloads,\vec\pi,\scoremeans,\scorecovs,\specificvariance\}$, and $\phi(\vecdata; \vec\theta)$
-describes the density of a multivariate gaussian distribution,
-and $\weight_\numcomponents$ describes the relative weighting of the $\numcomponents$th
-component in latent space and $\sum\vec\weight^K = 1$.
+describes the density of a multivariate Gaussian distribution,
+and $\weight_\numcomponents$ describes the relative weighting of the $\numcomponents^\mathrm{th}$
+component in latent space and $\sum\vec\weight_{\numcomponents} = 1$.
 The log likelihood is then given by
 \begin{equation}
-	\log\mathcal{L}(\vecdata|\vec\Psi) = \sum_{\numcomponents=1}^{\NumComponents}\log{f(\vecdata;\vec\Psi)} \quad . \label{eq:log-likelihood}
+	\log\mathcal{L}(\vecdata|\vec\Psi) = \sum_{\numdata=1}^{\NumData}\log{f(\vecdata;\vec\Psi)} \quad . \label{eq:log-likelihood}
 \end{equation}
@@
 matrix with a determinant of unity \citep{Stewart:1980}.
-We denote the $\NumLatentFactors \times \NumDimensions$ left-most region of this
-matrix to be $\mathbf{H}$, and by taking $\factorloads_\ast$ to be the Cholesky decomposition of $\mathbf{H}\transpose \mathbf{H}$, we initialise the factor loads as
-\begin{eqnarray}
- 	\factorloads = \mathbf{H}\left(\left(\factorloads_\ast\right)^{-1}\eye\right)
-\end{eqnarray}
-\noindent{}which ensures that $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$
-matrix of mutually orthogonal vectors.
+We denote the $\NumLatentFactors \times \NumDimensions$ left-most region\footnote{The region choice is arbitrary.
+All that is required is that the randomly-generated matrix have mutually orthogonal vectors.}
+ of this
+matrix to be our initial guess of $\factorloads$, which provides a set of mutually
+orthogonal vectors.
+%$\mathbf{H}$, and by taking $\factorloads_\ast$ to be the Cholesky decomposition of $\mathbf{H}\transpose \mathbf{H}$, we initialise the factor loads as
+%\begin{eqnarray}
+% 	\factorloads = \mathbf{H}\left(\left(\factorloads_\ast\right)^{-1}\eye\right)
+%\end{eqnarray}
+%\noindent{}which ensures that $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$
+%#matrix of mutually orthogonal vectors.
 %and that it is a valid rotation matrix
@@
 $\NumComponents$ components using the \texttt{k-means++} algorithm \citep{Arthur:2007}
-in the pseudo-latent space $\vecdata\factorloads$. Given the initial
+in the latent space. Given the initial
 factor loads and assignments, we then estimate the relative weights
@@
 \citep{Dempster:1977}. With each expectation step we evaluate the log likelihood 
-given the model parameters $\vec\Psi$ and we calculate the $\NumData \times \NumComponents$ responsibility 
+given the model parameters\footnote{When evaluating the log likelihood we use the precision (sparse inverse) matrix of the Cholesky decomposition of the covariance matrix for computational efficiency and stability.} $\vec\Psi$, the message length, and the $\NumData \times \NumComponents$ responsibility
 matrix $\vec\tau$ whose entries are the posterior probability that the 
@@
 practice it is sufficient to run expectation-maximization from multiple
-initialisations (as we do) in order to ensure that the global minima is reached.
+initialisations (as we do) in order to ensure that the global minimum is reached.
 At the maximization step we first update our estimate of the relative weights 
@@
 \begin{eqnarray}
-	\vec{W} &=& (\scorecovs_\numcomponents\thisstep)^{-1}\eye \\
+	\vec{W} &=& (\scorecovs_\numcomponents\thisstep)^{-1} \\
 	\vec{V} &=& \left(\specificvariance\thisstep\right)^{-1} \\
-	\vec{C} &=& (\vec{W} + (\factorloads\thisstep)\transpose\vec{V}\factorloads\thisstep)^{-1}\eye \\
+	\vec{C} &=& (\vec{W} + (\factorloads\thisstep)\transpose\vec{V}\factorloads\thisstep)^{-1} \\
 	\vec{G} &=& \left[\vec{V} - \vec{V}\factorloads\thisstep\vec{C}\left(\vec{V}\factorloads\thisstep\right)\transpose\right]\factorloads\thisstep\scorecovs_k\thisstep \quad .
@@
 
-\noindent{}where $\odot$ denotes is the entry-wise (Hadamard) product. 
+\noindent{}where $\odot$ denotes is the entry-wise product. 
 %We repeat
@@
 we impute zeros for missing data entries in $\vec\data$, and at each iteration we update
-these imputed valuee with our estimate of what the missing data values
+these imputed value with our estimate of what the missing data values
 are given the current model parameters. This ensures that the log-likelihood increases
@@
 One criterion commonly employed for evaluating a class of models is the 
-Bayesian Information Criterion \citep[BIC;][]{Schwarz:1978}, 
-\begin{equation}
-	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \quad , \label{eq:bic}
-\end{equation} 
-\noindent{}where $Q$ is the number of parameters in this model:
-\begin{equation}
-	Q = \frac{\NumLatentFactors}{2}\left[2\left(\NumDimensions - \NumLatentFactors\right) + \NumComponents\left(3 + \NumLatentFactors\right)\right] + \NumComponents + \NumDimensions - 1 \quad .
-\end{equation}
-\noindent{}While the BIC does include a penalisation term for the number of
+Bayesian Information Criterion \citep[BIC;][]{Schwarz:1978}
+\begin{equation}
+	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right)
+	 \label{eq:bic}
+\end{equation} 
+\noindent{}where $Q$ is the number of parameters in this model
+\begin{equation}
+	Q = \frac{\NumLatentFactors}{2}\left[2\left(\NumDimensions - \NumLatentFactors\right) + \NumComponents\left(3 + \NumLatentFactors\right)\right] + \NumComponents + \NumDimensions - 1
+\end{equation}
+\noindent{}which includes $\NumComponents - 1$ weights (as $\sum\weight_\numcomponents = 1$), $\NumDimensions$ specific variances, the $\NumComponents\NumLatentFactors - \NumLatentFactors^2$ free parameters needed to uniquely define the mutually orthogonal factor loads matrix $\factorloads$ \citep{Baek:2010}, $\NumComponents \times \NumLatentFactors$ parameters for the mean scores $\scoremeans$, and $\frac{\NumComponents\NumLatentFactors}{2}\left(\NumLatentFactors + 1\right)$ parameters to encode the $\NumComponents$ full ($\NumLatentFactors \times \NumLatentFactors$) covariance matrices $\scorecovs$.
+% weights: K - 1 model parameters.
+% psi: D model parameters
+% A: J * D - J^2 model parameters (Baek et al. 2010)
+% xi: J * K model parameters
+% omega: K * J*(J+1)/2
+While the BIC does include a penalisation term for the number of
 parameters (which scales with $\log{N}$), it does not describe for the
@@
 different models of varying complexity or explanatory power. MML has been 
-shown to perform well on a variety of empirical analyses (see, e.g., 
-\cite{viswanathan1999finding,fitzgibbon2004minimum}, with references to 
-further examples in \cite{Wallace:2005,dowe2007bayes,Dowe2008a,Dowe2011a}).
+shown to perform well on a variety of empirical analyses 
+\citep[see, e.g.,][]{WallaceDowe1994b,viswanathan1999finding,EdwardsDowe1998,WallaceDowe2000,fitzgibbon2004minimum,Wallace:2005,dowe2007bayes,Dowe2008a,Dowe2011a}.
 Arguments about the statistical consistency (i.e.,~as the number of data 
 points increases the distributions of the estimates become increasingly 
-concentrated near the true value) of MML are given in \cite{DoweWallace1997a,Dowe2011a}.
+concentrated near the true value) of MML are given in \citet{DoweWallace1997a,Dowe2011a}.
 The MML principle requires that we explicitly specify our prior beliefs on the
@@
 In practice calculating the message length can be a non-trivial task, 
-especially for models that are reasonably complex. This makes the strict MML
-principle intractable (or uncomputable) in many cases and necessitates
-approximations to the message length. Using a Taylor expansion, a generalised
+especially for models that are reasonably complex. This can make the strict MML
+principle intractable in many cases and necessitates
+approximations to the message length
+\citep[however see][]{Wallace:1987,WallaceDowe1999a,Wallace:2005}. Using a Taylor expansion, a generalised
 scheme can be calculated to estimate the parameter vector $\vec\Psi$ that
@@
 log likelihood, $p(\vec\Psi)$ is the joint prior density on $\vec\Psi$,
-$\mathcal{F}(\vec\Psi)$ is the negative second derivative of the log likelihood,
+$\mathcal{F}(\vec\Psi)$ is the matrix whose entries are the expected second order
+partial derivatives of the log likelihood,
 commonly referred to as the expected Fisher information matrix,
@@
 uncertainty in which the parameters $\vec\Psi$ are centred. The $\log\kappa_Q$
-term is
+term can be reasonably approximately by
 \begin{equation}
@@
 a conjugate inverted Wishart prior for the component covariance matrices
-$\scorecovs$ \citep[Section 5.2.3;][]{Knorr-Held:2000},
+$\scorecovs$ \citep{WallaceDowe1994b,WallaceDowe2000,Knorr-Held:2000},
 \begin{equation}
@@
 \end{equation}
+
 We approximate the determinate of the Fisher information of a multivariate normal $|\mathcal{F}(\scoremeans,\scorecovs)|$
@@
 addressed the indeterminacy between the factor loads and factor scores by
-placing a joint prior on the \emph{product} of factor loads and scores \citep{WallaceMLF}.
+placing a joint prior on the \emph{product} of factor loads and scores \citep{WallaceMLF,EdwardsDowe1998,Wallace:2005}.
 Adopting the same prior density in our model is not practical because 
@@
 %The Fisher information for $\factorloads$ \todo{needs to be double checked.}
+\begin{widetext}
 \noindent{}Thus the message length to encode $\factorloads$ is given by
 \begin{eqnarray}
-	I(\factorloads) &=& -\log\left(\frac{p(\factorloads)}{\sqrt{|\mathcal{F}(\factorloads)|}}\right) \nonumber \\
-I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} \nonumber \\
-	&& \cdots \,\, + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\ \label{eq:prior_L}%\nonumber \\
+%	I(\factorloads) &=& -\log\left(\frac{p(\factorloads)}{\sqrt{|\mathcal{F}(\factorloads)|}}\right) \nonumber \\
+I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \quad .
+\label{eq:prior_L}%\nonumber \\
 %	&& \cdots \todo{+\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \quad . \label{eq:prior_L}
@@
 
-Combining equations \ref{eq:prior_J}, \ref{eq:prior_K}, \ref{eq:prior_pi}, \ref{eq:prior_xi_omega}, and \ref{eq:prior_L} with equation \ref{eq:mml} leads to the full message length:\clearpage
-\begin{widetext}
+\noindent{}Combining equations \ref{eq:prior_J}, \ref{eq:prior_K}, \ref{eq:prior_pi}, \ref{eq:prior_xi_omega}, and \ref{eq:prior_L} with equation \ref{eq:mml} leads to the full message length:
 \begin{eqnarray}
@@
 			 message length. The white marker indicates the
-			 lowest value in each panel, showing the
-			 preferred number of latent factors and components.
-			 The black marker indicates the true value.}
+			 lowest value in each panel.
+			 The black marker indicates the true value. 
+			 In the top panel a line connects these markers to guide the eye.}
     \label{fig:experiment-1-gridsearch}
@@
 These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
-Unsurprisingly the negative log likelihood increases with increasing numbers of latent
+Unsurprisingly the negative log likelihood decreases with increasing numbers of latent
 factors $\NumLatentFactors$ and increasing numbers of components $\NumComponents$.
@@
 It is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
-and clustering in the latent space provides a better description of the (generated) data than a gaussian mixture model without latent factors.
+and clustering in the latent space provides a better description of the (generated) data than a Gaussian mixture model without latent factors.
 Adding components to the model does improves the log likelihood, even with a single latent factor,
@@
 to the true values. The agreement is excellent in all model parameters.
-
-\vspace{2em}
 
@@
 		 	 true data generating values
-		 	 for Experiment~2 (Section~\ref{sec:toy-model-missing-data}. Here 40\% of the data are missing at random.
+		 	 for Experiment~2 (Section~\ref{sec:toy-model-missing-data}). Here 40\% of the data are missing at random.
 			 The agreement remains excellent,
@@
 prefers higher components and latent factors than the extent of the grid (e.g., $J > 7$ and $K > 5$).
+Indeed, if we were to trial higher values of $J$ and $K$ then the negative log likelihood would continue to increase.
 The model with five latent factors and three components ($J = 5$, $K = 3$) is found to have the shortest message length, which we take as our preferred model for these data.
@@
 	\caption{The factor scores $\factorscores$ estimated in Experiment~3 (Section~\ref{sec:exp4} using $N =$ 1,072 stars in the \Galah\ data \citep{Buder:2018} that have 18 abundance measurements.
-Here each star is coloured by its inferred component.}
+Here each star is coloured by its inferred $K$th component.}
     \label{fig:exp3-latent-space}
@@
 	\caption{Detailed chemical abundances from the $N =$ 1,072 stars in \Galah\ \citep{Buder:2018} that have 18 chemical abundances (Section~\ref{sec:exp4}).
-Each star is coloured by its inferred component from the lower-dimensional latent space.}
+Each star is coloured by its $K$th inferred component from the lower-dimensional latent space, with the same colouring as per Figure~\ref{fig:exp3-latent-space}.}
     \label{fig:exp3-data-space}
@@
 For sample sizes up to $N \sim $2,000 we found that five latent factors were
-preferred, and these factors shared common features (Figure~\ref{fig:exp10-comparison-loads}). This illustrates that the
+preferred, and these factors shared common features (Figure~\ref{fig:exp10-comparison}).
+This illustrates that the
 first\footnote{`First' has no concept here in terms of factor load ordering, but
@@
 
-When $N \sim$ 100,000, the preferred number of latent factors rises to twelve. Of note among these factors is $\factorloads_9$, where
+When $N \sim$ 100,000, the preferred number of latent factors rises to twelve ($J = 12$). Of note among these factors is $\factorloads_9$, where
 no `target load' was prescribed\footnote{Although no `target load' was prescribed here, this statement should be interpreted with caution because the mutual orthogonality constraint exists and the remaining factors do have target loads.}, and the non-zero entries mimic what might be expected from a
@@
 
+\vspace{1em}
 \section{Discussion} \label{sec:discussion}
@@
 abundance pattern that currently lacks explanation \citep{Iliadis:2016,Kemp:2018}.
-In the \citep{Cohen:2012} sample their high [K/Fe] stars also tend
+In the \citet{Cohen:2012} sample their high [K/Fe] stars also tend
 to be high in heavier elements, but there
@@
 of K are perhaps not as well described by the latent factor model as other
-elements. Still, the specific model scatter is far less than the observed scatter,
-indicative that the model does have some predictive power.
+elements. This could in part be due to the non-trivial and significant effects
+that the assumption of local thermodynamic equilibrium (LTE) has on our inferred K
+abundances. These non-LTE effects are of order $0.5$~dex and will be accounted for in the upcoming \Galah\ data release (S. Buder, private communication).
+
+%Still, the specific model scatter is far less than the observed scatter,
+%indicative that the model does have some predictive power.
 

### 2019-07-22 article/mcfa.tex
notebooks etc

@@
 \usepackage{color}
-\usepackage{comment}
 \usepackage{layouts}
@@
 			 Canberra, ACT 2611, Australia}
-
-%\author{Janez Kos}
-%\affiliation{Sydney Institute for Astronomy, School of Physics,
-%			 A28, The University of Sydney, NSW 2006, Australia}
 
@@
 
-%There are further chemical tagging obstacles to overcome once all the chemical 
-%abundance information in a stellar spectrum can be extracted. Most notably is a 
 Chemical tagging experiments require a catalogue of precise chemical abundance
@@
 chemical abundances for a comparable number of stars.
-
-%These data provide an unparalleled view on the production
-%of chemical elements in the Milky Way.
 
@@
 
-%Alternative approaches are perhaps more challenging, as i
 It is not trivial to confidently estimate
@@
 stars as a contribution of yields from different systems \citep[however see][]{West:2013}.
-%e.g., 20\% of the
-%Ba was produced by 3 M$_\odot$ asymptotic giant branch stars).
 This is in part 
@@
 In this work the data $\vecdataunscaled$ is a 
-$\NumData \times \NumDimensions$ matrix where $\NumData$ is the number of 
+$\NumDimensions \times \NumData$ matrix where $\NumData$ is the number of 
 stars and $\NumDimensions$ is the number of chemical abundances measured 
@@
 
-\noindent{}where $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$ 
+\noindent{}where $\factorloads$ is a $\NumDimensions \times \NumLatentFactors$ 
 matrix of factor loads that is common to all data points, $\NumLatentFactors$ is
@@
 The factor scores for all data points $\factorscores$ is then a 
-$\NumData \times \NumLatentFactors$ matrix, where each data point has a partial
+$\NumLatentFactors \times \NumData$ matrix, where each data point has a partial
 association to the components in latent space. 
@@
 Without loss of generality the density of the mean-subtracted 
-data $\vecdataunscaled - \vec\mu$ (which we hereafter will refer to simply as $\vecdata$) can be described as \citep[Sec. 5.2.9, p236;][]{Wallace:2005}
+data $\vecdataunscaled - \vec\mu$ (which we hereafter will refer to simply as $\vecdata$) can be described as 
 \begin{equation}
 	f(\vecdata; \vec\Psi) = \sum_{\numcomponents=1}^{\NumComponents}\weight_\numcomponents\phi(\vecdata;\factorloads\scoremeans_\numcomponents, \factorloads\scorecovs_\numcomponents\factorloads\transpose + \diag{\specificvariance})
-	%\eye\specificvariance
 \end{equation}
@@
 algorithm \citep{Dempster:1977}.
-% TODO: say something about "..with a suitable number of initial guesses?"
 
@@
 orthogonal vectors.
-%$\mathbf{H}$, and by taking $\factorloads_\ast$ to be the Cholesky decomposition of $\mathbf{H}\transpose \mathbf{H}$, we initialise the factor loads as
-%\begin{eqnarray}
-% 	\factorloads = \mathbf{H}\left(\left(\factorloads_\ast\right)^{-1}\eye\right)
-%\end{eqnarray}
-%\noindent{}which ensures that $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$
-%#matrix of mutually orthogonal vectors.
-%and that it is a valid rotation matrix
-%$\factorloads\transpose \factorloads = \eye$.
+
 We then initially assign each data point as belonging to one of the
@@
 initialise from many points to avoid local minima.
-
-%Throughout this work we found that the optimised model parameters appear
-%to be insensitive to the choice of initialisation up until orthogonal
-%rotation of the latent space.
 
@@
 	\weight_\numcomponents\nextstep = \frac{1}{\NumData} \sum_{\numdata=1}^{\NumData}\tau_{\numdata\numcomponents}
-	% Include proof.
 \end{equation}
@@
 
-\noindent{}where $\odot$ denotes is the entry-wise product. 
-%We repeat
-%the expectation-maximization cycle for up to 10,000 steps or until the log likelihood
-%improved by less than $10^{-5}$ between successive iterations. 
-%Our software implementation allows for a positive regularisation term to be added
-%along the diagonal of covariance matrices in latent space $\scorecovs$, for all experiments
-%In all experiments 
+\noindent{}where $\odot$ denotes the entry-wise product. 
 Throughout this work we assume that the data are noiseless and
 we do not add any observed errors to the constructed covariance matrices.
-%Although our implementation
-%allows for a positive regularisation term to be added along the diagonal of the covariance
-%matrices in latent space $\scorecovs$, we set this regularisation term to zero for all
-%experiments performed here. Similarly, we assume that the data are noiseless and we do
-%not add any observed errors to the constructed covariance matrices.
 
@@
 \noindent{}which includes $\NumComponents - 1$ weights (as $\sum\weight_\numcomponents = 1$), $\NumDimensions$ specific variances, the $\NumComponents\NumLatentFactors - \NumLatentFactors^2$ free parameters needed to uniquely define the mutually orthogonal factor loads matrix $\factorloads$ \citep{Baek:2010}, $\NumComponents \times \NumLatentFactors$ parameters for the mean scores $\scoremeans$, and $\frac{\NumComponents\NumLatentFactors}{2}\left(\NumLatentFactors + 1\right)$ parameters to encode the $\NumComponents$ full ($\NumLatentFactors \times \NumLatentFactors$) covariance matrices $\scorecovs$.
-% weights: K - 1 model parameters.
-% psi: D model parameters
-% A: J * D - J^2 model parameters (Baek et al. 2010)
-% xi: J * K model parameters
-% omega: K * J*(J+1)/2
+
 While the BIC does include a penalisation term for the number of
@@
 uncertainty in which the parameters $\vec\Psi$ are centred. The $\log\kappa_Q$
-term can be reasonably approximately by
+term can be reasonably approximated by
 \begin{equation}
@@
 through the $\log\kappa_Q$ term. However, the model complexity is also 
-described through the priors and the Fisher information, which
+described through the priors and the Fisher information matrix, which
 describes the curvature of the log likelihood with respect to the model
@@
 parameters.
-
-%\todo{Dowe said something about MML reducing to BIC under some conditions, but I have not yet found the right paper to cite.}
-
-%\begin{figure*}
-%	\includegraphics[width=1.0\textwidth]{experiments/exp1-data-colour.pdf}
-%    \caption{The generated data for Experiment~1, where points are coloured by their
-%		     true associations to each component.}
-%    \label{fig:exp1-data}
-%\end{figure*}
 
@@
 \begin{equation}
-	p(\vec\weight) = (\NumComponents - 1)! \quad ,
+	p(\vec\weight) = (\NumComponents - 1)!
 \end{equation}
@@
 \begin{equation}
-	\mathcal{F}(\vec\weight) = \frac{\NumData^{\NumComponents - 1}}{\prod_{\numcomponents=1}^{\NumComponents}\weight_\numcomponents} \quad ,
+	\mathcal{F}(\vec\weight) = \frac{\NumData^{\NumComponents - 1}}{\prod_{\numcomponents=1}^{\NumComponents}\weight_\numcomponents}
 \end{equation}
@@
 
-% todo put line in equation.
 \begin{eqnarray}
@@
 
-%\todo{Fisher:}
-%\begin{widetext}
-%\begin{eqnarray}
-%	\log\likelihood(\vecdata|\vec\Psi_k) = -\frac{1}{2}(\vecdata - \factorloads\vec\xi_k)\transpose(\factorloads\omega_k\factorloads\transpose + \diag{D})^{-1}) - \frac{1}{2}\log|(\factorloads\omega_k\factorloads\transpose + \diag{D})| -\frac{1}{2}\log{2\pi}
-%	\log\likelihood(\vecdata|\vec\Psi) &=& -\frac{1}{2}(\vecdata - \vec{M})\transpose\vec\Sigma^{-1}(\vecdata - \vec{M}) - \frac{1}{2}\log|\vec\Sigma| - \frac{1}{2}\log{2\pi} \quad \mathrm{where} \\
-%	\vec{M} &=& \factorloads\scoremeans \\
-%	\vec\Sigma &=& \factorloads\scorecovs\factorloads\transpose + \diag{\specificvariance}
-%\end{eqnarray}
-%\end{widetext}
-
-%The Fisher information for $\factorloads$ \todo{needs to be double checked.}
 \begin{widetext}
@@
 \begin{eqnarray}
-%	I(\factorloads) &=& -\log\left(\frac{p(\factorloads)}{\sqrt{|\mathcal{F}(\factorloads)|}}\right) \nonumber \\
 I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \quad .
-\label{eq:prior_L}%\nonumber \\
-%	&& \cdots \todo{+\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \quad . \label{eq:prior_L}
+\label{eq:prior_L}
 \end{eqnarray}
@@
 \end{widetext}
-%\vspace{10em}\clearpage
-% \todo{\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}}
-
-% -\log\likelihood(\vec\data|\vec\Psi) \
-% + (J + K + \frac{KN}{2} + \frac{DJ}{2})\log{2}     --> \frac{1}{2}(3(J+K) + ND)\log{2}
-% 0.5*((K-1) logN - \sum\log\weight)  + \frac{1}{4}J(J+3)\sum\log(N\weight) --> 
-% -\log\gamma(K) - \Gamma(D/2)
-% -\frac{1}{2}(2J+3)\sum\log|\omega_k|
-% + factor loads stuff
-% + Q/2\log\kappa(q)
 
@@
 
-%\clearpage
-
 \section{Experiments} \label{sec:experiments}
@@
 $\NumLatentFactors = 5$ factor loads such that the vector $\factorloads$ has
-shape $\NumLatentFactors \times \NumDimensions$,
+shape $\NumDimensions \times \NumLatentFactors$,
 with $\NumComponents = 20$ clusters in the latent space. 
@@
 
-%\begin{figure*}
-%	\includegraphics[width=\textwidth]{experiments/exp1-latent.pdf}
-%    \caption{The inferred latent scores $\factorscores$ for our toy model
-%     with $\NumLatentFactors = 5$ latent factors and
-%		 	 $\NumComponents = 20$ components.
-%             Points are coloured by the inferred component.} 
-%    \label{fig:exp1-latent}
-%\end{figure*}
-
 \begin{figure*}[t!]
@@
 and   $\NumComponents_\textrm{max} = 40$ (e.g., twice the true values of $\NumLatentFactors_\textrm{true}$ and $\NumComponents_\textrm{true}$).
-%For each $\NumLatentFactors$, $\NumComponents$ permutation we initialised the factor
-%loads and component assignments randomly. 
-%We initialised the model as per Section~\ref{sec:initialisation} and performed expectation-maximization cycles until the relative log likelihood improved by less than $10^{-5}$.
 
@@
 clustering in the lower dimensional latent space.
-%The inferred clustering in latent space is shown in Figure~\ref{fig:exp1-latent},
-%where each point is coloured by the inferred associated component.
 
@@
 identifiable up to an orthogonal rotation. That is to say that
-if the data were truely generated by latent factors $\factorloads_\textrm{true}$,
+if the data were truly generated by latent factors $\factorloads_\textrm{true}$,
 then our estimates of those latent factors $\factorloads_\textrm{est}$ do not need
@@
 given the data and the estimated factor loads $\factorloads_\textrm{est}$ 
-%\citep[or Kullback-Leibler divergence;][]{Kullback:1951}
 would be identical to the log likelihood given the data and the true factor loads 
@@
 
-\begin{comment}
-\subsection{Experiment~2:\\Inferring astrophysically realistic latent factors}
-
-In this experiment we use detailed chemical abundances of metal-poor stars 
-from \citet{Barklem:2005} to illustrate the utility and limitations
-in interpreting the latent factors as nucleosynthetic processes. The data 
-can be described as 14 chemical abundance measurements from 61 stars ($\NumData = 61$, $\NumDimensions = 14$), and those 14 chemical elements trace multiple nucleosynthetic pathways.
-Specifically the chemical
-abundances measured include Al (light odd-Z), Mg and Ca ($\alpha$-process), Sc, Ti, Cr, Mn, Fe, Co, and Ni (Fe-peak elements), as well as Sr, Y, Ba, and Eu ($s$/$r$-process). 
-The data vector $\vecdata$ includes [Fe/H] abundance ratios and all other elements
-relative to Fe (e.g., [Ti/Fe]) after subtracting the mean value in each dimension.
-
-We performed a grid search for the optimal number of latent factors 
-$\NumLatentFactors$ and components $\NumComponents$, starting from 
-$\NumLatentFactors = 1$ and $K = 1$, and trialling up until 
-$\NumLatentFactors = 7$ and $\NumComponents = 3$. 
-% Although it is common
-%to add a regularization term to the diagonals of covariance matrices
-%when performing expectation-maximization, we found that our existing
-%grid search boundaries were sufficient to find the preferred $\NumLatentFactors$
-%and $\NumComponents$, and chose not to add any regularization term.
-%We initialised
-%the model parameters randomly for each permutation of $J$ and $K$, and continued
-%the expectation-maximization algorithm
-%until a convergence threshold of $10^{-5}$ was reached in the log likelihood. 
-In Figure~\ref{fig:exp2-gridsearch-contours} we show the negative log likelihood,
-the BIC, and the message length recorded for the optimised set of model parameters 
-for each permutation of $J$ and $K$. The negative log likelihood continues to
-increase with increasing numbers of clusters, and increasing numbers of latent
-factors.
-The lowest BIC value is found for $J = 4$ and $K = 2$, indicating
-four latent factors are preferred with two clusters. The shortest message length
-is also found with four latent factors, but with no clustering ($J = 4$, $K = 1$).
-We adopt the message length as our objective function and chose $J = 4$ and $K = 1$
-as the preferred model.
-
-\begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp2-\ExperimentHash-gridsearch-ll.pdf}
-	\includegraphics[width=0.45\textwidth]{experiments/exp2-\ExperimentHash-gridsearch-bic.pdf}
-	\includegraphics[width=0.45\textwidth]{experiments/exp2-\ExperimentHash-gridsearch-mml.pdf}
-    \caption{The top panel shows the negative log likelihood 
-			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
-			 evaluated at each combination of latent factors 
-			 $\NumLatentFactors$ and number of clusters 
-			 $\NumComponents$ using the \citet{Barklem:2005}
-			 data in Experiment~2.  The middle panel shows 
-			 the BIC (Eq.~\ref{eq:bic}) for those 
-			 combinations, and the lower panel shows the 
-			 message length. The white marker indicates the
-			 lowest value in each panel, showing the
-			 preferred number of latent factors and components.}
-    \label{fig:exp2-gridsearch-contours}
-\end{figure}
-
-%\todo{For example,
-%the first factor does show positive contributions to Al, like the target load,
-%but it also shows a slight negative contribution with respect to alpha elements
-%(Mg, Ca)}, and a gradual increase among Fe-peak elements up to Ni. Similarly while
-%Ca and Mg are indeed positive in the second factor, as targeted, there is a
-%weak negative contribution to Mn abundances, which was not part of the target
-%loads. 
-We applied an orthogonal rotation to the estimated latent factors such that we
-could make qualitative comparisons between experiments (e.g., so we can discuss
-the closest estimated latent factors between different experiments).
-Despite the caveats in interpreting latent factors as nucleosynthetic processes
-(e.g., see Sections~\ref{sec:exp-1} and \ref{sec:discussion}), here we will
-qualitatively describe the latent factors found in our selected model (Figure~\ref{fig:exp2-factor-loads}).
-The absolute\footnote{We discuss non-zero entries because a negative factor loads could have only negative factor scores, and their multiplication would produce a positive abundance.}
-entries of first factor load $\mathbf{L}_0$ has strong non-zero entries at Eu, Ba, Sr, and Al.
-The strong link with Eu is suggestive that this latent factor is consistent with being
-the rapid neutron-capture process.
-However, some of these factor load entries may be non-zero because we require the latent factors to
-be mutually orthogonal, and not because they truly contribute to the data. To try and
-disentangle these possibilities, we calculate the fractional contribution that factor
-load makes to the observed abundances relative to other factor loads. We define the
-fractional contribution of the $j$th factor load to the $d$th
-data dimension as:
-\begin{equation}
-% $|\mathbf{C}_{d}|$
-	\mathbf{C}_\textrm{d,j} = \frac{\sum^{N}|\factorloads_{j,d}\factorscores_{n,j}|}{\sum^J\sum^{N}|\factorloads_{j,d}\factorscores_{n,j}|} \quad .
-\end{equation}
-The fractional contributions to each element are shown in the right hand side of
-Figure~\ref{fig:exp2-factor-loads}.
-Here it is clear that despite that the first factor load $|\mathbf{L}_0|$ has support at Ba and Sr, it is not the dominant factor that contributes to those elements.
-However, we also find that Al abundances are primarily explained by this factor.
-
-The second factor load $\mathbf{L}_1$ closely represents the slow neutron capture process.
-Here the contributions to Ba, Y, and Sr are all highest for this factor load. We
-also find the Fe-peak element Ni to have the highest fractional contribution from this factor,
-but Ni has a near uniform mix of contributions from different factors.
-The third and fourth factors $\mathbf{L}_2$ show non-zero support for a mix of elements
-that are produced by $\alpha$-particle capture reactions. Mg, Ca, and Ti all have their
-highest contributions from $\mathbf{L}_3$, as do Sc and Co, whereas the iron-peak elements
-Cr, Mn, and Fe all have their largest contributions from $\mathbf{L}_2$.
-
-% Sr, Y, Ba, and to a
-%lesser extent, Eu, which is representative of slow neutron capture production.
-%Similarly, the last component shows a very strong positive contribution towards
-%Eu, and could be interpreted as production from the rapid nucleosynthesis process.
-
-\begin{figure*}[t!]
-	\includegraphics[width=\textwidth]{experiments/exp2-\ExperimentHash-latent-factors-and-contributions-with-targets.pdf}
-	\caption{Latent factors found in Experiment~2 using the \citet{Barklem:2005}
-			 data. The panels on the left show the absolute entries for each factor load. 
-			 On the right we show the absolute fractional contributions to
-			 each element, illustrating the loads that contribute most to each element.}
-    \label{fig:exp2-factor-loads}
-\end{figure*}
-
-After accounting for the contributions by the latent factors and scores,
-the specific scatter remaining in each dimension is shown in
-Figure~\ref{fig:exp2-specific-scatter}. The remaining scatter varies from as low 
-as 0.05~dex (Mg, Ca, Ti, Y) up to 0.25~dex (Ba). 
- A large relative scatter might suggest that there are other latent
-factors that contribute to those elements, but those latent factors were not included because
-the message length preferred a fewer number of factors, or it could be limitations in
-the assumptions of our model (e.g., we assume mutually orthogonal latent factors).
-%However, the scatter in 
-%Fe and Ni is somewhat surprising. Those elements usually have strong and numerous atomic
-%lines that can be easily measured in metal-poor stars.
-
-%The relatively wide uncertainty regions on these factors is perhaps not
-%surprising given that only 61 stars were used in this data set. However, this
-%experiment hopefully demonstrates that the rotated latent factors have some
-%interpretability.
-
-\begin{figure}
-	\includegraphics[width=0.5\textwidth]{../experiments/exp2-\ExperimentHash-specific-scatter.pdf}
-	\caption{Specific scatter (e.g., $\sqrt{\specificvariance}$) remaining in the \citet{Barklem:2005} data used in
-			 Experiment~2 after accounting for the contributions by all latent
-			 factors.}
-    \label{fig:exp2-specific-scatter}
-\end{figure}
-
-\end{comment}
-
 \subsection{A toy model with data missing at random}
@@
 up to 23 chemical abundances reported for 342,682 stars.
-%Some abundances are not reported for a variety of 
-%reasons related to astrophysics or the adopted analyses
-%(e.g., a star is too hot to estimate the abundance for a given
-%electronic transition, or the signal-to-noise of the spectrum is too low).
-%Our method as described cannot efficiently handle missing data 
-%entries for a given star, and as such we are forced to exclude stars that
-%do not have a full complement of abundances. However, we also
-%chose to exclude some elemental abundances from our chemical
 In this example we chose to restrict ourselves to stars with a
@@
 (i.e., no missing data entries).
-%tagging experiments. 
-%For example, our experiments
 For example, here we will exclude
-%The element subset should not 
-%with \Galah\ data do not include 
 lithium and carbon abundances because the
@@
 stars with no erroneous flags in all of the following abundances: Na, Al, Si, K, Ca, Sc, Ti, V, Mn, Fe, Ni, Cu, Zn, Y, Ba, La, and Eu.
-%These elements are produced from multiple nucleosynthetic
-%pathways, and 
-%It is this subsample that we will use for 
-%our initial chemical experiment with \Galah\ data.
 These elements were chosen because they trace multiple nucleosynthetic pathways, and they are more commonly reported in the \Galah\ data release, allowing for a larger number of stars with a complete abundance inventory.
 There are 1,072 stars that met these criteria.
-
-%["Mg", "Al", "Ca", "Sc", "Ti", "Cr", "Mn", "Fe", "Co", "Ni", "Y", "Ba", "Eu"]
-\begin{comment}
-We ordered the
-list of chemical abundances by the number of reliable measurements,
-which is shown in Figure~\ref{fig:exp4-abundance-counts}. The number of
-independent abundance measurements (e.g., where there is no reason
-to suspect a single abundance to be untrustworthy) is shown in black,
-and the cumulative number of trustworthy abundances is shown in blue.
-The cumulative line can be interpreted as the number of stars for
-which the full set of elemental abundances (up until and including
-the abundance on the x-axis) is reported. We chose to include the
-18 most commonly reported abundances (up to and including Al),
-leaving us with a sample of \todo{50,000} stars.
-\end{comment}
-
-\begin{comment}
-\begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp3-abundance-counts.pdf}
-	\caption{The number of abundance measurements in the second
-		     \Galah\ data release \citep{Buder:2018} where \texttt{flag\_cannon = 0}
-		     and \texttt{flag\_<element> = 0}. The black line indicates the number
-		     of reported measurements for a single element, and the blue line
-		     indicates the number of stars where all elements leftward (inclusive)
-		     are reported. The dashed vertical line indicates the top 18 reported
-		     elemental abundances, which we use for Experiment~4.}
-    \label{fig:exp3-abundance-counts} 
-\end{figure}
-\end{comment}
 
@@
 \begin{equation}
-% $|\mathbf{C}_{d}|$
 	\mathbf{C}_\textrm{d,j} = \frac{\sum^{N}|\factorloads_{j,d}\factorscores_{n,j}|}{\sum^J\sum^{N}|\factorloads_{j,d}\factorscores_{n,j}|} \quad .
@@
 The exception here is Zn, where a near-equal contribution comes from $\mathbf{L}_5$.
-% Recall that while $\mathbf{L}_5$ was described as our `light odd-Z element' factor load, 
-%it is also the dominant producer (just barely) of V, a Fe-peak element, and secondary contributor to K. 
 The third latent factor $\mathbf{L}_4$ is reasonably well behaved.
@@
 When projected to data space (Figure~\ref{fig:exp3-data-space}) the third component (dark green) is seen to have relatively higher abundance ratios of [K,Ba,Zn/Fe] at a given [Fe/H],
-and lower abundance ratios of [V/Fe].
-This is consistent with the clustering in latent space.%, where $\factorloads_5$ is the dominant contributor to K and Zn, and $\factorloads_{2,3}$ are the dominant contributor to Ba and V, respectively.
+and lower abundance ratios of [V/Fe]. This is consistent with the clustering in latent space.
 
@@
 missing. 
-%We assume those measurements are missing at random, and discuss the
-%impact of that assumption (and others) in Section~\ref{sec:discussion}.
 Specifically we started with the same subset of 1,072 stars in Section~\ref{sec:exp4}
@@
 
-% Something about the clustering in the large example.
-
 \vspace{1em}
 \section{Discussion} \label{sec:discussion}
-
-%Literature studies testing the limitations chemical tagging have shown
-%that the effective dimensionality of chemical abundance space is lower than
-%the number of reported abundances \citep[e.g.,][]{Ting:2012, Price-Jones:2018}. 
-%This is expected. Theoretical models of stellar evolution and nucleosynthesis
-%would suggest that there are of order at most ten important nucleosynthetic 
-%contributors that are responsible for the production of the $\sim$20--30
-%chemical abundances typically reported. As a consequence we can expect that 
-%the observed chemical abundances are correlated in some way (e.g., by some
-%linear combination of a fewer number of nucleosynthetic pathways).  However, 
-%most approaches to chemical tagging to date have assumed that the observed 
-%chemical abundances are independent of each other. This problem cannot be
-%resolved by simply lowering the number of abundances used for chemical tagging (e.g.,
-%down to something near the effective dimensionality), either, because those
-%chemical abundances are still produced by multiple pathways.
 
@@
 
-%Still, the specific model scatter is far less than the observed scatter,
-%indicative that the model does have some predictive power.
-
 A similar argument could be made for Sc, where $\mathbf{L}_1$ -- a factor load we
@@
 
-%The relatively small data set with complete abundances measured in \Galah\ both limits
-%the number of components we can infer, and the efficacy of chemical tagging. Only three
-%components were preferred from the data set of 1,072 stars, which we identified as
-%being stars with low- and high-[$\alpha$/Fe] abundance ratios, and a third component that
-%primarily differs in its abundances of K, Ba, Zn, and V (e.g., given some [Fe/H] and [$\alpha$/Fe]). A worthwhile extension
-%of this work would be to incorporate missing and noisy data, either by imputing the
-%missing data values or by additive variance in the constructed
-%covariance matrices. In doing so would allow for
-%blind chemical tagging on an industrial scale, allowing for the latent factors to
-%be efficiently estimated while chemical tagging occurs in the lower-dimensional
-%latent space.
-
 \section{Conclusions} \label{sec:conclusions}
@@
 those latent factors as being nucleosynthetic yields, our
-our model does provide the first data-driven approach to nucleosynthesis and chemical tagging.
+model does provide the first data-driven approach to nucleosynthesis and chemical tagging.
 We advocate that more data, and the inclusion of weakly informative priors -- joint probabilities using astrometry and a simplified model of the Milky Way -- would help in realising the full potential of chemical tagging.
 
-%With a larger sample of stars that have a complete abundance inventory we would expect to 
-%identify more latent factors and/or components.
-Alternatively, a useful extension of this work would be to allow for missing and/or noisy data, such that realistic data sets of chemical abundances can be used to perform chemical tagging experiments on a larger scale.
-
-% TODO:
 %\vspace{2em}
@@
 people, and pay our respects to elders past and present.
-%This work presents results from the European Space Agency (ESA) space mission Gaia.
-Gaia data is being processed by the Gaia Data Processing and Analysis Consortium (DPAC).
-Funding for the DPAC is provided by national institutions, in particular the institutions participating in the Gaia MultiLateral Agreement (MLA).
-The Gaia mission website is https://www.cosmos.esa.int/gaia .
-The Gaia archive website is https://archives.esac.esa.int/gaia .
 This research has made use of NASA's Astrophysics Data System.
@@
     \package{IPython} \citep{ipython},\,\,
+    \package{Jupyter Notebooks} \citep{jupyter-notebooks},\,\,
     \package{matplotlib} \citep{mpl},\,\,
@@
     \package{scipy} \citep{scipy},\,\,
-%    \package{Stan}\,\,\citep{stan},\,\,
-    \package{Jupyter Notebooks} \citep{jupyter-notebooks}
-}    % TOPCAT
+    \package{TOPCAT} \citep{Taylor:2005}
+}
 

### 2019-07-22 check_cookbook_eq88.py

@@ -1,233 +1,0 @@
-
-
-# Check eq 88 of matrix cookbook
-import numpy as np
-from scipy.optimize import check_grad
-from sklearn.utils import check_random_state
-
-from mcfa.mcfa import _initial_factor_loads_by_random, _initial_components_by_random, _factor_scores, _expectation
-
-N = 100
-D = 15
-J = 3
-K = 2
-
-random_seed = 0
-np.random.seed(random_seed)
-random_state = check_random_state(random_seed)
-
-
-
-
-X = np.random.uniform(-1, 1, size=(D, N))
-A = np.random.uniform(-1, 1, size=(D, J))
-s = np.random.uniform(-1, 1, size=(J, N))
-
-W = np.diag(np.random.uniform(0, 1, size=D))
-
-def func(X, A, s, W, idx):
-    diff = X[:, [idx]] - A @ s[:, [idx]]
-    return -0.5 * diff.T @ W @ diff
-
-def grad(X, A, s, W, idx):
-    diff = X[:, [idx]] - A @ s[:, [idx]]
-    return W @ diff @ s[:, [idx]].T
-
-
-
-def f(A, idx=0):
-    return func(X, A.reshape((D, J)), s, W, idx=idx)
-
-def g(A, idx=0):
-    return grad(X, A.reshape((D, J)), s, W, idx=idx).flatten()
-
-foo = check_grad(f, g, A.flatten())
-
-
-#bar = check_grad(f2, g2, A.flatten())
-
-A = _initial_factor_loads_by_random(X.T, J, random_state)
-pi, xi, omega = _initial_components_by_random(X.T, A, K, random_state)
-psi = np.random.uniform(0.5, 1.5, size=D)
-
-s = _factor_scores(X.T, pi, A, xi, omega, psi)[1].T
-
-def f_p1(A, n=0, k=0):
-    #S = s[:, [n]]
-    S = xi[:, [k]]
-    diff = X[:, [n]] - A @ S
-    O = omega[:, :, k]
-    return -0.5 * diff.T @ np.linalg.inv(A @ O @ A.T + np.diag(psi)) @ diff
-
-
-def g_p1(A, n=0, k=0):
-
-    D_ = np.diag(psi)
-    #S = s[:, [n]]
-    S = xi[:, [k]]
-
-    T0 = X[:, [n]] - A @ S
-    O = omega[:, :, k]
[…]

### 2019-07-22 mcfa/mcfa.py

@@ -323,7 +323,6 @@ class MCFA(object):
         J, K = self.n_latent_factors, self.n_components
         return int((K - 1) + D + J*(D + K) + (K*J*(J+1))/2 - J**2)
 
-
     def expectation(self, X, pi, A, xi, omega, psi, **kwargs):
         r"""
         Compute the conditional expectation of the complete-data log-likelihood

### 2019-07-22 prove_d_ll_wrt_A.py

@@ -1,223 +1,0 @@
-
-import sys
-import numpy as np
-import os
-import matplotlib
-import matplotlib.pyplot as plt
-from scipy import (linalg, optimize as op)
-from matplotlib.ticker import MaxNLocator
-from time import time
-
-sys.path.insert(0, "../../")
-
-from mcfa import (mcfa, mpl_utils, utils)
-
-matplotlib.style.use(mpl_utils.mpl_style)
-colors = plt.rcParams['axes.prop_cycle'].by_key()['color']
-
-np.random.seed(0)
-
-n_features = 10
-n_components = 8
-n_latent_factors = 3
-n_samples = 100
-
-omega_scale = 1
-noise_scale = 1
-random_seed = 1
-
-data_kwds = dict(n_features=n_features,
-                 n_components=n_components,
-                 n_latent_factors=n_latent_factors,
-                 n_samples=n_samples,
-                 omega_scale=omega_scale,
-                 noise_scale=noise_scale,
-                 random_seed=random_seed)
-
-
-Y, truth = utils.generate_data(**data_kwds)
-
-
-
-model = mcfa.MCFA(n_components=n_components, n_latent_factors=n_latent_factors)
-
-pi, A, xi, omega, psi = model._initial_parameters(Y)
-
-
-N, D = Y.shape
-J, K = (model.n_latent_factors, model.n_components)
-
-
-def f(A_flat, Y, pi, xi, omega, psi):
-
-    N, D = Y.shape
-    ll, tau = model.expectation(Y, pi, A_flat.reshape((D, -1)), xi, omega, psi)
-    return np.sum(ll)
-
-
-
-
-def g(A_flat, Y, pi, xi, omega, psi):
-
-    N, D = Y.shape
-    J, _, K = omega.shape
-
-    A = A_flat.reshape((D, J))
-
-    I_psi = np.eye(D) * psi
-    
-    v = np.ones((N, K, D, J))
-    
-    ll, tau = model.expectation(Y, pi, A, xi, omega, psi)
-    
-    for i in range(N):
-
[…]

### 2019-07-22 article/experiments/8761b-exp3.py
things

@@ -28,23 +28,23 @@ matplotlib.style.use(mpl_utils.mpl_style)
 
 here = os.path.dirname(os.path.realpath(__file__))
 
-with open("config.yml") as fp:
+with open("8761b.yml") as fp:
     config = yaml.load(fp)
 
 print(f"Config: {config}")
 
 np.random.seed(config["random_seed"])
 
-prefix = os.path.basename(__file__)[:-3]
+prefix = os.path.basename(__file__)[:-3].split("-")[1]
 
 unique_hash = md5((f"{config}").encode("utf-8")).hexdigest()[:5]
 
-unique_config_path = f"{unique_hash}.yml"
-if os.path.exists(unique_config_path):
-    print(f"Warning: this configuration already exists: {unique_config_path}")
+#unique_config_path = f"{unique_hash}.yml"
+#if os.path.exists(unique_config_path):
+#    print(f"Warning: this configuration already exists: {unique_config_path}")
 
-with open(unique_config_path, "w") as fp:
-    yaml.dump(config, fp)
+#with open(unique_config_path, "w") as fp:
+#    yaml.dump(config, fp)
 
 with open(__file__, "r") as fp:
     code = fp.read()
@@ -308,7 +308,7 @@ fig_fac = mpl_utils.plot_factor_loads_and_contributions(model, X,
                                                         label_names=latex_label_names, colors=colors)
 savefig(fig_fac, "latent-factors-and-contributions")
 
-
+raise a
 
 # Plot clustering in data space and latent space.
 

### 2019-07-22 article/experiments/exp-missing-non-random.py

@@ -1,0 +1,29 @@
+
+
+# Experiment 1 missing not at random.
+
+
+def non_random_missing_mask(N, D, f_total=0.1, f_per_dimension=None):
+
+    missing = np.zeros((N, D), dtype=bool)
+
+    if f_per_dimension is None:
+        f_per_dimension = np.random.uniform(0.1, 0.7, size=D)
+
+
+    M = int(f_total * N * D) # number of missing entries.
+    M_per_D = int(M*f_per_dimension/np.sum(f_per_dimension))
+
+    for d, M_d in enumerate(M_per_D):
+        idx = np.random.choice(N, M_d, replace=False)
+        missing[idx, d] = True
+
+    return missing
+
+
+
+
+
+
+
+

### 2019-07-22 article/experiments/exp-missing-not-at-random.py

@@ -19,6 +19,7 @@ from mcfa import (mcfa, grid_search, mpl_utils, utils)
 matplotlib.style.use(mpl_utils.mpl_style)
 colors = plt.rcParams['axes.prop_cycle'].by_key()['color']
 
+here = os.path.dirname(__file__)
 
 n_features = 30
 n_components = 10
@@ -164,19 +165,19 @@ for missing_data_fraction in (0, 0.05, 0.1, 0.15, 0.2, 0.25, 0.3, 0.35, 0.4, 0.4
     xlabels = [
         r"$\mathbf{L}_\textrm{true}$",
         r"$\mathbf{S}_\textrm{true}$",
-        r"$\mathbf{\Psi}_\textrm{true}$"
+        r"$\mathbf{D}_\textrm{true}$"
     ]
 
     ylabels = [
         r"$\mathbf{L}_\textrm{est}$",
         r"$\mathbf{S}_\textrm{est}$",
-        r"$\mathbf{\Psi}_\textrm{est}$"
+        r"$\mathbf{D}_\textrm{est}$"
     ]
 
     delta_labels = [
         r"$\Delta\mathbf{L}$",
         r"$\Delta\mathbf{S}$",
-        r"$\Delta\mathbf{\Psi}$"
+        r"$\Delta\mathbf{D}$"
     ]
 
     residual_ax_y_limits = [
@@ -249,5 +250,5 @@ for missing_data_fraction in (0, 0.05, 0.1, 0.15, 0.2, 0.25, 0.3, 0.35, 0.4, 0.4
 
     fig.tight_layout()
 
-    fig.savefig(f"exp-missing-data-not-at-random-inflated-{100*missing_data_fraction:.0f}percent.png")
+    fig.savefig(os.path.join(here, f"exp-missing-data-not-at-random-inflated-{100*missing_data_fraction:.0f}percent.png"))
     #fig.savefig(f"exp-missing-data-not-at-random-inflated-{100*missing_data_fraction:.0f}percent.pdf", dpi=300)

### 2019-07-22 article/experiments/exp-missing.py

@@ -109,19 +109,19 @@ for missing_data_fraction in (0, 0.01, 0.05, 0.1, 0.2, 0.3, 0.4, 0.5, 0.75, 0.9)
     xlabels = [
         r"$\mathbf{L}_\textrm{true}$",
         r"$\mathbf{S}_\textrm{true}$",
-        r"$\mathbf{\Psi}_\textrm{true}$"
+        r"$\mathbf{D}_\textrm{true}$"
     ]
 
     ylabels = [
         r"$\mathbf{L}_\textrm{est}$",
         r"$\mathbf{S}_\textrm{est}$",
-        r"$\mathbf{\Psi}_\textrm{est}$"
+        r"$\mathbf{D}_\textrm{est}$"
     ]
 
     delta_labels = [
         r"$\Delta\mathbf{L}$",
         r"$\Delta\mathbf{S}$",
-        r"$\Delta\mathbf{\Psi}$"
+        r"$\Delta\mathbf{D}$"
     ]
 
     idx = 0

### 2019-07-22 article/experiments/exp1.py

@@ -267,9 +267,9 @@ ax.yaxis.set_major_locator(MaxNLocator(5))
 ax_residual.set_xticks([])
 
 
-ax.set_xlabel(r"$\mathbf{\Psi}_\textrm{true}$")
-ax.set_ylabel(r"$\mathbf{\Psi}_\textrm{est}$")
-ax_residual.set_ylabel(r"$\Delta\mathbf{\Psi}$")
+ax.set_xlabel(r"$\mathbf{D}_\textrm{true}$")
+ax.set_ylabel(r"$\mathbf{D}_\textrm{est}$")
+ax_residual.set_ylabel(r"$\Delta\mathbf{D}$")
 
 fig.tight_layout()
 
@@ -302,19 +302,19 @@ ys = [
 xlabels = [
     r"$\mathbf{L}_\textrm{true}$",
     r"$\mathbf{S}_\textrm{true}$",
-    r"$\mathbf{\Psi}_\textrm{true}$"
+    r"$\mathbf{D}_\textrm{true}$"
 ]
 
 ylabels = [
     r"$\mathbf{L}_\textrm{est}$",
     r"$\mathbf{S}_\textrm{est}$",
-    r"$\mathbf{\Psi}_\textrm{est}$"
+    r"$\mathbf{D}_\textrm{est}$"
 ]
 
 delta_labels = [
     r"$\Delta\mathbf{L}$",
     r"$\Delta\mathbf{S}$",
-    r"$\Delta\mathbf{\Psi}$"
+    r"$\Delta\mathbf{D}$"
 ]
 
 idx = 0
@@ -447,8 +447,8 @@ kwds = dict(converged=converged,
             cmap="Spectral_r",
             truth=(J_true, K_true))
 
-fig_ll = mpl_utils.plot_filled_contours(J_grid, K_grid, ll,
-                                        colorbar_label=r"$-\log\mathcal{L}(\boldsymbol{\mathbf{Y}}|\boldsymbol{\mathbf{\Psi}})$", 
+fig_ll = mpl_utils.plot_filled_contours(J_grid, K_grid, -ll,
+                                        colorbar_label=r"$-\log\mathcal{L}(\boldsymbol{\mathbf{Y}}|\boldsymbol{\mathbf{D}})$", 
                                         **kwds)
 fig_ll.axes[0].set_yticks([1, 5, 10, 15, 20])
 
@@ -459,7 +459,7 @@ fig_bic.axes[0].set_yticks([1, 5, 10, 15, 20])
 
 
[…]

### 2019-09-09 article/mcfa.tex
things
foo

@@
 
+\newcommand{\ExpThreeNumAbundances}{17}
+
 \received{2019}
@@
 			 Sydney, NSW 2052, Australia}
+
+\author[0000-0002-0920-809X]{Sanjib Sharma}						 
+\affiliation{Sydney Institute for Astronomy, School of Physics,
+			 A28, The University of Sydney, NSW 2006, Australia}
 
@@
 Using the second \Galah\ data release we find that five latent factors are
-preferred to explain $N =$ 1,072 stars with 18 chemical abundances. 
+preferred to explain $N =$ 1,072 stars with \ExpThreeNumAbundances\ chemical abundances. 
 We identify the rapid- and slow-neutron capture 
@@
 improving upon this precision \citep{Ness:2015,Ness:2018a,Ness:2018b,
-Casey:2016,Casey:2017,Ho:2017b,Ho:2017a,Leung;Bovy:2018}, but more work is
+Casey:2016,Casey:2017,Ho:2017b,Ho:2017a,Leung;Bovy:2018,Ting:2019}, but more work is
 needed: astronomers have not yet developed unbiased estimators of chemical
@@
 quantities, and the effective dimensionality of stellar abundance datasets has been shown
-to be lower than the actual number of abundance dimensions \citep{Ting:2012,Price-Jones:2018}.
+to be lower than the actual number of abundance dimensions \citep{Ting:2012,Price-Jones:2018,Milosavljevic:2018}.
 Any clustering approach that treats each new elemental abundance as an 
@@
 dimensionality reduction tool to identify latent factors that contribute to
-the chemical abundances of stars \citep[e.g.,][]{Ting:2012,Price-Jones:2018}, factor
+the chemical abundances of stars \citep[e.g.,][]{Ting:2012,Price-Jones:2018,Milosavljevic:2018}, factor
 analysis cannot describe clustering in the data (or latent) space. As a result,
@@
 shape $\NumDimensions \times \NumLatentFactors$,
-with $\NumComponents = 20$ clusters in the latent space. 
+with $\NumComponents = 10$ clusters in the latent space. 
 We generated the random factor loads in the same way that we initialise the optimisation (Section~\ref{sec:initialisation}). The relative weights $\vec\weight$
@@
 until $\NumLatentFactors_\textrm{max} = 10$
-and   $\NumComponents_\textrm{max} = 40$ (e.g., twice the true values of $\NumLatentFactors_\textrm{true}$ and $\NumComponents_\textrm{true}$).
+and   $\NumComponents_\textrm{max} = 20$ (e.g., twice the true values of $\NumLatentFactors_\textrm{true}$ and $\NumComponents_\textrm{true}$).
 
@@
 The lowest BIC value and message length is found at $\NumLatentFactors = 5$
-and $\NumComponents = 20$, identical to the true values.
+and $\NumComponents = 10$, identical to the true values.
 It is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
@@
 We took the model with the preferred number of latent factors and components found
-from a grid search ($\NumComponents = 20$, $\NumLatentFactors = 5$; which are also
+from a grid search ($\NumComponents = 10$, $\NumLatentFactors = 5$; which are also
 the true values) and applied an orthogonal rotation to the latent space to be as
@@
 			 The agreement remains excellent,
-			 despite the large fraction of missing data entries.}
+			 despite the large fraction of missing data entries. Note that the scales on the top panels are 2-10 times larger than those in Figure~\ref{fig:exp1-compare}.}
     \label{fig:exp1-missing-data}
@@
 despite that 40\% of the data entries are missing.
+
+\vspace{2em}
 
@@
 	\caption{Latent factors inferred from 1,072 stars in \Galah\
-			 \citep[][thick lines]{Buder:2018} with 17 abundance measurements. Left panels show the absolute entries for each
+			 \citep[][thick lines]{Buder:2018} with \ExpThreeNumAbundances\ abundance measurements. Left panels show the absolute entries for each
 			 factor load, where the thin lines indicate the target latent factors (see Section~\ref{sec:exp4}). On the right we show the absolute fractional contributions
@@
 	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-latent-space.pdf}
-	\caption{The factor scores $\factorscores$ estimated in Experiment~3 (Section~\ref{sec:exp4} using $N =$ 1,072 stars in the \Galah\ data \citep{Buder:2018} that have 18 abundance measurements.
+	\caption{The factor scores $\factorscores$ estimated in Experiment~3 (Section~\ref{sec:exp4} using $N =$ 1,072 stars in the \Galah\ data \citep{Buder:2018} that have \ExpThreeNumAbundances\ abundance measurements.
 Here each star is coloured by its inferred $K$th component.}
@@
 	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-data-space.pdf}
-	\caption{Detailed chemical abundances from the $N =$ 1,072 stars in \Galah\ \citep{Buder:2018} that have 18 chemical abundances (Section~\ref{sec:exp4}).
+	\caption{Detailed chemical abundances from the $N =$ 1,072 stars in \Galah\ \citep{Buder:2018} that have \ExpThreeNumAbundances\ chemical abundances (Section~\ref{sec:exp4}).
 Each star is coloured by its $K$th inferred component from the lower-dimensional latent space, with the same colouring as per Figure~\ref{fig:exp3-latent-space}.}
@@
 	         of \Galah\ data \citep{Buder:2018}. The thin solid line shows the result from
-	         Section~\ref{sec:exp4} with $N=1,072$ stars with 18 abundances and no
+	         Section~\ref{sec:exp4} with $N=1,072$ stars with \ExpThreeNumAbundances\ abundances and no
 	         missing data. Increasing thickness indicates larger samples, up to the solid thick line with $N \sim$ 100,000 stars, where twelve latent factors are preferred.}
@@
 orthogonal, there are scenarios that one can imagine where there is a limited amount
-we can say with confidence about that new nucleosynthetic process.
+we can say with confidence about that new nucleosynthetic process \citep[see also][]{Milosavljevic:2018}.
+
+There are similar limitations that arise due to our assumption about the clustering in latent space.
+There is no justified reason why the factor scores should be well-described by multivariate normal
+distributions. If the \emph{true} underlying scores were not distributed as multivariate normals
+then one can imagine similar outcomes when directly fitting data with a mixture of gaussian distributions:
+additional components are required to describe complex (non-gaussian) shapes in data space. 
+This situation of model mismatch is more extreme when fitting only data rather than the model
+described here because some of the data complexity will be described by the orthogonal latent
+factors. However, qualitatively the picture is the same: when the \emph{true} underlying distribution
+in factor scores are not described by multivariate normals, additional components will likely be
+introduced in order to describe non-gaussian features.
 
@@
 These components can be described as those with (1) low- and (2) high-[$\alpha$/Fe] abundance ratios, and another (3) primarily differing in K, Ba, Zn, and V abundances at a given [Fe/H] and [$\alpha$/Fe] abundance ratio.
-When we include $\sim$100,000 stars with up to 18 abundances, and assume the incomplete abundances are missing at random, we find that 16 components in latent space are preferred to explain the data.
+When we include $\sim$100,000 stars with up to \ExpThreeNumAbundances\ abundances, and assume the incomplete abundances are missing at random, we find that 16 components in latent space are preferred to explain the data.
 By construction these  components

### 2019-09-09 article/experiments/exp13.py

@@ -1,0 +1,420 @@
+
+"""
+Experiment using all GALAH data.
+"""
+
+from __future__ import division # Just in case. Use Python 3.
+
+import os
+import sys
+import pickle
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import yaml
+from matplotlib.ticker import MaxNLocator
+from collections import Counter
+from scipy import linalg
+from hashlib import md5
+
+sys.path.insert(0, "../../")
+
+from mcfa import (mcfa, grid_search, mpl_utils, utils)
+
+import galah_dr2 as galah
+
+
+matplotlib.style.use(mpl_utils.mpl_style)
+
+here = os.path.dirname(os.path.realpath(__file__))
+
+with open("config.yml") as fp:
+    config = yaml.load(fp)
+
+print(f"Config: {config}")
+
+np.random.seed(config["random_seed"])
+
+prefix = os.path.basename(__file__)[:-3]
+
+unique_hash = md5((f"{config}").encode("utf-8")).hexdigest()[:5]
+
+unique_config_path = f"{unique_hash}.yml"
+if os.path.exists(unique_config_path):
+    print(f"Warning: this configuration already exists: {unique_config_path}")
+
+with open(unique_config_path, "w") as fp:
+    yaml.dump(config, fp)
+
+with open(__file__, "r") as fp:
+    code = fp.read()
+
+with open(f"{unique_hash}-{__file__}", "w") as fp:
+    fp.write(code)
+
+def savefig(fig, suffix):
+    here = os.path.dirname(os.path.realpath(__file__))
+    filename = os.path.join(here, f"{prefix}-{unique_hash}-{suffix}")
+    fig.savefig(f"{filename}.png", dpi=150)
+    fig.savefig(f"{filename}.pdf", dpi=300)
+
+
+import os
+os.system("rm -f *.pkl")
+
+N_elements = 20
+use_galah_flags = config["use_galah_flags"]
+
+mcfa_kwds = dict()
+mcfa_kwds.update(config["mcfa_kwds"])
+
+elements = config[prefix]["elements"]
+if config[prefix]["ignore_elements"] is not None:
+    elements = [el for el in elements if el not in config[prefix]["ignore_elements"]]
+
+print(elements)
+
+
+mask = galah.get_abundance_mask(elements, use_galah_flags=use_galah_flags)
+
+
[…]

### 2019-09-09 article/reviewer-response-v0.txt
ref

DONE

1, 2, 3, 5, 7, 8, 10, 11

EASY ITEMS:

4, 6, 9, 12, 13, 14

#4 - Diagram showing how things work.
#6 - Do toy model with 1,000 stars and see what's found.
    -> Running. 
#9 ->
#12 ->
#13 - update Figure 8 to show all abundances.
    -> To be fixed when experiments are re-run.

#14 - Check whether using giant stars in 1,072 sample. If so, remove V.
    -> Running

REVIEWER'S SUMMARY:

This paper presents a new data-driven model of nucleosynthesis which allows the authors to simultaneously 'tag' stars using their chemistry and quantify the contributions of nucleosynthetic events to each stars. The authors achieve this by building a model using a mixture of common factor analyzer (which is a variant of latent factor analysis). The model is essentially built form a set of latent factors (which can be thought of as mean nucleosynthetic yields from specific events), which contribute different amounts (score) to each star. The model also includes clustering lower dimensional latent space as families of elements (e.g. Mg, Si, Ca) can be produced from the same nucleosynthetic event.

The authors first apply their method to a toy model of N=100,000 data points drawn from 20 clusters in a 5-D latent space. The show that their method works. They also show that it continue to work with ~40% of missing data. Once they show it works, they then move on to fitting their model using N=1,072 stars in the GALAH survey which have a full set of 17 elemental abundances. Finally, they build up to a sample of N=100,246 stars within the GALAH survey, where there is missing data. The have a good discussion in Section 4, where they not only talk about how to interpret their result but caution the pitfalls of the method.

The paper is well written and is, in my opinion, certainly worth publication. However, there are some concerns I have which I list below. Hopefully these comments can serve as constructive criticism of the authors work.

1.  REVIEWER:

    Abstract:
    The authors mention they use the 1,072 stars GALAH dataset with 18 chemical abundances yet in section 3.2 (where they discuss the dataset) they actually only list 17 elements. In addition only 17 elements are also shown in Fig. 5, 6, 9. That is to say are there 18 or 17 elements that the authors consider?

    RESPONSE:

    Good catch. There are 17 elements. Corrected throughout.

2. REVIEWER:

    Section 1, end Paragraph 2:
    The authors cite data-driven or other machine learning approaches for modeling stellar spectra but miss that of Ting et al. 2018 (The Payne)

    RESPONSE:

    Citation added.

3. REVIEWER:

    Section 1 Paragraph 4:
    The authors discuss abundance dimensionality in the context of reducing the set of D-abundance to a representative set of abundance dimensions which may group by nucleosynthetic families. It would be good for the authors to also consider discussing the work from Milosavljevic et al., 2018 (arXiv:1809.02660). While the methods and data are different, those authors also attempt to build a data-driven method for understanding nucleosynthetic archetypes in chemical abundance data and is relevant for the introduction and discussion.

    RESPONSE:

    Agreed. We have added references to Milosavljevic et al. (2018) in the introduction, methods, and discussion.

4.  REVIEWER:

    Section 2 : 
    It would be highly beneficial for section 2 to potentially have a schematic (figure) of the method the authors outline. Especially in terms of the structure of how the code might be set up.

    RESPONSE:

5.  REVIEWER:

    Section 2: Paragraph 2:
    The authors state "... but the scoring (or extent) of those factors is different for each data point, and the data can be modelled as a mixture of multivariate normal distributions in the latent space (factor scores)."

    Can the authors comment on what would happen if the "true" underlying distribution of in latent space is not in fact Gaussian?

    RESPONSE:

    We have added a paragraph that addresses this point in Section 4 (Discussion). That paragraph reads:

        There are similar limitations that arise due to our assumption about the clustering in latent space.
        There is no justified reason why the factor scores should be well-described by multivariate normal
        distributions. If the \emph{true} underlying scores were not distributed as multivariate normals
        then one can imagine similar outcomes when directly fitting data with a mixture of gaussian distributions:
        additional components are required to describe complex (non-gaussian) shapes in data space. 
        This situation of model mismatch is more extreme when fitting only data rather than the model
        described here because some of the data complexity will be described by the orthogonal latent
        factors. However, qualitatively the picture is the same: when the \emph{true} underlying distribution
        in factor scores are not described by multivariate normals, additional components will likely be
        introduced in order to describe non-gaussian features.

6.  REVIEWER:

    Section 2:
    The authors state : "... there is a lower dimensional latent space in which the data are clustered, and that clustering is projected into real space by common factor loads" 
    The authors also mention that factor loads can be thought of as the mean yields of a nucleosynthetic event.

    Though we also know that not all Type Ia or Type II nucleosynthetic events can be thought of as the same (i.e. the nucleosynthetic events may not actually be clustered at all). Thus does one expect, given a predicted set yields, that there should be clustering in this latent space?

    RESPONSE:

7.  REVIEWER:

    Section 2.2, Footnote 5:
    The authors are discussing a statistical inconsistency , but note this inconsistency only
    becomes serious with small N. Can they quantify here was is defined as small N. Is it 10s, 100s, 1000s of stars? Particularly important because they only use ~1000 stars in a later section.

    RESPONSE:

    We have amended the footnote to note that here N ~ 30 or fewer.

8.  REVIEWER:

    Section 2.2 : 
    The authors note one of the underlying assumptions of the work is that the data are noiseless. This is, of course, not the case when using the GALAH observations. Therefore, I was surprised in Section 3.3 where the authors use a signal-to-noise ratio (SNR) cut of SNR > 40. For many high-resolution abundance studies SNR > 60 or even as high as 100 are preferred for precise abundances. What motivates the selection of SNR > 40 and more importantly what happens if the authors consider high SNR cuts. This is especially important as they assume the data are noiseless.

    RESPONSE:

    Good point. A spectroscopist might state that you can consider the abundances to be approximately noiseless for
    anything with S/N > 100. Another might say S/N > 500. Another will say > 50. 
    All of these choices are, of course, arbitrary.

    In this work we are introducing a new method that may be suitable for chemical tagging and understanding
    nucleosynthetic yields throughout the Galaxy. Exact numbers vary, but for chemical tagging to be successful it is generally accepted that large numbers of stars (and measured abundances) are necessary. For this reason we wanted to chose S/N (and other constraints) that would leave us with a large enough sample (~100k) such that we can demonstrate the
    computational efficiency of our method, and leave us with a sample where we could reasonably justify noiseless abundances.

    If we consider higher S/N cuts then the sample size reduces. As a consequence, fewer latent factors and components are required to explain those data. Of the factors that are found, they are quantitatively similar to those that are found from using the larger sample of stars. In short: the results stay the same, but using a larger sample of stars means we recover more factors.

    Indeed, one of the reasons we started with small numbers of stars (of high quality) and increased to a larger sample was to demonstrate that similar latent factors are found (Figure 9).

    As a general note: We have updated our code to allow for observational errors, but in its current
    implementation this greatly increases the computational cost (by orders of magnitude). We are
    investigating ways that this can be improved, but this is outside the scope of this paper.

9.  REVIEWER:

    Section 3.1 :
    The authors build a toy model in which they sample 100,000 'stars' using 20 clusters in a 5-D factor load space. The toy model seems to work (as per Fig.1). However when they move on to real data in section 3.3 they first only use a sample size of 1,000 (1% the sample size in the toy model) stars. It would be good for the authors to recompute their toy model sampling N=100,000 and N=1,000 'stars' and add this to Figure 1 to better illustrate the effect of lowering the sample size.

    RESPONSE:

10. REVIEWER:

    Figure 1 :
    The text notes that the toy model has 5 factor loads (i.e. J=5) with 20 clusters (i.e. K=20). However, in Figure 1, the 'true value' in K vs J (i.e. their black point in Fig 1) seems to be at J=5 and K= 10. This is contrary to K=20 as described in the text. Can the authors state why the true value of K in Fig. 1 does not match the value from the text?

    RESPONSE:

    We had previously used J = 10 and K = 20 as the true values for the toy experiment. We failed to update the text to show that J = 5 and K = 10 are the true values. We have corrected this.

11. REVIEWER:

    Figure 3 : 
    To properly assess the impact of 40% vs 0 % (randomly) missing data, it would be good if the scales in each of the 3 panels matched between Fig 2 and Fig 3. For example, currently the left panel of Fig 2 has scales from -0.025 < deltaL < +0.025 while in Fig 3, these scales are a factor of 10 larger (-0.2 < deltaL < +0.25). Either the same scale should be used or the authors should clarify in the text that the scales are different.

    RESPONSE:

    The scales are 2-10 times larger in Figure 3 than Figure 2. Context is lost when we scale Figure 2 to have the same range as Figure 3.

    We have added to the caption in Figure 3:

    "Note that the scales on the top panels are 2-10 times larger than those in Figure~\ref{fig:exp1-compare}."

12. REVIEWER:

    Section 3.2 : 
    The authors specify target latent factors in Section 3.2. The fourth latent factor has been identify as alpha tracers which include Si, Ca, and Ti. However, Ti is not apart of the alpha elements from a nucleosynthetic lens (e.g. see most recently Curtis S.,et al., 2019,ApJ, 870,2) for a discussion on this). Ti is more of an Fe-peak element rather than an alpha element. I would suggest removing Ti and adding it to the Fe-peak elements.

    Additionally, Sc, V, and Cu represent odd-Z (typically not the same as Fe-peak) elements.

    Since interpretation of the results rely somewhat on the target latent factors, it would be good to physically motivate the choices of the target latent factors.

    RESPONSE:

13. REVIEWER:

    Fig. 8:
    It would be good to show all 17 elements rather than only the 15 elements shown by the authors.

    RESPONSE:

14. REVIEWER:

    Fig 8 :
    The authors use Vanadium in their analysis. From Buder et al. 2018 (see their Fig. 23), the [V/Fe] behaves very differently when comparing dwarfs and giant stars. The [V/Fe] in this paper (Fig. 8) to me most resembles the [V/Fe] vs [Fe/H] for the giant stars in GALAH and not the dwarfs.

    However, Buder et al. 2018 states "For V, we see a significant disagreement with both the dwarf abundances and APOGEE DR14. We report this element anyway, because of the useful abundances in dwarfs, but advise not to use of [V/Fe] in giants."

    Are the authors largely using giant stars in the 1,072 stars from GALAH or is this sample dominated by dwarfs? If they there are any giant stars within these 1,072 stars the authors should disregard the V abundances as per Buder et al. 2018.

    RESPONSE:

### 2019-09-11 article/reviewer-response-v0.txt
break all the things

@@
-DONE
+RESPONSE SUMMARY:
 
-1, 2, 3, 5, 7, 8, 10, 11
-
-EASY ITEMS:
-
-4, 6, 9, 12, 13, 14
-
-#4 - Diagram showing how things work.
-#6 - Do toy model with 1,000 stars and see what's found.
-    -> Running. 
-#9 ->
-#12 ->
-#13 - update Figure 8 to show all abundances.
-    -> To be fixed when experiments are re-run.
-
-#14 - Check whether using giant stars in 1,072 sample. If so, remove V.
-    -> Running
+ 1. Done!
+ 2. Done!
+ 3. Done!
+ 4. Need to make a schematic diagram and write response.
+ 5. Done!
+ 6. Done!
+ 7. Done!
+ 8. Need to justify S/N choice --> text will depend on #12.
+ 9. Need to add text about N ~ 1000 stars in the toy model (awaiting figure to finish)
+10. Done!
+11. Done!
+12. Waiting for eval-5.py to finish, then consider re-running with S/N > 40.
+13. Done!
+14. Done!
 
@@
 
-    Good catch. There are 17 elements. Corrected throughout.
+    There were, and are now, 17 elements. We have dropped V (as per #14) and added Mg. Updated text throughout.
 
@@
 
-    Agreed. We have added references to Milosavljevic et al. (2018) in the introduction, methods, and discussion.
+    We have added references to Milosavljevic et al. (2018) in the introduction, methods, and discussion.
 
@@
         then one can imagine similar outcomes when directly fitting data with a mixture of gaussian distributions:
-        additional components are required to describe complex (non-gaussian) shapes in data space. 
+        additional components would be required to describe complex (non-gaussian) shapes in data space. 
         This situation of model mismatch is more extreme when fitting only data rather than the model
@@
     RESPONSE:
+
+    Conditioned on our other assumptions explicitly described in the manuscript, we would expect there to be 
+    clustering in latent space unless the yields produced by every nucleosynthetic mechanism (including Type Ia supernovae, Type II supernovae, AGB stars, etc) are entirely stochastic.
+
+    Even if Type Ia and Type II supernovae differ in the yields per event, there will still be some mean yield from each type of event and some intrinsic scatter around that mean.
+The differing means and intrinsic scatter would appear as clustering in latent space.
+If the intrinsic scatter among each type of event were so large that event types could not be distinguished then we would recover only one (broad) component in latent space.
 
@@
 
+    %%% ARC also note this is S/N > 40 in blue. what does that correspond to elsewhere?
+
 9.  REVIEWER:
@@
     RESPONSE:
+
+    We have updated Figure 1 to show the grid search using N = 100,000 'stars' and N = 1,000 'stars'. We have also updated the text (see PDF changes).
+
+    % ARC comment on changes
 
@@
 
-    The scales are 2-10 times larger in Figure 3 than Figure 2. Context is lost when we scale Figure 2 to have the same range as Figure 3.
-
-    We have added to the caption in Figure 3:
+    The scales are 2-10 times larger in Figure 3 than Figure 2. Context is lost when we scale Figure 2 to have the same range as Figure 3. Instead we have added to the caption in Figure 3:
 
@@
 
+    %%% ARC UPDATE AS REQUESTED.
+
 13. REVIEWER:
@@
     RESPONSE:
+
+    This was a bug in our plotting code. Figure 8 has been updated to include all 17 elements (16 as [X/Fe] on y-axis and [Fe/H] as common x-axis).
 
@@
     RESPONSE:
+
+    The sample included many giant stars. We have removed V from the set of abundances used and updated the text accordingly.

### 2019-09-11 article/mcfa.tex
break all the things

@@
 \usepackage{color}
-\usepackage{layouts}
+%\usepackage{layouts}
+
+\usepackage{minted}
+\usepackage{graphicx}
+%\usepackage{subfig}
 
@@
 \newcommand{\ExperimentHash}{89dab}
+\newcommand{\EvaluationHash}{c7d68}
 
@@
 of the log likelihood (Eq.~\ref{eq:log-likelihood}) to zero and solving for
-the parameter values.\footnote{Strictly this introduces a statistical inconsistency in that we should update our parameter estimates by setting the second derivative of our information-theoretic objective function (Eq.~\ref{eq:message-length}) to zero instead of the log likelihood, but this inconsistency only becomes serious with small $N$ -- precisely the opposite situation of chemical tagging!}
+the parameter values.\footnote{Strictly this introduces a statistical inconsistency in that we should update our parameter estimates by setting the second derivative of our information-theoretic objective function (Eq.~\ref{eq:message-length}) to zero instead of the log likelihood, but this inconsistency only becomes serious with small $N$ (e.g., $\approx 30$) -- precisely the opposite situation of chemical tagging!}
 In doing so this guarantees that every updated
@@
 
-\begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-ll-contours.pdf}
-	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-bic-contours.pdf}
-	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-mml-contours.pdf}
-    \caption{The top panel shows the negative log likelihood 
-			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
-			 evaluated at each combination of latent factors 
-			 $\NumLatentFactors$ and number of clusters 
-			 $\NumComponents$ using the generated data in our toy model.
-			 The middle panel shows 
-			 the BIC (Eq.~\ref{eq:bic}) for those 
-			 combinations, and the lower panel shows the 
-			 message length. The white marker indicates the
-			 lowest value in each panel.
-			 The black marker indicates the true value. 
-			 In the top panel a line connects these markers to guide the eye.}
-    \label{fig:experiment-1-gridsearch}
-\end{figure}
+%\begin{figure}
+%	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-ll-contours.pdf}
+%	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-bic-contours.pdf}
+%	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-mml-contours.pdf}
+%    \caption{The top panel shows the negative log likelihood 
+%			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
+%			 evaluated at each combination of latent factors 
+%			 $\NumLatentFactors$ and number of clusters 
+%			 $\NumComponents$ using the generated data in our toy model.
+%			 The middle panel shows 
+%			 the BIC (Eq.~\ref{eq:bic}) for those 
+%			 combinations, and the lower panel shows the 
+%			 message length. The white marker indicates the
+%			 lowest value in each panel.
+%			 The black marker indicates the true value. 
+%			 In the top panel a line connects these markers to guide the eye.}
+%    \label{fig:experiment-1-gridsearch}
+%\end{figure}
 
@@
 
-\begin{figure*}[t!]
-	\includegraphics[width=\textwidth]{experiments/exp1-compare-all.pdf}
-    \caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
-    		 and specific variances $\specificvariance$ (right) compared to the 
-		 	 true data generating values
-		 	 for Experiment~1 (Section~\ref{sec:exp-1}). The agreement is excellent.}
-    \label{fig:exp1-compare}
-\end{figure*}
-
 We treat the generated data set as if the number of latent factors
@@
 and   $\NumComponents_\textrm{max} = 20$ (e.g., twice the true values of $\NumLatentFactors_\textrm{true}$ and $\NumComponents_\textrm{true}$).
+
+\begin{figure*}
+	\centering
+	\begin{tabular}[b]{@{}p{0.45\textwidth}@{}}
+		\centering\includegraphics[width=\linewidth]{experiments/eval-figs/eval-1-gridsearch-ll-contours.pdf} \\
+	\end{tabular}
+	\begin{tabular}[b]{@{}p{0.45\textwidth}@{}}
+		\centering\includegraphics[width=\linewidth]{experiments/eval-figs/eval-2-gridsearch-ll-contours.pdf} \\
+	\end{tabular}
+
+	\begin{tabular}[b]{@{}p{0.45\textwidth}@{}}
+		\centering\includegraphics[width=\linewidth]{experiments/eval-figs/eval-1-gridsearch-bic-contours.pdf} \\
+	\end{tabular}
+	\begin{tabular}[b]{@{}p{0.45\textwidth}@{}}
+		\centering\includegraphics[width=\linewidth]{experiments/eval-figs/eval-2-gridsearch-bic-contours.pdf} \\
+	\end{tabular}
+
+	\begin{tabular}[b]{@{}p{0.45\textwidth}@{}}
+		\centering\includegraphics[width=\linewidth]{experiments/eval-figs/eval-1-gridsearch-mml-contours.pdf} \\
+	\end{tabular}
+	\begin{tabular}[b]{@{}p{0.45\textwidth}@{}}
+		\centering\includegraphics[width=\linewidth]{experiments/eval-figs/eval-2-gridsearch-mml-contours.pdf} \\
+	\end{tabular}
+    \caption{Metrics from our grid search for the toy model for two sample sizes: 
+    		 $N = 100,000$ (left) and $N = 5,000$ (right). 
+    		 The top panels show the negative log likelihood
+			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
+			 evaluated at each combination of latent factors 
+			 $\NumLatentFactors$ and number of clusters 
+			 $\NumComponents$ using the generated data in our toy model.
+			 The middle panels shows 
+			 the BIC (Eq.~\ref{eq:bic}) for those 
+			 combinations, and the lower panel shows the 
+			 message length. The white marker indicates the
+			 lowest value in each panel. A line connects to the true value (black point) to guide the eye.}
+    \label{fig:experiment-1-gridsearch}
+\end{figure*}
 
@@
 
+\begin{figure*}[t!]
+	%\includegraphics[width=\textwidth]{experiments/exp1-compare-all.pdf}
+	\includegraphics[width=\textwidth]{experiments/eval-figs/eval-1-compare-all.pdf}
+	\caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
+    		 and specific variances $\specificvariance$ (right) compared to the 
+		 	 true data generating values
+		 	 for Experiment~1 (Section~\ref{sec:exp-1}). The agreement is excellent.}
+    \label{fig:exp1-compare}
+\end{figure*}
+
 Some technical background is warranted before we compare our estimated model
@@
 \begin{figure*}[t!]
-	\includegraphics[width=\textwidth]{experiments/exp-missing-data-not-at-random-inflated-40percent.png}
+	%\includegraphics[width=\textwidth]{experiments/exp-missing-data-not-at-random-inflated-40percent.png}
+	\includegraphics[width=\textwidth]{experiments/eval-figs/eval-3-missing-data-40.pdf}
 	\caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
@@
 \begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp3-\ExperimentHash-gridsearch-ll.pdf}
-	\includegraphics[width=0.45\textwidth]{experiments/exp3-\ExperimentHash-gridsearch-bic.pdf}
-	\includegraphics[width=0.45\textwidth]{experiments/exp3-\ExperimentHash-gridsearch-mml.pdf}
+%	\includegraphics[width=0.45\textwidth]{experiments/exp3-\ExperimentHash-gridsearch-ll.pdf}
+	%\includegraphics[width=0.45\textwidth]{experiments/exp3-\ExperimentHash-gridsearch-bic.pdf}
+	%\includegraphics[width=0.45\textwidth]{experiments/exp3-\ExperimentHash-gridsearch-mml.pdf}
+
+	%TODO: Re-do this image and show the correct colorbar to be consistent with Fig 1
+	\includegraphics[width=0.45\textwidth]{experiments/eval-figs/eval-4-\EvaluationHash-gridsearch-mml.pdf}
     \caption{The top panel shows the negative log likelihood 
@@
 \begin{figure*}
-	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-latent-factors-and-contributions-with-targets.pdf}
+	%\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-latent-factors-and-contributions-with-targets.pdf}
+	\includegraphics[width=\textwidth]{experiments/eval-figs/eval-4-\EvaluationHash-latent-factors-and-contributions-with-targets.pdf}
 	\caption{Latent factors inferred from 1,072 stars in \Galah\
@@
 \begin{figure}
-	\includegraphics[width=0.45\textwidth]{experiments/exp3-\ExperimentHash-specific-scatter.pdf}
+	%\includegraphics[width=0.45\textwidth]{experiments/exp3-\ExperimentHash-specific-scatter.pdf}
+	\includegraphics[width=0.45\textwidth]{experiments/eval-figs/eval-4-\EvaluationHash-specific-scatter.pdf}
 	\caption{Specific scatter (e.g., $\sqrt{\specificvariance}$) remaining in the \Galah\ data \citep{Buder:2018}
@@
 \begin{figure*}
-	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-latent-space.pdf}
+	%\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-latent-space.pdf}
+	\includegraphics[width=\textwidth]{experiments/eval-figs/eval-4-\EvaluationHash-latent-space.pdf}
 	\caption{The factor scores $\factorscores$ estimated in Experiment~3 (Section~\ref{sec:exp4} using $N =$ 1,072 stars in the \Galah\ data \citep{Buder:2018} that have \ExpThreeNumAbundances\ abundance measurements.
@@
 \begin{figure*}
-	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-data-space.pdf}
+	%\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-data-space.pdf}
+	\includegraphics[width=\textwidth]{experiments/eval-figs/eval-4-\EvaluationHash-data-space.pdf}
 	\caption{Detailed chemical abundances from the $N =$ 1,072 stars in \Galah\ \citep{Buder:2018} that have \ExpThreeNumAbundances\ chemical abundances (Section~\ref{sec:exp4}).
@@
 then one can imagine similar outcomes when directly fitting data with a mixture of gaussian distributions:
-additional components are required to describe complex (non-gaussian) shapes in data space. 
+additional components would be required to describe complex (non-gaussian) shapes in data space. 
 This situation of model mismatch is more extreme when fitting only data rather than the model
@@
 model does provide the first data-driven approach to nucleosynthesis and chemical tagging.
-We advocate that more data, and the inclusion of weakly informative priors -- joint probabilities using astrometry and a simplified model of the Milky Way -- would help in realising the full potential of chemical tagging.
-
-%\vspace{2em}
+We advocate that more data, and the inclusion of weakly informative priors -- joint probabilities using astrometry and a simplified model of the Milky Way -- would help in realising the full potential of chemical tagging.\\
+
+\vspace{2em}
+
 %This \emph{Article} was compiled from the \texttt{git} commit \todo{\texttt{HASH}}.
@@
 
+\appendix
+
+Documentation for the software that accompanies this paper is available at \url{https://mcfa.rtfd.io}. 
+Below we provide code that generates ficticious data from a toy model and fits it.\\
+
+\begin{minted}[mathescape,linenos,numbersep=5pt]{python}
+import numpy as np
+from mcfa import (mcfa, grid_search, mpl_utils, utils)
+
+np.random.seed(42)
+
+do_grid_search = (np.random.uniform() > 0.5)
+
+# Generate data.
+X, true_theta = utils.generate_data(n_samples=1000,
+                                    n_features=15,
+                                    n_components=10,
+                                    n_latent_factors=5)
+
+# The data, X, has shape (1000, 15)
+assert X.shape == (1000, 15)
+
+if do_grid_search:
+    # Perform a grid search for the number of components and latent factors.
+    J_trial = np.arange(1, 11) # trial from 1 to 10 latent factors 
+    K_trial = np.arange(1, 21) # trial from 1 to 20 components
+    J_grid, K_grid, converged, metrics = grid_search.grid_search(J_trial, K_trial, X)
+
+    # Return the model with the smallest message length
+    model = metrics["best_models"]["mml"]
+
+else:
+    # Or just fit the data given a number of components and latent factors.
+    model = mcfa.MCFA(n_components=10, n_latent_factors=5)
+    model.fit(X)
+
+# Fitting quantities have a _ suffix
+tau = model.tau_ # responsibility matrix
+theta = dict(zip(model.parameter_names, model.theta_)) # model parameters
+
+# Plot the factors.
+fig_factors = mpl_utils.plot_factor_loads_and_contributions(model, X)
+
+# Plot latent space.
+fig_latent = mpl_utils.plot_latent_space(model, X)
+
+# Plot data space, coloured by most probable component.
+fig_data = mpl_utils.plot_data_space(model, X)
+\end{minted}
+
 \bibliographystyle{aasjournal}

### 2019-09-11 mcfa/grid_search.py

@@ -7,7 +7,7 @@ from .mcfa import MCFA
 
 logger = logging.getLogger(__name__)
 
-def grid_search(trial_n_latent_factors, trial_n_components, X, N_inits=1,
+def grid_search(trial_n_latent_factors, trial_n_components, X, N_inits=25,
                 suppress_exceptions=True, mcfa_kwds=None, pseudo_bic_kwds=None):
 
     Js = np.array(trial_n_latent_factors)

### 2019-09-11 mcfa/mcfa.py

@@ -174,8 +174,8 @@ class MCFA(object):
 
         Y = np.atleast_2d(X).copy()
 
-        if not np.all(np.isfinite(Y)):
-            logger.warn("Non-finite data points will be treated as missing data at random.")
+        #if not np.all(np.isfinite(Y)):
+        #    logger.warn("Non-finite data points will be treated as missing data at random.")
     
         N, D = Y.shape
         if D > N:
@@ -363,8 +363,10 @@ class MCFA(object):
             component in the mixture.
         """
 
-        ll, tau = _expectation(X, pi, A, xi, omega, psi, verbose=self.verbose,
-                               covariance_regularization=self.covariance_regularization, **kwargs)
+        kw = dict(verbose=self.verbose,
+                  covariance_regularization=self.covariance_regularization)
+        
+        ll, tau = _expectation(X, pi, A, xi, omega, psi, **{**kw, **kwargs})
         self.log_likelihoods_.append(ll)
 
         return (ll, tau)
@@ -415,11 +417,11 @@ class MCFA(object):
             and the variance in each dimension :math:`\psi`.
         """
 
-        return _maximization(X, tau, pi, A, xi, omega, psi, 
-                             X2=self._check_precomputed_X2(X, **kwargs),
-                             covariance_regularization=self.covariance_regularization,
-                             **kwargs)
+        kw = dict(X2=self._check_precomputed_X2(X, **kwargs),
+                  covariance_regularization=self.covariance_regularization)
 
+        return _maximization(X, tau, pi, A, xi, omega, psi, **{**kw, **kwargs})
+        
 
     def _check_convergence(self, previous, current):
         r"""
@@ -439,14 +441,11 @@ class MCFA(object):
             the current cost relative to the previous cost.
         """
 
-        #assert np.isfinite(current)
         if not np.isfinite(current):
-            logger.warn("Non-finite log-likelihood.")
+            logger.warn("Non-finite log likelihood.")
 
         if previous > current:
-            logger.warn(f"Log-likelihood *decreased* by {previous-current:.2e}")
-
-        #assert current > previous # depends on objective function
+            logger.warn(f"Log likelihood *decreased* by {previous-current:.2e}")
 
         ratio = abs((current - previous)/current)
         converged = (self.tol >= ratio) and previous < current
@@ -491,7 +490,7 @@ class MCFA(object):
             initial_lls.append(ll)
[…]

### 2019-09-11 mcfa/utils.py

@@ -7,7 +7,7 @@ def whiten(X, axis=0):
 
 
 def generate_data(n_samples, n_features, n_latent_factors, n_components,
-                  omega_scale, noise_scale, random_seed=0, force=None):
+                  omega_scale=1, noise_scale=1, random_seed=0, force=None):
 
     rng = np.random.RandomState(random_seed)
 

### 2019-09-11 article/experiments/eval-1.py
foo

@@ -1,0 +1,476 @@
+
+import sys
+import numpy as np
+import os
+import matplotlib
+import matplotlib.pyplot as plt
+from matplotlib.ticker import MaxNLocator
+from time import time
+from scipy import linalg
+import yaml
+
+sys.path.insert(0, "../../")
+
+from mcfa import (mcfa, grid_search, mpl_utils, utils)
+
+matplotlib.style.use(mpl_utils.mpl_style)
+colors = plt.rcParams['axes.prop_cycle'].by_key()['color']
+
+
+with open("config.yml") as fp:
+    config = yaml.load(fp)
+
+prefix = os.path.basename(__file__)[:-3]
+this_config = config.get(prefix, dict())
+data_kwds = this_config.get("data_kwds", dict())
+mcfa_kwds = this_config.get("mcfa_kwds", dict())
+
+D = data_kwds["n_features"]
+K = data_kwds["n_components"]
+J = data_kwds["n_latent_factors"]
+N = data_kwds["n_samples"]
+
+print(f"Data kwds: {data_kwds}")
+print(f"MCFA kwds: {mcfa_kwds}")
+
+def savefig(fig, suffix):
+    prefix = os.path.basename(__file__)[:-3]
+    here = os.path.dirname(os.path.realpath(__file__))
+    filename = os.path.join(here, "eval-figs", f"{prefix}-{suffix}")
+    fig.savefig(f"{filename}.pdf", dpi=300)
+    fig.savefig(f"{filename}.png", dpi=150)
+    print(f"Created figures {filename}.png and {filename}.pdf")
+
+
+Y, truth = utils.generate_data(**data_kwds)
+truth_packed = (truth["pi"], truth["A"], truth["xi"], truth["omega"], truth["psi"])
+
+
+# Fit with true number of latent factors and components.
+model = mcfa.MCFA(n_components=data_kwds["n_components"],
+                  n_latent_factors=data_kwds["n_latent_factors"],
+                  **mcfa_kwds)
+tick = time()
+model.fit(Y)
+tock = time()
+
+model.message_length(Y)
+
+print(f"Model took {tock - tick:.1f} seconds")
+
+
[…]

### 2019-09-11 article/experiments/eval-2.py

@@ -1,0 +1,477 @@
+
+import sys
+import numpy as np
+import os
+import matplotlib
+import matplotlib.pyplot as plt
+from matplotlib.ticker import MaxNLocator
+from time import time
+from scipy import linalg
+import yaml
+
+
+sys.path.insert(0, "../../")
+
+from mcfa import (mcfa, grid_search, mpl_utils, utils)
+
+matplotlib.style.use(mpl_utils.mpl_style)
+colors = plt.rcParams['axes.prop_cycle'].by_key()['color']
+
+
+
+with open("config.yml") as fp:
+    config = yaml.load(fp)
+
+prefix = os.path.basename(__file__)[:-3]
+this_config = config.get(prefix, dict())
+data_kwds = this_config.get("data_kwds", dict())
+mcfa_kwds = this_config.get("mcfa_kwds", dict())
+
+D = data_kwds["n_features"]
+K = data_kwds["n_components"]
+J = data_kwds["n_latent_factors"]
+N = data_kwds["n_samples"]
+
+print(f"Data kwds: {data_kwds}")
+print(f"MCFA kwds: {mcfa_kwds}")
+
+
+def savefig(fig, suffix):
+    prefix = os.path.basename(__file__)[:-3]
+    here = os.path.dirname(os.path.realpath(__file__))
+    filename = os.path.join(here, "eval-figs",  f"{prefix}-{suffix}")
+    fig.savefig(f"{filename}.pdf", dpi=300)
+    fig.savefig(f"{filename}.png", dpi=150)
+    print(f"Created figures {filename}.png and {filename}.pdf")
+
+
+Y, truth = utils.generate_data(**data_kwds)
+truth_packed = (truth["pi"], truth["A"], truth["xi"], truth["omega"], truth["psi"])
+
+
+# Fit with true number of latent factors and components.
+model = mcfa.MCFA(n_components=data_kwds["n_components"],
+                  n_latent_factors=data_kwds["n_latent_factors"],
+                  **mcfa_kwds)
+tick = time()
+model.fit(Y)
+tock = time()
+
+model.message_length(Y)
+
+print(f"Model took {tock - tick:.1f} seconds")
+
+
[…]

### 2019-09-11 article/experiments/eval-3.py

@@ -1,0 +1,246 @@
+
+# Experiment with missing data when the data are not missing at random.
+
+import sys
+import numpy as np
+import os
+import matplotlib
+import matplotlib.pyplot as plt
+from scipy import linalg
+from matplotlib.ticker import MaxNLocator
+from matplotlib import gridspec
+import yaml
+
+from time import time
+
+sys.path.insert(0, "../../")
+
+from mcfa import (mcfa, grid_search, mpl_utils, utils)
+
+matplotlib.style.use(mpl_utils.mpl_style)
+colors = plt.rcParams['axes.prop_cycle'].by_key()['color']
+
+here = os.path.dirname(__file__)
+
+
+
+with open("config.yml") as fp:
+    config = yaml.load(fp)
+
+prefix = os.path.basename(__file__)[:-3]
+this_config = config.get(prefix, dict())
+data_kwds = this_config.get("data_kwds", dict())
+mcfa_kwds = this_config.get("mcfa_kwds", dict())
+
+D = data_kwds["n_features"]
+K = data_kwds["n_components"]
+J = data_kwds["n_latent_factors"]
+N = data_kwds["n_samples"]
+
+print(f"Data kwds: {data_kwds}")
+print(f"MCFA kwds: {mcfa_kwds}")
+
+
+
+uniform_missing_probabilities = False
+fit_kwds = dict(n_inits=20)
+
+
+Y, truth = utils.generate_data(**data_kwds)
+truth_packed = (truth["pi"], truth["A"], truth["xi"], truth["omega"], truth["psi"])
+
+
+def savefig(fig, suffix):
+    prefix = os.path.basename(__file__)[:-3]
+    here = os.path.dirname(os.path.realpath(__file__))
+    filename = os.path.join(here, "eval-figs", f"{prefix}-{suffix}")
+    fig.savefig(f"{filename}.pdf", dpi=300)
+    fig.savefig(f"{filename}.png", dpi=150)
+    print(f"Created figures {filename}.png and {filename}.pdf")
+
+
+if uniform_missing_probabilities:
+    # Make uniform probabilities
+    fractional_prob_missing = np.ones(data_kwds["n_features"])
+    fractional_prob_missing /= np.sum(fractional_prob_missing)
+    print("Doing uniform missing probabilities per dimension")
+
+else:
+
+    # Sort the missing data probabilities by the psi values.
[…]

### 2019-09-11 article/experiments/eval-4.py

@@ -1,0 +1,374 @@
+
+"""
+Experiment using all GALAH data.
+"""
+
+from __future__ import division # Just in case. Use Python 3.
+
+import os
+import sys
+import pickle
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import yaml
+from matplotlib.ticker import MaxNLocator
+from collections import Counter
+from scipy import linalg
+from hashlib import md5
+
+sys.path.insert(0, "../../")
+
+from mcfa import (mcfa, grid_search, mpl_utils, utils)
+
+import galah_dr2 as galah
+
+
+matplotlib.style.use(mpl_utils.mpl_style)
+
+here = os.path.dirname(os.path.realpath(__file__))
+
+with open("config.yml") as fp:
+    config = yaml.load(fp)
+
+print(f"Config: {config}")
+
+np.random.seed(config["random_seed"])
+
+prefix = os.path.basename(__file__)[:-3]
+
+unique_hash = md5((f"{config}").encode("utf-8")).hexdigest()[:5]
+
+unique_config_path = f"{unique_hash}.yml"
+if os.path.exists(unique_config_path):
+    print(f"Warning: this configuration already exists: {unique_config_path}")
+
+with open(unique_config_path, "w") as fp:
+    yaml.dump(config, fp)
+
+with open(__file__, "r") as fp:
+    code = fp.read()
+
+with open(f"{unique_hash}-{__file__}", "w") as fp:
+    fp.write(code)
+
+def savefig(fig, suffix):
+    here = os.path.dirname(os.path.realpath(__file__))
+    filename = os.path.join(here, "eval-figs", f"{prefix}-{unique_hash}-{suffix}")
+    fig.savefig(f"{filename}.png", dpi=150)
+    fig.savefig(f"{filename}.pdf", dpi=300)
+
+
+import os
+os.system("rm -f *.pkl")
+
+N_elements = 20
+use_galah_flags = config["use_galah_flags"]
+
+mcfa_kwds = dict()
+mcfa_kwds.update(config["mcfa_kwds"])
+
+elements = config[prefix]["elements"]
+if config[prefix].get("ignore_elements", None) is not None:
+    elements = [el for el in elements if el not in config[prefix]["ignore_elements"]]
+
+print(elements)
+
+
+mask = galah.get_abundance_mask(elements, use_galah_flags=use_galah_flags)
+
+
[…]

### 2019-09-11 article/experiments/eval-5.py

@@ -1,0 +1,327 @@
+
+
+
+
+
+from __future__ import division # Just in case. Use Python 3.
+
+import os
+import sys
+import pickle
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import yaml
+from matplotlib.ticker import MaxNLocator
+from collections import Counter
+from scipy import linalg
+from hashlib import md5
+
+sys.path.insert(0, "../../")
+
+from mcfa import (mcfa, grid_search, mpl_utils, utils)
+
+import galah_dr2 as galah
+
+
+
+matplotlib.style.use(mpl_utils.mpl_style)
+
+here = os.path.dirname(os.path.realpath(__file__))
+
+with open(f"config.yml") as fp:
+    config = yaml.load(fp)
+
+print(f"Config: {config}")
+
+np.random.seed(config["random_seed"])
+
+prefix = os.path.basename(__file__)[:-3]
+unique_hash = "none"
+
+def savefig(fig, suffix):
+    here = os.path.dirname(os.path.realpath(__file__))
+    filename = os.path.join(here, "eval-figs", f"{prefix}-{unique_hash}-{suffix}")
+    fig.savefig(f"{filename}.png", dpi=150)
+    fig.savefig(f"{filename}.pdf", dpi=300)
+
+
+# Get the same original data from the experiment hash:
+
+import os
+
+N_elements = 20
+use_galah_flags = config["use_galah_flags"]
+
+mcfa_kwds = dict()
+mcfa_kwds.update(config["mcfa_kwds"])
+
+elements = config[prefix]["elements"]
+if config[prefix].get("ignore_elements", None) is not None:
+    elements = [el for el in elements if el not in config[prefix]["ignore_elements"]]
+
+print(elements)
+
+
+mask = galah.get_abundance_mask(elements, use_galah_flags=use_galah_flags)
+
+
+galah_cuts = config[prefix]["galah_cuts"]
+if galah_cuts is not None:
+    print(f"Applying cuts: {galah_cuts}")
+    for k, (lower, upper) in galah_cuts.items():
+        mask *= (upper >= galah.data[k]) * (galah.data[k] >= lower)
+
+
+# This will be the default mask. 
+possible_additions_mask = np.zeros(len(mask), dtype=bool)
+# Start off by selecting everything that isn't in mask
+possible_additions_mask = ~mask
+
+galah_cuts = config[prefix]["galah_cuts"]
[…]

### 2019-09-17 article/mcfa.tex
add schematic

@@
 
+\begin{figure*}
+	\includegraphics[width=\textwidth]{experiments/schematic.png}
+	\caption{A schematic that visualises the model components.
+			 The data are shown on the left in greyscale, and magnified on the
+			 right where each star (row) is coloured by its identified
+			 component in latent space.
+			 For each chemical abundance (column) there is a mean value and
+			 variance that is independent of the latent space.
+			 The latent factors $\factorloads$ are analogous to nucleosynthetic
+			 yields and are common to all stars. The factor scores $\factorscores$
+			 have an entry for each yield, for each star (row). The latent scores are modelled
+			 by a mix of multivariate normal distributions of $\NumComponents$
+			 components, which are coloured accordingly. The matrices of factor 
+			 scores $\factorscores$ and factor loads $\factorloads$ are 
+			 visualised at the bottom. For clarity here we show transposed
+			 matrices (e.g., see Eq. \ref{eq:generative-model}).}
+	\label{fig:schematic}
+\end{figure*}
+
 Including latent factors in the model description allows us to account for 
@@
 
+\vspace{1em}
+
 \subsection{Initialisation} \label{sec:initialisation}
@@
 
+\noindent{}Thus the message length to encode $\factorloads$ is given by
 \begin{widetext}
-\noindent{}Thus the message length to encode $\factorloads$ is given by
 \begin{eqnarray}
@@
 \end{eqnarray}
-
-\noindent{}Combining equations \ref{eq:prior_J}, \ref{eq:prior_K}, \ref{eq:prior_pi}, \ref{eq:prior_xi_omega}, and \ref{eq:prior_L} with equation \ref{eq:mml} leads to the full message length:
+%\end{widetext}
+Combining equations \ref{eq:prior_J}, \ref{eq:prior_K}, \ref{eq:prior_pi}, \ref{eq:prior_xi_omega}, and \ref{eq:prior_L} with equation \ref{eq:mml} leads to the full message length:
+%\begin{widetext}
 \begin{eqnarray}
@@
 \end{widetext}
+
+%\hphantom{asdfasdf\\ asdf \\}
+%\vspace{5em}
+%\hphantom{asdfasdf\\ asdf \\}
+%\vspace{5em}
+
+%\newpage
+
+%\vspace{2em}
 
@@
 \end{figure*}
+\vspace{-50em}
 
@@
 
-\begin{figure*}[t!]
-	%\includegraphics[width=\textwidth]{experiments/exp1-compare-all.pdf}
-	\includegraphics[width=\textwidth]{experiments/eval-figs/eval-1-compare-all.pdf}
-	\caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
-    		 and specific variances $\specificvariance$ (right) compared to the 
-		 	 true data generating values
-		 	 for Experiment~1 (Section~\ref{sec:exp-1}). The agreement is excellent.}
-    \label{fig:exp1-compare}
-\end{figure*}
-
 Some technical background is warranted before we compare our estimated model
@@
 values, but they only differ by an orthogonal rotation. We discuss the impact of this limitation on real data in more detail in Section~\ref{sec:discussion}.
+
+\begin{figure*}
+	%\includegraphics[width=\textwidth]{experiments/exp1-compare-all.pdf}
+	\includegraphics[width=\textwidth]{experiments/eval-figs/eval-1-compare-all.pdf}
+	\caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
+    		 and specific variances $\specificvariance$ (right) compared to the 
+		 	 true data generating values
+		 	 for Experiment~1 (Section~\ref{sec:exp-1}). The agreement is excellent.}
+    \label{fig:exp1-compare}
+\end{figure*}
 
@@
 
+\bibliographystyle{aasjournal}
+\bibliography{mcfa}
+
 \appendix
@@
 
-\bibliographystyle{aasjournal}
-\bibliography{mcfa}
-
 \end{document}

### 2019-09-17 article/experiments/schematic.py

@@ -312,7 +312,7 @@ ax.text(-2.5 - 2, 0, r"$\textrm{stars}$", fontsize=10, rotation=90)
 text_kwds = dict(horizontalalignment="center", verticalalignment="center")
 data_text_y = ys + h + 3
 ax.text(xs + w/2, data_text_y, r"$\textrm{data}$", **text_kwds)
-ax.text(xs + w/2, equation_y_text, r"$\mathbf{Y}$", **text_kwds)
+ax.text(xs + w/2, equation_y_text, r"$\mathbf{X}^\top$", **text_kwds)
 
 
 # draw lines connecting
@@ -334,7 +334,7 @@ edge_kwds = dict(lw=0.75, c="#000000", zorder=10)
 draw_matrix_values(ax, mean, D + space, N-1, D, 1, edge_kwds=edge_kwds, vmin=-0.5, vmax=2)
 
 ax.text(D + space + D/2, data_text_y, r"$\textrm{mean}$", **text_kwds)
-ax.text(D + space + D/2, equation_y_text, r"$\boldsymbol{\mu}$", **text_kwds)
+ax.text(D + space + D/2, equation_y_text, r"$\boldsymbol{\mu}^\top$", **text_kwds)
 
 #boldsymbol{\mathbf{#1}}
 ax.text(D + space + D + space/2., N/2, r"$+$", **text_kwds)
@@ -398,7 +398,7 @@ ax.text(x + 0.5 + 2.2, y - 1, r"$3$", fontsize=6, **text_kwds)
 
 #scores, x, y, D, J)
 ax.text(x + D/2, data_text_y, "$\\textrm{latent}$\n$\\textrm{scores}$", **text_kwds)
-ax.text(x + D/2, equation_y_text, r"$\mathbf{S}$", **text_kwds)
+ax.text(x + D/2, equation_y_text, r"$\mathbf{S}^\top$", **text_kwds)
 
 #ax.text(x + D/2, data_text_y - 1.2, r"$\textrm{scores}$", **text_kwds)
 
@@ -432,7 +432,7 @@ ax.text(x - 1, y, r"$3$", fontsize=6, horizontalalignment="center", verticalalig
 latent_factors_y = -16
 draw_latent_factors(ax, true_theta["A"], x, latent_factors_y, D, 3)
[…]

### 2019-09-17 article/reviewer-response-v0.txt
foo
things

@@
  3. Done!
- 4. Need to make a schematic diagram and write response.
+ 4. Done!
  5. Done!
@@
  8. Need to justify S/N choice --> text will depend on #12.
- 9. Need to add text about N ~ 1000 stars in the toy model (awaiting figure to finish)
+ 9. Done!
 10. Done!
@@
     RESPONSE:
+
+    We have included a schematic to try to help visualise the method (new Figure 1).
+
+    We have also added an Appendix that references the online documentation, and includes a code example. This code will generate fake data, fit those data, and produce some figures.
 
@@
 
-    We have updated Figure 1 to show the grid search using N = 100,000 'stars' and N = 1,000 'stars'. We have also updated the text (see PDF changes).
+    We have updated Figure 1 to show the grid search using N = 100,000 'stars' and N = 5,000 'stars'. We have also added the following paragraphs to Section 2:
 
-    % ARC comment on changes
+    "We repeated this toy model experiment using a smaller sample size ($N = 5,000$) to be more representative of the sample sizes in later \Galah\ experiments (Section~\ref{sec:exp4}).
+The results of the grid search are also shown in Figure~\ref{fig:experiment-1-gridsearch}.
+Here BIC estimates the true number of latent factors correctly, but tends to underestimate the true number of clusters, more so than the message length.
+Although the difference between the true number of components and that given by the shortest message length is not large, this does serve to illustrate that in this example a larger number of data points are required to `resolve' the true number of components in latent space."
 

### 2019-09-17 article/mcfa.tex
things

@@
 are using latent factor analysis as a form of dimensionality reduction and
-simultaneously performing clustering in latent space.
+simultaneously performing clustering in latent space (Figure~\ref{fig:schematic}).
 
@@
 \vspace{1em}
-
 \subsection{Initialisation} \label{sec:initialisation}
@@
 
+\begin{widetext}
 \noindent{}Thus the message length to encode $\factorloads$ is given by
-\begin{widetext}
 \begin{eqnarray}
@@
  \end{eqnarray}
+ \newpage
 \end{widetext}
-
-%\hphantom{asdfasdf\\ asdf \\}
-%\vspace{5em}
-%\hphantom{asdfasdf\\ asdf \\}
-%\vspace{5em}
-
-%\newpage
-
-%\vspace{2em}
 
@@
     \label{fig:experiment-1-gridsearch}
+    \vspace{-10em}
 \end{figure*}
-\vspace{-50em}
 
@@
 and $\NumComponents = 10$, identical to the true values.
+
+We repeated this toy model experiment using a smaller sample size ($N = 5,000$) to be more representative of the sample sizes in later \Galah\ experiments (Section~\ref{sec:exp4}).
+The results of the grid search are also shown in Figure~\ref{fig:experiment-1-gridsearch}.
+Here BIC estimates the true number of latent factors correctly, but tends to underestimate the true number of clusters, more so than the message length.
+Although the difference between the true number of components and that given by the shortest message length is not large, this does serve to illustrate that in this example a larger number of data points are required to `resolve' the true number of components in latent space.
+
 It is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
@@
 
-\begin{figure*}
-	%\includegraphics[width=\textwidth]{experiments/exp1-compare-all.pdf}
-	\includegraphics[width=\textwidth]{experiments/eval-figs/eval-1-compare-all.pdf}
-	\caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
-    		 and specific variances $\specificvariance$ (right) compared to the 
-		 	 true data generating values
-		 	 for Experiment~1 (Section~\ref{sec:exp-1}). The agreement is excellent.}
-    \label{fig:exp1-compare}
-\end{figure*}
-
 We took the model with the preferred number of latent factors and components found
@@
 
-\begin{figure*}[t!]
-	%\includegraphics[width=\textwidth]{experiments/exp-missing-data-not-at-random-inflated-40percent.png}
-	\includegraphics[width=\textwidth]{experiments/eval-figs/eval-3-missing-data-40.pdf}
-	\caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
-    		 and specific variances $\specificvariance$ (right) compared to the 
-		 	 true data generating values
-		 	 for Experiment~2 (Section~\ref{sec:toy-model-missing-data}). Here 40\% of the data are missing at random.
-			 The agreement remains excellent,
-			 despite the large fraction of missing data entries. Note that the scales on the top panels are 2-10 times larger than those in Figure~\ref{fig:exp1-compare}.}
-    \label{fig:exp1-missing-data}
-\end{figure*}
-
 In Figure~\ref{fig:exp1-missing-data} we show the results of this
@@
 despite that 40\% of the data entries are missing.
-
-\vspace{2em}
 
@@
 
+We first selected stars with \texttt{flag\_cannon = 0} to exclude
+stars where there is reason to suspect that the stellar parameters
+(e.g., $\teff$, $\logg$) are unreliable, and as a result the 
+detailed chemical abundances would be untrustworthy. We then took all
+stars with a signal-to-noise ratio exceeding 30 in the blue arm (\texttt{snr\_c1 > 30}), and
+stars with no erroneous flags in all of the following abundances: 
+Mg, Na, Al, Si, K, Ca, Sc, Ti, Mn, Fe, Ni, Cu, Zn, Y, Ba, La, and Eu.
+These elements were chosen because they trace multiple nucleosynthetic pathways, and they are more commonly reported in the \Galah\ data release, allowing for a larger number of stars with a complete abundance inventory.
+There are 2,566 stars that met these criteria.
+
+We executed a grid search for the number of latent factors $\NumLatentFactors$
+and the number of components $\NumComponents$ that were preferred by the data.
+Starting with $\NumLatentFactors = 1$ and $\NumComponents = 1$, we trialled each
+permutation of $\NumLatentFactors$ and $\NumComponents$ up until $\NumLatentFactors = 7$
+and $\NumComponents = 5$. 
+The results of this grid search are shown in Figure~\ref{fig:exp3-gridsearch},
+where we show the negative log likelihood, the BIC, and message length found for each permutation.
+The behaviour of the BIC and the message length are very different here, unlike what was observed
+in our toy model. Here the BIC behaviour appears similar to the negative log likelihood in that the BIC
+prefers higher components and latent factors than the extent of the grid (e.g., $J > 7$ and $K > 5$).
+Indeed, if we were to trial higher values of $J$ and $K$ then the negative log likelihood would continue to increase.
+The model with five latent factors and three components ($J = 6$, $K = 3$) is found to have the shortest message length, which we take as our preferred model for these data.
+
+Earlier we described how the latent factors we estimate can only be identified up until an orthogonal rotation.
+If we want to interpret the latent factors estimated from \Galah\ data, then we must specify
+some target factor loads such that we can identify which factors are most similar to the yields we expect.
+We specified the following target latent factors where:
+\begin{itemize}
+	\item The first factor load should have non-zero entries in Eu and La (e.g., the $r$-process).
+	\item The second factor load should have non-zero entries in Ba, Y, and La (e.g., the $s$-process).
+	\item The third factor load should have non-zero entries in Fe-peak elements V, Mn, Fe, Ni, Zn, and Ti.
+	\item The fourth factor load should have non-zero entries in the odd-Z elements Na, Al, K, Sc, and Cu.
+	\item The fifth factor load should have non-zero entries in the $\alpha$-element tracers Si, Ca, and Mg.
+%	\item The first factor load should have non-zero entries in Eu and La (e.g., the $r$-process).
+%	\item The second factor load should have non-zero entries in Ba, Y, and La (e.g., the $s$-process).
+%	\item The third factor load should have non-zero entries in Fe-peak elements Sc, V, Mn, Fe, Ni, Cu, and Zn.
+%	\item The fourth factor load should have non-zero entries in the $\alpha$-element tracers Si, Ca, and Ti.
+%	\item The fifth factor load should have non-zero entries in the light odd-Z elements Na, Al, and K.
+\end{itemize}
+
+\begin{figure*}[t!]
+	%\includegraphics[width=\textwidth]{experiments/exp1-compare-all.pdf}
+	\includegraphics[width=\textwidth]{experiments/eval-figs/eval-1-compare-all.pdf}
+	\caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
+    		 and specific variances $\specificvariance$ (right) compared to the 
+		 	 true data generating values
+		 	 for Experiment~1 (Section~\ref{sec:exp-1}). The agreement is excellent.}
+    \label{fig:exp1-compare}
+\end{figure*}
+
+We initially set each non-zero entry in these target factor loads $\factorloads_\textrm{target}$ to $E^{-\frac{1}{2}}$, where $E$ is the number of non-zero entries in that factor load, to ensure that $\factorloads_\textrm{target}$ is mutually orthogonal.
+We solved for the $\NumLatentFactors$ unknown angles to produce a valid rotation matrix $\vec{R}$ that would make our estimated loads $\factorloads$ as close as possible to the target loads $\factorloads_\textrm{target}$, and then applied that rotation to the model.
+The target loads and (rotated) estimated loads are shown in Figure~\ref{fig:exp3-factor-loads}.
+Note that the purpose of this procedure is not to `find' the target loads that we expect, but to provide as little information needed in order to identify and describe all factor loads within an astrophysical context.
+This procedure still requires that the factors be mutually orthogonal and that they describe the data.
+For these reasons, we will not always recover the exact target loads we seek: we will only be able to identify factor loads that are closest to the target loads.
+
+This is demonstrated in Figure~\ref{fig:exp3-factor-loads}, where some estimated factor loads match
+closely to the target load (e.g., $\mathbf{L}_2$ which we identify as the s-process), and some barely
+match at all (e.g., $\mathbf{L}_5$). Here we show the absolute entry of the factor loads because even
+if an entry is negative, the corresponding factor scores could also be negative, and their product will contribute to the observed abundances. For this reason the sign does not matter here.
+
+Some of these factor load entries may be non-zero because we require the latent factors to
+be mutually orthogonal, and not because they truly contribute to the data. To try and
+disentangle these possibilities, we calculate the fractional contribution that factor
+load makes to the observed abundances relative to other factor loads. We define the
+fractional contribution of the $j$th factor load to the $d$th
+data dimension as:
+\begin{equation}
+	\mathbf{C}_\textrm{d,j} = \frac{\sum^{N}|\factorloads_{j,d}\factorscores_{n,j}|}{\sum^J\sum^{N}|\factorloads_{j,d}\factorscores_{n,j}|} \quad .
+\end{equation}
+The fractional contributions to each element are shown in the right hand side of
+Figure~\ref{fig:exp3-factor-loads}. We identify the first factor $\mathbf{L}_1$ 
+as being most similar to the r-process, and here it is the dominant contributor to Eu, a typical r-process tracer.
+Surprisingly we also find that this factor load is a reasonable contributor to the odd-Z element Sc.
+The specific scatter in Sc is 0.03~dex (Figure~\ref{fig:exp3-specific-scatter}), suggesting that the Sc abundances are well-described by this latent factor model.
+
+The second latent factor $\mathbf{L}_2$ here is most representative of the slow neutron capture process (s-process), with dominant contributions to Ba, and Y.
+This factor has some support at other elements, notably K.
+$\mathbf{L}_3$ is the primary contributor to nearly all Fe-peak elements, with close to negligible contributions from other factors.
+The exception here is Cu, where a near-equal contribution comes from $\mathbf{L}_4$.
+The fifth latent factor $\mathbf{L}_5$ is the dominant contributor to the $\alpha$-element tracers Si, Ca, and Ti, and surprisingly, Al. 
+The specific scatter after accounting for these latent factors is smallest
+for Fe (0.01~dex) and largest for K (0.13~dex; Figure~\ref{fig:exp3-specific-scatter}). The typical scatter in most elements is about 0.05~dex.
+
+In Figure~\ref{fig:exp3-latent-space} we show the inferred clustering in latent space,
+where the separation between components is arguably best seen in the splitting between $\factorscores_6$ with respect to $\factorscores_2$ or $\factorscores_3$.
+When projected to data space (Figure~\ref{fig:exp3-data-space}) the third component (light green) is seen to have relatively higher abundance ratios of [K,Ba,Zn/Fe] at a given [Fe/H].
+This is consistent with the clustering in latent space.
+
+\begin{figure*}[t!]
+	%\includegraphics[width=\textwidth]{experiments/exp-missing-data-not-at-random-inflated-40percent.png}
+	\includegraphics[width=\textwidth]{experiments/eval-figs/eval-3-missing-data-40.pdf}
+	\caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
+    		 and specific variances $\specificvariance$ (right) compared to the 
+		 	 true data generating values
+		 	 for Experiment~2 (Section~\ref{sec:toy-model-missing-data}). Here 40\% of the data are missing at random.
+			 The agreement remains excellent,
+			 despite the large fraction of missing data entries. Note that the scales on the top panels are 2-10 times larger than those in Figure~\ref{fig:exp1-compare}.}
+    \label{fig:exp1-missing-data}
+\end{figure*}
+
+\subsection{\Galah\ survey data with an increasing number of stars with missing data entries} \label{sec:exp10}
+Here we extend our experiment in Section~\ref{sec:exp4} to progressively
+include more stars, even though those stars have some abundance measurements
+missing. 
+Specifically we started with the same subset of 1,072 stars in Section~\ref{sec:exp4}
+and added a random set of stars that met our criteria of \texttt{flag\_cannon = 0} and
+\texttt{snr\_c1 > 40}.
+
+We initially added 100 stars to give a sample of $N =$ 1,172, then repeated
+the grid search for the number of latent factors and components, and recorded
+the model with the lowest message length. We then repeated this
+procedure using 1,000 stars ($N =$ 2,072), again with 10,000 ($N =$ 11,072),
+and finally using all 99,174 stars that met the criteria of \texttt{flag\_cannon = 0} and
+\texttt{snr\_c1 > 40} to give a total sample size of $N =$ 100,246 stars.
+
+For sample sizes up to $N \sim $2,000 we found that five latent factors were
+preferred, and these factors shared common features (Figure~\ref{fig:exp10-comparison}).
+This illustrates that the
+first\footnote{`First' has no concept here in terms of factor load ordering, but
+for the purposes of comparing inferred loads from different data sets we have ordered
+the loads to be as close to those inferred in Section~\ref{sec:exp4}.} set of inferred factor
+loads inferred from a smaller, complete data set, remain largely unchanged despite the increasing sample size and the increasing number of missing data entries.
+When the sample size reaches $N = $11,072 we find another three latent factors are
+required to best explain the data.
+
+When $N \sim$ 100,000, the preferred number of latent factors rises to twelve ($J = 12$). Of note among these factors is $\factorloads_9$, where
+no `target load' was prescribed\footnote{Although no `target load' was prescribed here, this statement should be interpreted with caution because the mutual orthogonality constraint exists and the remaining factors do have target loads.}, and the non-zero entries mimic what might be expected from a
+light s-process production. Similarly, $\factorloads_{10}$ has near zero contributions
+everywhere except among the light elements Al, K, Ca, and Sc. 
+With this data set we find that $K = 16$ components are preferred in latent space. None of these components appear coherently structured in their positions or motions.
+
 \begin{figure}
@@
 
-We first selected stars with \texttt{flag\_cannon = 0} to exclude
-stars where there is reason to suspect that the stellar parameters
-(e.g., $\teff$, $\logg$) are unreliable, and as a result the 
-detailed chemical abundances would be untrustworthy. We then took all
-stars with a signal-to-noise ratio exceeding 40 in the blue arm (\texttt{snr\_c1 > 40}), and
-stars with no erroneous flags in all of the following abundances: Na, Al, Si, K, Ca, Sc, Ti, V, Mn, Fe, Ni, Cu, Zn, Y, Ba, La, and Eu.
-These elements were chosen because they trace multiple nucleosynthetic pathways, and they are more commonly reported in the \Galah\ data release, allowing for a larger number of stars with a complete abundance inventory.
-There are 1,072 stars that met these criteria.
-
-We executed a grid search for the number of latent factors $\NumLatentFactors$
-and the number of components $\NumComponents$ that were preferred by the data.
-Starting with $\NumLatentFactors = 1$ and $\NumComponents = 1$, we trialled each
-permutation of $\NumLatentFactors$ and $\NumComponents$ up until $\NumLatentFactors = 7$
-and $\NumComponents = 5$. 
-The results of this grid search are shown in Figure~\ref{fig:exp3-gridsearch},
-where we show the negative log likelihood, the BIC, and message length found for each permutation.
-The behaviour of the BIC and the message length are very different here, unlike what was observed
-in our toy model. Here the BIC behaviour appears similar to the negative log likelihood in that the BIC
-prefers higher components and latent factors than the extent of the grid (e.g., $J > 7$ and $K > 5$).
-Indeed, if we were to trial higher values of $J$ and $K$ then the negative log likelihood would continue to increase.
-The model with five latent factors and three components ($J = 5$, $K = 3$) is found to have the shortest message length, which we take as our preferred model for these data.
-
 \begin{figure}
@@
 
-Earlier we described how the latent factors we estimate can only be identified up until an orthogonal rotation.
-If we want to interpret the latent factors estimated from \Galah\ data, then we must specify
-some target factor loads such that we can identify which factors are most similar to the yields we expect.
-We specified the following target latent factors where:
-\begin{itemize}
-	\item The first factor load should have non-zero entries in Eu and La (e.g., the $r$-process).
-	\item The second factor load should have non-zero entries in Ba, Y, and La (e.g., the $s$-process).
-	\item The third factor load should have non-zero entries in Fe-peak elements Sc, V, Mn, Fe, Ni, Cu, and Zn.
-	\item The fourth factor load should have non-zero entries in the $\alpha$-element tracers Si, Ca, and Ti.
-	\item The fifth factor load should have non-zero entries in the light odd-Z elements Na, Al, and K.
-\end{itemize}
-
-We initially set each non-zero entry in these target factor loads $\factorloads_\textrm{target}$ to $E^{-\frac{1}{2}}$, where $E$ is the number of non-zero entries in that factor load, to ensure that $\factorloads_\textrm{target}$ is mutually orthogonal.
-We solved for the $\NumLatentFactors$ unknown angles to produce a valid rotation matrix $\vec{R}$ that would make our estimated loads $\factorloads$ as close as possible to the target loads $\factorloads_\textrm{target}$, and then applied that rotation to the model.
-The target loads and (rotated) estimated loads are shown in Figure~\ref{fig:exp3-factor-loads}.
-Note that the purpose of this procedure is not to `find' the target loads that we expect, but to provide as little information needed in order to identify and describe all factor loads within an astrophysical context.
-This procedure still requires that the factors be mutually orthogonal and that they describe the data.
-For these reasons, we will not always recover the exact target loads we seek: we will only be able to identify factor loads that are closest to the target loads.
-
-This is demonstrated in Figure~\ref{fig:exp3-factor-loads}, where some estimated factor loads match
-closely to the target load (e.g., $\mathbf{L}_2$ which we identify as the s-process), and some barely
-match at all (e.g., $\mathbf{L}_5$). Here we show the absolute entry of the factor loads because even
-if an entry is negative, the corresponding factor scores could also be negative, and their product will contribute to the observed abundances. For this reason the sign does not matter here.
-
-Some of these factor load entries may be non-zero because we require the latent factors to
-be mutually orthogonal, and not because they truly contribute to the data. To try and
-disentangle these possibilities, we calculate the fractional contribution that factor
-load makes to the observed abundances relative to other factor loads. We define the
-fractional contribution of the $j$th factor load to the $d$th
-data dimension as:
-\begin{equation}
-	\mathbf{C}_\textrm{d,j} = \frac{\sum^{N}|\factorloads_{j,d}\factorscores_{n,j}|}{\sum^J\sum^{N}|\factorloads_{j,d}\factorscores_{n,j}|} \quad .
-\end{equation}
-The fractional contributions to each element are shown in the right hand side of
-Figure~\ref{fig:exp3-factor-loads}. We identify the first factor $\mathbf{L}_1$ 
-as being most similar to the r-process, and here it is the dominant contributor to Eu, a typical r-process tracer.
-Surprisingly we also find that this factor load is a reasonable  contributor to the Fe-peak element Sc.
-The specific scatter in Sc is 0.03~dex (Figure~\ref{fig:exp3-specific-scatter}), suggesting that the Sc abundances are well-described by this latent factor model.
-
-The second latent factor $\mathbf{L}_2$ here is most representative of the slow neutron capture process (s-process), with dominant contributions to Ba, Y, and La.
-This factor has some support at other elements, notably K.
-Most Sc is contributed by $\mathbf{L}_3$, our Fe-peak latent factor.
-In fact $\mathbf{L}_3$ is the primary contributor to nearly all Fe-peak elements, with close to negligible contributions from other factors.
-The exception here is Zn, where a near-equal contribution comes from $\mathbf{L}_5$.
-The third latent factor $\mathbf{L}_4$ is reasonably well behaved.
-It is the dominant contributor to the $\alpha$-element tracers Si, Ca, and Ti, and surprisingly, Al.
-It also seems to contribute a non-zero fraction to Eu, with negligible contributions elsewhere.
-The specific scatter after accounting for these latent factors is smallest
-for Fe (0.01~dex) and largest for K (0.13~dex; Figure~\ref{fig:exp3-specific-scatter}). The typical scatter in most elements is about 0.05~dex.
-
-In Figure~\ref{fig:exp3-latent-space} we show the inferred clustering in latent space,
-where the separation between components is arguably best seen in the splitting between $\factorscores_5$ with respect to $\factorscores_2$ or $\factorscores_3$.
-When projected to data space (Figure~\ref{fig:exp3-data-space}) the third component (dark green) is seen to have relatively higher abundance ratios of [K,Ba,Zn/Fe] at a given [Fe/H],
-and lower abundance ratios of [V/Fe]. This is consistent with the clustering in latent space.
-
-\subsection{\Galah\ survey data with an increasing number of stars with missing data entries} \label{sec:exp10}
-Here we extend our experiment in Section~\ref{sec:exp4} to progressively
-include more stars, even though those stars have some abundance measurements
-missing. 
-Specifically we started with the same subset of 1,072 stars in Section~\ref{sec:exp4}
-and added a random set of stars that met our criteria of \texttt{flag\_cannon = 0} and
-\texttt{snr\_c1 > 40}.
-
-We initially added 100 stars to give a sample of $N =$ 1,172, then repeated
-the grid search for the number of latent factors and components, and recorded
-the model with the lowest message length. We then repeated this
-procedure using 1,000 stars ($N =$ 2,072), again with 10,000 ($N =$ 11,072),
-and finally using all 99,174 stars that met the criteria of \texttt{flag\_cannon = 0} and
-\texttt{snr\_c1 > 40} to give a total sample size of $N =$ 100,246 stars.
-
-For sample sizes up to $N \sim $2,000 we found that five latent factors were
-preferred, and these factors shared common features (Figure~\ref{fig:exp10-comparison}).
-This illustrates that the
-first\footnote{`First' has no concept here in terms of factor load ordering, but
-for the purposes of comparing inferred loads from different data sets we have ordered
-the loads to be as close to those inferred in Section~\ref{sec:exp4}.} set of inferred factor
-loads inferred from a smaller, complete data set, remain largely unchanged despite the increasing sample size and the increasing number of missing data entries.
-When the sample size reaches $N = $11,072 we find another three latent factors are
-required to best explain the data.
-
-When $N \sim$ 100,000, the preferred number of latent factors rises to twelve ($J = 12$). Of note among these factors is $\factorloads_9$, where
-no `target load' was prescribed\footnote{Although no `target load' was prescribed here, this statement should be interpreted with caution because the mutual orthogonality constraint exists and the remaining factors do have target loads.}, and the non-zero entries mimic what might be expected from a
-light s-process production. Similarly, $\factorloads_{10}$ has near zero contributions
-everywhere except among the light elements Al, K, Ca, and Sc. 
-With this data set we find that $K = 16$ components are preferred in latent space. None of these components appear coherently structured in their positions or motions.
-
 \begin{figure*}[t!]
@@
 
-\vspace{1em}
 \section{Discussion} \label{sec:discussion}
@@
 model does provide the first data-driven approach to nucleosynthesis and chemical tagging.
-We advocate that more data, and the inclusion of weakly informative priors -- joint probabilities using astrometry and a simplified model of the Milky Way -- would help in realising the full potential of chemical tagging.\\
-
-\vspace{2em}
-
-%This \emph{Article} was compiled from the \texttt{git} commit \todo{\texttt{HASH}}.
-%The analysis, results, and figures presented in this \emph{Article} can be reproduced in full 
-%using the following commands in a modern terminal:\vspace{1em}\\
-%\texttt{git clone https://github.com/andycasey/mcfa}\\
-%\texttt{cd mcfa/}\\
-%\todo{\texttt{sh article/build.sh}}
+We advocate that more data, and the inclusion of weakly informative priors -- joint probabilities using astrometry and a simplified model of the Milky Way -- would help in realising the full potential of chemical tagging.
 

### 2019-09-30 article/experiments/eval-6.py
add eval-6 fig

@@ -1,0 +1,397 @@
+
+"""
+Plot results from exp10
+"""
+
+import os
+import pickle
+import sys
+import matplotlib.pyplot as plt
+import yaml
+import numpy as np
+from scipy import linalg
+
+sys.path.insert(0, "../../")
+
+from mcfa import (mcfa, grid_search, mpl_utils, utils)
+
+np.random.seed(0)
+
+
+with open("config.yml") as fp:
+    config = yaml.load(fp)
+
+"""
+# Get original model.
+with open("89dab-exp3-model.pkl", "rb") as fp:
+    original_model = pickle.load(fp)
+
+
+# Get various models from exp10 for comparison.
+comparison_model_paths = [
+    "exp10-models-size-100.pkl",
+    "exp10-models-size-1000.pkl",
+    "exp10-models-size-10000.pkl",
+    "exp10-models-size-99174.pkl"
+]
+
+# 89dab is for all intents and purposes identical to 8761b
+with open("8761b.yml") as fp:
+    config = yaml.load(fp)
+
+
+
+label_names = [
+    "na_h",
+    "al_h",
+    "si_h",
+    "k_h",
+    "ca_h",
+    "sc_h",
+    "ti_h",
+    "v_h",
+    "mn_h",
+    "fe_h",
+    "ni_h",
+    "cu_h",
+    "zn_h",
+    "y_h",
+    "ba_h",
+    "la_h",
+    "eu_h"
+]
+"""
+
+comparison_model_paths = [
+    "eval-figs/eval-5-models-size-1000.pkl",
+    "eval-figs/eval-5-models-size-10000.pkl",
+    "eval-figs/eval-5-models-size-157242.pkl"
+]
+
+label_names = ['na_h', 'mg_h', 'al_h', 'si_h', 'k_h', 'ca_h', 'sc_h', 'ti_h', 'mn_h', 'fe_h', 'ni_h', 'cu_h', 'zn_h', 'y_h', 'ba_h', 'la_h', 'eu_h']
+
+def _get_model(path):
+    with open(path, "rb") as fp:
+        contents = pickle.load(fp)
+
+    Js, Ks, gJs, gKs, converged, meta, X_H, label_names = contents
+    #pickle.dump((Js, Ks, gJs, gKs, converged, meta, X_H, label_names), fp)
+
+    print(f"label_names from {path}: {label_names}")
+
+    # Get best model.
[…]
